### 2023-02-28 src/tex/ms.tex
[showyourwork] first commit

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}

% Begin!
\begin{document}

% Title
\title{An open source scientific article}

% Author list
\author{@afarah18}

% Abstract with filler text
\begin{abstract}
    Lorem ipsum dolor sit amet, consectetuer adipiscing elit.
    Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis.
    Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna.
    Donec vehicula augue eu neque, morbi tristique senectus et netus et.
    Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum.
    Phasellus eu tellus sit amet tortor gravida placerat.
    Integer sapien est, iaculis in, pretium quis, viverra ac, nunc.
    Praesent eget sem vel leo ultrices bibendum.
    Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac.
    Curabitur auctor semper nulla donec varius orci eget risus.
    Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam.
    Duis eget orci sit amet orci dignissim rutrum.
\end{abstract}

% Main body with filler text
\section{Introduction}
\label{sec:intro}

Lorem ipsum dolor sit amet, consectetuer adipiscing elit.
Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis.
Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna.
Donec vehicula augue eu neque, morbi tristique senectus et netus et.
Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum.
Phasellus eu tellus sit amet tortor gravida placerat.
Integer sapien est, iaculis in, pretium quis, viverra ac, nunc.
Praesent eget sem vel leo ultrices bibendum.
Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac.
Curabitur auctor semper nulla donec varius orci eget risus.
Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam.
Duis eget orci sit amet orci dignissim rutrum.

Nam dui ligula, fringilla a, euismod sodales, sollici- tudin vel, wisi.
Morbi auctor lorem non justo, nam lacus libero, pretium at, lobortis vitae.
Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna.
Morbi ac orci et nisl hendrerit mollis, suspendisse ut massa, cras nec ante.
Pellentesque a nulla cum sociis natoque penatibus et magnis dis parturient.
Aliquam tincidunt urna, nulla ullamcorper vestibulum turpis.
Pellentesque cursus luctus mauris \citep{Luger2021}.

\bibliography{bib}

\end{document}

### 2023-03-01 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 % Title
-\title{An open source scientific article}
+\title{Nonparametric Spectral Siren Cosmology}
 
@@
 
-% Main body with filler text
 \section{Introduction}
@@
 
-Lorem ipsum dolor sit amet, consectetuer adipiscing elit.
-Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis.
-Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna.
-Donec vehicula augue eu neque, morbi tristique senectus et netus et.
-Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum.
-Phasellus eu tellus sit amet tortor gravida placerat.
-Integer sapien est, iaculis in, pretium quis, viverra ac, nunc.
-Praesent eget sem vel leo ultrices bibendum.
-Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac.
-Curabitur auctor semper nulla donec varius orci eget risus.
-Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam.
-Duis eget orci sit amet orci dignissim rutrum.
-
-Nam dui ligula, fringilla a, euismod sodales, sollici- tudin vel, wisi.
-Morbi auctor lorem non justo, nam lacus libero, pretium at, lobortis vitae.
-Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna.
-Morbi ac orci et nisl hendrerit mollis, suspendisse ut massa, cras nec ante.
-Pellentesque a nulla cum sociis natoque penatibus et magnis dis parturient.
-Aliquam tincidunt urna, nulla ullamcorper vestibulum turpis.
-Pellentesque cursus luctus mauris \citep{Luger2021}.
+\section{Motivating Example: Fitting an incorrect functional form to the BBH mass distribution yields biased inference of cosmological parameters.}
+\section{Methods}
+\subsection{Data simulation}
+\begin{itemize}
+    \item \texttt{GWMockCat} with specific PSD and parameter uncertainty choices
+    \item explain how we estimate selection effects (injections, Pdet, etc.)
+\end{itemize}
+\subsection{Population inference}
+(the order of these points is probably not the best way to explain things, should rethink it)
+\begin{itemize}
+    \item explain spectral sirens in the generative way: choose one H0 draw, transform the masses using DL, use the source frame masses as inputs to the mass distribution, evaluate the likelihood of that mass distribution and that H0 draw, repeat.
+    \item explain HBA with selection effects
+    \item in the parametric way, $p(\theta|\Lambda)$ is evaluated with a specific model and $\Lambda$ is a small set of hyperparameters dictating the models exact shape. 
+    \item Now, hyperparameters are the rate at each PE sample's source frame mass value, subject to a hyper-prior set by the Gaussian process draw.
+Instead of a few hyperparameters, we have $N_{\text{evs}}\times M + N_{\text{inj}}$ hyperparameters, where $N$ is the number of events, $M$ is the number of PE samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xsi$.
+    \item note this is not the usual use case for GPs (regression, cite some examples in GW data analysis).
+We are instead using GPs as a prior family of functions so that the mass distribution is smooth.
+Cite Tom's AR paper
+    \item describe what a GP and kernel are, \citep{}
+    \item Kernel choice
+    \item Explain priors on hyper-hyper params: smoothness assumptions, PC priors~\citep{simpson_penalising_2017} (if used)
+    \item cite \texttt{tinygp}, \texttt{numpyro}, HMC/NUTS~\citep{hoffman_no-u-turn_2011}, etc.
+    \item mention any computational tricks: Quasi-separable kernels, sparse kernels, etc.
+\end{itemize}
 

### 2023-03-01 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 % Define document class
-\documentclass[twocolumn]{aastex631}
+\documentclass[]{aastex631}
 \usepackage{showyourwork}
@@
 
-\section{Motivating Example: Fitting an incorrect functional form to the BBH mass distribution yields biased inference of cosmological parameters.}
+\section{Motivating Example}
+Fitting an incorrect functional form to the BBH mass distribution yields biased inference of cosmological parameters: Draw events from a Gaussian distribution, fit it with (A) a Gaussian distribution and (B) a truncated power law, show that you get the wrong H0 in case B but the right one in case A.
 \section{Methods}
+\label{sec:methods}
 \subsection{Data simulation}
@@
     \item Now, hyperparameters are the rate at each PE sample's source frame mass value, subject to a hyper-prior set by the Gaussian process draw.
-Instead of a few hyperparameters, we have $N_{\text{evs}}\times M + N_{\text{inj}}$ hyperparameters, where $N$ is the number of events, $M$ is the number of PE samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xsi$.
+Instead of a few hyperparameters, we have $N_{\text{evs}}M + N_{\text{inj}}$ hyperparameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of PE samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
     \item note this is not the usual use case for GPs (regression, cite some examples in GW data analysis).
@@
 Cite Tom's AR paper
-    \item describe what a GP and kernel are, \citep{}
+    \item describe what a GP and kernel are, \citep{rasmussen_gaussian_2006}
     \item Kernel choice
@@
 \end{itemize}
+\section{Results}
+\label{sec:results}
+\begin{itemize}
+    \item Figure of mass distribution fit
+    \item Figure of posterior on \Ho, \Omm
+    \item Can get constraints on \Ho to X\% by the end of O5
+    \item Comparison to fitting with correct parametric model with the same data: hopefully same central value but smaller error bars
+    \item Comparison to fitting with \emph{incorrect} parametric model with the same data: hopefully different central value    
+\end{itemize}
+\section{Discussion}
+\label{sec:discussion}
+\begin{itemize}
+    \item we are able to get constraints on comsological parameters with a very flexible model for the mass distribution
+    \item This is because the information in spectral sirens comes from assuming events come from the same overall mass distribution across redshift, and that mass distribution doesn't evolve with redshift in the same exact way as would be mimiced by cosmological redshifting.
+    \item future/in prep.
+work will do the same thing but for a mass distribution that is allowed to evolve with redshift, similarly to the parametric analysis done in~\cite{ezquiaga_spectral_2022}, but with the evolution with redshift allowed to be arbitrary so long as it does not mimic cosmology.
+\end{itemize}
 

### 2023-10-06 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2023-02-28 src/tex/ms.tex
@@
 % Define document class
-\documentclass[twocolumn]{aastex631}
+\documentclass[]{aastex631}
 \usepackage{showyourwork}
+\usepackage{macros}
+\usepackage{amsmath}
 
@@
 
-% Main body with filler text
 \section{Introduction}
 \label{sec:intro}
+Like light, gravitational waves (GWs) are redshifted as they propagate across the universe.
+Unlike light, however, the selection effects of GWs are extremely well understood, allowing for a precise estimate of each catalogs’ completeness and an unbiased measurement of the true GW population.
+This allows GW signals to be used as relatively clean probes of cosmological parameters \citep[e.g.][]{bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide}.
 
-Lorem ipsum dolor sit amet, consectetuer adipiscing elit.
-Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis.
-Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna.
-Donec vehicula augue eu neque, morbi tristique senectus et netus et.
-Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum.
-Phasellus eu tellus sit amet tortor gravida placerat.
-Integer sapien est, iaculis in, pretium quis, viverra ac, nunc.
-Praesent eget sem vel leo ultrices bibendum.
-Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac.
-Curabitur auctor semper nulla donec varius orci eget risus.
-Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam.
-Duis eget orci sit amet orci dignissim rutrum.
+GW signals provide direct measurements of each source’s luminosity distance and redshifted, or detector frame, masses, $m_{\det} = m_{\source}(1+z)$.
+Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the GW signal was emitted, $H(z)$.
+In practice, the source frame mass is unknown, but it is possible to know the location of features in the source frame mass distribution.
+The whole mass distribution therefore acts similarly to an electromagnetic spectrum, where the location of spectral features provides a redshift measurement.
+The method of using the mass distribution of GW sources to measure cosmological parameters has therefore been coined ``spectral sirens'' \citep{ezquiaga_spectral_2022}.
+Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff+fin} using the BNS mass distribution, and extended to the BBH mass distribution by \cite{farr_future_2019}.
 
-Nam dui ligula, fringilla a, euismod sodales, sollici- tudin vel, wisi.
-Morbi auctor lorem non justo, nam lacus libero, pretium at, lobortis vitae.
-Donec aliquet, tortor sed accumsan bibendum, erat ligula aliquet magna.
-Morbi ac orci et nisl hendrerit mollis, suspendisse ut massa, cras nec ante.
-Pellentesque a nulla cum sociis natoque penatibus et magnis dis parturient.
-Aliquam tincidunt urna, nulla ullamcorper vestibulum turpis.
-Pellentesque cursus luctus mauris \citep{Luger2021}.
+Perhaps because of the analogy to electromagnetic spectra in which the location of the source frame spectral features is well-known, or because of the examples given in \cite{farr_future_2019} and \cite{chernoff+fin}, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to do this measurement, from e.g.
+theoretical expectations of the location of a pulsational-pair instability pileup, upper mass gap, or maximum neutron star mass.
+However, the information in the spectral siren measurement comes solely from the assumption that, at a given luminosity distance, all CBC mergers follow the same mass distribution.
+Therefore the location, shape, and existence of all features in the mass distribution can be inferred simultaneously with $H(z)$.
+
+In this work, we explicitly demonstrate that no \emph{a priori} knowledge about the shape of the CBC mass spectrum is necessary to use the spectral siren methodology by inferring \Ho with a non-parametric model for the mass distribution of CBCs.
+This flexible model makes minimal prior assumptions for the shape of the mass distribution and is therefore capable of accurately inferring a wide range of mass spectrum morphologies.
+Despite its flexibility, it is able to consistently obtaining an unbiased measurement for \Ho, showing that nonparametric methods are not only sufficient for a spectral siren measurement, they can also mitigate systematic effects in the measurement caused by model misspecification.
+
+\section{Motivating Example}
+\comment{Maybe this should just be part of results. Like just do this for O5 data and show how bad it can be. if its not that bad, we can do it for even more events to show when it will be a problem.}
+
+In this Section, we show that fitting an incorrect functional form to the mass distribution of CBCs biases the inference of cosmological parameters when using the spectral siren methodology.
+To demonstrate this, we generate a catalog of \result{1,000} GW events from an underlying mass distribution described by the \plp{} model and its maximum \emph{a posteriori} hyperparameters from \citet{o3b_pop}.
+We then use this catalog to infer \Ho{} using three different models for the source-frame mass distribution:
+\begin{enumerate}
+    \item the \plp{} model, which includes the true mass distribution within its hyperprior
+    \item a single power law, which does not include the true mass distribution within its hyperprior, and 
+    \item the flexible, Gaussian process-based model that we use for the remainder of this work, which can closely approximate the morphology of the true mass distribution, along with several other morphologies.
+\end{enumerate}
+The results of each of these fits are shown in Figure~\ref{fig:motivation}. 
+The \result{top/left} panel shows the inferred source frame mass distribution for each of the considered models, and the \result{bottom/right} panel shows the corresponding posteriors on \Ho{}.
+
+With \result{1,000} events, we find that using the incorrect parametric form of the mass distribution (a power law) causes a biased inferred value of \Ho{} at the \result{X}-$\sigma$ level, whereas using the correct functional form (\plp) provides an unbiased inference of \Ho.
+However, this is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the Gaussian process-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution.
+The Gaussian process-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters.
+We have repeated this analysis with \result{50} separate simulated catalogs, and found that the power law model causes us to infer a \result{higher/lower} value of \Ho{} at greater than \result{X}-$\sigma$, \result{Y}\% of the time, whereas the Gaussian process-based model reaches that level of bias only \result{Z}\% of the time, and the \plp{} model reaches the same level of bias \result{A}\% of the time.
+This indicates that prior knowledge of the shape of the mass distribution is not required to perform a spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/diff_mass_dists_example.pdf}
+    \caption{Caption}
+    \label{fig:motivation}
+\end{figure}
+
+\section{Methods}
+\label{sec:methods}
+\subsection{Data simulation}
+We generate a mock catalog of GW events characteristic of that expected from the LVK's fifth observing run (O5). 
+These events are drawn from an underlying population described by
+\begin{equation}
+\label{eq:underlying population}
+    \frac{\diff N}{\diff m_1 \diff m_2 \diff z} \propto p(m_1, m_2|\bar{\Lambda}_m) p(z|H_0, \Omega_M),
+\end{equation}
+where 
+\begin{align}
+    p(m_1,m_2|\bar{\Lambda}_m) &\propto \\
+    p(z|H_0, \Omega_M) &\propto \frac{\diff V_C}{\diff z} \frac{1}{1+z},
+\label{eq:underlying mass and redshift dist}
+\end{align}
+, $V_C(H_0, \Omega_M)$ is the comoving volume for given cosmological parameters \Ho{} and \Omm{}, and $\bar{\Lambda_m} = \alpha$...
+are the parameters describing the power law in primary mass, ...
+respectively.
+When generating the simulated events, we have fixed $\alpha= $... and used cosmological parameters \Ho$=68.7$, \Omm$=0.3$, $\Omega_\Lambda=1-\Omega_M$.
+These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \_ model \cite{o3b_pop}, which is morphologically similar to that described in Equations~\ref{eq:underlying population}--\ref{eq:underlying mass and redshift dist}.
+We neglect spins entirely, as they do not carry cosmological information.
+\comment{TODO: fill in details of mass distribuiton.
+Should we use PDB and include BNSs, which will give us a nice sharp feature to work with, or should we use PLP which has a bump?
+Could also do PLP and then put a uniform dist or power law for NS masses at the LMG, and leave an empty mass gap.
+Would need to get relative heights correct though.}
+
+After passing the simulated events through projected detector selection effects, the resulting catalog has \result{X} events, including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} BBH systems, consistent with the numbers projected for O5 by \citet{kiendrebogo_observing_2023}.
+We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from O5 detectors.
+\texttt{GWMockCat} also simulates a set of software injections, which we use to estimate selection effects in the inference.
+To determine the detectability of both injections and simulated events in O5, we use the projected O5 LIGO power spectral density \citep{dcc page} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
+The full procedure for this mock data generation process is described in \citet{fishbach_where, farah_things_2023, essick_dagnabbit_2023}, and the exact settings used for \texttt{GWMockCat} are made available in the accompanying data release.
+\comment{TODO:put a showyourwork link here.}
+
+\subsection{Spectral sirens}
+\label{sec:ss}
+\begin{itemize}
+    \item explain the basis of spectral sirens
+    \item explain HBA with selection effects (or don't and just reference papers).
+    \item ``Therefore, another way to think about it is...
+/ Algorithmically, this is equivalent to..." explain spectral sirens in the generative way: choose one H0 draw, transform the masses using DL, use the source frame masses as inputs to the mass distribution, evaluate the likelihood of that mass distribution and that H0 draw, repeat.
+\end{itemize}
+
+\subsection{Gaussian process-based mass distribution}
+\label{sec:model}
+Gaussian processes (GPs) are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
+Their smoothness properties allow them to be widely useful in GW data analysis and beyond for regression, such as modeling time-domain waveforms \citep{zoheyr, others} or density estimation, such as estimate posterior densities of single-event parameters from parameter estimation samples ~\citep{dangelo}, and as a prior on the bin heights of histograms describing the CBC mass distribution \citep{ray_2023}.
+Our use case is slightly different.
+We will utilize a GP as a prior on the functions that describe the primary mass distribution of CBCs.
+This choice adds very little prior information about the shape of the mass distribution, besides enforcing that it must be smooth.
+Practically, the only difference in the inference of the population when using a GP vs other modeling choices is that $p(m_1|\Lambda)$ in Equation~\ref{eq:} is determined directly by a draw from the GP, rather than by drawing a handful of hyperparameters $\Lambda$ and then evaluating them on some analytical function.
+
+In other words, when using parametric models, $p(\theta|\Lambda)$ in Equation~\ref{eq:} is calculated by evaluating a specific functional form described by a small set of hyperparameters, $\Lambda$. 
+With the GP approach, our hyperparameters are the rate at each event-level posterior sample's source frame mass value, and the rate at each found injection's source frame mass.
+The source frame mass is different for each possible value of \Ho{} and \Omm{}, so the hyperparemeters are defined in different locations for each draw from the hyperposterior.
+% can remove this sentence.
+We therefore have $N_{\text{evs}}M + N_{\text{inj}}$ hyperparameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of PE samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
+In this way, our GP-based mass distribution is similar to the autoregressive population models used in ~\citet{callister_ar_paper}.
+Indeed, an autoregressive model is a Gaussian process with a particular choice of kernel (described below).
+
+\begin{itemize}
+    \item Normally it would be bad to have more hyperparameters than datapoints, but that's where the kernel comes in!
+it greatly limits the number of thigns we actually have to fit (cite someone, can't remember where I originally saw this)
+    \item Kernel choice   
+    \item Explain priors on hyper-hyper params: smoothness assumptions, PC priors~\citep{simpson_penalising_2017} (if used)
+    \item for computational efficiency, we don't actually evaluate the GP at every data point, and instead use an interpolant.
+Future work will develop faster computational methods for evaluating GPs on each individual data point.
+    \item We also use Quasi-separable kernels (Cite DFM) within the software package \texttt{tinygp} (cite) for further computational efficiency. 
+    \item cite \texttt{tinygp}, \texttt{numpyro}, HMC/NUTS~\citep{hoffman_no-u-turn_2011}, etc.
+\end{itemize}
+\section{Results}
+\label{sec:results}
+We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated GW catalog of \result{1,000} GW events from an underlying mass distribution described by the \plp{} model and its maximum \emph{a posteriori} hyperparameters from \citet{o3b_pop}.
+
+The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
+The inferred mass distribution (left panel) closely resembles the true, simulated distribution, and the injected value of \Ho is consistent with its inferred posterior distribution (right panel).
+Additionally, the posterior on \Ho is distinct from its prior distribution, indicating that the data is informative despite the flexibility of the population model.
+This measurement represents
+
+We repeat this analysis with two parametric models: the \plp{} model, which includes the true mass distribution within its hyperprior, as well as a single power law, which does not include the true mass distribution within its hyperprior.
+The results of each of these fits is shown in Figure~\ref{fig:O5_parametric}, and their performance is summarized in Table~\ref{tab:bias}.
+When using a parametric model of the same form as the true simulated distribution (\plp), we recover the injected value of \Ho{} within \result{X-$\sigma$}.
+However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by \result{X}-$\sigma$.
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/O5_GP.pdf}
+    \caption{Two-panel figure of mass distribution and \Ho{} posterior. If we end up fitting \Omm{} we can put a corner plot instead.}
+    \label{fig:O5_GP}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics{figures/O5_parametric.pdf}
+    \caption{Same as Figure~\ref{fig:O5_GP} but for parametric models. We can probably combine these two figures.}
+    \label{fig:O5_parametric}
+\end{figure}
+
+\begin{table}[]
+    \centering
+    \begin{tabular}{c|c c}
+         Model & $\sigma_{H_0}$ [km/s/Mpc] & bias [km/s/Mpc]\\
+         \hline
+         & 
+    \end{tabular}
+    \caption{Performance of various mass distribution models used for measuring the Hubble constant, \Ho.
+    We consider the correct parametric model, an incorrect parametric model, and the non-parametric, Gaussian process-based model presented in this work.
+    Bias is defined as the number of standard deviations between the true injected value of \Ho and the recovered posterior mean. 
+    The parametric models yield more precise constraints on \Ho{}, as shown by smaller standard deviations ($\sigma_{H_0}$), but the incorrect parametric model is less accurate than the non-parametric model.
+    In reality, it is impossible to know the true functional form of the mass distribution, so the additional statistical uncertainty introduced by non-parametric approaches may be desirable over the systematic errors associated with choosing a parametric model.
+    }
+    \label{tab:bias}
+\end{table}
+
+\begin{itemize}
+    \item Can get constraints on \Ho{} to X\% by the end of O5
+    \item If possible, could do constraints on H0 as a function of \# of events.
+    \item could also plot H(z) using posterior draws from \Ho{} and \Omm{}- can see at which redshift you get the best constraints.
+Then plot posterior of $H(z=z_{\text{best measured}})$ rather than \Ho
+\end{itemize}
+\section{Discussion}
+\label{sec:discussion}
+
+\begin{itemize}
+    \item We are able to get constraints on comsological parameters with a very flexible model for the mass distribution. 
+    \item The true population of compact objects in the Universe is unknown, so we will never have access to a measurement of \Ho{} made with a model of the same form as the true mass distribution when using real data.
+Systematic uncertainties arising from an incorrect choice for the functional form of the mass distribution are therefore inevitable.
+With current detectors, these systematic effects are dwarfed by statistical uncertainties.
+However, next-generation detectors will herald in enough GW observations to substantially decrease statistical uncertainty in these measurements.
+Non-parametric approaches such as the ones presented here will aid in mitigating systematic effects.
+    \item hopefully these will be less biased than using a parametric model
+    \item Also, this shows that you don't need to know the shape of the mass distribution a priori to do spectral sirens.
+    \item This is because the information in spectral sirens comes from assuming events come from the same overall mass distribution across redshift, and that mass distribution doesn't evolve with redshift in the same exact way as would be mimicked by cosmological redshifting.
+This assumption is very basic, not an assumption that we know the exact astrophyisics of BBH formation (i.e.
+PPISN feature location)
+    \item future/in prep.
+work will do the same thing but for a mass distribution that is allowed to evolve with redshift, similarly to the parametric analysis done in~\cite{ezquiaga_spectral_2022}, but with the evolution with redshift allowed to be arbitrary so long as it does not mimic cosmology.
+- GPs are good bc they can do correlations/generalize to multiple dimensions naturally.
+Our choice of kernel allows us to do this, but its not possible with the autoregressive kernel.
+\end{itemize}
+\acknowlegements{The authors thank Reed Essick, Utkarsh Mali, Ben Farr, Maya Fishbach, ... for helpful conversations}
 
 \bibliography{bib}
+\appendix
+\section{Details of data simulation}
 

### 2023-10-06 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 Unlike light, however, the selection effects of GWs are extremely well understood, allowing for a precise estimate of each catalogs’ completeness and an unbiased measurement of the true GW population.
-This allows GW signals to be used as relatively clean probes of cosmological parameters \citep[e.g.][]{bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide}.
+This allows GW signals to be used as relatively clean probes of cosmological parameters %\citep[e.g.][]{bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide}.
 

### 2023-10-06 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2023-10-06 src/tex/ms.tex
@@
 Like light, gravitational waves (GWs) are redshifted as they propagate across the universe.
-Unlike light, however, the selection effects of GWs are extremely well understood, allowing for a precise estimate of each catalogs’ completeness and an unbiased measurement of the true GW population.
+Unlike light, however, the selection effects of GWs are extremely well understood, allowing for a precise estimate of each catalogs' completeness and an unbiased measurement of the true GW population.
 This allows GW signals to be used as relatively clean probes of cosmological parameters \citep[e.g.][]{bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide}.
 
-GW signals provide direct measurements of each source’s luminosity distance and redshifted, or detector frame, masses, $m_{\det} = m_{\source}(1+z)$.
+GW signals provide direct measurements of each source's luminosity distance and redshifted, or detector frame, masses, $m_{\det} = m_{\source}(1+z)$.
 Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the GW signal was emitted, $H(z)$.

### 2023-10-09 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 \usepackage{showyourwork}
-\usepackage{macros}
+% \usepackage{macros}
 \usepackage{amsmath}
+
+\newcommand{\Ho}{$H_0$}
+\newcommand{\Omm}{$\Omega_M$}
+\newcommand{\source}{\text{source}}
+\newcommand{\diff}{\text{d}}
+
+\newcommand{\plp}{\textsc{Power Law + Peak}}
+
+\newcommand{\result}[1]{\textcolor{red}{#1}}
+\newcommand{\comment}[1]{\textcolor{orange}{#1}}
 
@@
 
-\subsection{Spectral sirens}
+\subsection{Spectral sirens and statistical framework}
 \label{sec:ss}
-\begin{itemize}
-    \item explain the basis of spectral sirens
-    \item explain HBA with selection effects (or don't and just reference papers).
-    \item ``Therefore, another way to think about it is...
-/ Algorithmically, this is equivalent to..." explain spectral sirens in the generative way: choose one H0 draw, transform the masses using DL, use the source frame masses as inputs to the mass distribution, evaluate the likelihood of that mass distribution and that H0 draw, repeat.
-\end{itemize}
+\begin{enumerate}
+    \item Briefly explain HBA with selection effects.
+\end{enumerate}
+\begin{equation}
+    p(\Lambda|data) = p(\Lambda)\exp{-Nexp} \Pi L_i
+    \label{eq:inhomog-poisson}
+\end{equation}
+We infer this posterior using the no-u-turn sampler for Hamiltonian Monte Carlo within `numpyro` \citep{hoffman_no-u-turn_2011, numpyro}.
+
+The spectral siren method of measuring cosmological parameters relies on the assumption that all compact binaries follow the same mass distribution, regardless of redshift, or if the mass distribution evolves with redshift in any way besides a monotonic rightwards shift.
+The likelihood in Equation~\ref{eq:inhomog-poisson} is maximized when the relationship between redshift and luminosity distance makes all source frame masses to follow the same distribution\footnote{\citet{ezquiaga_spectral_2022} show that the source frame masses can follow different distributions at different redshifts, so long as this redshift evolution of the mass distribution does not exactly mimic cosmology (i.e.
+all features cannot exhibit a monotonic rightward shift with redshift).}.
+This is the basis of the spectral siren method of measuring cosmological parameters; we can simultaneously infer the relationship between redshift and luminosity distance with the parameters of the source frame mass distribution.
+The full set of hyper-parameters therefore includes the cosmological parameters that dictate the $D_L$-$z$ relation: \Ho, \Omm, $\Omega_\Lambda, \Omega_r,$ and $w$.
+In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and $w=1$ and use uniform priors on \Ho{} and \Omm{}.
+Algorithmically, this is equivalent to drawing a value of \Ho, using it to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to $m_\det (1+z(D_L)) = m_{\text{source}}$, evaluating the source-frame mass distribution at these transformed values, and then evaluating the likelihood in Equation~\ref{eq:inhomog-poisson}.
+The full process is outlined in Section~\ref{sec:model}.
 
@@
 Gaussian processes (GPs) are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
-Their smoothness properties allow them to be widely useful in GW data analysis and beyond for regression, such as modeling time-domain waveforms \citep{zoheyr, others} or density estimation, such as estimate posterior densities of single-event parameters from parameter estimation samples ~\citep{dangelo}, and as a prior on the bin heights of histograms describing the CBC mass distribution \citep{ray_2023}.
+Their smoothness properties make them widely useful in GW data analysis and beyond for regression problems, such as modeling time-domain waveforms \citep{zoheyr, huerta}, density estimation problems, such as estimating posterior densities of single-event parameters from parameter estimation samples ~\citep{dangelo}, and as a prior on histogram bin heights \citep{ray_2023,li_flexible_2021}.
 Our use case is slightly different.
@@
 This choice adds very little prior information about the shape of the mass distribution, besides enforcing that it must be smooth.
-Practically, the only difference in the inference of the population when using a GP vs other modeling choices is that $p(m_1|\Lambda)$ in Equation~\ref{eq:} is determined directly by a draw from the GP, rather than by drawing a handful of hyperparameters $\Lambda$ and then evaluating them on some analytical function.
+Practically, the only difference in the inference of the population when using a GP versus other modeling choices is that $p(m_1|\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is determined directly by a realization of the GP, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
 
-In other words, when using parametric models, $p(\theta|\Lambda)$ in Equation~\ref{eq:} is calculated by evaluating a specific functional form described by a small set of hyperparameters, $\Lambda$. 
-With the GP approach, our hyperparameters are the rate at each event-level posterior sample's source frame mass value, and the rate at each found injection's source frame mass.
-The source frame mass is different for each possible value of \Ho{} and \Omm{}, so the hyperparemeters are defined in different locations for each draw from the hyperposterior.
-% can remove this sentence.
-We therefore have $N_{\text{evs}}M + N_{\text{inj}}$ hyperparameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of PE samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
-In this way, our GP-based mass distribution is similar to the autoregressive population models used in ~\citet{callister_ar_paper}.
-Indeed, an autoregressive model is a Gaussian process with a particular choice of kernel (described below).
+In other words, when using parametric models, $p(\theta|\Lambda)$ in Equation~\ref{eq:} is calculated by evaluating a specific functional form described by a small set of hyper-parameters, $\Lambda$. 
+With the GP approach, the hyper-parameters describing the mass distribution are the rate at each event-level posterior sample's source frame mass, and the rate at each found injection's source frame mass.
+% The source frame mass is different for each possible value of \Ho{} and \Omm{}, so the hyper-parameters are defined in different locations for each draw from the hyper-posterior.
+We therefore have $N_{\text{evs}}M + N_{\text{inj}}$ mass hyper-parameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of PE samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
+We label these $\{\mathcal{R}_ij,\mathcal{R}_k\}$, and our full set of hyper-parameters is $\Lambda=\{\H_0, \Omega_M, \mathcal{R}_ij,\mathcal{R}_k\}$. 
+In this way, our GP-based mass distribution is similar to the autoregressive population models used in ~\citet{callister_ar}.
+Indeed, an autoregressive model is a Gaussian process with a particular choice of kernel.
 
-\begin{itemize}
-    \item Normally it would be bad to have more hyperparameters than datapoints, but that's where the kernel comes in!
-it greatly limits the number of thigns we actually have to fit (cite someone, can't remember where I originally saw this)
-    \item Kernel choice   
-    \item Explain priors on hyper-hyper params: smoothness assumptions, PC priors~\citep{simpson_penalising_2017} (if used)
-    \item for computational efficiency, we don't actually evaluate the GP at every data point, and instead use an interpolant.
-Future work will develop faster computational methods for evaluating GPs on each individual data point.
-    \item We also use Quasi-separable kernels (Cite DFM) within the software package \texttt{tinygp} (cite) for further computational efficiency. 
-    \item cite \texttt{tinygp}, \texttt{numpyro}, HMC/NUTS~\citep{hoffman_no-u-turn_2011}, etc.
-\end{itemize}
+A kernel is a function that defines the covariance between input points in the GP (in our case, two source frame mass values). 
+It defines the notion of similarity between adjacent points and thereby encodes our assumptions about the smoothness of the source frame mass distribution \citep{rasmussen_gaussian_2006}.
+We use a Mat\'ern kernel \citep{handcock_and_stein,stein_1999} with $\nu = 5/2$, but have repeated the analysis with $\nu=3/2$ and $\infty$, finding little impact on the results, except that the $\nu=\infty$ case (also called the ``squared exponential'' kernel) produces a slightly more jagged mass distribution.
+Mat\'ern kernels have two parameters besides the mean variance that determine their properties: a length scale $l$ and variance $s$.
+In our use case, these are one level further removed from hyper-parameters, so we adopt the terminology used in \citet{callister_ar} and call them ``hyper-hyper-parameters.''
+We fit these hyper-hyper-parameters along with the hyper-parameters $\Lambda$ to minimize prior assumptions about the form of the mass distribution.
+We use ``penalized-complexity'' priors on the hyper-hyper-parameters to enforce that the model does not create small-scale structure uninformed by data, thereby avoiding over-fitting \citep{simpson_penalising_2017,simpson_garcpas_2022}.
+% remove this if we don't use it.
+Explicitly, the priors on $l$ and $s$ have less than 5\% support for correlation lengths smaller than the \result{minimum} spacing between event-level posterior samples.
+
+GP evaluations notoriously scale as the cube of the number of data points, making them unwieldy with large data sets such as the ones expected for O5.
+We therefore make two approximations to a full GP to increase computational efficiency.
+First, for each likelihood evaluation, we evaluate a full GP on a regular grid between $0\Msun$ and $250\Msun$ and then interpolate it at each data point.
+Second, we use the quasi-separability of Mat\'ern kernels to analytically perform the transformation between covariance matrix and GP draw.
+This second step is done using the \texttt{QuasisepSolver} module \citep{foreman-mackey_fast_2017} in the \texttt{tinygp} code base, and requires data to be sort-able (i.e. one-dimensional).
+
+Algorithmically, each hyper-likelihood evaluation contains the following steps:
+\begin{enumerate}
+    \item Draw cosmological parameters \Ho{} and \Omm{} from the prior distributions described in Equation~\ref{eq:cosmo-priors}.
+    \item Transform the luminosity distances and detector frame masses of each event posterior sample and injection into redshifts and source frame masses according to the cosmology specified by step 1.
+    \item Draw hyper-hyper-parameters $l$ and $s$ from the penalized-complexity priors described above.
+    \item Draw a single GP realization with a kernel defined by $l$ and $s$. This is defined on a regular grid of source-frame masses and evaluated using the \texttt{QuasisepSolver}.
+    \item Interpolate the GP at each event posterior sample and injection source frame mass (from step 2)
+    \item Calculate the population likelihood according to Equation~\ref{eq:inhomog-poisson}.
+\end{enumerate}
+
 \section{Results}
@@
     \label{fig:O5_GP}
+    \script{nonparametric_plots.py}
 \end{figure}
@@
 \end{itemize}
-\acknowlegements{The authors thank Reed Essick, Utkarsh Mali, Ben Farr, Maya Fishbach, ... for helpful conversations}
+\begin{acknowledgments}
+    The authors thank Reed Essick, Utkarsh Mali, Ben Farr, Maya Fishbach,  for helpful conversations.
+\end{acknowledgments}
 

### 2023-10-09 src/scripts/data_generation.py
modularize data generation from inference and add dependencies in makefiles

@@ -1,0 +1,99 @@
+import gwcosmo
+import gwpop
+import paths
+Clight = gwcosmo.Clight
+from utils import inverse_transform_sample
+from mock_posteriors import gen_snr_scaled_PE
+
+import numpy as np
+from GWMockCat.vt_utils import draw_thetas, interpolate_optimal_snr_grid 
+
+# Constants
+H0_FID = 67.66
+DH_FID = (Clight/1.0e3) / H0_FID
+OM0_FID = 0.3096
+ZMAX = 10.
+ZMIN = 1e-6 # can't be zero since we will eventually take the log of it
+
+N_SOURCES = 10000
+N_SAMPLES_PER_EVENT = 100
+SIGMA_M = 0.5
+SIGMA_DL = 1000
+SNR_THRESH=8.
+NUM_INJ=N_SOURCES*50
+
+# random number generators
+np_rng = np.random.default_rng(4242)
+
+def true_vals_feature_full(rng, num_ridges=3, mmin=5, mmax=100,
+                        n_sources=N_SOURCES,zmax=ZMAX):
+    assert N_SOURCES % num_ridges == 0 , "num_ridges must be a divisor of N_SOURCES"
+    num_sources_per_ridge = int(N_SOURCES/num_ridges)
+    ridge_width = ((mmax - mmin)/num_ridges )/2
+    ridge_leftedge = np.linspace(mmin,mmax-ridge_width,num=num_ridges)
+    mass_samples = []
+    for ridge in range(num_ridges):
+        mass_samples.append(rng.uniform(
+            low=ridge_leftedge[ridge],
+            high=ridge_leftedge[ridge]+ridge_width,
+            size=num_sources_per_ridge
+            )
+        )
+    m1s_true = np.array(mass_samples).flatten()
+    zt = inverse_transform_sample(gwpop.unif_comoving_rate, [1e-3,ZMAX], rng,
+                              N=N_SOURCES, H0=H0_FID,Om0=OM0_FID)
+    m1z_true = m1s_true * (1+zt)
+    dL_true = gwcosmo.dL_approx(zt,H0_FID,OM0_FID)
+    return m1s_true, zt, m1z_true, dL_true
+
+def make_injections(rng, alpha, mmax_inj, mmin_inj, zmax_inj=ZMAX,num_inj=NUM_INJ):
[…]

### 2023-10-09 src/scripts/nonparametric_plots.py

@@ -1,0 +1,16 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import arviz as az
+
+import paths
+import gwpop
+
+# load in inference results
+# mcmc = az.InferenceData.from_netcdf(paths.data / "mcmc_nonparametric.nc4")
+# samples = mcmc.posterior
+
+# plot true underlying disribution
+# import it from gwpop
+
+plt.plot(1,2)
+plt.savefig(paths.figures / "O5_GP.pdf")

### 2023-10-09 src/scripts/parametric_inference.py

@@ -10,8 +10,6 @@ import jax
 import jax.numpy as jnp
 import numpyro
 import numpyro.distributions as dist
-from tinygp import kernels, GaussianProcess
-from tinygp.solvers import QuasisepSolver
 import arviz as az
 
 # data
@@ -40,11 +38,8 @@ def model(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,remove_low_Neff=Fal
     H0 = numpyro.sample("H0",dist.Uniform(H0_PRIOR_MIN,H0_PRIOR_MAX))
     Om0=OM0_FID
 
-    # construct GP in the source frame
-    kernel = sigma**2 * kernels.quasisep.Matern52(rho) # can change kernel type
-    gp = GaussianProcess(kernel,TEST_M1S,mean=mean,diag=0.001,
-                         solver=QuasisepSolver,assume_sorted=True)
-    log_rate_test = numpyro.sample("log_rate_test",gp.numpyro_dist())
+    # construct mass dist in the source frame
+    log_rate = 
 
     # convert event data to source frame
     z = jgwcosmo.z_at_dl_approx(dL,H0,Om0,zmin=ZMIN,zmax=ZMAX+8.)
@@ -60,11 +55,7 @@ def model(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,remove_low_Neff=Fal
     z_taper = - jnp.log(1.+(z/z_injs.min())**(-15.))
     p_z = jgwpop.unif_comoving(z,H0,Om0)
 
-    # interpolate GP
-    log_rate = numpyro.deterministic("log_rate", jnp.interp(m1source,
-                    TEST_M1S,log_rate_test, left=-jnp.inf, right=-jnp.inf)
-                    )
-    # total population is interpolated GP times redshift distribution 
+    # total population is mass dist times redshift distribution 
     # times jacobians for the transformation
     # this is the only line that is different for the hierarchical problem
     single_event_logL = jax.scipy.special.logsumexp(log_rate + jnp.log(p_z) + z_taper
@@ -73,8 +64,7 @@ def model(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,remove_low_Neff=Fal
     numpyro.factor("logp",jnp.sum(single_event_logL))
     
     # evaluate the population for injections 
[…]

### 2023-10-09 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and $w=1$ and use uniform priors on \Ho{} and \Omm{}.
-Algorithmically, this is equivalent to drawing a value of \Ho, using it to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to $m_\det (1+z(D_L)) = m_{\text{source}}$, evaluating the source-frame mass distribution at these transformed values, and then evaluating the likelihood in Equation~\ref{eq:inhomog-poisson}.
+Algorithmically, this is equivalent to drawing a value of \Ho, using it to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to $m_{\det} (1+z(D_L)) = m_{\text{source}}$, evaluating the source-frame mass distribution at these transformed values, and then evaluating the likelihood in Equation~\ref{eq:inhomog-poisson}.
 The full process is outlined in Section~\ref{sec:model}.

### 2023-10-09 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2023-10-09 src/tex/ms.tex
@@
 \usepackage{amsmath}
+\usepackage{enumerate}
 
@@
 \end{equation}
-We infer this posterior using the no-u-turn sampler for Hamiltonian Monte Carlo within `numpyro` \citep{hoffman_no-u-turn_2011, numpyro}.
+We infer this posterior using the no-u-turn sampler for Hamiltonian Monte Carlo within \texttt{numpyro} \citep{hoffman_no-u-turn_2011, numpyro}.
 
@@
 In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and $w=1$ and use uniform priors on \Ho{} and \Omm{}.
-Algorithmically, this is equivalent to drawing a value of \Ho, using it to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to $m_\det (1+z(D_L)) = m_{\text{source}}$, evaluating the source-frame mass distribution at these transformed values, and then evaluating the likelihood in Equation~\ref{eq:inhomog-poisson}.
+Algorithmically, this is equivalent to drawing a value of \Ho, using it to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to $m_{\det} (1+z(D_L)) = m_{\source}$, evaluating the source-frame mass distribution at these transformed values, and then evaluating the likelihood in Equation~\ref{eq:inhomog-poisson}.
 The full process is outlined in Section~\ref{sec:model}.
@@
 We therefore have $N_{\text{evs}}M + N_{\text{inj}}$ mass hyper-parameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of PE samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
-We label these $\{\mathcal{R}_ij,\mathcal{R}_k\}$, and our full set of hyper-parameters is $\Lambda=\{\H_0, \Omega_M, \mathcal{R}_ij,\mathcal{R}_k\}$. 
+We label these $\{\mathcal{R}_ij,\mathcal{R}_k\}$, and our full set of hyper-parameters is $\Lambda=\{H_0, \Omega_M, \mathcal{R}_ij,\mathcal{R}_k\}$. 
 In this way, our GP-based mass distribution is similar to the autoregressive population models used in ~\citet{callister_ar}.

### 2023-10-09 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-\newcommand{\Ho}{$H_0$}
-\newcommand{\Omm}{$\Omega_M$}
-\newcommand{\source}{\text{source}}
-\newcommand{\diff}{\text{d}}
+% \newcommand{\Ho}{$H_0$}
+% \newcommand{\Omm}{$\Omega_M$}
+% \newcommand{\source}{\text{source}}
+% \newcommand{\diff}{\text{d}}
+% \newcommand{\Msun}{\,\text{M}_{\odot}}
 
-\newcommand{\plp}{\textsc{Power Law + Peak}}
+% \newcommand{\plp}{\textsc{Power Law + Peak}}
 
-\newcommand{\result}[1]{\textcolor{red}{#1}}
-\newcommand{\comment}[1]{\textcolor{orange}{#1}}
+% \newcommand{\result}[1]{\textcolor{red}{#1}}
+% \newcommand{\comment}[1]{\textcolor{orange}{#1}}
 
@@
 
-GW signals provide direct measurements of each source's luminosity distance and redshifted, or detector frame, masses, $m_{\det} = m_{\source}(1+z)$.
+GW signals provide direct measurements of each source's luminosity distance and redshifted, or detector frame, masses, $m_{\det} $%$= m_{\source}(1+z)$.
 Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the GW signal was emitted, $H(z)$.

### 2023-10-09 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-GW signals provide direct measurements of each source's luminosity distance and redshifted, or detector frame, masses, $m_{\det} $%$= m_{\source}(1+z)$.
+GW signals provide direct measurements of each source's luminosity distance and redshifted, or detector frame, masses, $m_{\source} $%$= m_{\source}(1+z)$.
 Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the GW signal was emitted, $H(z)$.

### 2023-10-09 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2023-10-09 src/tex/ms.tex
@@
 \usepackage{showyourwork}
-% \usepackage{macros}
+\usepackage{macros}
 \usepackage{amsmath}
@@
 
-GW signals provide direct measurements of each source's luminosity distance and redshifted, or detector frame, masses, $m_{\det} $%$= m_{\source}(1+z)$.
+GW signals provide direct measurements of each source's luminosity distance and redshifted, or detector frame, masses, %$m_{\det}$%$= m_{\source}(1+z)$.
 Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the GW signal was emitted, $H(z)$.

### 2023-10-09 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-GW signals provide direct measurements of each source's luminosity distance and redshifted, or detector frame, masses, %$m_{\det}$%$= m_{\source}(1+z)$.
+GW signals provide direct measurements of each source's luminosity distance and redshifted, or detector frame, masses, $m_{\det}= m_{\source}(1+z)$.
 Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the GW signal was emitted, $H(z)$.

### 2023-10-09 src/scripts/nonparametric_plots.py
dbugging

@@ -1,6 +1,6 @@
 import matplotlib.pyplot as plt
 import numpy as np
-import arviz as az
+# import arviz as az
 
 import paths
 import gwpop

### 2024-01-22 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 \begin{abstract}
-    Lorem ipsum dolor sit amet, consectetuer adipiscing elit.
-    Ut purus elit, vestibulum ut, placerat ac, adipiscing vitae, felis.
-    Curabitur dictum gravida mauris, consectetuer id, vulputate a, magna.
-    Donec vehicula augue eu neque, morbi tristique senectus et netus et.
-    Mauris ut leo, cras viverra metus rhoncus sem, nulla et lectus vestibulum.
-    Phasellus eu tellus sit amet tortor gravida placerat.
-    Integer sapien est, iaculis in, pretium quis, viverra ac, nunc.
-    Praesent eget sem vel leo ultrices bibendum.
-    Aenean faucibus, morbi dolor nulla, malesuada eu, pulvinar at, mollis ac.
-    Curabitur auctor semper nulla donec varius orci eget risus.
-    Duis nibh mi, congue eu, accumsan eleifend, sagittis quis, diam.
-    Duis eget orci sit amet orci dignissim rutrum.
+    \textcolor{red}{Showyourwork Guidelines:}
+    \begin{itemize}
+        \item Always edit \texttt{ms.tex} in Overleaf
+        \item Always edit \texttt{bib.bib} in Overleaf
+        \item Always edit \texttt{macros.sty} in Overleaf
+        \item Always edit anything in the \texttt{figures} directory using git (with a python script that should generate the figures)
+        \item Always edit anything in the \texttt{output} directory using git (with a python script that should generate the output)
+    \end{itemize}
 \end{abstract}
@@
 
+\jme{[Add some numbers about expected precision (e.g.
+few percent with O5, sub-percent 3G).
+Emphasize spectral sirens allow you to measure expansion rate at high redshift - inaccesible by other means!]}
+
 \section{Motivating Example}
 \comment{Maybe this should just be part of results. Like just do this for O5 data and show how bad it can be. if its not that bad, we can do it for even more events to show when it will be a problem.}
+\jme{[I agree.
+I'd follow: 1) Intro 2) Spectral sirens (intro to the methods) 3) Non-parametric spectral sirens (new methods part) 4) Results (O5 forecast comparing non-parametric, right parametric, wrong parametric) 5) Conclusions and future prospects (discuss going to non-parametric 2D) + Appendices (any technical part regarding details on the analysis)]}
 
@@
 
-\begin{figure}
-    \centering
-    \includegraphics{figures/diff_mass_dists_example.pdf}
-    \caption{Caption}
-    \label{fig:motivation}
-\end{figure}
+% \begin{figure}
+%     \centering
+%     \includegraphics{figures/diff_mass_dists_example.pdf}
+%     \caption{Caption}
+%     \label{fig:motivation}
+% \end{figure}
 
@@
 \begin{align}
-    p(m_1,m_2|\bar{\Lambda}_m) &\propto \\
-    p(z|H_0, \Omega_M) &\propto \frac{\diff V_C}{\diff z} \frac{1}{1+z},
-\label{eq:underlying mass and redshift dist}
+    p(m_1,m_2|\bar{\Lambda}_m) &\propto ,\\
+\label{eq:underlying mass dist}
 \end{align}
-, $V_C(H_0, \Omega_M)$ is the comoving volume for given cosmological parameters \Ho{} and \Omm{}, and $\bar{\Lambda_m} = \alpha$...
+\begin{equation}
+    p(z|H_0, \Omega_M) \propto \frac{\diff V_C}{\diff z} \frac{1}{1+z},
+    \label{eq:underlying redshift dist}
+\end{equation}
+$V_C(H_0, \Omega_M)$ is the comoving volume for given cosmological parameters \Ho{} and \Omm{}, and $\bar{\Lambda_m} = \alpha$...
 are the parameters describing the power law in primary mass, ...
@@
 Would need to get relative heights correct though.}
+\jme{[For O5 BNSs will be subdominant. I would focus on BBHs for simplicity]}
 
@@
 
-\subsection{Spectral sirens and statistical framework}
+\subsection{Spectral sirens and hierarchical Bayesian Inference}
 \label{sec:ss}
@@
 \end{enumerate}
+Given GW data $\{d\}$ that contains $N_{\text{evs}}$ detected GW signals, each with parameters $\theta_i$, the posterior probability of the GW sources being drawn from an underlying population and background cosmology described by hyper-parameters $\Lambda$ is \citep{loredo, taylor, mandel}
 \begin{equation}
-    p(\Lambda|data) = p(\Lambda)\exp{-Nexp} \Pi L_i
+    p(\Lambda|d) \propto p(\Lambda) e^{-N_{\text{exp}}(\Lambda)} \prod_i^{N_{\text{evs}}} \frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda) .
     \label{eq:inhomog-poisson}
 \end{equation}
-We infer this posterior using the no-u-turn sampler for Hamiltonian Monte Carlo within \texttt{numpyro} \citep{hoffman_no-u-turn_2011, numpyro}.
+Here, $N_{\text{exp}(\Lambda)}$ is the expected number of detections given $\Lambda$ and the GW detector sensitivity, calculated using a Monte Carlo sum over $N_{\text{inj}}$ found signals injected into the data stream \citep[see][for a detailed explanation of this process]{essick}.
+$\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ is the detector frame merger rate density.
+This quantity is the population model conditioned on hyper-parameters $\Lambda$.
+Following \citet{callister_parameter_2023}, we use a semicolon to explicitly indicate that this is a function of $\Lambda$, not a density over $\Lambda$.
 
-The spectral siren method of measuring cosmological parameters relies on the assumption that all compact binaries follow the same mass distribution, regardless of redshift, or if the mass distribution evolves with redshift in any way besides a monotonic rightwards shift.
-The likelihood in Equation~\ref{eq:inhomog-poisson} is maximized when the relationship between redshift and luminosity distance makes all source frame masses to follow the same distribution\footnote{\citet{ezquiaga_spectral_2022} show that the source frame masses can follow different distributions at different redshifts, so long as this redshift evolution of the mass distribution does not exactly mimic cosmology (i.e.
+Since the parameters of each event are not perfectly measured, we marginalize $\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ over the event parameters' likelihood given the data.
+Practically, this is done by a Monte Carlo average over each event's posterior samples $\{\theta_j\}_i$, dividing out the prior used when inferring those posterior samples, $\pi_{PE}(\theta)$:
+\begin{equation}
+    \frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda) = \frac{1}{N_{\text{samples}}}\sum_j^{N_{\text{samples}}} \frac{\diff N}{\diff t_{\det} \diff \theta_{i,j}} (\theta_{i,j};\Lambda) \frac{1}{\pi_{PE}(\theta_{i,j})} .
+    \label{eq:single-event-likelihood}
+\end{equation}
+Finally, $p(\Lambda)$ is the prior on the population and parameters on the background cosmology, which we discuss in further detail below.
+We infer the posterior in Equation~\ref{eq:inhomog-poisson} using the no-u-turn sampler for Hamiltonian Monte Carlo within \texttt{numpyro} \citep{hoffman_no-u-turn_2011, numpyro}.
+
+We only consider masses and redshifts, neglecting spins as they do not contain cosmological information.
+Thus, $\theta = \{m_1,m_2,z\}$.
+We assume a uniform-in-comoving volume redshift distribution, as in Equation~\ref{eq:underlying redshift dist}.
+We assume $m_1=m_2$ and model the distribution of primary masses using a Gaussian process (described in Section~\ref{sec:model}.
+
+The spectral siren method of measuring cosmological parameters relies on the assumption that all compact binaries follow the same mass distribution regardless of redshift\footnote{\citet{ezquiaga_spectral_2022} show that the source frame masses can follow different distributions at different redshifts, so long as this redshift evolution of the mass distribution does not exactly mimic cosmology (i.e.
 all features cannot exhibit a monotonic rightward shift with redshift).}.
+The likelihood in Equation~\ref{eq:inhomog-poisson} is therefore maximized when the relationship between redshift and luminosity distance makes all source frame masses follow the same distribution.
 This is the basis of the spectral siren method of measuring cosmological parameters; we can simultaneously infer the relationship between redshift and luminosity distance with the parameters of the source frame mass distribution.
@@
 \label{sec:model}
+\jme{[Suggestion to bump up this to its own section to gain visibility - see proposed structure in previous comment]}
 Gaussian processes (GPs) are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
@@
 We will utilize a GP as a prior on the functions that describe the primary mass distribution of CBCs.
+It will therefore replace $p(\Lambda)$ in Equation~\ref{eq:inhomog-poisson}.
 This choice adds very little prior information about the shape of the mass distribution, besides enforcing that it must be smooth.
-Practically, the only difference in the inference of the population when using a GP versus other modeling choices is that $p(m_1|\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is determined directly by a realization of the GP, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
+Practically, the only difference in the inference of the population when using a GP versus other modeling choices is that $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is determined directly by a realization of the GP, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
 
-In other words, when using parametric models, $p(\theta|\Lambda)$ in Equation~\ref{eq:} is calculated by evaluating a specific functional form described by a small set of hyper-parameters, $\Lambda$. 
+In other words, when using parametric models, $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is calculated by evaluating a specific functional form described by a small set of hyper-parameters.
 With the GP approach, the hyper-parameters describing the mass distribution are the rate at each event-level posterior sample's source frame mass, and the rate at each found injection's source frame mass.
@@
 
+\jme{[advertise code is public!]}
+
 \section{Results}
@@
 
-\begin{figure}
-    \centering
-    \includegraphics{figures/O5_parametric.pdf}
-    \caption{Same as Figure~\ref{fig:O5_GP} but for parametric models. We can probably combine these two figures.}
-    \label{fig:O5_parametric}
-\end{figure}
+% \begin{figure}
+%     \centering
+%     \includegraphics{figures/O5_parametric.pdf}
+%     \caption{Same as Figure~\ref{fig:O5_GP} but for parametric models. We can probably combine these two figures.}
+%     \label{fig:O5_parametric}
+% \end{figure}
 
@@
     \item could also plot H(z) using posterior draws from \Ho{} and \Omm{}- can see at which redshift you get the best constraints.
-Then plot posterior of $H(z=z_{\text{best measured}})$ rather than \Ho
+Then plot posterior of $H(z=z_{\text{best measured}})$ rather than \Ho \jme{[I like this option!]}
 \end{itemize}
+
+\jme{[I think it would be relevant to compare the forecast of the non-parametric and the right parametric model. Do we loose constrianing power by being flexible (probably)? how much?]}
+
 \section{Discussion}
@@
 \begin{itemize}
-    \item We are able to get constraints on comsological parameters with a very flexible model for the mass distribution. 
+    \item We are able to get constraints on cosmological parameters with a very flexible model for the mass distribution. 
     \item The true population of compact objects in the Universe is unknown, so we will never have access to a measurement of \Ho{} made with a model of the same form as the true mass distribution when using real data.
@@
 Our choice of kernel allows us to do this, but its not possible with the autoregressive kernel.
+    \item in XG we will get most bang for our buck with spectral sirens
 \end{itemize}

### 2024-01-22 src/scripts/data_generation.py
generate data from PLP, add options to generate m2 and save true values

@@ -16,7 +16,13 @@ OM0_FID = 0.3096
 ZMAX = 10.
 ZMIN = 1e-6 # can't be zero since we will eventually take the log of it
 
-N_SOURCES = 10000
+TRUEVALS = dict(
+    H0=H0_FID, OM0=OM0_FID,
+    alpha=-2.7,f_peak=0.05,mmax=78.0,mmin=10.0, mu_m1=30.0,sig_m1=7.0,
+    zp=2.4,alpha_z=1.,beta_z=3.4
+)
+
+N_SOURCES = 15000
 N_SAMPLES_PER_EVENT = 100
 SIGMA_M = 0.5
 SIGMA_DL = 1000
@@ -26,31 +32,22 @@ NUM_INJ=N_SOURCES*50
 # random number generators
 np_rng = np.random.default_rng(4242)
 
-def true_vals_feature_full(rng, num_ridges=3, mmin=5, mmax=100,
-                        n_sources=N_SOURCES,zmax=ZMAX):
-    assert N_SOURCES % num_ridges == 0 , "num_ridges must be a divisor of N_SOURCES"
-    num_sources_per_ridge = int(N_SOURCES/num_ridges)
-    ridge_width = ((mmax - mmin)/num_ridges )/2
-    ridge_leftedge = np.linspace(mmin,mmax-ridge_width,num=num_ridges)
-    mass_samples = []
-    for ridge in range(num_ridges):
-        mass_samples.append(rng.uniform(
-            low=ridge_leftedge[ridge],
-            high=ridge_leftedge[ridge]+ridge_width,
-            size=num_sources_per_ridge
-            )
-        )
-    m1s_true = np.array(mass_samples).flatten()
-    zt = inverse_transform_sample(gwpop.unif_comoving_rate, [1e-3,ZMAX], rng,
-                              N=N_SOURCES, H0=H0_FID,Om0=OM0_FID)
-    m1z_true = m1s_true * (1+zt)
+def true_vals_PLP(rng, n_sources=N_SOURCES):
+    m1s_true = inverse_transform_sample(gwpop.powerlaw_peak,[1,400],rng,N=n_sources,
+                                        alpha=TRUEVALS['alpha'],f_peak=TRUEVALS['f_peak'],
+                                        mMax=TRUEVALS['mmax'],mMin=TRUEVALS['mmin'],
[…]

### 2024-01-22 src/scripts/mock_posteriors.py

@@ -20,20 +20,19 @@ def gen_jittered_PE(rng, true_vals,sigma,bounds,n_samples,n_detections=None,retu
     else:
         return mock_samples
     
-def gen_snr_scaled_PE(rng, true_m1s, true_dl, osnr_interp, reference_distance, n_samples,
-                      H0, Om0, mc_sigma=0.04,eta_sigma=0.01,snr_thresh=8):
+def gen_snr_scaled_PE(rng, true_m1s, true_m2s, true_dl, osnr_interp, reference_distance, n_samples,
+                      H0, Om0, mc_sigma=0.04,eta_sigma=0.01,snr_thresh=8, return_og=False):
     """ Wrapper function for GWMockCat functions"""
     from GWMockCat import posterior_utils, cosmo_utils
     import astropy.cosmology as ap_cosmology 
     from astropy import units as u
     import numpy as np
-    from data_generation import ZMAX
 
     uncert_gwmc = {'threshold_snr': snr_thresh, 'snr': 1.0, 'mc': mc_sigma, 'Theta': 0.15, 'eta': eta_sigma}
     cosmo = ap_cosmology.FlatLambdaCDM(H0=H0, Om0=Om0)
-    cosmo_gwmc = cosmo_utils.interp_cosmology(zmin=0.,zmax=ZMAX+1.,cosmology=cosmo,dist_unit=u.Unit(reference_distance[-3:]))
+    cosmo_gwmc = cosmo_utils.interp_cosmology(zmin=0.,zmax=11.,cosmology=cosmo,dist_unit=u.Unit(reference_distance[-3:]))
     true_z = cosmo_gwmc['z_at_dL'](true_dl)
-    observed, detected_dict = posterior_utils.generate_obs_from_true_list(m1t=true_m1s,m2t=true_m1s,
+    observed, detected_dict = posterior_utils.generate_obs_from_true_list(m1t=true_m1s,m2t=true_m2s,
                                                                                     zt=true_z,osnr_interp=osnr_interp,
                                                                                     cosmo_dict=cosmo_gwmc,rng=rng,
                                                                                     PEsamps=n_samples,uncert=uncert_gwmc,
@@ -49,7 +48,10 @@ def gen_snr_scaled_PE(rng, true_m1s, true_dl, osnr_interp, reference_distance, n
[…]

### 2024-01-22 src/scripts/gwcosmo.py
update populations and cosmology functions, delete unused functions

@@ -71,7 +71,7 @@ def diff_comoving_volume_approx(z,H0,Om0):
     dL = dL_approx(z,H0,Om0) #Mpc
     Ez_i = Ez_inv(z,Om0)
     D_H = (Clight/1e3)  / H0 #Mpc
-    return 1.0e-9 * (4.*np.pi) * np.power(dL,2) * D_H * Ez_i / np.power(1.+z,2.)
+    return 1.0e-9 * (4.*np.pi) * np.power(dL,2) * D_H * Ez_i / np.power(1.+z,2.) # Mpc**3
 
 def dLdH_approx(z,Om0):    
     return 2. * (1.+z) * (Phi(xx(0.,Om0)) - Phi(xx(z,Om0))/np.sqrt(1.+z))/np.sqrt(Om0)

### 2024-01-22 src/scripts/gwpop.py

@@ -1,21 +1,18 @@
 import numpy as np
-from utils import powerlaw, truncnorm, butterworthFilter
+from utils import powerlaw, truncnorm
 from gwcosmo import diff_comoving_volume_approx
 
 """ Mass distribution """
-maximum_mass_considered = 150
-minimum_mass_considered = 1
-
 
 def powerlaw_peak(m1,mMin,mMax,alpha,sig_m1,mu_m1,f_peak):
     tmp_min = 2.
-    tmp_max = 150.
+    tmp_max = 400.
     dmMax = 2
     dmMin = 1
 
     # Define power-law and peak
     p_m1_pl = powerlaw(m1,low=tmp_min,high=tmp_max,alpha=alpha)
-    p_m1_peak = np.exp(-(m1-mu_m1)**2/(2.*sig_m1**2))/np.sqrt(2.*np.pi*sig_m1**2)
+    p_m1_peak = truncnorm(m1, mu=mu_m1, sigma=sig_m1, high=tmp_max, low=tmp_min) #
 
     # Compute low- and high-mass filters
     low_filter = np.exp(-(m1-mMin)**2/(2.*dmMin**2))
@@ -42,20 +39,6 @@ def powerlaw_smooth(m,mMin,mMax,alpha):
 
     return pm_pl*low_filter*high_filter
 
-def pl_peak(m1,alpha,mu_m1,sig_m1,f_peak,mMax,mMin,low_smoothing, high_smoothing):
-    tmp_min = minimum_mass_considered
-    tmp_max = maximum_mass_considered
-    # Define power-law and peak
-    p_m1_pl = powerlaw(m1, alpha=alpha, high=tmp_max, low=tmp_min)
-    p_m1_peak = truncnorm(m1, mu=mu_m1, sigma=sig_m1, high=tmp_max, low=tmp_min)
-
-    # Compute low- and high-mass filters
-    low_filter = butterworthFilter(m1,mMin,-1*low_smoothing)
-    high_filter =  butterworthFilter(m1,mMax,high_smoothing)
-
-    # Apply filters to combined power-law and peak
-    return (f_peak*p_m1_peak + (1.-f_peak)*p_m1_pl)*low_filter*high_filter
-
 """ Redshift distribution """
 maximum_redshift_considered = 10
 minimum_redshift_considered = 0
[…]

### 2024-01-22 src/scripts/jgwcosmo.py

@@ -1,7 +1,5 @@
-#import numpyro
 import jax.numpy as jnp
 from jax import jit
-#from constants import *
 Clight = 3e8
 
 def Ez_inv(z,Om0):
@@ -44,13 +42,6 @@ def dDLdz(z, H0, Om0):
     D_H = (Clight/1e3) / H0 #Mpc
     return jnp.abs(dL/(1.+z) + (1.+z)*D_H * Ez_i)
 
-@jit
-def dDLdz_approx(z, H0, Om0):
-    dL = dL_approx(z,H0,Om0)#Mpc
-    Ez_i = Ez_inv(z,Om0)
-    D_H = (Clight/1e3) / H0 #Mpc
-    return jnp.abs(dL/(1.+z) + (1.+z)*D_H * Ez_i)
-
 
 """Approximate luminosity distance"""
 #https://arxiv.org/pdf/1111.6396.pdf
@@ -67,6 +58,13 @@ def xx(z,Om0):
 def dL_approx(z,H0,Om0):
     D_H = (Clight/1.0e3)  / H0 #Mpc
     return 2.*D_H * (1.+z) * (Phi(xx(0.,Om0)) - Phi(xx(z,Om0))/jnp.sqrt(1.+z))/jnp.sqrt(Om0)
+
+@jit
+def dDLdz_approx(z, H0, Om0):
+    dL = dL_approx(z,H0,Om0)#Mpc
+    Ez_i = Ez_inv(z,Om0)
+    D_H = (Clight/1e3) / H0 #Mpc
+    return jnp.abs(dL/(1.+z) + (1.+z)*D_H * Ez_i)
 
 @jit
 def z_at_dl_approx(dl,H0,Om0,zmin=1e-3,zmax=100):
@@ -86,6 +84,13 @@ def diff_comoving_volume_approx(z,H0,Om0):
     Ez_i = Ez_inv(z,Om0)
     D_H = (Clight/1e3)  / H0 #Mpc
     return 1.0e-9 * (4.*jnp.pi) * jnp.power(dL,2) * D_H * Ez_i / jnp.power(1.+z,2.)
+
+@jit
+def log_diff_comoving_volume_approx(z,H0,Om0):
+    dL = dL_approx(z,H0,Om0) #Mpc
+    Ez_i = Ez_inv(z,Om0)
+    D_H = (Clight/1e3)  / H0 #Mpc
[…]

### 2024-01-22 src/scripts/jgwpop.py

@@ -1,12 +1,8 @@
 import jax.numpy as jnp
-from jutils import powerlaw, truncnorm, butterworthFilter
-from jgwcosmo import diff_comoving_volume_approx
+from jutils import powerlaw, truncnorm
+from jgwcosmo import diff_comoving_volume_approx, log_diff_comoving_volume_approx
 
 """ Mass distribution """
-maximum_mass_considered = 150
-minimum_mass_considered = 1
-
-
 def powerlaw_peak(m1,mMin,mMax,alpha,sig_m1,mu_m1,f_peak):
     tmp_min = 2.
     tmp_max = 150.
@@ -42,23 +38,49 @@ def powerlaw_smooth(m,mMin,mMax,alpha):
 
     return pm_pl*low_filter*high_filter
 
-def pl_peak(m1,alpha,mu_m1,sig_m1,f_peak,mMax,mMin,low_smoothing, high_smoothing):
-    tmp_min = minimum_mass_considered
-    tmp_max = maximum_mass_considered
+def logpowerlaw(m,mMin,mMax,alpha):
+    lognorm = jnp.log(1. + alpha) - jnp.log(mMax**(alpha+1.) - mMin**(alpha+1.))
+    return lognorm + alpha * jnp.log(m)
+
+def logbroken_powerlaw(m1,mMin,mMax,alpha1,alpha2,b_m1):
+    dmMax = 2
+    dmMin = 1
+    
     # Define power-law and peak
-    p_m1_pl = powerlaw(m1, alpha=alpha, high=tmp_max, low=tmp_min)
-    p_m1_peak = truncnorm(m1, mu=mu_m1, sigma=sig_m1, high=tmp_max, low=tmp_min)
+    # p_m1_pl1 = powerlaw(m1/b_m1,low=tmp_min,high=b_m1,alpha=alpha1)
+    # p_m1_pl2 = powerlaw(m1/b_m1,low=b_m1,high=tmp_max,alpha=alpha2)
+    # logpl=jnp.log(jnp.where(m1<b_m1,p_m1_pl1,p_m1_pl2))
+    logpl = jnp.where(m1<b_m1,alpha1*(jnp.log(m1)-jnp.log(b_m1)), alpha2*(jnp.log(m1)-jnp.log(b_m1)))
 
     # Compute low- and high-mass filters
-    low_filter = butterworthFilter(m1,mMin,-1*low_smoothing)
[…]

### 2024-01-22 src/scripts/utils.py

@@ -3,10 +3,6 @@ import numpy as np
 from scipy.special import erf
 from scipy.integrate import cumtrapz
 
-def butterworthFilter(x,loc,smooth):
-    # if smooth <0, its a low-pass. if smooth<0, its a high-pass
-    return 1./(1. + (x/loc)**smooth)
-
 def powerlaw(xx, alpha, high, low):
     r"""
     Normalized truncated power-law probability
@@ -94,24 +90,3 @@ def inverse_transform_sample(pdf, x_limits, rng, N=1000, **pdf_kwargs):
     samples = np.interp(y, cdf, xs)
 
     return samples
-
-def contour_plot(samples,theory,grid_m,grid_z, lognorm=True,
-                 contour_kwargs=dict(colors='k')):
-    import matplotlib.pyplot as plt
-    from scipy.stats import gaussian_kde
-    from matplotlib.colors import LogNorm, Normalize
-
-    if lognorm:
-        n = LogNorm()
-    else:
-        n = Normalize()
-
-    fig, ax = plt.subplots(figsize=(5,5))
-#     k = gaussian_kde(samples)
-#     positions = np.vstack([grid_m.ravel(), grid_z.ravel()])
-#     Z = np.reshape(k(positions).T, grid_m.shape)
-#     ax.contourf(grid_m,grid_z,Z,norm=n)
-    ax.hexbin(samples[0],samples[1],norm=n)
-    CS = ax.contour(grid_m, grid_z, theory,norm=n, **contour_kwargs)
-
-    return ax

### 2024-01-22 src/scripts/nonparametric_plots.py
add true distribution line, clean up code

@@ -1,39 +1,45 @@
 import matplotlib.pyplot as plt
 import numpy as np
 import arviz as az
+from scipy.stats import gaussian_kde
 
 from nonparametric_inference import TEST_M1S
 from data_generation import H0_FID
+from gwpop import powerlaw_peak
+
 import paths
 
 plt.style.use(paths.scripts / "matplotlibrc")
 
-# load in inference results
-mcmc = az.InferenceData.from_netcdf(paths.data / "mcmc_nonparametric.nc4")
-samples = mcmc.posterior
-r = np.exp(samples['log_rate'][0])
-rt = np.exp(samples['log_rate_test'][0])
+def two_panel(path):
+    id = az.InferenceData.from_netcdf(path)
+    samples = id.posterior
+    try:
+        r = np.nan_to_num(np.exp(samples['log_rate_test'][0]))
+    except KeyError :
+        r = np.nan_to_num(np.exp(samples['log_rate'][0]))
 
-# plot them
-fig, axes = plt.subplots(ncols=2,figsize=(6.5,3))
-for i in range(200):
-    gp_norm = np.trapz(rt[i],TEST_M1S)
-    axes[0].plot(TEST_M1S, rt[i]/gp_norm, lw=0.2, c='b',alpha=0.1)
+    fig, axes = plt.subplots(ncols=2,figsize=(20,8),facecolor='none',gridspec_kw={'width_ratios': [1.2, 1]})
+    for i in range(200):
+        PLP_norm = np.trapz(r[i],TEST_M1S)*1.15
+        axes[0].plot(TEST_M1S, r[i]/PLP_norm, lw=0.2, c="blue",alpha=0.1)
+        
+    axes[0].plot(TEST_M1S, 7*powerlaw_peak(TEST_M1S,alpha=-2.7,f_peak=0.05,mMax=78.0,mMin=10.0,mu_m1=30.0,sig_m1=7.0),
+                c='k',lw=3,label='Underlying distribution')
+    axes[0].set_yscale('log')
+    axes[0].set_ylim(1e-5,1e-1)
+    axes[0].set_xlim(1,80)
 
-axes[0].plot([],linewidth=0.5,alpha=0.5, label='Gaussian process fit')
+    prior=np.linspace(55,120,num=200)
+    ho_kde=gaussian_kde(samples['H0'])
+    axes[1].plot(prior,ho_kde(prior))
 
-# plot true underlying disribution
[…]

### 2024-01-22 src/scripts/nonparametric_inference.py
fewer samples to make testing faster

@@ -108,7 +108,7 @@ if  __name__ == "__main__":
     kwargs = dict(m1det=m1z_PE,dL=dL_PE, m1det_inj=m1zinj_det,dL_inj=dLinj_det,
                     log_pinj=log_pinj_det, log_PE_prior=log_PE_prior,
                     remove_low_Neff=remove_low_Neff)
-    mcmc = numpyro.infer.MCMC(nuts_kernel,num_warmup=100,num_samples=200,
+    mcmc = numpyro.infer.MCMC(nuts_kernel,num_warmup=10,num_samples=20,
                               num_chains=1,progress_bar=True)   
     mcmc.run(jax_rng,**kwargs)
 

### 2024-02-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-% \newcommand{\Ho}{$H_0$}
-% \newcommand{\Omm}{$\Omega_M$}
-% \newcommand{\source}{\text{source}}
-% \newcommand{\diff}{\text{d}}
-% \newcommand{\Msun}{\,\text{M}_{\odot}}
-
-% \newcommand{\plp}{\textsc{Power Law + Peak}}
-
-% \newcommand{\result}[1]{\textcolor{red}{#1}}
-% \newcommand{\comment}[1]{\textcolor{orange}{#1}}
-
 % Begin!
@@
 % Title
-\title{Nonparametric Spectral Siren Cosmology}
+\title{No need to know: astrophysics-agnostic spectral siren cosmology}
+%\title{No need to know: gravitational-wave cosmology with an astrophysics-agnostic mass distribution}
+%\title{No need to know: spectral siren cosmology with an astrophysics-agnostic mass distribution of gravitational-wave sources}
+%\title{No need to know: gravitational-wave cosmology while astrophysically ignorant/for the astrophysically-ignorant}
 
 % Author list
-\author{@afarah18}
+%\author{@afarah18}
+\author{Amanda M. Farah}
+\email{afarah@uchicago.edu}
+\affiliation{Department of Physics, University of Chicago, Chicago, IL 60637, USA}
 
-% Abstract with filler text
+\author{Thomas A. Callister}
+\affiliation{Kavli Institute for Cosmological Physics, The University of Chicago, Chicago, IL 60637, USA}
+
+\author{Jose Mar\'ia Ezquiaga}
+%\email{jose.ezquiaga@nbi.ku.dk}
+\affiliation{Niels Bohr International Academy, Niels Bohr Institute, Blegdamsvej 17, DK-2100 Copenhagen, Denmark}
+
+\author{Michael Zevin}
+\affiliation{The Adler Planetarium, 1300 South DuSable Lake Shore Drive, Chicago, 60605, IL, USA}
+
+\author{Daniel E. Holz}
+\affiliation{Kavli Institute for Cosmological Physics and Enrico Fermi Institute, The University of Chicago, Chicago, IL 60637, USA}
+\affiliation{Department of Physics, Department of Astronomy \& Astrophysics, The University of Chicago, Chicago, IL 60637, USA}
+
+% Abstract
 \begin{abstract}
+Gravitational waves (GWs) are unique transients in which they encode direct information about their luminosity distance. 
+Adding redshift, a GW standard siren Hubble diagram can be constructed to probe the standard cosmological model.  
+Although individual signals do not have any reference scale imprinting a redshift timestamp, as a collective there could be signatures in their mass spectrum that would break the mass-redshift degeneracy.
+This spectral siren method thus provides a GW data-only approach to constrain the Universe's expansion rate from the population of compact binary coalescences. 
+So far, this technique has only been applied with simple parametric representations of the mass distribution. 
+However, the use of a wrong mass model would inevitably lead to biases in the cosmological inference, an acute problem given our current lack of understanding of the true astrophysical model.
+We demonstrate that the spectral siren method is purely data-driven, in which the right cosmology can be inferred without a given model, using a flexible Gaussian process reconstruction of the mass spectrum.
+In particular, we analyze an observing scenario with XX GW signals and detectors at XX sensitivity. 
+We find that both the source mass model and the cosmological parameters are correctly reconstructed. 
+This astrophysics-agnostic spectral siren technique will be key to deliver precise and unbiased cosmological constraints with future observations in the presence of possible redshift evolution of the mass spectrum.
+\end{abstract}
+
+\section{}
 \textcolor{red}{Showyourwork Guidelines:}
@@
 \end{itemize}
-\end{abstract}
 
@@
 
-In this work, we explicitly demonstrate that no \emph{a priori} knowledge about the shape of the CBC mass spectrum is necessary to use the spectral siren methodology by inferring \Ho with a non-parametric model for the mass distribution of CBCs.
-This flexible model makes minimal prior assumptions for the shape of the mass distribution and is therefore capable of accurately inferring a wide range of mass spectrum morphologies.
-Despite its flexibility, it is able to consistently obtaining an unbiased measurement for \Ho, showing that nonparametric methods are not only sufficient for a spectral siren measurement, they can also mitigate systematic effects in the measurement caused by model misspecification.
+In this work, we explicitly demonstrate that no \emph{a priori} knowledge about the shape of the CBC mass spectrum is necessary to use the spectral siren methodology. 
+We do this by inferring \Ho with a flexible, non-parametric model for the mass distribution of CBCs. 
+This model makes minimal prior assumptions about the shape of the mass distribution, enabling it to accurately infer a wide range of morphologies.
+Despite its flexibility, it is able to consistently obtain an unbiased measurement for \Ho, showing that nonparametric methods are not only sufficient for a spectral siren measurement, they can also mitigate systematic effects in the measurement caused by model misspecification.
 
-\jme{[Add some numbers about expected precision (e.g.
-few percent with O5, sub-percent 3G).
-Emphasize spectral sirens allow you to measure expansion rate at high redshift - inaccesible by other means!]}
+We find that using this model allows for a \result{X}\% measurement of \Ho by the end of the LVK's fifth observing run, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
+While constraints on \Ho are used in this work to quantify the accuracy and precision of our non-parametric method, we emphasize that spectral sirens are able to constrain $H(z)$ out large redshifts that are inaccessible by electromagnetic observations, especially with next-generation detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn}\footnote{See example calculations here: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}}.
 
-\section{Motivating Example}
-\comment{Maybe this should just be part of results. Like just do this for O5 data and show how bad it can be. if its not that bad, we can do it for even more events to show when it will be a problem.}
-\jme{[I agree.
-I'd follow: 1) Intro 2) Spectral sirens (intro to the methods) 3) Non-parametric spectral sirens (new methods part) 4) Results (O5 forecast comparing non-parametric, right parametric, wrong parametric) 5) Conclusions and future prospects (discuss going to non-parametric 2D) + Appendices (any technical part regarding details on the analysis)]}
+\jme{[cite joint multi-siren forecast \cite{Chen:2024gdn}. Cite code \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}?]}\\ 
+\comment{[Should I add a separate paragraph for other sirens or is it sufficient to just cite it above with other forecasts, as I did?]}
 
-In this Section, we show that fitting an incorrect functional form to the mass distribution of CBCs biases the inference of cosmological parameters when using the spectral siren methodology.
-To demonstrate this, we generate a catalog of \result{1,000} GW events from an underlying mass distribution described by the \plp{} model and its maximum \emph{a posteriori} hyperparameters from \citet{o3b_pop}.
-We then use this catalog to infer \Ho{} using three different models for the source-frame mass distribution:
-\begin{enumerate}
-    \item the \plp{} model, which includes the true mass distribution within its hyperprior
-    \item a single power law, which does not include the true mass distribution within its hyperprior, and 
-    \item the flexible, Gaussian process-based model that we use for the remainder of this work, which can closely approximate the morphology of the true mass distribution, along with several other morphologies.
-\end{enumerate}
-The results of each of these fits are shown in Figure~\ref{fig:motivation}. 
-The \result{top/left} panel shows the inferred source frame mass distribution for each of the considered models, and the \result{bottom/right} panel shows the corresponding posteriors on \Ho{}.
-
-With \result{1,000} events, we find that using the incorrect parametric form of the mass distribution (a power law) causes a biased inferred value of \Ho{} at the \result{X}-$\sigma$ level, whereas using the correct functional form (\plp) provides an unbiased inference of \Ho.
-However, this is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the Gaussian process-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution.
-The Gaussian process-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters.
-We have repeated this analysis with \result{50} separate simulated catalogs, and found that the power law model causes us to infer a \result{higher/lower} value of \Ho{} at greater than \result{X}-$\sigma$, \result{Y}\% of the time, whereas the Gaussian process-based model reaches that level of bias only \result{Z}\% of the time, and the \plp{} model reaches the same level of bias \result{A}\% of the time.
-This indicates that prior knowledge of the shape of the mass distribution is not required to perform a spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.
-
-% \begin{figure}
-%     \centering
-%     \includegraphics{figures/diff_mass_dists_example.pdf}
-%     \caption{Caption}
-%     \label{fig:motivation}
-% \end{figure}
-
-\section{Methods}
-\label{sec:methods}
-\subsection{Data simulation}
-We generate a mock catalog of GW events characteristic of that expected from the LVK's fifth observing run (O5). 
-These events are drawn from an underlying population described by
-\begin{equation}
-\label{eq:underlying population}
-    \frac{\diff N}{\diff m_1 \diff m_2 \diff z} \propto p(m_1, m_2|\bar{\Lambda}_m) p(z|H_0, \Omega_M),
-\end{equation}
-where 
-\begin{align}
-    p(m_1,m_2|\bar{\Lambda}_m) &\propto ,\\
-\label{eq:underlying mass dist}
-\end{align}
-\begin{equation}
-    p(z|H_0, \Omega_M) \propto \frac{\diff V_C}{\diff z} \frac{1}{1+z},
-    \label{eq:underlying redshift dist}
-\end{equation}
-$V_C(H_0, \Omega_M)$ is the comoving volume for given cosmological parameters \Ho{} and \Omm{}, and $\bar{\Lambda_m} = \alpha$...
-are the parameters describing the power law in primary mass, ...
-respectively.
-When generating the simulated events, we have fixed $\alpha= $... and used cosmological parameters \Ho$=68.7$, \Omm$=0.3$, $\Omega_\Lambda=1-\Omega_M$.
-These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \_ model \cite{o3b_pop}, which is morphologically similar to that described in Equations~\ref{eq:underlying population}--\ref{eq:underlying mass and redshift dist}.
-We neglect spins entirely, as they do not carry cosmological information.
-\comment{TODO: fill in details of mass distribuiton.
-Should we use PDB and include BNSs, which will give us a nice sharp feature to work with, or should we use PLP which has a bump?
-Could also do PLP and then put a uniform dist or power law for NS masses at the LMG, and leave an empty mass gap.
-Would need to get relative heights correct though.}
-\jme{[For O5 BNSs will be subdominant. I would focus on BBHs for simplicity]}
-
-After passing the simulated events through projected detector selection effects, the resulting catalog has \result{X} events, including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} BBH systems, consistent with the numbers projected for O5 by \citet{kiendrebogo_observing_2023}.
-We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from O5 detectors.
-\texttt{GWMockCat} also simulates a set of software injections, which we use to estimate selection effects in the inference.
-To determine the detectability of both injections and simulated events in O5, we use the projected O5 LIGO power spectral density \citep{dcc page} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
-The full procedure for this mock data generation process is described in \citet{fishbach_where, farah_things_2023, essick_dagnabbit_2023}, and the exact settings used for \texttt{GWMockCat} are made available in the accompanying data release.
-\comment{TODO:put a showyourwork link here.}
-
-\subsection{Spectral sirens and hierarchical Bayesian Inference}
+\section{Methods I: Spectral Siren Cosmology}
 \label{sec:ss}
@@
 
-\subsection{Gaussian process-based mass distribution}
+\section{Methods II: Gaussian process-based mass distribution}
 \label{sec:model}
-\jme{[Suggestion to bump up this to its own section to gain visibility - see proposed structure in previous comment]}
 Gaussian processes (GPs) are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
@@
 \label{sec:results}
-We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated GW catalog of \result{1,000} GW events from an underlying mass distribution described by the \plp{} model and its maximum \emph{a posteriori} hyperparameters from \citet{o3b_pop}.
+
+\subsection{Option 1: include parametric models}
+In this Section, we first show that fitting an incorrect functional form to the mass distribution of CBCs biases the inference of cosmological parameters when using the spectral siren methodology.
+We then demonstrate that our flexible model alleviates this bias without the need to know the mass distribution's morphology \emph{a priori}.
+To illustrate this, we generate a catalog of \result{1,000} GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
+This simulated catalog is designed to be representative of the gravitational wave transient catalog after the \ac{LVK}'s fifth observing run.
+The details of the data simulation are described in Appendix~\ref{ap:data generation}.
+We then use this catalog to infer \Ho{} using three different models for the source-frame mass distribution:
+\begin{enumerate}
+    \item the \plp{} model, which includes the true mass distribution within its hyperprior
+    \item a broken power law, which does not include the true mass distribution within its hyperprior, and 
+    \item the flexible, Gaussian process-based model described in Section~\ref{sec:model}, which can closely approximate the morphology of the true mass distribution, along with several other morphologies.
+\end{enumerate}
+The results of each of these fits are shown in Figure~\ref{fig:O5_GP}. 
+The \result{left} panel shows the inferred source frame mass distribution for each of the considered models, and the \result{right} panel shows the corresponding posteriors on \Ho{}.
+
+With \result{1,000} events, we find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within \result{X-$\sigma$}.
+However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by \result{X}-$\sigma$.
+This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the Gaussian process-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution.
+The Gaussian process-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters. 
+In reality, it is impossible to know the true functional form of the mass distribution, so the additional statistical uncertainty introduced by non-parametric approaches may be desirable over the systematic errors associated with choosing a parametric model.
+
+We have repeated this analysis with \result{50} separate simulated catalogs, and found that the power law model causes us to infer a \result{higher/lower} value of \Ho{} at greater than \result{X}-$\sigma$, \result{Y}\% of the time, whereas the Gaussian process-based model reaches that level of bias only \result{Z}\% of the time, and the \plp{} model reaches the same level of bias \result{A}\% of the time.
+This indicates that prior knowledge of the shape of the mass distribution is not required to perform a spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.
+
+\jme{[I think it would be relevant to compare the forecast of the non-parametric and the right parametric model.
+Do we loose constrianing power by being flexible (probably)?
+how much?]} \comment{[I have now included this in the discussion.]}
+
+\subsection{Option 2: GP only}
+We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated GW catalog of \result{1,000} GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
+This simulated catalog is designed to be representative of the gravitational wave transient catalog after the \ac{LVK}'s fifth observing run.
+The details of the data simulation are described in Appendix~\ref{ap:data generation}.
 
 The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
-The inferred mass distribution (left panel) closely resembles the true, simulated distribution, and the injected value of \Ho is consistent with its inferred posterior distribution (right panel).
+The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\result{67.66 km/s/Mpc}) is consistent with its inferred posterior distribution (\result{X}).
 Additionally, the posterior on \Ho is distinct from its prior distribution, indicating that the data is informative despite the flexibility of the population model.
-This measurement represents
 
-We repeat this analysis with two parametric models: the \plp{} model, which includes the true mass distribution within its hyperprior, as well as a single power law, which does not include the true mass distribution within its hyperprior.
-The results of each of these fits is shown in Figure~\ref{fig:O5_parametric}, and their performance is summarized in Table~\ref{tab:bias}.
-When using a parametric model of the same form as the true simulated distribution (\plp), we recover the injected value of \Ho{} within \result{X-$\sigma$}.
-However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by \result{X}-$\sigma$.
+This result also implies that we can expect to measure \Ho{} to \result{X\%} by the end of the \ac{LVK}'s fifth observing run using this method.
+
+\begin{itemize}
+    \item If possible, could do constraints on H0 as a function of \# of events.
+    \item could also plot H(z) using posterior draws from \Ho{} and \Omm{}- can see at which redshift you get the best constraints.
+Then plot posterior of $H(z=z_{\text{best measured}})$ rather than \Ho \jme{[I like this option!]}
+\end{itemize}
 
@@
     \centering
-    \includegraphics{figures/O5_GP.pdf}
-    \caption{Two-panel figure of mass distribution and \Ho{} posterior. If we end up fitting \Omm{} we can put a corner plot instead.}
+    \includegraphics[width=\textwidth]{figures/O5_GP.pdf}
+    \caption{Two-panel figure of mass distribution and \Ho{} posterior. 
+    Include prior on \Ho{} as alpha=0.5 line. 
+    If we end up fitting \Omm{} we can put a corner plot instead.}
     \label{fig:O5_GP}
-    \script{nonparametric_plots.py}
+    \script{nonparametric_twopanel.py}
 \end{figure}
 
-% \begin{figure}
-%     \centering
-%     \includegraphics{figures/O5_parametric.pdf}
-%     \caption{Same as Figure~\ref{fig:O5_GP} but for parametric models. We can probably combine these two figures.}
-%     \label{fig:O5_parametric}
-% \end{figure}
+\section{Discussion}
+\label{sec:discussion}
+
+\begin{itemize}
+    \item We are able to get constraints on cosmological parameters with a very flexible model for the mass distribution. 
+    \item The success of the flexible model in recovering an informative and accurate posterior on \Ho demonstrates that there is no need to know the CBC mass distribution's morphology in order to constrain cosmological parameters with it.
+    \item This is because the information in spectral sirens comes from assuming events come from the same overall mass distribution across redshift, and that mass distribution doesn't evolve with redshift in the same exact way as would be mimicked by cosmological redshifting.
+This assumption is very basic, not an assumption that we know the exact astrophyisics of BBH formation (i.e.
+PPISN feature location).
+    \item We have also predicted a \result{X\%} measurement of \Ho{} by the end of the \ac{LVK}'s fifth observing run using non-parametric spectral sirens.
+A similar study with parametric methods \citep{Chen:2024gdn} predicts an \result{X\%} measurement, demonstrating a \result{negligible/substantial/marginal} increase in precision.
+However, the true population of compact objects in the Universe is unknown, so it is not possible to know the true functional form of the mass distribution, so systematic uncertainties arising from an incorrect choice for the functional form of the mass distribution are therefore inevitable.
+With current detectors, these systematic effects are dwarfed by statistical uncertainties.
+However, next-generation detectors will herald in enough GW observations to substantially decrease statistical uncertainty in these measurements: \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to \result{$\sim3\sigma$} systematic biases in \Ho when catalogs contain 2,000 events.
+Thus, the additional statistical uncertainty introduced by non-parametric approaches may be preferable to the systematic errors associated with choosing a parametric model.
+    \item future/in prep.
+work will do the same thing but for a mass distribution that is allowed to evolve with redshift, similarly to the parametric analysis done in~\cite{ezquiaga_spectral_2022}, but with the evolution with redshift allowed to be arbitrary so long as it does not mimic cosmology.
+- GPs are good bc they can do correlations/generalize to multiple dimensions naturally.
+Our choice of kernel allows us to do this, but its not possible with the autoregressive kernel.
+    \item in XG we will get most bang for our buck with spectral sirens
+\end{itemize}
+\begin{acknowledgments}
+    The authors thank Reed Essick, Utkarsh Mali, Ben Farr, Maya Fishbach,  for helpful conversations.
+\end{acknowledgments}
+
+\bibliography{bib}
+\appendix
+\section{Details of data simulation}
+\label{ap:data generation}
+
+We generate a mock catalog of GW events characteristic of that expected from the LVK's fifth observing run (O5). 
+These events are drawn from an underlying population described by
+\begin{equation}
+\label{eq:underlying population}
+    \frac{\diff N}{\diff m_1 \diff m_2 \diff z} \propto p(m_1, m_2|\bar{\Lambda}_m) p(z|H_0, \Omega_M),
+\end{equation}
+where 
+\begin{align}
+    p(m_1,m_2|\bar{\Lambda}_m) &\propto ,\\
+\label{eq:underlying mass dist}
+\end{align}
+\begin{equation}
+    p(z|H_0, \Omega_M) \propto \frac{\diff V_C}{\diff z} \frac{1}{1+z},
+    \label{eq:underlying redshift dist}
+\end{equation}
+$V_C(H_0, \Omega_M)$ is the comoving volume for given cosmological parameters \Ho{} and \Omm{}, and $\bar{\Lambda_m} = \alpha$...
+are the parameters describing the power law in primary mass, ...
+respectively.
+When generating the simulated events, we have fixed $\alpha= $... and used cosmological parameters \Ho$=68.7$, \Omm$=0.3$, $\Omega_\Lambda=1-\Omega_M$.
+These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \_ model \cite{o3b_pop}, which is morphologically similar to that described in Equations~\ref{eq:underlying population}--\ref{eq:underlying mass and redshift dist}.
+We neglect spins entirely, as they do not carry cosmological information.
+\comment{TODO: fill in details of mass distribuiton.
+Should we use PDB and include BNSs, which will give us a nice sharp feature to work with, or should we use PLP which has a bump?
+Could also do PLP and then put a uniform dist or power law for NS masses at the LMG, and leave an empty mass gap.
+Would need to get relative heights correct though.}
+\jme{[For O5 BNSs will be subdominant. I would focus on BBHs for simplicity]}
+\comment{[Ok, so we should say that although a putuative lower mass gap would give and additional feature, it is not expected to be constrained well in O5, so we focus on the BBH spectrum for now.
+In XG, though, it will be important to consider the full spectrum \citep{ezquiaga_spectral_2022}.]}
+
+After passing the simulated events through projected detector selection effects, the resulting catalog has \result{X} events, including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} BBH systems, consistent with the numbers projected for O5 by \citet{kiendrebogo_observing_2023}.
+We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from O5 detectors.
+\texttt{GWMockCat} also simulates a set of software injections, which we use to estimate selection effects in the inference.
+To determine the detectability of both injections and simulated events in O5, we use the projected O5 LIGO power spectral density \citep{dcc page} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
+The full procedure for this mock data generation process is described in \citet{fishbach_where, farah_things_2023, essick_dagnabbit_2023}, and the exact settings used for \texttt{GWMockCat} are made available in the accompanying data release.
+\comment{TODO:put a showyourwork link here.}
+
+\section{Old}
+Old tex that we may end up needing.
 
@@
 
-\begin{itemize}
-    \item Can get constraints on \Ho{} to X\% by the end of O5
-    \item If possible, could do constraints on H0 as a function of \# of events.
-    \item could also plot H(z) using posterior draws from \Ho{} and \Omm{}- can see at which redshift you get the best constraints.
-Then plot posterior of $H(z=z_{\text{best measured}})$ rather than \Ho \jme{[I like this option!]}
-\end{itemize}
-
-\jme{[I think it would be relevant to compare the forecast of the non-parametric and the right parametric model. Do we loose constrianing power by being flexible (probably)? how much?]}
-
-\section{Discussion}
-\label{sec:discussion}
-
-\begin{itemize}
-    \item We are able to get constraints on cosmological parameters with a very flexible model for the mass distribution. 
-    \item The true population of compact objects in the Universe is unknown, so we will never have access to a measurement of \Ho{} made with a model of the same form as the true mass distribution when using real data.
-Systematic uncertainties arising from an incorrect choice for the functional form of the mass distribution are therefore inevitable.
-With current detectors, these systematic effects are dwarfed by statistical uncertainties.
-However, next-generation detectors will herald in enough GW observations to substantially decrease statistical uncertainty in these measurements.
-Non-parametric approaches such as the ones presented here will aid in mitigating systematic effects.
-    \item hopefully these will be less biased than using a parametric model
-    \item Also, this shows that you don't need to know the shape of the mass distribution a priori to do spectral sirens.
-    \item This is because the information in spectral sirens comes from assuming events come from the same overall mass distribution across redshift, and that mass distribution doesn't evolve with redshift in the same exact way as would be mimicked by cosmological redshifting.
-This assumption is very basic, not an assumption that we know the exact astrophyisics of BBH formation (i.e.
-PPISN feature location)
-    \item future/in prep.
-work will do the same thing but for a mass distribution that is allowed to evolve with redshift, similarly to the parametric analysis done in~\cite{ezquiaga_spectral_2022}, but with the evolution with redshift allowed to be arbitrary so long as it does not mimic cosmology.
-- GPs are good bc they can do correlations/generalize to multiple dimensions naturally.
-Our choice of kernel allows us to do this, but its not possible with the autoregressive kernel.
-    \item in XG we will get most bang for our buck with spectral sirens
-\end{itemize}
-\begin{acknowledgments}
-    The authors thank Reed Essick, Utkarsh Mali, Ben Farr, Maya Fishbach,  for helpful conversations.
-\end{acknowledgments}
-
-\bibliography{bib}
-\appendix
-\section{Details of data simulation}
-
 \end{document}

### 2024-02-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 To illustrate this, we generate a catalog of \result{1,000} GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
-This simulated catalog is designed to be representative of the gravitational wave transient catalog after the \ac{LVK}'s fifth observing run.
+This simulated catalog is designed to be representative of the gravitational wave transient catalog after the LVK's fifth observing run.
 The details of the data simulation are described in Appendix~\ref{ap:data generation}.
@@
 We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated GW catalog of \result{1,000} GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
-This simulated catalog is designed to be representative of the gravitational wave transient catalog after the \ac{LVK}'s fifth observing run.
+This simulated catalog is designed to be representative of the gravitational wave transient catalog after the LVK's fifth observing run.
 The details of the data simulation are described in Appendix~\ref{ap:data generation}.
@@
 
-This result also implies that we can expect to measure \Ho{} to \result{X\%} by the end of the \ac{LVK}'s fifth observing run using this method.
+This result also implies that we can expect to measure \Ho{} to \result{X\%} by the end of the LVK's fifth observing run using this method.
 
@@
 PPISN feature location).
-    \item We have also predicted a \result{X\%} measurement of \Ho{} by the end of the \ac{LVK}'s fifth observing run using non-parametric spectral sirens.
+    \item We have also predicted a \result{X\%} measurement of \Ho{} by the end of the LVK's fifth observing run using non-parametric spectral sirens.
 A similar study with parametric methods \citep{Chen:2024gdn} predicts an \result{X\%} measurement, demonstrating a \result{negligible/substantial/marginal} increase in precision.

### 2024-02-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 Unlike light, however, the selection effects of GWs are extremely well understood, allowing for a precise estimate of each catalogs' completeness and an unbiased measurement of the true GW population.
-This allows GW signals to be used as relatively clean probes of cosmological parameters \citep[e.g.][]{bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide}.
+% This allows GW signals to be used as relatively clean probes of cosmological parameters \citep[e.g.][]{bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide}.
 
@@
 \texttt{GWMockCat} also simulates a set of software injections, which we use to estimate selection effects in the inference.
-To determine the detectability of both injections and simulated events in O5, we use the projected O5 LIGO power spectral density \citep{dcc page} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
+% To determine the detectability of both injections and simulated events in O5, we use the projected O5 LIGO power spectral density \citep{dcc page} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
 The full procedure for this mock data generation process is described in \citet{fishbach_where, farah_things_2023, essick_dagnabbit_2023}, and the exact settings used for \texttt{GWMockCat} are made available in the accompanying data release.

### 2024-02-14 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2024-02-14 src/tex/ms.tex
@@
 Unlike light, however, the selection effects of GWs are extremely well understood, allowing for a precise estimate of each catalogs' completeness and an unbiased measurement of the true GW population.
-This allows GW signals to be used as relatively clean probes of cosmological parameters \citep[e.g.][]{bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide}.
+This allows GW signals to be used as relatively clean probes of cosmological parameters \citep[e.g.][]{}%bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide.
 
@@
 \texttt{GWMockCat} also simulates a set of software injections, which we use to estimate selection effects in the inference.
-To determine the detectability of both injections and simulated events in O5, we use the projected O5 LIGO power spectral density \citep{dcc page} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
+To determine the detectability of both injections and simulated events in O5, we use the projected O5 LIGO power spectral density \citep{dccpage} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
 The full procedure for this mock data generation process is described in \citet{fishbach_where, farah_things_2023, essick_dagnabbit_2023}, and the exact settings used for \texttt{GWMockCat} are made available in the accompanying data release.

### 2024-02-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 We then demonstrate that our flexible model alleviates this bias without the need to know the mass distribution's morphology \emph{a priori}.
-To illustrate this, we generate a catalog of \result{1,000} GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
+To illustrate this, we generate a catalog of \variable{output/num_found_events.txt}GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
 This simulated catalog is designed to be representative of the gravitational wave transient catalog after the LVK's fifth observing run.
@@
 
-With \result{1,000} events, we find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within \result{X-$\sigma$}.
+With \variable{output/num_found_events.txt}events, we find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within \result{X-$\sigma$}.
 However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by \result{X}-$\sigma$.
@@
 \subsection{Option 2: GP only}
-We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated GW catalog of \result{1,000} GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
+We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated GW catalog of \variable{output/num_found_events.txt}GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
 This simulated catalog is designed to be representative of the gravitational wave transient catalog after the LVK's fifth observing run.
@@
 
-After passing the simulated events through projected detector selection effects, the resulting catalog has \result{X} events, including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} BBH systems, consistent with the numbers projected for O5 by \citet{kiendrebogo_observing_2023}.
+After passing the simulated events through projected detector selection effects, the resulting catalog has \variable{output/num_found_events.txt}events, including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} BBH systems, consistent with the numbers projected for O5 by \citet{kiendrebogo_observing_2023}.
 We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from O5 detectors.

### 2024-02-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 The results of each of these fits are shown in Figure~\ref{fig:O5_GP}. 
-The \result{left} panel shows the inferred source frame mass distribution for each of the considered models, and the \result{right} panel shows the corresponding posteriors on \Ho{}.
+The left panel shows the inferred source frame mass distribution for each of the considered models, and the right panel shows the corresponding posteriors on \Ho{}.
 
@@
 
-We have repeated this analysis with \result{50} separate simulated catalogs, and found that the power law model causes us to infer a \result{higher/lower} value of \Ho{} at greater than \result{X}-$\sigma$, \result{Y}\% of the time, whereas the Gaussian process-based model reaches that level of bias only \result{Z}\% of the time, and the \plp{} model reaches the same level of bias \result{A}\% of the time.
+We have repeated this analysis with \result{50} separate simulated catalogs, and found that the broken power law model causes us to infer a \result{higher/lower} value of \Ho{} at greater than \result{X}-$\sigma$, \result{Y}\% of the time, whereas the Gaussian process-based model reaches that level of bias only \result{Z}\% of the time, and the \plp{} model reaches the same level of bias \result{A}\% of the time.
 This indicates that prior knowledge of the shape of the mass distribution is not required to perform a spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.
@@
 The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
-The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\result{67.66 km/s/Mpc}) is consistent with its inferred posterior distribution (\result{X}).
+The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\variable km/s/Mpc) is consistent with its inferred posterior distribution (\result{X}).
 Additionally, the posterior on \Ho is distinct from its prior distribution, indicating that the data is informative despite the flexibility of the population model.

### 2024-02-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
-The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\variable km/s/Mpc) is consistent with its inferred posterior distribution (\result{X}).
+The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\variable{output/H0_FID.txt} km/s/Mpc) is consistent with its inferred posterior distribution (\result{X}).
 Additionally, the posterior on \Ho is distinct from its prior distribution, indicating that the data is informative despite the flexibility of the population model.

### 2024-02-14 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2024-02-14 src/tex/ms.tex
@@
 The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
-The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\variable km/s/Mpc) is consistent with its inferred posterior distribution (\result{X}).
+The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\variable{output/H0_FID.txt} km/s/Mpc) is consistent with its inferred posterior (\result{X}).
 Additionally, the posterior on \Ho is distinct from its prior distribution, indicating that the data is informative despite the flexibility of the population model.

### 2024-02-14 src/scripts/data_generation.py
add variables to reference in ms

@@ -87,6 +87,7 @@ if  __name__ == "__main__":
     np.save(paths.data / "gw_data/m1zinj_det.npy",m1zinj_det)
     np.save(paths.data / "gw_data/dLinj_det.npy",dLinj_det)
     np.save(paths.data / "gw_data/log_pinj_det.npy",log_pinj_det)
+    
     ## find events and generate mock PE
     m1z_PE, _, dL_PE, log_PE_prior = gen_snr_scaled_PE(np_rng,m1s_true,m1s_true,dL_true/1000,osnr_interp,
                                                        reference_distance,N_SAMPLES_PER_EVENT,H0_FID,OM0_FID,
@@ -97,4 +98,7 @@ if  __name__ == "__main__":
     np.save(paths.data / "gw_data/dL_PE.npy",dL_PE)
     np.save(paths.data / "gw_data/log_PE_prior.npy",log_PE_prior)
 
-    
+    with open(paths.output / "num_found_events.txt", "w") as f:
+        print(len(dL_PE), file=f)
+    with open(paths.output / "H0_FID.txt", "w") as f:
+        print(H0_FID, file=f)

### 2024-02-14 src/scripts/nonparametric_numbers.py

@@ -1,0 +1,12 @@
+import arviz as az
+import numpy as np
+import paths
+
+id = az.InferenceData.from_netcdf(paths.data / "mcmc_nonparametric.nc4")
+h0samps = id.posterior['H0'][0]
+print(len(h0samps))
+
+with open(paths.output / "nonparh0percent.txt.txt", "w") as f:
+    print(f"{np.std(h0samps)/np.mean(h0samps):.1f}", file=f)
+with open(paths.output / "nonparh0CI.txt.txt", "w") as f:
+    print(f"{np.mean(h0samps):.1f}"+"^{+"+f"{np.percentile(h0samps,95)}"+"}"+"_{-"+f"{np.percentile(h0samps,95)}"+"}", file=f)

### 2024-02-14 src/scripts/data_generation.py
move data-related numbers to separate script

@@ -97,8 +97,3 @@ if  __name__ == "__main__":
     np.save(paths.data / "gw_data/m1z_PE.npy",m1z_PE)
     np.save(paths.data / "gw_data/dL_PE.npy",dL_PE)
     np.save(paths.data / "gw_data/log_PE_prior.npy",log_PE_prior)
-
-    with open(paths.output / "num_found_events.txt", "w") as f:
-        print(len(dL_PE), file=f)
-    with open(paths.output / "H0_FID.txt", "w") as f:
-        print(H0_FID, file=f)

### 2024-02-14 src/scripts/data_numbers.py

@@ -1,0 +1,10 @@
+from data_generation import H0_FID
+import paths
+import numpy as np
+
+dL_PE = np.load(paths.data / "gw_data/dL_PE.npy")
+
+with open(paths.output / "num_found_events.txt", "w") as f:
+    print(len(dL_PE), file=f)
+with open(paths.output / "H0_FID.txt", "w") as f:
+    print(H0_FID, file=f)

### 2024-02-14 src/scripts/nonparametric_numbers.py

@@ -6,7 +6,7 @@ id = az.InferenceData.from_netcdf(paths.data / "mcmc_nonparametric.nc4")
 h0samps = id.posterior['H0'][0]
 print(len(h0samps))
 
-with open(paths.output / "nonparh0percent.txt.txt", "w") as f:
+with open(paths.output / "nonparh0percent.txt", "w") as f:
     print(f"{np.std(h0samps)/np.mean(h0samps):.1f}", file=f)
-with open(paths.output / "nonparh0CI.txt.txt", "w") as f:
+with open(paths.output / "nonparh0CI.txt", "w") as f:
     print(f"{np.mean(h0samps):.1f}"+"^{+"+f"{np.percentile(h0samps,95)}"+"}"+"_{-"+f"{np.percentile(h0samps,95)}"+"}", file=f)

### 2024-02-14 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2024-02-14 src/tex/ms.tex
@@
 The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
-The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\variable km/s/Mpc) is consistent with its inferred posterior distribution (\result{X}).
+The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\variable{output/H0_FID.txt} km/s/Mpc) is consistent with its inferred posterior (\variable{output/nonparh0CI.txt} \,km/s/Mpc).
 Additionally, the posterior on \Ho is distinct from its prior distribution, indicating that the data is informative despite the flexibility of the population model.

### 2024-02-14 src/scripts/nonparametric_numbers.py
add output as dependency of ms

@@ -9,4 +9,4 @@ print(len(h0samps))
 with open(paths.output / "nonparh0percent.txt", "w") as f:
     print(f"{np.std(h0samps)/np.mean(h0samps):.1f}", file=f)
 with open(paths.output / "nonparh0CI.txt", "w") as f:
-    print(f"{np.mean(h0samps):.1f}"+"^{+"+f"{np.percentile(h0samps,95)}"+"}"+"_{-"+f"{np.percentile(h0samps,95)}"+"}", file=f)
+    print(f"${np.mean(h0samps):.1f}"+"^{+"+f"{np.percentile(h0samps,95)}"+"}"+"_{-"+f"{np.percentile(h0samps,95)}"+"}$", file=f)

### 2024-02-14 src/scripts/nonparametric_numbers.py
formatting and algebra fix

@@ -4,9 +4,10 @@ import paths
 
 id = az.InferenceData.from_netcdf(paths.data / "mcmc_nonparametric.nc4")
 h0samps = id.posterior['H0'][0]
-print(len(h0samps))
 
 with open(paths.output / "nonparh0percent.txt", "w") as f:
     print(f"{np.std(h0samps)/np.mean(h0samps):.1f}", file=f)
+lower = np.mean(h0samps)-np.percentile(h0samps,5)
+upper = np.percentile(h0samps,95)-np.mean(h0samps)
 with open(paths.output / "nonparh0CI.txt", "w") as f:
-    print(f"${np.mean(h0samps):.1f}"+"^{+"+f"{np.percentile(h0samps,95)}"+"}"+"_{-"+f"{np.percentile(h0samps,95)}"+"}$", file=f)
+    print(f"${np.mean(h0samps):.1f}"+"^{+"+f"{lower:.1f}"+"}"+"_{-"+f"{upper:.1f}"+"}$", file=f)

### 2024-02-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\variable{output/H0_FID.txt} km/s/Mpc) is consistent with its inferred posterior (\variable{output/nonparh0CI.txt} \,km/s/Mpc).
-Additionally, the posterior on \Ho is distinct from its prior distribution, indicating that the data is informative despite the flexibility of the population model.
+Additionally, the posterior on \Ho{} is distinct from its prior distribution, indicating that the data is informative despite the flexibility of the population model.
 

### 2024-02-14 src/scripts/nonparametric_numbers.py

@@ -6,7 +6,7 @@ id = az.InferenceData.from_netcdf(paths.data / "mcmc_nonparametric.nc4")
 h0samps = id.posterior['H0'][0]
 
 with open(paths.output / "nonparh0percent.txt", "w") as f:
-    print(f"{np.std(h0samps)/np.mean(h0samps):.1f}", file=f)
+    print(f"{np.std(h0samps)/np.mean(h0samps)*100:.1f}", file=f)
 lower = np.mean(h0samps)-np.percentile(h0samps,5)
 upper = np.percentile(h0samps,95)-np.mean(h0samps)
 with open(paths.output / "nonparh0CI.txt", "w") as f:

### 2024-02-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-This result also implies that we can expect to measure \Ho{} to \result{X\%} by the end of the LVK's fifth observing run using this method.
+This result also implies that we can expect to measure \Ho{} to \result{\variable{output/nonparh0percent.txt}\%} by the end of the LVK's fifth observing run using this method.
 

### 2024-03-09 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2024-02-14 src/tex/ms.tex
@@
 \usepackage{enumerate}
+\usepackage{acro}
+
+\DeclareAcronym{\ac{LVK}}{short=\ac{LVK}, long=LIGO Virgo KAGRA}
+\DeclareAcronym{O5}{short=O5, long=the fifth LIGO Virgo KAGRA observing run}
+
+\DeclareAcronym{GW}{short=GW, long=gravitational wave}
+\DeclareAcronym{CBC}{short=CBC, long=compact binary coalescence}
+\DeclareAcronym{BBH}{short=BBH, long=binary black hole}
+
+\DeclareAcronym{GP}{short=GP, long=Gaussian process, long-plural=es}
 
@@
 % Title
-\title{No need to know: astrophysics-agnostic spectral siren cosmology}
-%\title{No need to know: gravitational-wave cosmology with an astrophysics-agnostic mass distribution}
+\title{No need to know: astrophysics-agnostic gravitational wave cosmology}
+%\title{No need to know: astrophysics-agnostic spectral siren cosmology}
 %\title{No need to know: spectral siren cosmology with an astrophysics-agnostic mass distribution of gravitational-wave sources}
-%\title{No need to know: gravitational-wave cosmology while astrophysically ignorant/for the astrophysically-ignorant}
+%\title{No need to know: gravitational-wave cosmology for the astrophysically-ignorant}
 
@@
 \begin{abstract}
-Gravitational waves (GWs) are unique transients in which they encode direct information about their luminosity distance. 
-Adding redshift, a GW standard siren Hubble diagram can be constructed to probe the standard cosmological model.  
-Although individual signals do not have any reference scale imprinting a redshift timestamp, as a collective there could be signatures in their mass spectrum that would break the mass-redshift degeneracy.
-This spectral siren method thus provides a GW data-only approach to constrain the Universe's expansion rate from the population of compact binary coalescences. 
+\Acp{GW} from merging compact objects are the only transient astrophysical sources that encode direct information about their luminosity distances. 
+Paired with their redshifts, a Hubble diagram can be constructed to probe the standard cosmological model.  
+Although individual signals do not have any reference scale from which a redshift can be inferred, an ensemble of events can provide such a scale via their distribution of source masses. 
+This ``spectral siren'' method thus constrains the Universe's expansion rate from \ac{GW} observations alone, without the need for accompanying electromagnetic observations. 
 So far, this technique has only been applied with simple parametric representations of the mass distribution. 
-However, the use of a wrong mass model would inevitably lead to biases in the cosmological inference, an acute problem given our current lack of understanding of the true astrophysical model.
-We demonstrate that the spectral siren method is purely data-driven, in which the right cosmology can be inferred without a given model, using a flexible Gaussian process reconstruction of the mass spectrum.
-In particular, we analyze an observing scenario with XX GW signals and detectors at XX sensitivity. 
-We find that both the source mass model and the cosmological parameters are correctly reconstructed. 
+However, the use of an incorrect mass model inevitably leads to biases in the cosmological inference, an acute problem given our current lack of understanding of the true astrophysical model.
+Furthermore, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to obtain unbiased measurements of cosmological parameters with the spectral siren methodology.
+We demonstrate that unbiased inference of cosmological parameters is possible without any assumptions for the mass distribution's morphology.
+We do this by constructing a flexible, Gaussian-process based model for \ac{GW} source population to analyze simulated \ac{GW} data consistent with expectations for the next \acl{LVK} observing run.
+We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a \variable{output/nonparh0percent.txt}\% measurement.
 This astrophysics-agnostic spectral siren technique will be key to deliver precise and unbiased cosmological constraints with future observations in the presence of possible redshift evolution of the mass spectrum.
@@
 \label{sec:intro}
-Like light, gravitational waves (GWs) are redshifted as they propagate across the universe.
-Unlike light, however, the selection effects of GWs are extremely well understood, allowing for a precise estimate of each catalogs' completeness and an unbiased measurement of the true GW population.
-This allows GW signals to be used as relatively clean probes of cosmological parameters \citep[e.g.][]{}%bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide.
+Like light, \acp{GW} are redshifted as they propagate across the universe.
+Unlike light, however, the selection effects of \acp{GW} are extremely well understood, allowing for a precise estimate of each catalogs' completeness and an unbiased measurement of the true \ac{GW} population.
+This allows \ac{GW} signals to be used as relatively clean probes of cosmological parameters \citep[e.g.][]{}%bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide.
 
 GW signals provide direct measurements of each source's luminosity distance and redshifted, or detector frame, masses, $m_{\det}= m_{\source}(1+z)$.
-Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the GW signal was emitted, $H(z)$.
+Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the \ac{GW} signal was emitted, $H(z)$.
 In practice, the source frame mass is unknown, but it is possible to know the location of features in the source frame mass distribution.
 The whole mass distribution therefore acts similarly to an electromagnetic spectrum, where the location of spectral features provides a redshift measurement.
-The method of using the mass distribution of GW sources to measure cosmological parameters has therefore been coined ``spectral sirens'' \citep{ezquiaga_spectral_2022}.
-Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff+fin} using the BNS mass distribution, and extended to the BBH mass distribution by \cite{farr_future_2019}.
+The method of using the mass distribution of \ac{GW} sources to measure cosmological parameters has therefore been coined ``spectral sirens'' \citep{ezquiaga_spectral_2022}.
+Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff_gravitational_1993} using the binary neutron star mass distribution, and extended to the \ac{BBH} mass distribution by \cite{farr_future_2019}.
 
-Perhaps because of the analogy to electromagnetic spectra in which the location of the source frame spectral features is well-known, or because of the examples given in \cite{farr_future_2019} and \cite{chernoff+fin}, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to do this measurement, from e.g.
+Perhaps because of the analogy to electromagnetic spectra in which the location of the source frame spectral features is well-known, or because of the examples given in \cite{farr_future_2019} and \cite{chernoff_gravitational_1993}, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to do this measurement, from e.g.
 theoretical expectations of the location of a pulsational-pair instability pileup, upper mass gap, or maximum neutron star mass.
-However, the information in the spectral siren measurement comes solely from the assumption that, at a given luminosity distance, all CBC mergers follow the same mass distribution.
+However, the information in the spectral siren measurement comes solely from the assumption that, at a given luminosity distance, all \ac{CBC} mergers follow the same mass distribution.
 Therefore the location, shape, and existence of all features in the mass distribution can be inferred simultaneously with $H(z)$.
 
-In this work, we explicitly demonstrate that no \emph{a priori} knowledge about the shape of the CBC mass spectrum is necessary to use the spectral siren methodology. 
-We do this by inferring \Ho with a flexible, non-parametric model for the mass distribution of CBCs. 
+In this work, we explicitly demonstrate that no \emph{a priori} knowledge about the shape of the \ac{CBC} mass spectrum is necessary to use the spectral siren methodology. 
+We do this by inferring \Ho with a flexible, non-parametric model for the mass distribution of \acp{CBC}. 
 This model makes minimal prior assumptions about the shape of the mass distribution, enabling it to accurately infer a wide range of morphologies.
@@
 
-We find that using this model allows for a \result{X}\% measurement of \Ho by the end of the LVK's fifth observing run, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
+We find that using this model allows for a \variable{output/nonparh0percent.txt}\% measurement of \Ho{} by the end of \ac{O5}, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
 While constraints on \Ho are used in this work to quantify the accuracy and precision of our non-parametric method, we emphasize that spectral sirens are able to constrain $H(z)$ out large redshifts that are inaccessible by electromagnetic observations, especially with next-generation detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn}\footnote{See example calculations here: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}}.
+Applied to data from next-generation detectors, this method will therefore be able to constrain the dark energy equation of state parameter, $w$, at redshifts $\gtrsim 2$.
 
@@
 
+This paper is organized as follows: \comment{insert outline here.}
+
+The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL under an MIT license, and was enabled by the \showyourwork package \citep{Luger2021}.
+
 \section{Methods I: Spectral Siren Cosmology}
 \label{sec:ss}
-\begin{enumerate}
-    \item Briefly explain HBA with selection effects.
-\end{enumerate}
-Given GW data $\{d\}$ that contains $N_{\text{evs}}$ detected GW signals, each with parameters $\theta_i$, the posterior probability of the GW sources being drawn from an underlying population and background cosmology described by hyper-parameters $\Lambda$ is \citep{loredo, taylor, mandel}
+
+To simultaneously infer cosmological parameters and the population of GW sources, we employ a hierarchical Bayesian analysis.
+This allows us to ``undo'' the selection effects of \ac{GW} detectors to obtain a true, or astrophysical, population.
+
+Given \ac{GW} data $\{d\}$ that contains $N_{\text{evs}}$ detected \ac{GW} signals, each with parameters $\theta_i$, the posterior probability of the \ac{GW} sources being drawn from an underlying population and background cosmology described by hyper-parameters $\Lambda$ is \citep{loredo, taylor, mandel}
 \begin{equation}
-    p(\Lambda|d) \propto p(\Lambda) e^{-N_{\text{exp}}(\Lambda)} \prod_i^{N_{\text{evs}}} \frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda) .
+    p(\Lambda|d) \propto p(\Lambda) e^{-N_{\text{exp}}(\Lambda)} \prod_i^{N_{\text{evs}}} \frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda) \, .
     \label{eq:inhomog-poisson}
 \end{equation}
-Here, $N_{\text{exp}(\Lambda)}$ is the expected number of detections given $\Lambda$ and the GW detector sensitivity, calculated using a Monte Carlo sum over $N_{\text{inj}}$ found signals injected into the data stream \citep[see][for a detailed explanation of this process]{essick}.
-$\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ is the detector frame merger rate density.
-This quantity is the population model conditioned on hyper-parameters $\Lambda$.
-Following \citet{callister_parameter_2023}, we use a semicolon to explicitly indicate that this is a function of $\Lambda$, not a density over $\Lambda$.
+Here, $\Lambda$ is the set of population parameters, or hyper-parameters, and $p(\Lambda)$ is the prior on $\Lambda$.
+$\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ is the detector frame merger rate density conditioned on hyper-parameters $\Lambda$, this quantity is also referred to as the population model.
+Following \citet{callister_parameter-free_2023}, we use a semicolon to explicitly indicate that this is a function of $\Lambda$, not a density over $\Lambda$.
+$N_{\text{exp}}(\Lambda)$ is the expected number of detections given $\Lambda$ and the \ac{GW} detector sensitivity, and is calculated using a Monte Carlo sum over $N_{\text{inj}}$ found signals injected into the data stream \citep[see][for a detailed explanation of this process]{essick}.
 
-Since the parameters of each event are not perfectly measured, we marginalize $\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ over the event parameters' likelihood given the data.
+Since the event parameters are not perfectly measured, we marginalize $\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ over the likelihood of each event's parameters given the data.
 Practically, this is done by a Monte Carlo average over each event's posterior samples $\{\theta_j\}_i$, dividing out the prior used when inferring those posterior samples, $\pi_{PE}(\theta)$:
@@
 We assume a uniform-in-comoving volume redshift distribution, as in Equation~\ref{eq:underlying redshift dist}.
-We assume $m_1=m_2$ and model the distribution of primary masses using a Gaussian process (described in Section~\ref{sec:model}.
+We assume $m_1=m_2$ and model the distribution of primary masses using a Gaussian process (described in Section~\ref{sec:model}).
 
-The spectral siren method of measuring cosmological parameters relies on the assumption that all compact binaries follow the same mass distribution regardless of redshift\footnote{\citet{ezquiaga_spectral_2022} show that the source frame masses can follow different distributions at different redshifts, so long as this redshift evolution of the mass distribution does not exactly mimic cosmology (i.e.
-all features cannot exhibit a monotonic rightward shift with redshift).}.
+The spectral siren method of measuring cosmological parameters relies on the assumption that all compact binaries follow the same mass distribution at all redshifts\footnote{This assumption does not strictly have to hold, as \citet{ezquiaga_spectral_2022} show that the source frame masses can follow different distributions at different redshifts, so long as this redshift evolution of the mass distribution does not exactly mimic cosmology (i.e.
+all features cannot exhibit a monotonic rightward shift with redshift).
+However, for the redshift range considered in this work, we assume the mass distribution does not evolve at all with redshift.}.
 The likelihood in Equation~\ref{eq:inhomog-poisson} is therefore maximized when the relationship between redshift and luminosity distance makes all source frame masses follow the same distribution.
@@
 In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and $w=1$ and use uniform priors on \Ho{} and \Omm{}.
-Algorithmically, this is equivalent to drawing a value of \Ho, using it to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to $m_{\det} (1+z(D_L)) = m_{\source}$, evaluating the source-frame mass distribution at these transformed values, and then evaluating the likelihood in Equation~\ref{eq:inhomog-poisson}.
+
+Algorithmically, we evaluate the likelihood in Equation~\ref{eq:inhomog-poisson} by drawing \Ho{} and \Omm{} from their prior distributions (contained in $p(\Lambda)$), using them to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to $m_{\det} (1+z(D_L)) = m_{\source}$, evaluating the source-frame mass distribution at these transformed values ($\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ for events and $N_{\text{exp}}(\Lambda)$ for injections), and then taking the product of these terms to evaluate the expression in Equation~\ref{eq:inhomog-poisson}.
 The full process is outlined in Section~\ref{sec:model}.
@@
 \label{sec:model}
-Gaussian processes (GPs) are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
-Their smoothness properties make them widely useful in GW data analysis and beyond for regression problems, such as modeling time-domain waveforms \citep{zoheyr, huerta}, density estimation problems, such as estimating posterior densities of single-event parameters from parameter estimation samples ~\citep{dangelo}, and as a prior on histogram bin heights \citep{ray_2023,li_flexible_2021}.
+\Acp{GP} are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
+Their smoothness properties make them widely useful in \ac{GW} data analysis and beyond for regression problems, such as modeling time-domain waveforms \citep{doctor_statistical_2017, huerta_eccentric_2018} and the neutron star equation of state \citep{landry_nonparametric_2019}, density estimation problems, such as estimating posterior densities of single-event parameters from parameter estimation samples ~\citep{demilio_density_2021}, and as a prior on histogram bin heights for population inference \citep{mandel_model-independent_2017, li_flexible_2021, ray_non-parametric_2023}.
 Our use case is slightly different.
-We will utilize a GP as a prior on the functions that describe the primary mass distribution of CBCs.
-It will therefore replace $p(\Lambda)$ in Equation~\ref{eq:inhomog-poisson}.
+We will utilize a \ac{GP} as a prior on the functions that describe the primary mass distribution of \acp{CBC}.
+It will therefore replace $p(\Lambda)$ in Equation~\ref{eq:inhomog-poisson}, except in the case of cosmological parameters, which all have uniform priors.
 This choice adds very little prior information about the shape of the mass distribution, besides enforcing that it must be smooth.
-Practically, the only difference in the inference of the population when using a GP versus other modeling choices is that $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is determined directly by a realization of the GP, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
+Practically, the only difference in the inference of the population when using a GP versus other modeling choices is that $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is determined directly by a realization of the \ac{GP}, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
 
@@
 We therefore have $N_{\text{evs}}M + N_{\text{inj}}$ mass hyper-parameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of PE samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
-We label these $\{\mathcal{R}_ij,\mathcal{R}_k\}$, and our full set of hyper-parameters is $\Lambda=\{H_0, \Omega_M, \mathcal{R}_ij,\mathcal{R}_k\}$. 
-In this way, our GP-based mass distribution is similar to the autoregressive population models used in ~\citet{callister_ar}.
+We label these $\{\mathcal{R}_{ij},\mathcal{R}_k\}$, and our full set of hyper-parameters is $\Lambda=\{H_0, \Omega_M, \mathcal{R}_{ij},\mathcal{R}_k\}$. 
+In this way, our GP-based mass distribution is similar to the autoregressive population models used in ~\citet{callister_parameter-free_2023}.
 Indeed, an autoregressive model is a Gaussian process with a particular choice of kernel.
@@
 We use a Mat\'ern kernel \citep{handcock_and_stein,stein_1999} with $\nu = 5/2$, but have repeated the analysis with $\nu=3/2$ and $\infty$, finding little impact on the results, except that the $\nu=\infty$ case (also called the ``squared exponential'' kernel) produces a slightly more jagged mass distribution.
-Mat\'ern kernels have two parameters besides the mean variance that determine their properties: a length scale $l$ and variance $s$.
-In our use case, these are one level further removed from hyper-parameters, so we adopt the terminology used in \citet{callister_ar} and call them ``hyper-hyper-parameters.''
+Mat\'ern kernels have two parameters besides the mean that determine their properties: a length scale $l$ and variance $s$.
+In our use case, these are one level further removed from hyper-parameters, so we adopt the terminology used in \citet{callister_parameter-free_2023} and call them ``hyper-hyper-parameters.''
 We fit these hyper-hyper-parameters along with the hyper-parameters $\Lambda$ to minimize prior assumptions about the form of the mass distribution.
 We use ``penalized-complexity'' priors on the hyper-hyper-parameters to enforce that the model does not create small-scale structure uninformed by data, thereby avoiding over-fitting \citep{simpson_penalising_2017,simpson_garcpas_2022}.
-% remove this if we don't use it.
-Explicitly, the priors on $l$ and $s$ have less than 5\% support for correlation lengths smaller than the \result{minimum} spacing between event-level posterior samples.
+Explicitly, the priors on $l$ and $s$ have less than 5\% support for correlation lengths smaller than the average spacing between event-level posterior means.
 
-GP evaluations notoriously scale as the cube of the number of data points, making them unwieldy with large data sets such as the ones expected for O5.
-We therefore make two approximations to a full GP to increase computational efficiency.
-First, for each likelihood evaluation, we evaluate a full GP on a regular grid between $0\Msun$ and $250\Msun$ and then interpolate it at each data point.
-Second, we use the quasi-separability of Mat\'ern kernels to analytically perform the transformation between covariance matrix and GP draw.
-This second step is done using the \texttt{QuasisepSolver} module \citep{foreman-mackey_fast_2017} in the \texttt{tinygp} code base, and requires data to be sort-able (i.e. one-dimensional).
+\ac{GP} evaluations notoriously scale as the cube of the number of data points, making them unwieldy with large data sets such as the ones expected for \ac{O5}.
+We therefore make two approximations to a full \ac{GP} to increase computational efficiency.
+First, for each likelihood evaluation, we evaluate a full \ac{GP} on a regular grid between $0\Msun$ and $250\Msun$ and then interpolate it at each data point.
+Second, we use the quasi-separability of Mat\'ern kernels to analytically perform the transformation between covariance matrix and \ac{GP} draw.
+This second step is done using the \texttt{QuasisepSolver} module \citep{foreman-mackey_fast_2017} in the \texttt{tinygp} code base \citep{foreman-mackey_tinygp_2021}, and requires data to be sortable (i.e.
+one-dimensional).
 
@@
 \begin{enumerate}
-    \item Draw cosmological parameters \Ho{} and \Omm{} from the prior distributions described in Equation~\ref{eq:cosmo-priors}.
+\script{nonparametric_inference.py}
+    \item Draw cosmological parameters \Ho{} and \Omm{} from uniform prior distributions.
     \item Transform the luminosity distances and detector frame masses of each event posterior sample and injection into redshifts and source frame masses according to the cosmology specified by step 1.
     \item Draw hyper-hyper-parameters $l$ and $s$ from the penalized-complexity priors described above.
-    \item Draw a single GP realization with a kernel defined by $l$ and $s$. This is defined on a regular grid of source-frame masses and evaluated using the \texttt{QuasisepSolver}.
-    \item Interpolate the GP at each event posterior sample and injection source frame mass (from step 2)
+    \item Draw a single \ac{GP} realization with a kernel defined by $l$ and $s$.
+    This is defined on a regular grid of source-frame masses and evaluated using the \texttt{QuasisepSolver}.
+    \item Interpolate the \ac{GP} at each event posterior sample and injection source frame mass (from step 3)
     \item Calculate the population likelihood according to Equation~\ref{eq:inhomog-poisson}.
 \end{enumerate}
-
-\jme{[advertise code is public!]}
 
@@
 \subsection{Option 1: include parametric models}
-In this Section, we first show that fitting an incorrect functional form to the mass distribution of CBCs biases the inference of cosmological parameters when using the spectral siren methodology.
+In this Section, we first show that fitting an incorrect functional form to the mass distribution of \acp{CBC} biases the inference of cosmological parameters when using the spectral siren methodology.
 We then demonstrate that our flexible model alleviates this bias without the need to know the mass distribution's morphology \emph{a priori}.
 To illustrate this, we generate a catalog of \variable{output/num_found_events.txt}GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
-This simulated catalog is designed to be representative of the gravitational wave transient catalog after the LVK's fifth observing run.
+This simulated catalog is designed to be representative of the  \ac{GW}  transient catalog after \ac{O5}.
 The details of the data simulation are described in Appendix~\ref{ap:data generation}.
@@
 \subsection{Option 2: GP only}
-We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated GW catalog of \variable{output/num_found_events.txt}GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
-This simulated catalog is designed to be representative of the gravitational wave transient catalog after the LVK's fifth observing run.
+We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated \ac{GW} catalog of \variable{output/num_found_events.txt}GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
+This simulated catalog is designed to be representative of the  \ac{GW}  transient catalog after \ac{O5}.
 The details of the data simulation are described in Appendix~\ref{ap:data generation}.
@@
 The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\variable{output/H0_FID.txt} km/s/Mpc) is consistent with its inferred posterior (\variable{output/nonparh0CI.txt} \,km/s/Mpc).
+At very high and very low masses, there is both a lack of data and little detector sensitivity.
+The mass distribution is therefore uninformed in this region and the GP reverts to its prior, which resembles random scatter around a central value.
+Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
+On the other hand, the lack of data just below the minimum BH mass and just above the maximum BH mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the GP to learn a relatively low merger rate at the edges of the mass distribution.
+Indeed, these features are very informative to the \Ho{} constraint \citep{the_ligo_scientific_collaboration_constraints_2021}.
+The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
 Additionally, the posterior on \Ho{} is distinct from its prior distribution, indicating that the data is informative despite the flexibility of the population model.
 
-This result also implies that we can expect to measure \Ho{} to \result{X\%} by the end of the LVK's fifth observing run using this method.
+\subsection{Projections for Future Measurements}
+This result implies that we can expect to measure \Ho{} to \variable{output/nonparh0percent.txt}\% by the end of \ac{O5} using the method presented in this work.
+However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\sim 3$, with most sources lying near redshift $\sim 1.2$ \citep{chen_distance_2021}.
+Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 40$.
+This means that future observations of \acp{GW} will be more sensitive $H(z>1)$ than to \Ho, and can therefore constrain several additional cosmological parameters.
+We demonstrate this by repeating the same analysis while also fitting for the local matter density, \Omm.
+The result is shown in Figure~\ref{fig:O5_corner}.
+We find \ac{O5} observations to be most sensitive to \result{$H(z=X)$}, which results in similarly informative constraints on \Omm{} as on \Ho.
+\begin{itemize}
+    \item There is a strong degeneracy in these two cosmological parameters, caused by the sensitivity at z=1
+    \item explain why - balance between slope and intercept.
+\end{itemize}
 
-\begin{itemize}
-    \item If possible, could do constraints on H0 as a function of \# of events.
-    \item could also plot H(z) using posterior draws from \Ho{} and \Omm{}- can see at which redshift you get the best constraints.
-Then plot posterior of $H(z=z_{\text{best measured}})$ rather than \Ho \jme{[I like this option!]}
-\end{itemize}
+% BONUS
+To demonstrate the capabilities of next-generation detectors, we repeat the full analysis with \result{1000} events indicative of Cosmic Explorer's projected best-measured events within the first month alone.
+We fit the dark energy equation of state parameter, $w$, along with \Ho, and show results in Figure~\ref{fig:XG_corner}.
+We recover unbiased measurements of each parameter and find a \result{X}\% measurement of $w$ and a \result{X}\% measurement of \Ho.
+In comparison, the most recent constraints on these parameters with \result{X experiment} were \result{X}\% and \result{X}\% , respectively.
+This illustrates the unique ability of \acp{GW} to probe the high-redshift universe.
 
@@
     \centering
-    \includegraphics[width=\textwidth]{figures/O5_GP.pdf}
+    \includegraphics[width=\textwidth]{figures/O5_GP_pm.pdf}
     \caption{Two-panel figure of mass distribution and \Ho{} posterior. 
-    Include prior on \Ho{} as alpha=0.5 line. 
-    If we end up fitting \Omm{} we can put a corner plot instead.}
+    Include prior on \Ho{} as alpha=0.5 line.}
     \label{fig:O5_GP}
     \script{nonparametric_twopanel.py}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figures/O5_GP_corner.pdf}
+    \caption{Two-panel figure of (A) \Omm{} and \Ho{} corner plot and (B)$H(z)$ for O5.
+We can include a posterior of $H(z=z_{\text{best measured}})$ as an inset to the $H(z)$ plot.
+Explain the inset and comment on $z_{\text{best measured}}$}
+    \label{fig:O5_corner}
+    \script{nonparametric_corner.py}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\textwidth]{figures/CE_GP.pdf}
+    \caption{Two-panel figure of (A) $w$, and \Ho{} corner plot and (B)$H(z)$ for CE. We can include a posterior of $H(z=z_{\text{best measured}})$ as an inset to the $H(z)$ plot.}
+    \label{fig:XG_corner}
+    \script{XG_corner.py}
 \end{figure}
@@
     \item We are able to get constraints on cosmological parameters with a very flexible model for the mass distribution. 
-    \item The success of the flexible model in recovering an informative and accurate posterior on \Ho demonstrates that there is no need to know the CBC mass distribution's morphology in order to constrain cosmological parameters with it.
+    \item The success of the flexible model in recovering an informative and accurate posterior on \Ho{} demonstrates that there is no need to know the \ac{CBC} mass distribution's morphology in order to constrain cosmological parameters with it.
     \item This is because the information in spectral sirens comes from assuming events come from the same overall mass distribution across redshift, and that mass distribution doesn't evolve with redshift in the same exact way as would be mimicked by cosmological redshifting.
-This assumption is very basic, not an assumption that we know the exact astrophyisics of BBH formation (i.e.
+This assumption is very basic, not an assumption that we know the exact astrophyisics of \ac{BBH} formation (i.e.
 PPISN feature location).
-    \item We have also predicted a \result{X\%} measurement of \Ho{} by the end of the LVK's fifth observing run using non-parametric spectral sirens.
+    \item We have also predicted a \variable{output/nonparh0percent.txt}\% measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens.
 A similar study with parametric methods \citep{Chen:2024gdn} predicts an \result{X\%} measurement, demonstrating a \result{negligible/substantial/marginal} increase in precision.
@@
 With current detectors, these systematic effects are dwarfed by statistical uncertainties.
-However, next-generation detectors will herald in enough GW observations to substantially decrease statistical uncertainty in these measurements: \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to \result{$\sim3\sigma$} systematic biases in \Ho when catalogs contain 2,000 events.
+However, next-generation detectors will herald in enough \ac{GW} observations to substantially decrease statistical uncertainty in these measurements: \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to \result{$\sim3\sigma$} systematic biases in \Ho when catalogs contain 2,000 events.
 Thus, the additional statistical uncertainty introduced by non-parametric approaches may be preferable to the systematic errors associated with choosing a parametric model.
+    \item additionally, in \ac{O5} we will be most sensitive to \result{$z=X$}.
+This is larger than we expect to see EM counterparts to, and larger than we expect galaxy surveys to be complete to (is this true?) so upgrades to current facilities will allow spectral sirens to probe H(z) to relatively unexplored redshifts.
+This is not unique to our nonparametric method, but to spectral sirens in general.
     \item future/in prep.
 work will do the same thing but for a mass distribution that is allowed to evolve with redshift, similarly to the parametric analysis done in~\cite{ezquiaga_spectral_2022}, but with the evolution with redshift allowed to be arbitrary so long as it does not mimic cosmology.
-- GPs are good bc they can do correlations/generalize to multiple dimensions naturally.
-Our choice of kernel allows us to do this, but its not possible with the autoregressive kernel.
+Because of ~\citet{ezquiaga_spectral_2022}, we think we could get a useful measurement if we include evolving peak shapes/positions in the mass distribution, provided there are multiple features present (which we know there are) and that they do not evolve in the same exact way as cosmology would - i.e.
+all features do not monotonically move to higher source frame mass with increasing redshift, all at the same rate.
+That scenario would be astrophysically unlikely, especially since the processes changing the features are all distinct from one another.
+    \item GPs are good bc they can do correlations/generalize to multiple dimensions naturally. Our choice of kernel allows us to do this, but its not possible with the autoregressive kernel.
     \item in XG we will get most bang for our buck with spectral sirens
@@
 
-We generate a mock catalog of GW events characteristic of that expected from the LVK's fifth observing run (O5). 
+We generate a mock catalog of \ac{GW} events characteristic of that expected from \acf{O5}. 
 These events are drawn from an underlying population described by
@@
 Would need to get relative heights correct though.}
-\jme{[For O5 BNSs will be subdominant. I would focus on BBHs for simplicity]}
-\comment{[Ok, so we should say that although a putuative lower mass gap would give and additional feature, it is not expected to be constrained well in O5, so we focus on the BBH spectrum for now.
+\jme{[For \ac{O5} BNSs will be subdominant. I would focus on \acp{BBH} for simplicity]}
+\comment{[Ok, so we should say that although a putuative lower mass gap would give and additional feature, it is not expected to be constrained well in O5, so we focus on the \ac{BBH} spectrum for now.
 In XG, though, it will be important to consider the full spectrum \citep{ezquiaga_spectral_2022}.]}
 
-After passing the simulated events through projected detector selection effects, the resulting catalog has \variable{output/num_found_events.txt}events, including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} BBH systems, consistent with the numbers projected for O5 by \citet{kiendrebogo_observing_2023}.
-We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from O5 detectors.
+After passing the simulated events through projected detector selection effects, the resulting catalog has \variable{output/num_found_events.txt}events, including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} \ac{BBH} systems, consistent with the numbers projected for \ac{O5} by \citet{kiendrebogo_observing_2023}.
+We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from \ac{O5} detectors.
 \texttt{GWMockCat} also simulates a set of software injections, which we use to estimate selection effects in the inference.
-To determine the detectability of both injections and simulated events in O5, we use the projected O5 LIGO power spectral density \citep{dccpage} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
+To determine the detectability of both injections and simulated events in O5, we use the projected \ac{O5} LIGO power spectral density \citep{dccpage} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
 The full procedure for this mock data generation process is described in \citet{fishbach_where, farah_things_2023, essick_dagnabbit_2023}, and the exact settings used for \texttt{GWMockCat} are made available in the accompanying data release.

### 2024-03-09 src/scripts/nonparametric_twopanel.py
try plots with new run

@@ -8,10 +8,11 @@ from data_generation import H0_FID
 from gwpop import powerlaw_peak
 
 import paths
+NSAMPS = 500
 
 plt.style.use(paths.scripts / "matplotlibrc")
 
-def two_panel(path):
+def two_panel(path, hyperparam='H0'):
     id = az.InferenceData.from_netcdf(path)
     samples = id.posterior
     try:
@@ -19,29 +20,41 @@ def two_panel(path):
     except KeyError :
         r = np.nan_to_num(np.exp(samples['log_rate'][0]))
 
-    fig, axes = plt.subplots(ncols=2,figsize=(20,8),facecolor='none',gridspec_kw={'width_ratios': [1.2, 1]})
-    for i in range(5):
-        PLP_norm = np.trapz(r[i],TEST_M1S)*1.15
-        axes[0].plot(TEST_M1S, r[i]/PLP_norm, lw=0.2, c="blue",alpha=0.1)
-    axes[0].plot([], lw=0.2, c="blue",alpha=0.1,label='Gaussian process fit')
-        
-    axes[0].plot(TEST_M1S, 7*powerlaw_peak(TEST_M1S,alpha=-2.7,f_peak=0.05,mMax=78.0,mMin=10.0,mu_m1=30.0,sig_m1=7.0),
-                c='k',lw=3,label='injected distribution')
+    fig, axes = plt.subplots(ncols=2,figsize=(7.5,4*.75),facecolor='none',gridspec_kw={'width_ratios': [1.2, 1]})
+    logtestm1s = np.log(TEST_M1S)
+    rate = 0.
+    for i in range(NSAMPS):
+        rate += np.trapz(r[i][2:200],logtestm1s[2:200])
+        axes[0].plot(TEST_M1S, r[i], lw=0.2, c="blue",alpha=0.05)
+    rate /= NSAMPS
+    axes[0].plot([], lw=0.2, c="blue",alpha=0.5,label='Gaussian process fit')
+    
+    norm = np.trapz(powerlaw_peak(TEST_M1S,alpha=-2.7,f_peak=0.05,mMax=78.0,mMin=10.0,mu_m1=30.0,sig_m1=7.0),TEST_M1S)
+    axes[0].plot(TEST_M1S, rate*powerlaw_peak(TEST_M1S,alpha=-2.7,f_peak=0.05,mMax=78.0,mMin=10.0,mu_m1=30.0,sig_m1=7.0)/norm*TEST_M1S,
[…]

### 2024-03-09 src/scripts/nonparametric_corner.py
add corner plot of Omega_m and H0, plus H(z) plot

@@ -1,0 +1,101 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import arviz as az
+from scipy.stats import gaussian_kde
+
+from data_generation import H0_FID, OM0_FID
+from nonparametric_inference import TEST_Z
+from gwcosmo import E_of_z
+
+import paths
+plt.style.use(paths.scripts / "matplotlibrc")
+
+def calc_Hz(z,H0,Om0):
+    E = E_of_z(z=z,Om0=Om0)
+    return H0 * E
+
+def Hz(id, ax=None,save=False,inset=True):
+    samples = id.posterior
+    try:
+        Om0=samples['Om0'][0].values
+        H0=samples['H0'][0].values
+        nsamps=len(H0)
+        if ax is None:
+            fig, ax = plt.subplots(figsize=(8,8))
+        H_z = np.zeros((nsamps,len(TEST_Z)))
+        for i in range(nsamps):
+            H_z[i] = calc_Hz(TEST_Z,H0[i],Om0[i])
+            ax.plot(TEST_Z,H_z[i],lw=0.2, c="blue",alpha=0.1)
+        ax.plot(TEST_Z, calc_Hz(TEST_Z,H0_FID,OM0_FID),c='k', lw=2,label="injected value")
+        ax.set_xlabel("$z$")
+        ax.set_ylabel("$H(z)$ [km/s/Mpc]")
+        ax.set_xlim(0,5)
+        ax.set_ylim(0,750)
+        ax.legend(framealpha=0,loc='lower right')
+        
+        if inset:
+            # find the best-measured redshift
+            a = np.argmin(H_z.std(axis=0))
+            zbest=TEST_Z[a]
+            H_of_zbest = H_z[:,a]
+
+            # calculate the truth there
+            H_of_zbest_true = calc_Hz(zbest,H0_FID,OM0_FID)            
+
+            # make the inset
+            iax = ax.inset_axes(bounds=[0.5,450,1.75,250],transform=ax.transData)
+            iax.set_xlabel("$H(z=%1.1f)$ [km/s/Mpc]"%zbest)
+            # arrow from the inset to zbest
+            # midpoint_x = 1.75/2 + 0.5
+            # offset_y= 50
+            # lowpoint_y = 450-offset_y
+            # ax.arrow(x=zbest,y=H_of_zbest+offset_y,dx=midpoint_x-zbest,dy = lowpoint_y-(H_of_zbest+offset_y+10),
[…]

### 2024-03-10 src/scripts/nonparametric_corner.py
cosmetic plot changes

@@ -44,7 +44,9 @@ def Hz(id, ax=None,save=False,inset=True):
 
             # make the inset
             iax = ax.inset_axes(bounds=[0.5,450,1.75,250],transform=ax.transData)
-            iax.set_xlabel("$H(z=%1.1f)$ [km/s/Mpc]"%zbest)
+            iax.set_xlabel("$H(z=%1.1f)$\n[km/s/Mpc]"%zbest,fontsize=10)
+            iax.set_xticks([100,125,150])
+            # iax.set_ylabel('posterior density',fontsize=8)
             # arrow from the inset to zbest
             # midpoint_x = 1.75/2 + 0.5
             # offset_y= 50
@@ -57,7 +59,6 @@ def Hz(id, ax=None,save=False,inset=True):
             kde=gaussian_kde(H_of_zbest)
             iax.plot(prior,kde(prior))
             iax.axvline(H_of_zbest_true,color='k',label="True value")
-            iax.set_ylabel('posterior density')
             iax.tick_params(left=False,labelleft=False)
 
         if save:
@@ -82,20 +83,23 @@ def H0_Om_corner(id):
     az.plot_pair(id,var_names=['H0','Om0'],marginals=True,kind='kde', ax=axsLeft,
                  kde_kwargs={'plot_kwargs':{'color':'b'},'contourf_kwargs':{'cmap':'Blues'}},
                  reference_values={'Om0':OM0_FID,'H0':H0_FID},
-                 reference_values_kwargs={'marker':'+','color':'k','ms':20,'mew':5})
+                 reference_values_kwargs={'marker':'+','color':'k','ms':10,'mew':2})
     # un-rotate Om0 plot
     axsLeft[1,1].cla()
+    axsLeft[0,0].tick_params(bottom=False,labelbottom=False)
     az.plot_dist(samples['Om0'][0],ax=axsLeft[1,1])
     axsLeft[1,1].set_xlabel("$\Omega_m$")
+    axsLeft[1,1].set_xticks([0,0.5,1])
+    axsLeft[1,0].set_xticks([50,75,100])
     # label axes
     axsLeft[1,0].set_xlabel("$H_0$ [km/s/Mpc]")
     axsLeft[1,0].set_ylabel("$\Omega_m$")
-    axsLeft[1,0].plot([],[],**{'marker':'+','color':'k','ms':15,'mew':3,'lw':0},label='injected value')
[…]

### 2024-03-10 src/scripts/nonparametric_twopanel.py

@@ -49,12 +49,12 @@ def two_panel(path, hyperparam='H0'):
     axes[1].plot(prior,kde(prior))
     axes[1].set_xlabel(hyperparam)
     if hyperparam == 'H0':
-        axes[1].axvline(H0_FID,color='k',ls='--',label="True value")
+        axes[1].axvline(H0_FID,color='k',label="True value")
         axes[1].set_xlabel('$H_0$ [km/s/Mpc]')
-        axes[1].legend(framealpha=0)
+        # axes[1].legend(framealpha=0, loc="upper right")
     axes[1].set_ylabel('posterior density')
     plt.tight_layout()
     fig.savefig(paths.figures / "O5_GP_pm.pdf")
     plt.clf()
 
-two_panel(paths.data / "logm_gwmc_O5_errs_oldPEprior_516_samples_interpolant.nc4")
+two_panel(paths.data / "mcmc_nonparametric.nc4")

### 2024-03-10 src/scripts/nonparametric_corner.py
plot layouts

@@ -74,7 +74,7 @@ def H0_Om_corner(id):
         samples['Om0']
     except KeyError:
         pass
-    fig = plt.figure(layout='constrained',figsize=(7.5,4*.75),facecolor='none')
+    fig = plt.figure(figsize=(7.5,4*.75),facecolor='none')
     subfigs = fig.subfigures(1,2,wspace=0.07,width_ratios=[1.2, 1])
     for sf in subfigs:
         sf.set_facecolor('none')

### 2024-03-10 src/scripts/nonparametric_corner.py
save best measured z

@@ -38,6 +38,8 @@ def Hz(id, ax=None,save=False,inset=True):
             a = np.argmin(H_z.std(axis=0))
             zbest=TEST_Z[a]
             H_of_zbest = H_z[:,a]
+            with open(paths.output / "mostsensitivez.txt", "w") as f:
+                print(f"{zbest:.1f}", file=f)
 
             # calculate the truth there
             H_of_zbest_true = calc_Hz(zbest,H0_FID,OM0_FID)            
@@ -94,8 +96,8 @@ def H0_Om_corner(id):
     # label axes
     axsLeft[1,0].set_xlabel("$H_0$ [km/s/Mpc]")
     axsLeft[1,0].set_ylabel("$\Omega_m$")
-    axsLeft[1,0].plot([],[],**{'marker':'+','color':'k','ms':10,'mew':2,'lw':0},label='injected value')
-    axsLeft[1,0].legend(framealpha=0,handletextpad=0.1)
+    # axsLeft[1,0].plot([],[],**{'marker':'+','color':'k','ms':10,'mew':2,'lw':0},label='injected value')
+    # axsLeft[1,0].legend(framealpha=0,handletextpad=0.1)
     
     axsRight=subfigs[1].subplots(1,1)
     Hz(id,ax=axsRight,save=False)

### 2024-03-10 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 \usepackage{acro}
+\acsetup{patch/longtable=false}
 
-\DeclareAcronym{\ac{LVK}}{short=\ac{LVK}, long=LIGO Virgo KAGRA}
+\DeclareAcronym{LVK}{short=LVK, long=LIGO Virgo KAGRA}
 \DeclareAcronym{O5}{short=O5, long=the fifth LIGO Virgo KAGRA observing run}
@@
 % The source frame mass is different for each possible value of \Ho{} and \Omm{}, so the hyper-parameters are defined in different locations for each draw from the hyper-posterior.
-We therefore have $N_{\text{evs}}M + N_{\text{inj}}$ mass hyper-parameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of PE samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
+We therefore have $N_{\text{evs}}M + N_{\text{inj}}$ mass hyper-parameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of posterior samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
 We label these $\{\mathcal{R}_{ij},\mathcal{R}_k\}$, and our full set of hyper-parameters is $\Lambda=\{H_0, \Omega_M, \mathcal{R}_{ij},\mathcal{R}_k\}$. 
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{figures/CE_GP.pdf}
-    \caption{Two-panel figure of (A) $w$, and \Ho{} corner plot and (B)$H(z)$ for CE. We can include a posterior of $H(z=z_{\text{best measured}})$ as an inset to the $H(z)$ plot.}
-    \label{fig:XG_corner}
-    \script{XG_corner.py}
-\end{figure}
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\textwidth]{figures/CE_GP.pdf}
+%     \caption{Two-panel figure of (A) $w$, and \Ho{} corner plot and (B)$H(z)$ for CE. We can include a posterior of $H(z=z_{\text{best measured}})$ as an inset to the $H(z)$ plot.}
+%     \label{fig:XG_corner}
+%     \script{XG_corner.py}
+% \end{figure}
 

### 2024-03-10 src/scripts/data_generation.py
update to O5 errors

@@ -89,11 +89,32 @@ if  __name__ == "__main__":
     np.save(paths.data / "gw_data/log_pinj_det.npy",log_pinj_det)
     
     ## find events and generate mock PE
-    m1z_PE, _, dL_PE, log_PE_prior = gen_snr_scaled_PE(np_rng,m1s_true,m1s_true,dL_true/1000,osnr_interp,
-                                                       reference_distance,N_SAMPLES_PER_EVENT,H0_FID,OM0_FID,
-                                                       return_og=False)
+    m1z_PE, m2z_PE, dL_PE, log_PE_prior = gen_snr_scaled_PE(np_rng,m1s_true,m1s_true,dL_true/1000,osnr_interp,
+                                                            reference_distance,N_SAMPLES_PER_EVENT,H0_FID,OM0_FID,
+                                                            # errors taken from Jose's "gwutils.py" for O5
+                                                            mc_sigma=3.0e-2,eta_sigma=5.0e-3,theta_sigma=5.0e-2, 
+                                                            return_og=False)
 
     dL_PE *= 1000 # unit matching
+
+    # cut out samples below SNR interpolation range. 
+    # It is very unlikely that any samples are down there, but we do this just in case.
+    # this does not change the prior shape, only the normalization
+    n_samps = np.inf
+    for i in range(len(m1z_PE)): # for each event
+        m1z_PE[i], m2z_PE[i], dL_PE[i], log_PE_prior[i] = m1z_PE[i,m2z_PE[i]>0.05], \
+            m2z_PE[i,m2z_PE[i]>0.05], dL_PE[i,m2z_PE[i]>0.05], log_PE_prior[i,m2z_PE[i]>0.05]
+        if len(m1z_PE[i]) < n_samps:
+            n_samps = len(m1z_PE[i])
+    print(n_samps)
+    print(np.mean(np.diff(np.sort(np.log(m1z_PE).mean(axis=1)))))
+    # make all events have the same # of samples
+    if n_samps < N_SAMPLES_PER_EVENT:
+        for i in range(len(m1z_PE)):
+            m1z_PE[i], m2z_PE[i], dL_PE[i], log_PE_prior[i] = m1z_PE[i,:n_samps], \
+                m2z_PE[i,:n_samps], dL_PE[i,:n_samps], log_PE_prior[i,:n_samps]
+        N_SAMPLES_PER_EVENT = n_samps
+    
[…]

### 2024-03-10 src/scripts/data_generation.py
try making numbers from data_generation script

@@ -106,8 +106,6 @@ if  __name__ == "__main__":
             m2z_PE[i,m2z_PE[i]>0.05], dL_PE[i,m2z_PE[i]>0.05], log_PE_prior[i,m2z_PE[i]>0.05]
         if len(m1z_PE[i]) < n_samps:
             n_samps = len(m1z_PE[i])
-    print(n_samps)
-    print(np.mean(np.diff(np.sort(np.log(m1z_PE).mean(axis=1)))))
     # make all events have the same # of samples
     if n_samps < N_SAMPLES_PER_EVENT:
         for i in range(len(m1z_PE)):
@@ -118,3 +116,7 @@ if  __name__ == "__main__":
     np.save(paths.data / "gw_data/m1z_PE.npy",m1z_PE)
     np.save(paths.data / "gw_data/dL_PE.npy",dL_PE)
     np.save(paths.data / "gw_data/log_PE_prior.npy",log_PE_prior)
+    with open(paths.output / "num_found_events.txt", "w") as f:
+        print(len(dL_PE), file=f)
+    with open(paths.output / "H0_FID.txt", "w") as f:
+        print(H0_FID, file=f)

### 2024-03-10 src/scripts/data_generation.py
Revert "try making numbers from data_generation script"
This reverts commit fb72aafb1f4150a6b03a9e106232483f46dc6117.

@@ -106,6 +106,8 @@ if  __name__ == "__main__":
             m2z_PE[i,m2z_PE[i]>0.05], dL_PE[i,m2z_PE[i]>0.05], log_PE_prior[i,m2z_PE[i]>0.05]
         if len(m1z_PE[i]) < n_samps:
             n_samps = len(m1z_PE[i])
+    print(n_samps)
+    print(np.mean(np.diff(np.sort(np.log(m1z_PE).mean(axis=1)))))
     # make all events have the same # of samples
     if n_samps < N_SAMPLES_PER_EVENT:
         for i in range(len(m1z_PE)):
@@ -116,7 +118,3 @@ if  __name__ == "__main__":
     np.save(paths.data / "gw_data/m1z_PE.npy",m1z_PE)
     np.save(paths.data / "gw_data/dL_PE.npy",dL_PE)
     np.save(paths.data / "gw_data/log_PE_prior.npy",log_PE_prior)
-    with open(paths.output / "num_found_events.txt", "w") as f:
-        print(len(dL_PE), file=f)
-    with open(paths.output / "H0_FID.txt", "w") as f:
-        print(H0_FID, file=f)

### 2024-03-10 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 The result is shown in Figure~\ref{fig:O5_corner}.
-We find \ac{O5} observations to be most sensitive to \result{$H(z=X)$}, which results in similarly informative constraints on \Omm{} as on \Ho.
+We find \ac{O5} observations to be most sensitive to \result{$H(z=\variable{output/mostsensitivez.txt})$}, which results in similarly informative constraints on \Omm{} as on \Ho.
 \begin{itemize}
@@
 Thus, the additional statistical uncertainty introduced by non-parametric approaches may be preferable to the systematic errors associated with choosing a parametric model.
-    \item additionally, in \ac{O5} we will be most sensitive to \result{$z=X$}.
+    \item additionally, in \ac{O5} we will be most sensitive to $z=\variable{output/mostsensitivez.txt}$.
 This is larger than we expect to see EM counterparts to, and larger than we expect galaxy surveys to be complete to (is this true?) so upgrades to current facilities will allow spectral sirens to probe H(z) to relatively unexplored redshifts.
@@
 
-After passing the simulated events through projected detector selection effects, the resulting catalog has \variable{output/num_found_events.txt}events, including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} \ac{BBH} systems, consistent with the numbers projected for \ac{O5} by \citet{kiendrebogo_observing_2023}.
+After passing the simulated events through projected detector selection effects, the resulting catalog has \variable{output/num_found_events.txt}events, %including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} \ac{BBH} systems,
+consistent with the numbers projected for \ac{O5} by \citet{kiendrebogo_observing_2023}.
 We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from \ac{O5} detectors.

### 2024-03-10 src/scripts/mock_posteriors.py
let sky angle errors be input

@@ -21,14 +21,14 @@ def gen_jittered_PE(rng, true_vals,sigma,bounds,n_samples,n_detections=None,retu
         return mock_samples
     
 def gen_snr_scaled_PE(rng, true_m1s, true_m2s, true_dl, osnr_interp, reference_distance, n_samples,
-                      H0, Om0, mc_sigma=0.04,eta_sigma=0.01,snr_thresh=8, return_og=False):
+                      H0, Om0, mc_sigma=0.04,eta_sigma=0.01,theta_sigma=0.15,snr_thresh=8, return_og=False):
     """ Wrapper function for GWMockCat functions"""
     from GWMockCat import posterior_utils, cosmo_utils
     import astropy.cosmology as ap_cosmology 
     from astropy import units as u
     import numpy as np
 
-    uncert_gwmc = {'threshold_snr': snr_thresh, 'snr': 1.0, 'mc': mc_sigma, 'Theta': 0.15, 'eta': eta_sigma}
+    uncert_gwmc = {'threshold_snr': snr_thresh, 'snr': 1.0, 'mc': mc_sigma, 'Theta': theta_sigma, 'eta': eta_sigma}
     cosmo = ap_cosmology.FlatLambdaCDM(H0=H0, Om0=Om0)
     cosmo_gwmc = cosmo_utils.interp_cosmology(zmin=0.,zmax=11.,cosmology=cosmo,dist_unit=u.Unit(reference_distance[-3:]))
     true_z = cosmo_gwmc['z_at_dL'](true_dl)
@@ -46,8 +46,10 @@ def gen_snr_scaled_PE(rng, true_m1s, true_m2s, true_dl, osnr_interp, reference_d
     # transform to useful parameters
     m1det_PE, m2det_PE = posterior_utils.m1m2_from_mceta(mc_det,eta)
     dl_PE = theta * osnr_interp(m1det_PE,m2det_PE,grid=False)/rho
-    log_jacobian = - np.log(posterior_utils.dm1m2_dMceta(m1det_PE,m2det_PE)) -np.log(posterior_utils.ddL_drho(osnr_interp,m1det_PE,m2det_PE,rho,theta))
-    log_prior_PE = np.log(0.25*mc_det.max()*rho.max()) + log_jacobian 
+    # transform prior to be represented in terms of det frame m1, m2 and dl
[…]

### 2024-03-10 src/scripts/data_generation.py
test different random seed

@@ -30,7 +30,7 @@ SNR_THRESH=8.
 NUM_INJ=N_SOURCES*50
 
 # random number generators
-np_rng = np.random.default_rng(4242)
+np_rng = np.random.default_rng(516)
 
 def true_vals_PLP(rng, n_sources=N_SOURCES):
     m1s_true = inverse_transform_sample(gwpop.powerlaw_peak,[1,400],rng,N=n_sources,

### 2024-03-10 src/scripts/nonparametric_corner.py
consistent number of hyperparam samples

@@ -4,7 +4,7 @@ import arviz as az
 from scipy.stats import gaussian_kde
 
 from data_generation import H0_FID, OM0_FID
-from nonparametric_inference import TEST_Z
+from nonparametric_inference_fitOm0 import TEST_Z, NSAMPS
 from gwcosmo import E_of_z
 
 import paths
@@ -19,56 +19,56 @@ def Hz(id, ax=None,save=False,inset=True):
     try:
         Om0=samples['Om0'][0].values
         H0=samples['H0'][0].values
-        nsamps=len(H0)
-        if ax is None:
-            fig, ax = plt.subplots(figsize=(8,8))
-        H_z = np.zeros((nsamps,len(TEST_Z)))
-        for i in range(nsamps):
-            H_z[i] = calc_Hz(TEST_Z,H0[i],Om0[i])
-            ax.plot(TEST_Z,H_z[i],lw=0.2, c="blue",alpha=0.1)
-        ax.plot(TEST_Z, calc_Hz(TEST_Z,H0_FID,OM0_FID),c='k', lw=2,label="injected value")
-        ax.set_xlabel("$z$")
-        ax.set_ylabel("$H(z)$ [km/s/Mpc]")
-        ax.set_xlim(0,5)
-        ax.set_ylim(0,750)
-        ax.legend(framealpha=0,loc='lower right')
-        
-        if inset:
-            # find the best-measured redshift
-            a = np.argmin(H_z.std(axis=0))
-            zbest=TEST_Z[a]
-            H_of_zbest = H_z[:,a]
-            with open(paths.output / "mostsensitivez.txt", "w") as f:
-                print(f"{zbest:.1f}", file=f)
-
-            # calculate the truth there
-            H_of_zbest_true = calc_Hz(zbest,H0_FID,OM0_FID)            
-
-            # make the inset
-            iax = ax.inset_axes(bounds=[0.5,450,1.75,250],transform=ax.transData)
-            iax.set_xlabel("$H(z=%1.1f)$\n[km/s/Mpc]"%zbest,fontsize=10)
-            iax.set_xticks([100,125,150])
-            # iax.set_ylabel('posterior density',fontsize=8)
-            # arrow from the inset to zbest
-            # midpoint_x = 1.75/2 + 0.5
-            # offset_y= 50
-            # lowpoint_y = 450-offset_y
[…]

### 2024-03-10 src/scripts/nonparametric_inference_fitOm0.py

@@ -3,7 +3,8 @@ import jgwcosmo # for inference
 import jgwpop
 import paths
 Clight = jgwcosmo.Clight
-from data_generation import NUM_INJ, OM0_FID, ZMAX, ZMIN, N_SAMPLES_PER_EVENT
+from data_generation import NUM_INJ, ZMAX, ZMIN, N_SAMPLES_PER_EVENT
+from nonparametric_inference import TEST_Z, TEST_M1S, H0_PRIOR_MIN, H0_PRIOR_MAX, NSAMPS, remove_low_Neff
 
 # inference
 import jax
@@ -18,21 +19,10 @@ import arviz as az
 import numpy as np
 
 jax.config.update("jax_enable_x64", True)
-
-# Constants
-TEST_Z = jnp.linspace(ZMIN,ZMAX,num=100)
-TEST_M1S = jnp.linspace(0,250.,num=500)
-
-H0_PRIOR_MIN=30.
-H0_PRIOR_MAX=120.
-
-NSAMPS=5
-
 # random number generators
 jax_rng = jax.random.PRNGKey(425)
 
 # options
-remove_low_Neff=False
 
 def model(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,remove_low_Neff=False):
     mean = numpyro.sample("mean",dist.Normal(0,3))
@@ -40,7 +30,7 @@ def model(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,remove_low_Neff=Fal
     rho = numpyro.deterministic("rho",3.)
 
     H0 = numpyro.sample("H0",dist.Uniform(H0_PRIOR_MIN,H0_PRIOR_MAX))
-    Om0=OM0_FID
+    Om0 = numpyro.sample("Om0",dist.Uniform(0.01,0.99))
 
     # construct GP in the source frame
     kernel = sigma**2 * kernels.quasisep.Matern52(rho) # can change kernel type
@@ -116,4 +106,4 @@ if  __name__ == "__main__":
 
     # save results
     id = az.from_numpyro(mcmc)
-    id.to_netcdf(paths.data / "mcmc_nonparametric.nc4")
+    id.to_netcdf(paths.data / "mcmc_nonparametric_fitOm0.nc4")

### 2024-03-10 src/scripts/nonparametric_twopanel.py

@@ -3,12 +3,11 @@ import numpy as np
 import arviz as az
 from scipy.stats import gaussian_kde
 
-from nonparametric_inference import TEST_M1S
+from nonparametric_inference import TEST_M1S, NSAMPS
 from data_generation import H0_FID
 from gwpop import powerlaw_peak
 
 import paths
-NSAMPS = 500
 
 plt.style.use(paths.scripts / "matplotlibrc")
 

### 2024-03-10 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 We do this by constructing a flexible, Gaussian-process based model for \ac{GW} source population to analyze simulated \ac{GW} data consistent with expectations for the next \acl{LVK} observing run.
-We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a \variable{output/nonparh0percent.txt}\% measurement.
+We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a \result{X\%} measurement of $H(z=\variable{output/mostsensitivez.txt})$.
 This astrophysics-agnostic spectral siren technique will be key to deliver precise and unbiased cosmological constraints with future observations in the presence of possible redshift evolution of the mass spectrum.
@@
 We assume a uniform-in-comoving volume redshift distribution, as in Equation~\ref{eq:underlying redshift dist}.
-We assume $m_1=m_2$ and model the distribution of primary masses using a Gaussian process (described in Section~\ref{sec:model}).
+We assume $m_1=m_2$ and model the distribution of primary masses using a \acl{Gaussian process} (described in Section~\ref{sec:model}).
 
@@
 In this way, our GP-based mass distribution is similar to the autoregressive population models used in ~\citet{callister_parameter-free_2023}.
-Indeed, an autoregressive model is a Gaussian process with a particular choice of kernel.
+Indeed, an autoregressive model is a \ac{GP} with a particular choice of kernel.
 
@@
 We then demonstrate that our flexible model alleviates this bias without the need to know the mass distribution's morphology \emph{a priori}.
-To illustrate this, we generate a catalog of \variable{output/num_found_events.txt}GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
+To illustrate this, we generate a catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
 This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after \ac{O5}.
@@
 \begin{enumerate}
-    \item the \plp{} model, which includes the true mass distribution within its hyperprior
+    \item the \plp{} model, which includes the true mass distribution within its hyperprior,
     \item a broken power law, which does not include the true mass distribution within its hyperprior, and 
-    \item the flexible, Gaussian process-based model described in Section~\ref{sec:model}, which can closely approximate the morphology of the true mass distribution, along with several other morphologies.
+    \item the flexible, \ac{GP}-based model described in Section~\ref{sec:model}, which is able to closely approximate the morphology of the true mass distribution, along with many other morphologies.
 \end{enumerate}
@@
 
-With \variable{output/num_found_events.txt}events, we find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within \result{X-$\sigma$}.
+We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within \result{X-$\sigma$}.
 However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by \result{X}-$\sigma$.
-This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the Gaussian process-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution.
-The Gaussian process-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters. 
-In reality, it is impossible to know the true functional form of the mass distribution, so the additional statistical uncertainty introduced by non-parametric approaches may be desirable over the systematic errors associated with choosing a parametric model.
+This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the \ac{GP}-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution.
+The \ac{GP}-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters. 
+In reality, it is impossible to know the true functional form of the mass distribution, so the additional statistical uncertainty introduced by non-parametric approaches may be desirable over the systematic errors introduced by choosing a parametric model.
 
-We have repeated this analysis with \result{50} separate simulated catalogs, and found that the broken power law model causes us to infer a \result{higher/lower} value of \Ho{} at greater than \result{X}-$\sigma$, \result{Y}\% of the time, whereas the Gaussian process-based model reaches that level of bias only \result{Z}\% of the time, and the \plp{} model reaches the same level of bias \result{A}\% of the time.
+The mass distribution inferred with the \ac{GP}-based model closely resembles the true, simulated distribution.%, and the injected value of \Ho{} ($\variable{output/H0_FID.txt}\Hunits$) is consistent with its inferred posterior ($\variable{output/nonparh0CI.txt}\Hunits$).
+At very high and very low masses, there is both a lack of data and little detector sensitivity.
+The mass distribution is therefore uninformed in this region and the \ac{GP} reverts to its prior distribution, which resembles random scatter around the mean differential merger rate.
+Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
+On the other hand, the lack of data just below the minimum \ac{BH} mass and just above the maximum \ac{BH} mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the \ac{GP} to learn a relatively low merger rate at the edges of the mass distribution.
+Indeed, these features are very informative to the \Ho{} constraint \citep{the_ligo_scientific_collaboration_constraints_2021}.
+The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
+Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,60\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
+
+We have repeated this analysis with \result{50} separate simulated catalogs, and found that the broken power law model produces an \result{overestimate} of \Ho{} at greater than \result{X}-$\sigma$, \result{Y}\% of the time, whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and the \plp{} model reaches the same level of bias \result{A}\% of the time.
 This indicates that prior knowledge of the shape of the mass distribution is not required to perform a spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.
 
-\jme{[I think it would be relevant to compare the forecast of the non-parametric and the right parametric model.
-Do we loose constrianing power by being flexible (probably)?
-how much?]} \comment{[I have now included this in the discussion.]}
-
 \subsection{Option 2: GP only}
-We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated \ac{GW} catalog of \variable{output/num_found_events.txt}GW events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
+We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated \ac{GW} catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
 This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after \ac{O5}.
@@
 The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
-The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho (\variable{output/H0_FID.txt} km/s/Mpc) is consistent with its inferred posterior (\variable{output/nonparh0CI.txt} \,km/s/Mpc).
+The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho{} ($\variable{output/H0_FID.txt}\Hunits$) is consistent with its inferred posterior ($\variable{output/nonparh0CI.txt}\Hunits$).
 At very high and very low masses, there is both a lack of data and little detector sensitivity.
-The mass distribution is therefore uninformed in this region and the GP reverts to its prior, which resembles random scatter around a central value.
+The mass distribution is therefore uninformed in this region and the \ac{GP} reverts to its prior distribution, which resembles random scatter around the mean differential merger rate.
 Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
-On the other hand, the lack of data just below the minimum BH mass and just above the maximum BH mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the GP to learn a relatively low merger rate at the edges of the mass distribution.
+On the other hand, the lack of data just below the minimum \ac{BH} mass and just above the maximum \ac{BH} mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the GP to learn a relatively low merger rate at the edges of the mass distribution.
 Indeed, these features are very informative to the \Ho{} constraint \citep{the_ligo_scientific_collaboration_constraints_2021}.
 The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
-Additionally, the posterior on \Ho{} is distinct from its prior distribution, indicating that the data is informative despite the flexibility of the population model.
+Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,60\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
 
 \subsection{Projections for Future Measurements}
-This result implies that we can expect to measure \Ho{} to \variable{output/nonparh0percent.txt}\% by the end of \ac{O5} using the method presented in this work.
+The fit shown in Figure~\ref{fig:O5_GP} implies that we can expect to measure \Ho{} to \variable{output/nonparh0percent.txt}\% by the end of \ac{O5} using the \ac{GP}-based spectral siren method.
 However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\sim 3$, with most sources lying near redshift $\sim 1.2$ \citep{chen_distance_2021}.
 Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 40$.
-This means that future observations of \acp{GW} will be more sensitive $H(z>1)$ than to \Ho, and can therefore constrain several additional cosmological parameters.
+This means that future \acp{GW} observations will be more sensitive $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters.
 We demonstrate this by repeating the same analysis while also fitting for the local matter density, \Omm.
 The result is shown in Figure~\ref{fig:O5_corner}.
-We find \ac{O5} observations to be most sensitive to \result{$H(z=\variable{output/mostsensitivez.txt})$}, which results in similarly informative constraints on \Omm{} as on \Ho.
+We find \ac{O5} observations to be most sensitive to \result{$H(z=\variable{output/mostsensitivez.txt})$}, which results in similarly informative constraints on \Omm{} and \Ho.
+The left panel of Figure~\ref{fig:O5_corner} demonstrates a strong degeneracy between the measurement of \Omm{} and \Ho, which can be explained by the 
 \begin{itemize}
@@
     \includegraphics[width=\textwidth]{figures/O5_GP_pm.pdf}
-    \caption{Two-panel figure of mass distribution and \Ho{} posterior. 
-    Include prior on \Ho{} as alpha=0.5 line.}
+    \caption{Spectral siren measurement of \ac{O5}-like catalog with the ``correct'' parametric model (\plp, green), an ``incorrect'' parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{Gaussian process}, blue).
+    The left panel shows the recovered source frame primary mass distribution for each model, and the simultaneously-inferred posteriors on \Ho{} are shown in the right panel.
+    The mass distribution and \Ho{} value used to generate the data are shown by a solid black line in each panel. 
+    The ``incorrect'' parametric model fails to recover the true mass distribution and therefore produces a biased estimate of \Ho{}, whereas both the ``correct'' and non-parametric models recover  the mass distribution and \Ho{}.
+    As the true mass distribution is unknown, using a non-parametric model mitigates the systematic uncertainty from mismodeling the \ac{CBC} population, though it does introduce additional statistical uncertainty.
+    }
     \label{fig:O5_GP}
@@
     \includegraphics[width=\textwidth]{figures/O5_GP_corner.pdf}
-    \caption{Two-panel figure of (A) \Omm{} and \Ho{} corner plot and (B)$H(z)$ for O5.
+    \caption{
+
+    Two-panel figure of (A) \Omm{} and \Ho{} corner plot and (B)$H(z)$ for O5.
 We can include a posterior of $H(z=z_{\text{best measured}})$ as an inset to the $H(z)$ plot.
@@
     \caption{Performance of various mass distribution models used for measuring the Hubble constant, \Ho.
-    We consider the correct parametric model, an incorrect parametric model, and the non-parametric, Gaussian process-based model presented in this work.
+    We consider the correct parametric model, an incorrect parametric model, and the non-parametric, \ac{GP}-based model presented in this work.
     Bias is defined as the number of standard deviations between the true injected value of \Ho and the recovered posterior mean. 

### 2024-03-11 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 \author{Daniel E. Holz}
+\affiliation{Department of Physics, University of Chicago, Chicago, IL 60637, USA}
 \affiliation{Kavli Institute for Cosmological Physics and Enrico Fermi Institute, The University of Chicago, Chicago, IL 60637, USA}
-\affiliation{Department of Physics, Department of Astronomy \& Astrophysics, The University of Chicago, Chicago, IL 60637, USA}
+\affiliation{Department of Astronomy \& Astrophysics, The University of Chicago, Chicago, IL 60637, USA}
 
@@
 We demonstrate that unbiased inference of cosmological parameters is possible without any assumptions for the mass distribution's morphology.
-We do this by constructing a flexible, Gaussian-process based model for \ac{GW} source population to analyze simulated \ac{GW} data consistent with expectations for the next \acl{LVK} observing run.
-We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a \result{X\%} measurement of $H(z=\variable{output/mostsensitivez.txt})$.
+We do this by constructing a flexible, Gaussian process-based model for \ac{GW} source population to analyze simulated \ac{GW} data consistent with expectations for the next \acl{LVK} observing run.
+We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a \result{X\%} measurement of $H(z=\variable{output/mostsensitivez.txt})$ ($1-\sigma$).
 This astrophysics-agnostic spectral siren technique will be key to deliver precise and unbiased cosmological constraints with future observations in the presence of possible redshift evolution of the mass spectrum.
@@
 \label{sec:intro}
-Like light, \acp{GW} are redshifted as they propagate across the universe.
-Unlike light, however, the selection effects of \acp{GW} are extremely well understood, allowing for a precise estimate of each catalogs' completeness and an unbiased measurement of the true \ac{GW} population.
-This allows \ac{GW} signals to be used as relatively clean probes of cosmological parameters \citep[e.g.][]{}%bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide.
+Like light, \acp{GW} are redshifted as they propagate across the universe\jme{, receiving an imprint of the cosmic expansion history}.
+Unlike light, however, \jme{GW signals are known from first principles and their} selection effects of \acp{GW} are extremely well understood, allowing for a precise estimate of each catalogs' completeness and an unbiased measurement of the true \ac{GW} population\jme{, including the absolute distance to the sources}.
+This allows \ac{GW} signals to be used as \jme{``standard sirens'', becoming powerful} %relatively clean 
+probes of cosmological parameters. 
+\jme{When complemented with redshift information about the host galaxy, as in the multi-messenger GW170817, ... }
+\citep[e.g.][]{}%bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide.
 
-GW signals provide direct measurements of each source's luminosity distance and redshifted, or detector frame, masses, $m_{\det}= m_{\source}(1+z)$.
+\ac{GW} signals provide direct measurements of each source's luminosity distance \jme{$d_L$} and redshifted, or detector frame, masses, $m_{\det}= m_{\source}(1+z)$.
 Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the \ac{GW} signal was emitted, $H(z)$.
@@
 The method of using the mass distribution of \ac{GW} sources to measure cosmological parameters has therefore been coined ``spectral sirens'' \citep{ezquiaga_spectral_2022}.
-Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff_gravitational_1993} using the binary neutron star mass distribution, and extended to the \ac{BBH} mass distribution by \cite{farr_future_2019}.
+Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff_gravitational_1993,Taylor:2011fs} using the binary neutron star mass distribution, and extended to the \ac{BBH} mass distribution by \cite{farr_future_2019}.
+\jme{Spectral siren analyses have now been implemented by the LIGO--Virgo--KAGRA Collaborations using the latest catalog \cite{LIGOScientific:2021aug}.}
 
@@
 theoretical expectations of the location of a pulsational-pair instability pileup, upper mass gap, or maximum neutron star mass.
-However, the information in the spectral siren measurement comes solely from the assumption that, at a given luminosity distance, all \ac{CBC} mergers follow the same mass distribution.
+\jme{This would certainly be problematic, as mismodeling of the specific feature would hinder the accuracy of the method and introduce biases in the cosmological inference \cite{ezquiaga_spectral_2022,Mukherjee:2021rtw,Pierra:2023deu}.}
+However, the information in the spectral siren measurement comes solely from the assumption that all \acp{CBC} follow the same mass distribution.\footnote{
+The spectral siren method is also valid if \acp{CBC} have different mass distributions at different redshifts, so long as the evolution of the mass distribution does not exactly mimic cosmology, and the model for the mass distribution allows for evolution \citep{ezquiaga_spectral_2022}.
+\jme{[maybe is worth having this comment in the main text?]}
+}
 Therefore the location, shape, and existence of all features in the mass distribution can be inferred simultaneously with $H(z)$. 
+\jme{[do we want to comment somewhere about the observed features in the data? and perhaps quickly mention about their robustenss?]}
 
@@
 We find that using this model allows for a \variable{output/nonparh0percent.txt}\% measurement of \Ho{} by the end of \ac{O5}, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
-While constraints on \Ho are used in this work to quantify the accuracy and precision of our non-parametric method, we emphasize that spectral sirens are able to constrain $H(z)$ out large redshifts that are inaccessible by electromagnetic observations, especially with next-generation detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn}\footnote{See example calculations here: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}}.
+While constraints on \Ho{} are used in this work to quantify the accuracy and precision of our non-parametric method, we emphasize that spectral sirens are able to constrain $H(z)$ out to redshifts that are inaccessible by electromagnetic observations, especially with next-generation detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn,You:2020wju}\footnote{See example \jme{spectral siren cosmological inference using parametric models for the mass spectrum here}: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}.
 Applied to data from next-generation detectors, this method will therefore be able to constrain the dark energy equation of state parameter, $w$, at redshifts $\gtrsim 2$.
 
-\jme{[cite joint multi-siren forecast \cite{Chen:2024gdn}. Cite code \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}?]}\\ 
-\comment{[Should I add a separate paragraph for other sirens or is it sufficient to just cite it above with other forecasts, as I did?]}
+%\jme{[cite joint multi-siren forecast \cite{Chen:2024gdn}. Cite code \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}?]}\\ 
+\comment{[Should I add a separate paragraph for other sirens or is it sufficient to just cite it above with other forecasts, as I did?]} \jme{[I suggest to do it quickly in the first paragraph.
+Then in the discussion maybe we could emphasize that future analysis will be joint as we discussed in \cite{Chen:2024gdn}]}
 
-This paper is organized as follows: \comment{insert outline here.}
+This paper is organized as follows: Section~\ref{sec:ss} explains how cosmological parameters are inferred from the mass distribution of \ac{GW} sources (i.e. the spectral siren method).
+Section~\ref{sec:model} describes the non-parametric mass distribution we develop for use within the spectral siren method.
+We show the results of using parametric and non-parametric mass distributions in Section~\ref{sec:results}, as well as projections for future constraints on $H(z)$.
+In Section~\ref{sec:discussion} we discuss the implications of our results and outline future work.
 
@@
 We assume a uniform-in-comoving volume redshift distribution, as in Equation~\ref{eq:underlying redshift dist}.
-We assume $m_1=m_2$ and model the distribution of primary masses using a \acl{Gaussian process} (described in Section~\ref{sec:model}).
+We assume $m_1=m_2$ and model the distribution of primary masses using a \acl{GP} (described in Section~\ref{sec:model}).
 
@@
 \label{sec:model}
-\Acp{GP} are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
+\Acfp{GP} are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
 Their smoothness properties make them widely useful in \ac{GW} data analysis and beyond for regression problems, such as modeling time-domain waveforms \citep{doctor_statistical_2017, huerta_eccentric_2018} and the neutron star equation of state \citep{landry_nonparametric_2019}, density estimation problems, such as estimating posterior densities of single-event parameters from parameter estimation samples ~\citep{demilio_density_2021}, and as a prior on histogram bin heights for population inference \citep{mandel_model-independent_2017, li_flexible_2021, ray_non-parametric_2023}.
@@
 \end{enumerate}
+\jme{[comment: is there a way to summarize this list in a simple diagram?]}
 
@@
 
-\subsection{Option 1: include parametric models}
 In this Section, we first show that fitting an incorrect functional form to the mass distribution of \acp{CBC} biases the inference of cosmological parameters when using the spectral siren methodology.
@@
 
-\subsection{Option 2: GP only}
-We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated \ac{GW} catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
-This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after \ac{O5}.
-The details of the data simulation are described in Appendix~\ref{ap:data generation}.
-
-The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
-The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho{} ($\variable{output/H0_FID.txt}\Hunits$) is consistent with its inferred posterior ($\variable{output/nonparh0CI.txt}\Hunits$).
-At very high and very low masses, there is both a lack of data and little detector sensitivity.
-The mass distribution is therefore uninformed in this region and the \ac{GP} reverts to its prior distribution, which resembles random scatter around the mean differential merger rate.
-Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
-On the other hand, the lack of data just below the minimum \ac{BH} mass and just above the maximum \ac{BH} mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the GP to learn a relatively low merger rate at the edges of the mass distribution.
-Indeed, these features are very informative to the \Ho{} constraint \citep{the_ligo_scientific_collaboration_constraints_2021}.
-The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
-Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,60\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
-
-\subsection{Projections for Future Measurements}
-The fit shown in Figure~\ref{fig:O5_GP} implies that we can expect to measure \Ho{} to \variable{output/nonparh0percent.txt}\% by the end of \ac{O5} using the \ac{GP}-based spectral siren method.
-However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\sim 3$, with most sources lying near redshift $\sim 1.2$ \citep{chen_distance_2021}.
-Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 40$.
-This means that future \acp{GW} observations will be more sensitive $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters.
-We demonstrate this by repeating the same analysis while also fitting for the local matter density, \Omm.
-The result is shown in Figure~\ref{fig:O5_corner}.
-We find \ac{O5} observations to be most sensitive to \result{$H(z=\variable{output/mostsensitivez.txt})$}, which results in similarly informative constraints on \Omm{} and \Ho.
-The left panel of Figure~\ref{fig:O5_corner} demonstrates a strong degeneracy between the measurement of \Omm{} and \Ho, which can be explained by the 
-\begin{itemize}
-    \item There is a strong degeneracy in these two cosmological parameters, caused by the sensitivity at z=1
-    \item explain why - balance between slope and intercept.
-\end{itemize}
-
-% BONUS
-To demonstrate the capabilities of next-generation detectors, we repeat the full analysis with \result{1000} events indicative of Cosmic Explorer's projected best-measured events within the first month alone.
-We fit the dark energy equation of state parameter, $w$, along with \Ho, and show results in Figure~\ref{fig:XG_corner}.
-We recover unbiased measurements of each parameter and find a \result{X}\% measurement of $w$ and a \result{X}\% measurement of \Ho.
-In comparison, the most recent constraints on these parameters with \result{X experiment} were \result{X}\% and \result{X}\% , respectively.
-This illustrates the unique ability of \acp{GW} to probe the high-redshift universe.
-
-\begin{figure}
+\begin{figure*}
     \centering
     \includegraphics[width=\textwidth]{figures/O5_GP_pm.pdf}
-    \caption{Spectral siren measurement of \ac{O5}-like catalog with the ``correct'' parametric model (\plp, green), an ``incorrect'' parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{Gaussian process}, blue).
+    \caption{Spectral siren measurement for an \ac{O5}-like catalog with the ``correct'' parametric model (\plp, green), an ``incorrect'' parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{Gaussian process}, blue).
     The left panel shows the recovered source frame primary mass distribution for each model, and the simultaneously-inferred posteriors on \Ho{} are shown in the right panel.
@@
     \script{nonparametric_twopanel.py}
-\end{figure}
+\end{figure*}
 
-\begin{figure}
+\subsection{Projections for Future Measurements}
+The fit shown in Figure~\ref{fig:O5_GP} implies that we can expect to measure \Ho{} to \variable{output/nonparh0percent.txt}\% ($1-\sigma$) by the end of \ac{O5} using the \ac{GP}-based spectral siren method.
+However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\sim 3$, with most sources lying near redshift $\sim 1.2$ \citep{chen_distance_2021}.
+Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 40$.
+This means that future \acp{GW} observations will be more sensitive $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters.
+We demonstrate this by repeating the same analysis while also fitting for the local matter density, \Omm.
+For \ac{O5}, this informs $H(z)$ up to $z \sim 3$.
+The result is shown in Figure~\ref{fig:O5_corner}.
+
+We find \ac{O5} observations to be most sensitive to $H(z=\variable{output/mostsensitivez.txt})$, which results in similarly informative constraints on \Omm{} and \Ho.
+The left panel of Figure~\ref{fig:O5_corner} demonstrates a strong anti-correlation between the \Omm{} and \Ho posteriors. 
+This is because \Ho{} controls the $y$-intercept of the $H(z)$ curves on the right panel, while \Omm{} informs the slope of those curves.
+The same measurement of $H(z\neq0)$ can be obtained by increasing the slope while decreasing the $y$-intercept, and vice versa.
+Similar behavior can be observed in measurements of the \ac{BBH} redshift distribution, which exhibit a ``waist'' at $z\sim0.2$ with current observations \citep{abbott_population_2023, callister_parameter-free_2023}.
+
+Next-generation detectors will be sensitive to a much larger range of redshifts \citep{something!} and will therefore break the degeneracy between cosmological parameters, allowing for tighter constraints on each.
+However, the low cosmological volume (and thus low number of mergers) at low redshift will always limit the constraining power of \acp{GW} at $z=0$, making spectral sirens relatively more sensitive to cosmological parameters that affect higher redshifts than to \Ho.
+
+\begin{figure*}
     \centering
@@
     \caption{
-
-    Two-panel figure of (A) \Omm{} and \Ho{} corner plot and (B)$H(z)$ for O5.
-We can include a posterior of $H(z=z_{\text{best measured}})$ as an inset to the $H(z)$ plot.
-Explain the inset and comment on $z_{\text{best measured}}$}
+    Projected constraints on multiple cosmological parameters at the end of \ac{O5} using the \acl{GP}-based spectral siren method.
+    The right panel shows the inferred expansion history of the universe, $H(z)$.
+    It will be measured most precisely at $z=\variable{output/mostsensitivez.txt}$, as can be seen by the narrowing of the inferred $H(z)$ curves there.
+    The inset shows the posterior on $H(z=\variable{output/mostsensitivez.txt})$.
+    Black solid lines indicate the true value of $H(z)$ in both the inset and main panel.
+    The left panel shows the two-dimensional posterior on \Ho{} and \Omm, with the true value indicated by a black ``+''.
+    The two parameters are strongly degenerate because of the multiple ways of measuring $z=\variable{output/mostsensitivez.txt}$.
+    Spectral sirens are particularly well-suited to measuring cosmological parameters that affect the mid-to-high redshift universe. 
+    }
     \label{fig:O5_corner}
     \script{nonparametric_corner.py}
-\end{figure}
+\end{figure*}
 
-% \begin{figure}
+\subsubsection{BONUS: XG}
+To demonstrate the capabilities of next-generation detectors, we repeat the full analysis with \result{1000} events indicative of Cosmic Explorer's projected best-measured events within the first month alone.
+We fit the dark energy equation of state parameter, $w$, along with \Ho, and show results in Figure~\ref{fig:XG_corner}.
+We recover unbiased measurements of each parameter and find a \result{X}\% measurement of $w$ and a \result{X}\% measurement of \Ho.
+In comparison, the most recent constraints on these parameters with \result{X experiment} were \result{X}\% and \result{X}\% , respectively.
+This illustrates the unique ability of \acp{GW} to probe the high-redshift universe.
+
+% \begin{figure*}
 %     \centering
@@
 %     \script{XG_corner.py}
-% \end{figure}
+% \end{figure*}
 
@@
 
-\begin{itemize}
-    \item We are able to get constraints on cosmological parameters with a very flexible model for the mass distribution. 
-    \item The success of the flexible model in recovering an informative and accurate posterior on \Ho{} demonstrates that there is no need to know the \ac{CBC} mass distribution's morphology in order to constrain cosmological parameters with it.
-    \item This is because the information in spectral sirens comes from assuming events come from the same overall mass distribution across redshift, and that mass distribution doesn't evolve with redshift in the same exact way as would be mimicked by cosmological redshifting.
-This assumption is very basic, not an assumption that we know the exact astrophyisics of \ac{BBH} formation (i.e.
-PPISN feature location).
-    \item We have also predicted a \variable{output/nonparh0percent.txt}\% measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens.
+\acp{GW} are unique messengers in that they carry both redshift and distance information, making them remarkably clean probes of the Universe's expansion history. 
+However, the current method of determining \ac{GW} redshifts via the distribution of their source masses employs an assumption on the shape of their population.
+This choice of mass distribution, often assumed to be necessary or fundamental to the method, introduces a systematic bias to an otherwise \comment{straightforward? clean? unbiased?} measurement.
+
+In this work, we show that such a choice is not in fact necessary.
+We do this by accurately measuring cosmological parameters with a highly flexible model for the mass distribution. 
+The success of the flexible model in recovering an informative and accurate posteriors on \Ho{} and \Omm{} demonstrates that there is no need to know the \ac{CBC} mass distribution's morphology in order to constrain cosmological parameters with it.
+This reinforces the notion that the information in the spectral siren measurement comes from the assumption that all \ac{GW} sources come from the same population, which is a much \comment{weaker/ more basic/ more defensible/ more reasonable/ more conservative/ more fundamental} statement than the assumption that we understand the exact astrophysical processes that give rise to that population (i.e.
+the physics governing compact binary formation and evolution).
+
+We have also predicted a \variable{output/nonparh0percent.txt}\% measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens.
 A similar study with parametric methods \citep{Chen:2024gdn} predicts an \result{X\%} measurement, demonstrating a \result{negligible/substantial/marginal} increase in precision.
-However, the true population of compact objects in the Universe is unknown, so it is not possible to know the true functional form of the mass distribution, so systematic uncertainties arising from an incorrect choice for the functional form of the mass distribution are therefore inevitable.
-With current detectors, these systematic effects are dwarfed by statistical uncertainties.
+However, the true population of compact objects in the Universe is unknown.
+It is therefore impossible to know the true functional form of the mass distribution, so systematic uncertainties arising from an incorrect choice for the form of the mass distribution are inevitable.
+With current observations, these systematic effects are dwarfed by statistical uncertainties.
 However, next-generation detectors will herald in enough \ac{GW} observations to substantially decrease statistical uncertainty in these measurements: \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to \result{$\sim3\sigma$} systematic biases in \Ho when catalogs contain 2,000 events.
 Thus, the additional statistical uncertainty introduced by non-parametric approaches may be preferable to the systematic errors associated with choosing a parametric model.
-    \item additionally, in \ac{O5} we will be most sensitive to $z=\variable{output/mostsensitivez.txt}$.
-This is larger than we expect to see EM counterparts to, and larger than we expect galaxy surveys to be complete to (is this true?) so upgrades to current facilities will allow spectral sirens to probe H(z) to relatively unexplored redshifts.
-This is not unique to our nonparametric method, but to spectral sirens in general.
-    \item future/in prep.
-work will do the same thing but for a mass distribution that is allowed to evolve with redshift, similarly to the parametric analysis done in~\cite{ezquiaga_spectral_2022}, but with the evolution with redshift allowed to be arbitrary so long as it does not mimic cosmology.
-Because of ~\citet{ezquiaga_spectral_2022}, we think we could get a useful measurement if we include evolving peak shapes/positions in the mass distribution, provided there are multiple features present (which we know there are) and that they do not evolve in the same exact way as cosmology would - i.e.
-all features do not monotonically move to higher source frame mass with increasing redshift, all at the same rate.
-That scenario would be astrophysically unlikely, especially since the processes changing the features are all distinct from one another.
-    \item GPs are good bc they can do correlations/generalize to multiple dimensions naturally. Our choice of kernel allows us to do this, but its not possible with the autoregressive kernel.
-    \item in XG we will get most bang for our buck with spectral sirens
-\end{itemize}
+
+Additionally, we find spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ at the end of \acf{O5}, .
+This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebogo_observing_2023}, meaning upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances.
+Proposed next generation detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to $z\sim40$.
+Galaxy catalogs are anticipated to be complete to \result{$z=X$} at that time, and type 1a supernovae are expected to be observable to $z\simY$ \comment{cite Rubin papers?}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
+While this sensitivity to high redshifts is a feature of the spectral siren method in general, non-parametric methods such as the one presented here will be imperative to avoid systematic biases as next-generation detectors provide additional data.
+
+Future work \comment{in preparation?} will extend the method developed here to mass distributions that are allowed to evolve with redshift.
+Non-parametric methods will allow for arbitrary redshift evolution, but must constrain that evolution to avoid exactly mimicking the effects of cosmological redshift.
+As demonstrated in \citet{ezquiaga_spectral_2022}, the degeneracy between an evolving mass distribution and the expansion of the universe can be broken provided there are multiple features present in the mass distribution -- this is known to be the case in current data \citep{abbott_population_2023} -- and all features do not identically and monotonically move to higher source frame mass with increasing redshift.
+The latter scenario would be astrophysically unlikely as the the locations of features in the mass distribution are each thought to be governed by different physical processes.
+\acl{GP}es naturally scale to multiple data dimensions \comment{cite something?
+or elaborate?
+I'm just worried its getting too wordy}, so the method presented here can be straightforwardly generalized to fitting the redshift dependence of the mass distribution.
+
+In this work, we have restricted our analysis to the \ac{BBH} mass distribution, \Ho, and \Omm.
+However, the method can be trivially extended to the full mass distribution of \acp{CBC} \citep[e.g.]{matter matters, ezquiaga_spectral_2022, bridging gap,ray_non-parametric_2023} and to constrain other cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$.
+
 \begin{acknowledgments}
     The authors thank Reed Essick, Utkarsh Mali, Ben Farr, Maya Fishbach,  for helpful conversations. 
+    %Amanda
+    A.M.F. is supported by the National Science Foundation Graduate Research Fellowship Program under Grant No. DGE-1746045.
+    %Jose
+    J.M.E.
+is supported by the European Union’s Horizon 2020 research and innovation program under the Marie Sklodowska-Curie grant agreement No.
+847523 INTERACTIONS, and by VILLUM FONDEN (grant no.
+53101 and 37766).
+    %Daniel
+    D.E.H is supported by NSF grants AST-2006645 and PHY-2110507, as well as by the Kavli Institute for Cosmological Physics through an endowment from the Kavli Foundation and its founder Fred Kavli.
+    %Cluster
+    The Tycho supercomputer hosted at the SCIENCE HPC center at the University of Copenhagen was used for supporting this work.
 \end{acknowledgments}
@@
 
+\subsection{Option 2: GP only}
+We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated \ac{GW} catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
+This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after \ac{O5}.
+The details of the data simulation are described in Appendix~\ref{ap:data generation}.
+
+The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
+The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho{} ($\variable{output/H0_FID.txt}\Hunits$) is consistent with its inferred posterior ($\variable{output/nonparh0CI.txt}\Hunits$).
+At very high and very low masses, there is both a lack of data and little detector sensitivity.
+The mass distribution is therefore uninformed in this region and the \ac{GP} reverts to its prior distribution, which resembles random scatter around the mean differential merger rate.
+Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
+On the other hand, the lack of data just below the minimum \ac{BH} mass and just above the maximum \ac{BH} mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the GP to learn a relatively low merger rate at the edges of the mass distribution.
+Indeed, these features are very informative to the \Ho{} constraint \citep{the_ligo_scientific_collaboration_constraints_2021}.
+The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
+Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,60\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
+
 \end{document}

### 2024-03-11 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
-On the other hand, the lack of data just below the minimum \ac{BH} mass and just above the maximum \ac{BH} mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the \ac{GP} to learn a relatively low merger rate at the edges of the mass distribution.
+On the other hand, the lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the \ac{GP} to learn a relatively low merger rate at the edges of the mass distribution.
 Indeed, these features are very informative to the \Ho{} constraint \citep{the_ligo_scientific_collaboration_constraints_2021}.
@@
 Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
-On the other hand, the lack of data just below the minimum \ac{BH} mass and just above the maximum \ac{BH} mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the GP to learn a relatively low merger rate at the edges of the mass distribution.
+On the other hand, the lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the GP to learn a relatively low merger rate at the edges of the mass distribution.
 Indeed, these features are very informative to the \Ho{} constraint \citep{the_ligo_scientific_collaboration_constraints_2021}.

### 2024-03-11 src/tex/ms.tex
[showyourwork] overleaf sync

@@
     \includegraphics[width=\textwidth]{figures/O5_GP_pm.pdf}
-    \caption{Spectral siren measurement for an \ac{O5}-like catalog with the ``correct'' parametric model (\plp, green), an ``incorrect'' parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{Gaussian process}, blue).
+    \caption{Spectral siren measurement for an \ac{O5}-like catalog with the ``correct'' parametric model (\plp, green), an ``incorrect'' parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{GP}, blue).
     The left panel shows the recovered source frame primary mass distribution for each model, and the simultaneously-inferred posteriors on \Ho{} are shown in the right panel.

### 2024-03-11 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 Proposed next generation detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to $z\sim40$.
-Galaxy catalogs are anticipated to be complete to \result{$z=X$} at that time, and type 1a supernovae are expected to be observable to $z\simY$ \comment{cite Rubin papers?}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
+Galaxy catalogs are anticipated to be complete to \result{$z=X$} at that time, and type 1a supernovae are expected to be observable to $z\sim Y$ \comment{cite Rubin papers?}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
 While this sensitivity to high redshifts is a feature of the spectral siren method in general, non-parametric methods such as the one presented here will be imperative to avoid systematic biases as next-generation detectors provide additional data.

### 2024-03-11 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
-The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho{} ($\variable{output/H0_FID.txt}\Hunits$) is consistent with its inferred posterior ($\variable{output/nonparh0CI.txt}\Hunits$).
+The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho{} ($\variable{output/H0_FID.txt}\Hunits$) is consistent with its inferred posterior (\variable{output/nonparh0CI.txt}$\Hunits$).
 At very high and very low masses, there is both a lack of data and little detector sensitivity.

### 2024-03-11 src/scripts/nonparametric_corner.py
change nsamps and also change label on corner plot

@@ -29,11 +29,11 @@ def Hz(id, ax=None,save=False,inset=True):
     for i in range(NSAMPS):
         H_z[i] = calc_Hz(TEST_Z,H0[i],Om0[i])
         ax.plot(TEST_Z,H_z[i],lw=0.2, c="blue",alpha=0.1)
-    ax.plot(TEST_Z, calc_Hz(TEST_Z,H0_FID,OM0_FID),c='k', lw=2,label="injected value")
+    ax.plot(TEST_Z, calc_Hz(TEST_Z,H0_FID,OM0_FID),c='k', lw=2,label="True value")
     ax.set_xlabel("$z$")
     ax.set_ylabel("$H(z)$ [km/s/Mpc]")
-    ax.set_xlim(0,5)
-    ax.set_ylim(0,750)
+    ax.set_xlim(0,3)
+    ax.set_ylim(0,500)
     ax.legend(framealpha=0,loc='lower right')
     
     if inset:
@@ -52,7 +52,7 @@ def Hz(id, ax=None,save=False,inset=True):
         iax.set_xlabel("$H(z=%1.1f)$\n[km/s/Mpc]"%zbest,fontsize=10)
         iax.set_xticks([100,125,150])
         # iax.set_ylabel('posterior density',fontsize=8)
-        
+
         # arrow from the inset to zbest
         # midpoint_x = 1.75/2 + 0.5
         # offset_y= 50

### 2024-03-11 src/scripts/nonparametric_inference_fitOm0.py

@@ -1,6 +1,6 @@
 # custom
 import paths
-from nonparametric_inference import NSAMPS, remove_low_Neff
+from nonparametric_inference import remove_low_Neff
 from priors import get_ell_frechet_params, get_sigma_gamma_params, hyper_prior
 
 # inference
@@ -11,7 +11,7 @@ import numpyro
 # data
 import numpy as np
 
-NSAMPS=50
+NSAMPS = 50
 
 jax.config.update("jax_enable_x64", True)
 # random number generators

### 2024-03-11 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 % Define document class
-\documentclass[]{aastex631}
+\documentclass[twocolumn]{aastex631}
 \usepackage{showyourwork}
@@
 With current observations, these systematic effects are dwarfed by statistical uncertainties.
-However, next-generation detectors will herald in enough \ac{GW} observations to substantially decrease statistical uncertainty in these measurements: \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to \result{$\sim3\sigma$} systematic biases in \Ho when catalogs contain 2,000 events.
+However, next-generation detectors will herald in enough \ac{GW} observations to substantially decrease statistical uncertainty in these measurements: \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to \result{$\sim3\sigma$} systematic biases in \Ho when catalogs contain 2,000 events, though this bias may be an over-estimate as it does not include measurement uncertainty of \ac{GW} parameters.
 Thus, the additional statistical uncertainty introduced by non-parametric approaches may be preferable to the systematic errors associated with choosing a parametric model.

### 2024-03-11 src/scripts/submit.sh
add bash script for RCC

@@ -1,0 +1,16 @@
+#!/bin/bash
+#SBATCH --job-name=paper_fitOm0_gwmc
+#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_gwmc.out
+#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_gwmc.err
+#SBATCH --account=kicp
+#SBATCH --time=48:00:00
+#SBATCH --partition=kicp-gpu
+#SBATCH --nodes=1
+#SBATCH --mem=128G
+#SBATCH --gres=gpu:1
+
+module load cuda/11.5
+echo "loaded cuda"
+module load python/anaconda-2022.05
+echo "loaded anaconda"
+showyourwork build

### 2024-03-13 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2024-03-11 src/tex/ms.tex
@@
 \label{sec:intro}
-Like light, \acp{GW} are redshifted as they propagate across the universe\jme{, receiving an imprint of the cosmic expansion history}.
-Unlike light, however, \jme{GW signals are known from first principles and their} selection effects of \acp{GW} are extremely well understood, allowing for a precise estimate of each catalogs' completeness and an unbiased measurement of the true \ac{GW} population\jme{, including the absolute distance to the sources}.
-This allows \ac{GW} signals to be used as \jme{``standard sirens'', becoming powerful} %relatively clean 
-probes of cosmological parameters. 
-\jme{When complemented with redshift information about the host galaxy, as in the multi-messenger GW170817, ... }
-\citep[e.g.][]{}%bright sirens first paper, 170817 bright siren, dark siren first paper, dark siren with 0817, hithhikers guide.
+Like light, \acp{GW} are redshifted as they propagate across the universe, thereby becoming imprinted with the cosmic expansion history.
+Unlike light, however, the form of \ac{GW} signals are known from first principles and their selection effects are extremely well understood.
+This allows for a precise estimate of each \ac{GW} catalogs' completeness and an unbiased measurement of the true \ac{GW} population. 
+Additionally, the fact that \ac{GW} signals provide direct measurements of the distance to their sources makes them ``standard sirens'': powerful probes of cosmological parameters that circumvent the cosmological distance ladder~\citep{schutz_determining_1986,holz_using_2005}.
+This is particularly useful in the context of the current ``Hubble tension:'' a disagreement between multiple methods of measuring the local expansion rate of the universe with electromagnetic observations \citep{freedman_measurements_2021}.
 
-\ac{GW} signals provide direct measurements of each source's luminosity distance \jme{$d_L$} and redshifted, or detector frame, masses, $m_{\det}= m_{\source}(1+z)$.
+A well-known application of standard sirens was the multi-messenger event GW170817 \citep{abbott_multi-messenger_2017}, whose clear association with a host galaxy provided a precise redshift measurement, allowing for a direct measurement of the Hubble constant, \Ho{} \citep{ abbott_gravitational-wave_2017}.
+External redshift information can also come from galaxy catalogs, which provide redshift distributions for each \ac{GW} signal, allowing for a probabilistic measurement of \Ho{} when multiple \ac{GW} detections are combined \citep[e.g.][]{del_pozzo_inference_2012, chen_two_2018, fishbach_standard_2019, gray_cosmological_2020, gwtc3_cosmo, gair_hitchhikers_2023}.
+
+However, external information about \ac{GW} source redshifts need not be available in order to use them as standard sirens.
+\ac{GW} signals provide direct measurements of each source's luminosity distance, $d_L$, and redshifted, or detector frame, masses, $m_{\det}= m_{\source}(1+z)$ \citep[e.g.][]{chen_mass-redshift_2019}.
 Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the \ac{GW} signal was emitted, $H(z)$.
-In practice, the source frame mass is unknown, but it is possible to know the location of features in the source frame mass distribution.
+In practice, unless tidal information is available \citep{messenger_measuring_2012}, the source frame mass is unknown.
+However, it is possible to know the location of features in the source frame mass distribution.
 The whole mass distribution therefore acts similarly to an electromagnetic spectrum, where the location of spectral features provides a redshift measurement.
@@
 Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff_gravitational_1993,Taylor:2011fs} using the binary neutron star mass distribution, and extended to the \ac{BBH} mass distribution by \cite{farr_future_2019}.
-\jme{Spectral siren analyses have now been implemented by the LIGO--Virgo--KAGRA Collaborations using the latest catalog \cite{LIGOScientific:2021aug}.}
+Spectral siren analyses have since been implemented by the LIGO--Virgo--KAGRA Collaborations using the latest catalog \cite{gwtc3_cosmo}.
 
@@
 theoretical expectations of the location of a pulsational-pair instability pileup, upper mass gap, or maximum neutron star mass.
-\jme{This would certainly be problematic, as mismodeling of the specific feature would hinder the accuracy of the method and introduce biases in the cosmological inference \cite{ezquiaga_spectral_2022,Mukherjee:2021rtw,Pierra:2023deu}.}
-However, the information in the spectral siren measurement comes solely from the assumption that all \acp{CBC} follow the same mass distribution.\footnote{
-The spectral siren method is also valid if \acp{CBC} have different mass distributions at different redshifts, so long as the evolution of the mass distribution does not exactly mimic cosmology, and the model for the mass distribution allows for evolution \citep{ezquiaga_spectral_2022}.
-\jme{[maybe is worth having this comment in the main text?]}
+This would certainly be problematic, as mismodeling of the specific feature would hinder the accuracy of the method and introduce biases in the cosmological inference \citep{Mukherjee:2021rtw,mastrogiovanni_importance_2021,pierra_study_2023}.
+
+However, the information in the spectral siren measurement comes solely from the assumption that all \acp{CBC} follow the same mass distribution.\footnote{More precisely, the spectral siren method is also valid if \acp{CBC} have different mass distributions at different redshifts, so long as the evolution of the mass distribution does not exactly mimic cosmology, and the model for the mass distribution allows for evolution \citep{ezquiaga_spectral_2022}.
+%\jme{[maybe is worth having this comment in the main text?] \comment{True, but it kind of interrupts the flow and I think the current text sets up the question nicely.
+I discuss it more in the discussion.}} OK!
 }
@@
 In this work, we explicitly demonstrate that no \emph{a priori} knowledge about the shape of the \ac{CBC} mass spectrum is necessary to use the spectral siren methodology. 
-We do this by inferring \Ho with a flexible, non-parametric model for the mass distribution of \acp{CBC}. 
+We do this by inferring $H(z)$ with a flexible, non-parametric model for the mass distribution of \acp{CBC}. 
 This model makes minimal prior assumptions about the shape of the mass distribution, enabling it to accurately infer a wide range of morphologies.
-Despite its flexibility, it is able to consistently obtain an unbiased measurement for \Ho, showing that nonparametric methods are not only sufficient for a spectral siren measurement, they can also mitigate systematic effects in the measurement caused by model misspecification.
+Despite its flexibility, it is able to consistently obtain unbiased measurements of cosmological parameters, showing that nonparametric methods are not only sufficient for a spectral siren measurement, they can also mitigate systematic effects in the measurement caused by model misspecification.
 
 We find that using this model allows for a \variable{output/nonparh0percent.txt}\% measurement of \Ho{} by the end of \ac{O5}, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
-While constraints on \Ho{} are used in this work to quantify the accuracy and precision of our non-parametric method, we emphasize that spectral sirens are able to constrain $H(z)$ out to redshifts that are inaccessible by electromagnetic observations, especially with next-generation detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn,You:2020wju}\footnote{See example \jme{spectral siren cosmological inference using parametric models for the mass spectrum here}: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}.
+While constraints on \Ho{} are used in this work to quantify the accuracy and precision of our non-parametric method, we emphasize that spectral sirens are able to constrain $H(z)$ out to redshifts that are inaccessible by electromagnetic observations, especially with next-generation detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn, You:2020wju}\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}.
 Applied to data from next-generation detectors, this method will therefore be able to constrain the dark energy equation of state parameter, $w$, at redshifts $\gtrsim 2$.
-
-%\jme{[cite joint multi-siren forecast \cite{Chen:2024gdn}. Cite code \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}?]}\\ 
-\comment{[Should I add a separate paragraph for other sirens or is it sufficient to just cite it above with other forecasts, as I did?]} \jme{[I suggest to do it quickly in the first paragraph.
-Then in the discussion maybe we could emphasize that future analysis will be joint as we discussed in \cite{Chen:2024gdn}]}
 
@@
 
-Algorithmically, each hyper-likelihood evaluation contains the following steps:
+Algorithmically, each hyper-likelihood \comment{is that a word?} evaluation contains the following steps: \script{priors.py}
 \begin{enumerate}
@@
 However, the current method of determining \ac{GW} redshifts via the distribution of their source masses employs an assumption on the shape of their population.
-This choice of mass distribution, often assumed to be necessary or fundamental to the method, introduces a systematic bias to an otherwise \comment{straightforward? clean? unbiased?} measurement.
+This choice of mass distribution, often presumed to be necessary or fundamental to the method, introduces a systematic bias to a measurement that otherwise makes very few assumptions.
 
@@
 The success of the flexible model in recovering an informative and accurate posteriors on \Ho{} and \Omm{} demonstrates that there is no need to know the \ac{CBC} mass distribution's morphology in order to constrain cosmological parameters with it.
-This reinforces the notion that the information in the spectral siren measurement comes from the assumption that all \ac{GW} sources come from the same population, which is a much \comment{weaker/ more basic/ more defensible/ more reasonable/ more conservative/ more fundamental} statement than the assumption that we understand the exact astrophysical processes that give rise to that population (i.e.
+This reinforces the notion that the information in the spectral siren measurement comes from the assumption that all \ac{GW} sources come from the same population, a far less stringent statement than the assumption that we understand the exact astrophysical processes which give rise to that population (i.e.
 the physics governing compact binary formation and evolution).
 
-We have also predicted a \variable{output/nonparh0percent.txt}\% measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens.
-A similar study with parametric methods \citep{Chen:2024gdn} predicts an \result{X\%} measurement, demonstrating a \result{negligible/substantial/marginal} increase in precision.
+We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens and a \result{X\%} %$\variable{output/parh0percent.txt}\%$
+measurement with parametric spectral sirens, demonstrating a \result{negligible/substantial/marginal} increase in precision when using the correct parametric model.
 However, the true population of compact objects in the Universe is unknown.
@@
 With current observations, these systematic effects are dwarfed by statistical uncertainties.
-However, next-generation detectors will herald in enough \ac{GW} observations to substantially decrease statistical uncertainty in these measurements: \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to \result{$\sim3\sigma$} systematic biases in \Ho when catalogs contain 2,000 events.
+However, next-generation detectors will herald in enough \ac{GW} observations to substantially decrease statistical uncertainty in these measurements: \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to $\sim3\sigma$ systematic biases in \Ho{} when catalogs contain 2,000 events, though this bias may be an over-estimate as it does not include measurement uncertainty of \ac{GW} parameters.
 Thus, the additional statistical uncertainty introduced by non-parametric approaches may be preferable to the systematic errors associated with choosing a parametric model.
 
-Additionally, we find spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ at the end of \acf{O5}, .
-This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebogo_observing_2023}, meaning upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances.
+Additionally, we find spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ at the end of \acf{O5}.
+This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}, meaning upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances.
 Proposed next generation detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to $z\sim40$.
-Galaxy catalogs are anticipated to be complete to \result{$z=X$} at that time, and type 1a supernovae are expected to be observable to $z\sim Y$ \comment{cite Rubin papers?}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
-While this sensitivity to high redshifts is a feature of the spectral siren method in general, non-parametric methods such as the one presented here will be imperative to avoid systematic biases as next-generation detectors provide additional data.
+At that time, cosmological surveys such as those enabled by the Nancy Grace Roman Space Telescope and the Vera Rubin Observatory are expected to be able to precisely measure $H(z)$ to $z\sim3$ \citep{spergel_wide-field_2015}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
+When combined with other standard siren methods, the low-redshift expansion will also be well-constrained \citep{Chen:2024gdn}.
+While the sensitivity to high redshifts is a feature of the spectral siren method in general, non-parametric methods such as the one presented here will be imperative to avoid systematic biases as next-generation detectors provide additional data.
 
-Future work \comment{in preparation?} will extend the method developed here to mass distributions that are allowed to evolve with redshift.
+Work in preparation will extend the method developed here to mass distributions that are allowed to evolve with redshift.
 Non-parametric methods will allow for arbitrary redshift evolution, but must constrain that evolution to avoid exactly mimicking the effects of cosmological redshift.
-As demonstrated in \citet{ezquiaga_spectral_2022}, the degeneracy between an evolving mass distribution and the expansion of the universe can be broken provided there are multiple features present in the mass distribution -- this is known to be the case in current data \citep{abbott_population_2023} -- and all features do not identically and monotonically move to higher source frame mass with increasing redshift.
-The latter scenario would be astrophysically unlikely as the the locations of features in the mass distribution are each thought to be governed by different physical processes.
-\acl{GP}es naturally scale to multiple data dimensions \comment{cite something?
-or elaborate?
-I'm just worried its getting too wordy}, so the method presented here can be straightforwardly generalized to fitting the redshift dependence of the mass distribution.
+As demonstrated in \citet{ezquiaga_spectral_2022}, the degeneracy between an evolving mass distribution and the expansion of the universe can be broken provided there are multiple features present in the mass distribution and all features do not identically and monotonically move to higher source frame mass with increasing redshift.
+The first condition (multiple features) is known to be met in current data \citep{abbott_population_2023}, with three robust features \citep{farah_things_2023}: a maximum black hole mass and overdensities at $\sim10\Msun$ and $\sim35\Msun$.
+The latter scenario (features identically and monotonically moving to higher masses with increasing redshift) would be astrophysically unlikely as the the locations of features in the mass distribution are each thought to be governed by different physical processes \citep[e.g.][]{mapelli_binary_2020}.
+Since \aclp{GP} naturally scale to multiple data dimensions, \citep{rasmussen_gaussian_2006}, the method presented here can be straightforwardly generalized to fitting the redshift dependence of the mass distribution.
 
 In this work, we have restricted our analysis to the \ac{BBH} mass distribution, \Ho, and \Omm.
-However, the method can be trivially extended to the full mass distribution of \acp{CBC} \citep[e.g.]{matter matters, ezquiaga_spectral_2022, bridging gap,ray_non-parametric_2023} and to constrain other cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$.
+However, the method can be trivially extended to the full mass distribution of \acp{CBC} \citep[e.g.][]{fishbach_does_2020, ezquiaga_spectral_2022} and to constrain other cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$.
 
@@
 We generate a mock catalog of \ac{GW} events characteristic of that expected from \acf{O5}. 
-These events are drawn from an underlying population described by
+These events are drawn from an underlying population described by the \plp{} mass distribution presented in \citet{talbot} and used in \citet{abbot_1 abbot2} and the redshift distribution presented in \citet{callister_shouts_2020}.
 \begin{equation}
 \label{eq:underlying population}
-    \frac{\diff N}{\diff m_1 \diff m_2 \diff z} \propto p(m_1, m_2|\bar{\Lambda}_m) p(z|H_0, \Omega_M),
+    \frac{\diff N}{\diff m_1 \diff z} \propto p(m_1|\bar{\Lambda}_m) p(z|\bar{\Lambda}_z, H_0, \Omega_M),
 \end{equation}
@@
 \begin{align}
-    p(m_1,m_2|\bar{\Lambda}_m) &\propto ,\\
+    p(m_1,m_2|\bar{\Lambda}_m) &\propto \mathcal{S}(m_{\min},m_{\max})
+    \left( f_{\text{peak}}e^{-\frac{1}{2}(\frac{m_1-\mu}{\sigma})^2}\mathcal{N}_{\text{g}} +
+    (1-f_{\text{peak}})m_1^{\alpha}\mathcal{N}_{\text{pl}} \right) ,
 \label{eq:underlying mass dist}
@@
 \begin{equation}
-    p(z|H_0, \Omega_M) \propto \frac{\diff V_C}{\diff z} \frac{1}{1+z},
+    p(z|H_0, \Omega_M) \propto \frac{\diff V_C}{\diff z} \frac{1}{1+z} \frac{(1+z)^\alpha}{1+\left(\frac{1+z}{1+z_p}\right)^{\alpha+\beta}}.
     \label{eq:underlying redshift dist}
 \end{equation}
-$V_C(H_0, \Omega_M)$ is the comoving volume for given cosmological parameters \Ho{} and \Omm{}, and $\bar{\Lambda_m} = \alpha$...
-are the parameters describing the power law in primary mass, ...
-respectively.
-When generating the simulated events, we have fixed $\alpha= $... and used cosmological parameters \Ho$=68.7$, \Omm$=0.3$, $\Omega_\Lambda=1-\Omega_M$.
-These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \_ model \cite{o3b_pop}, which is morphologically similar to that described in Equations~\ref{eq:underlying population}--\ref{eq:underlying mass and redshift dist}.
-We neglect spins entirely, as they do not carry cosmological information.
-\comment{TODO: fill in details of mass distribuiton.
-Should we use PDB and include BNSs, which will give us a nice sharp feature to work with, or should we use PLP which has a bump?
-Could also do PLP and then put a uniform dist or power law for NS masses at the LMG, and leave an empty mass gap.
-Would need to get relative heights correct though.}
-\jme{[For \ac{O5} BNSs will be subdominant. I would focus on \acp{BBH} for simplicity]}
-\comment{[Ok, so we should say that although a putuative lower mass gap would give and additional feature, it is not expected to be constrained well in O5, so we focus on the \ac{BBH} spectrum for now.
-In XG, though, it will be important to consider the full spectrum \citep{ezquiaga_spectral_2022}.]}
+Here, $V_C(H_0, \Omega_M)$ is the comoving volume for given cosmological parameters \Ho{} and \Omm{}, and $\bar{\Lambda}_m = \{\alpha, m_{\min}, m_{\max}, \mu, \sigma, f_{\text{peak}}\}$ are the parameters describing the power law in primary mass, minimum and maximum black hole mass, Gaussian peak location and width, and fraction of events in the Gaussian peak, respectively.
+$\mathcal{S}$ is a smoothing function at low and high masses, and $\mathcal{N}_{\text{pl}}$ and $\mathcal{N}_{\text{g}}$ are the normalizations between  $m_{\min}$ and $m_{\max}$ for the power law component and truncated Gaussian component, respectively.
+$\bar{\Lambda}_z = \{z_p,\alpha_z,\beta_z\}$ are the parameters governing the redshift distribution's peak redshift, low-$z$ power law slope, and high-$z$ power law slope.
+When generating the simulated events, we have fixed $\alpha=-2.7,m_{\max}=78\Msun,m_{\min}=10\Msun, \mu=30\Msun,\sigma=7.0\Msun, f_{\text{peak}}=0.05,   z_p=2.4,\alpha_z=1.,\beta_z=3.4$.
+These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{o3b_pop}.
+We do not fit for secondary masses or spins, as they do not carry additional cosmological information.
+We use cosmological parameters \Ho$=\variable{output/H0_FID.txt}\Hunits$, \Omm$=0.3$, $\Omega_\Lambda=1-$\Omm, consistent with those found by \citet{planck_collaboration_planck_2016}.
+We emphasize, however, that the choice of cosmological parameters for data generation is arbitrary and does not impact the results, since we are concerned only with the ability of our method to recover the injected values.
 
-After passing the simulated events through projected detector selection effects, the resulting catalog has \variable{output/num_found_events.txt}events, %including \result{X} binary neutron star systems, \result{Y} NSBH systems, and \result{Z} \ac{BBH} systems,
-consistent with the numbers projected for \ac{O5} by \citet{kiendrebogo_observing_2023}.
+We do not include neutron stars in our simulation set, as their contribution to the spectral siren measurement is expected to be subdominant in \ac{O5}.
+However, if a lower mass gap between the heaviest neutron stars and lightest black holes exists, it will provide an additional feature with which to inform the measurement, and will be the most informative feature for spectral siren measurements with next generation detectors \citep{ezquiaga_spectral_2022}.
+
+After passing the simulated events through projected detector selection effects, the resulting catalog has \variable{output/num_found_events.txt}events, consistent with the numbers projected for \ac{O5} by \citet{kiendrebogo_observing_2023}.
 We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from \ac{O5} detectors.
@@
 The full procedure for this mock data generation process is described in \citet{fishbach_where, farah_things_2023, essick_dagnabbit_2023}, and the exact settings used for \texttt{GWMockCat} are made available in the accompanying data release.
-\comment{TODO:put a showyourwork link here.}
 

### 2024-03-13 src/scripts/nonparametric_inference_fitOm0.py
set nsamps=50 for testing

@@ -11,7 +11,7 @@ import numpyro
 # data
 import numpy as np
 
-NSAMPS = 1600
+NSAMPS = 50
 
 jax.config.update("jax_enable_x64", True)
 # random number generators
@@ -40,7 +40,7 @@ if  __name__ == "__main__":
     kwargs = dict(m1det=m1z_PE,dL=dL_PE, m1det_inj=m1zinj_det,dL_inj=dLinj_det,
                     log_pinj=log_pinj_det, log_PE_prior=log_PE_prior,
                     remove_low_Neff=remove_low_Neff,fit_Om0=True)
-    mcmc = numpyro.infer.MCMC(nuts_kernel,num_warmup=800,num_samples=NSAMPS,
+    mcmc = numpyro.infer.MCMC(nuts_kernel,num_warmup=NSAMPS,num_samples=NSAMPS,
                               num_chains=1,progress_bar=True)   
     mcmc.run(jax_rng,**kwargs)
 

### 2024-03-13 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 We do this by constructing a flexible, Gaussian process-based model for \ac{GW} source population to analyze simulated \ac{GW} data consistent with expectations for the next \acl{LVK} observing run.
-We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a \result{X\%} measurement of $H(z=\variable{output/mostsensitivez.txt})$ ($1-\sigma$).
+We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/nonparhzpercent.txt}\%$  measurement of $H(z=\variable{output/mostsensitivez.txt})$ ($1-\sigma$).
 This astrophysics-agnostic spectral siren technique will be key to deliver precise and unbiased cosmological constraints with future observations in the presence of possible redshift evolution of the mass spectrum.

### 2024-03-13 src/scripts/nonparametric_twopanel.py
typo

@@ -59,7 +59,7 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     axes[1].plot(prior,kde_PLP(prior),c='green',label=r'\textsc{Power Law + Peak}')
     axes[1].plot(prior,kde_BPL(prior),c='orange',label='Broken power law')
     axes[1].set_ylabel('posterior density')
-    fig.legend(ncols=4,framealpha=0,fontsize=9,loc="upper center")
+    fig.legend(ncol=4,framealpha=0,fontsize=9,loc="upper center")
     plt.tight_layout()
     fig.savefig(paths.figures / "O5_GP_pm.pdf")
     plt.clf()

### 2024-03-13 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2024-03-13 src/tex/ms.tex
@@
 \begin{abstract}
-\Acp{GW} from merging compact objects are the only transient astrophysical sources that encode direct information about their luminosity distances. 
-Paired with their redshifts, a Hubble diagram can be constructed to probe the standard cosmological model.  
-Although individual signals do not have any reference scale from which a redshift can be inferred, an ensemble of events can provide such a scale via their distribution of source masses. 
-This ``spectral siren'' method thus constrains the Universe's expansion rate from \ac{GW} observations alone, without the need for accompanying electromagnetic observations. 
+\Acp{GW} from merging compact objects are the only transient sources that encode direct information about their luminosity distances. 
+Paired with their redshifts, a Hubble diagram can be constructed to probe the Universe's expansion.  
+``Spectral sirens'' use features in the mass distribution of \ac{GW} sources to extract this redshift information without the need for accompanying electromagnetic observations.
 So far, this technique has only been applied with simple parametric representations of the mass distribution. 
-However, the use of an incorrect mass model inevitably leads to biases in the cosmological inference, an acute problem given our current lack of understanding of the true astrophysical model.
+However, the use of an incorrect mass model inevitably leads to biases in the cosmological inference, an acute problem given our current lack of understanding of the true source population.
 Furthermore, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to obtain unbiased measurements of cosmological parameters with the spectral siren methodology.
-We demonstrate that unbiased inference of cosmological parameters is possible without any assumptions for the mass distribution's morphology.
-We do this by constructing a flexible, Gaussian process-based model for \ac{GW} source population to analyze simulated \ac{GW} data consistent with expectations for the next \acl{LVK} observing run.
-We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a \result{X\%} measurement of $H(z=\variable{output/mostsensitivez.txt})$ ($1-\sigma$).
-This astrophysics-agnostic spectral siren technique will be key to deliver precise and unbiased cosmological constraints with future observations in the presence of possible redshift evolution of the mass spectrum.
+We demonstrate that prior knowledge of the \ac{GW} source mass distribution is not necessary in order to accurately infer cosmological parameters with spectral sirens.
+We do this by constructing a flexible, Gaussian process-based model for the \ac{GW} source population to analyze simulated \ac{GW} data consistent with expectations for the next \acl{LVK} observing run.
+We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hzpercent.txt}\%$  measurement of $H(z=\variable{output/mostsensitivez.txt})$ ($1\sigma$ uncertainties).
+This astrophysics-agnostic spectral siren technique will be key to deliver precise and unbiased cosmological constraints with future observations and in the presence of redshift evolution of the mass spectrum.
 \end{abstract}
@@
 To simultaneously infer cosmological parameters and the population of GW sources, we employ a hierarchical Bayesian analysis.
-This allows us to ``undo'' the selection effects of \ac{GW} detectors to obtain a true, or astrophysical, population.
+This allows us to ``undo'' the selection effects of \ac{GW} detectors to obtain a true, or astrophysical, population \jme{and constrain the cosmic expansion history}.
 
@@
 
-We only consider masses and redshifts, neglecting spins as they do not contain cosmological information.
+\jme{For our population model we focus on the} %We only consider 
+masses and redshifts, neglecting spins as they do not contain cosmological information.
 Thus, $\theta = \{m_1,m_2,z\}$. 
-We assume a uniform-in-comoving volume redshift distribution, as in Equation~\ref{eq:underlying redshift dist}.
-We assume $m_1=m_2$ and model the distribution of primary masses using a \acl{GP} (described in Section~\ref{sec:model}).
+\jme{The population can be modeled with different levels of complexity and flexibility. 
+As described in detail in Section~\ref{sec:model}, we model the distribution of primary masses in an agnostic way, using a \acl{GP}. 
+We also compare this non-parametric reconstruction with different parametric models.  
+For simplicity we assume that the population has equal masses, $m_1=m_2$, and that its redshift distribution is uniform-in-comoving volume, as in Equation~\ref{eq:underlying redshift dist}. }
+%We assume a uniform-in-comoving volume redshift distribution, as in Equation~\ref{eq:underlying redshift dist}.
+%We assume $m_1=m_2$ and model the distribution of primary masses using a \acl{GP} (described in Section~\ref{sec:model}).
 
@@
 This is the basis of the spectral siren method of measuring cosmological parameters; we can simultaneously infer the relationship between redshift and luminosity distance with the parameters of the source frame mass distribution.
-The full set of hyper-parameters therefore includes the cosmological parameters that dictate the $D_L$-$z$ relation: \Ho, \Omm, $\Omega_\Lambda, \Omega_r,$ and $w$.
-In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and $w=1$ and use uniform priors on \Ho{} and \Omm{}.
+The full set of hyper-parameters therefore includes the cosmological parameters that dictate the $D_L$-$z$ relation 
+\jme{such as the local expansion rate \Ho, the present fractional energy density of dark matter \Omm, dark energy $\Omega_Lambda$ and radiation $\Omega_r$, or the equation of state of dark energy $w$.} %: \Ho, \Omm, $\Omega_\Lambda, \Omega_r,$ and $w$.
+In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and \jme{$w=-1$} and use uniform priors on \Ho{} and \Omm{}. 
+\jme{This corresponds to a flat $\Lambda$CDM cosmology.}
 
-Algorithmically, we evaluate the likelihood in Equation~\ref{eq:inhomog-poisson} by drawing \Ho{} and \Omm{} from their prior distributions (contained in $p(\Lambda)$), using them to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to $m_{\det} (1+z(D_L)) = m_{\source}$, evaluating the source-frame mass distribution at these transformed values ($\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ for events and $N_{\text{exp}}(\Lambda)$ for injections), and then taking the product of these terms to evaluate the expression in Equation~\ref{eq:inhomog-poisson}.
+Algorithmically, we evaluate the likelihood in Equation~\ref{eq:inhomog-poisson} by drawing \Ho{} and \Omm{} from their prior distributions (contained in $p(\Lambda)$), using them to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to \jme{$m_{\det} /(1+z(D_L)) = m_{\source}$}, evaluating the source-frame mass distribution at these transformed values ($\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ for events and $N_{\text{exp}}(\Lambda)$ for injections), and then taking the product of these terms to evaluate the expression in Equation~\ref{eq:inhomog-poisson}.
 The full process is outlined in Section~\ref{sec:model}.
@@
 \label{sec:model}
-\Acfp{GP} are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
+
+\jme{Our goal is to model the population of \ac{GW} in an agnostic way. 
+We choose to do it with a \Acfp{GP}. 
+\ac{GP}} 
+%\Acfp{GP} 
+are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
 Their smoothness properties make them widely useful in \ac{GW} data analysis and beyond for regression problems, such as modeling time-domain waveforms \citep{doctor_statistical_2017, huerta_eccentric_2018} and the neutron star equation of state \citep{landry_nonparametric_2019}, density estimation problems, such as estimating posterior densities of single-event parameters from parameter estimation samples ~\citep{demilio_density_2021}, and as a prior on histogram bin heights for population inference \citep{mandel_model-independent_2017, li_flexible_2021, ray_non-parametric_2023}.
-Our use case is slightly different.
+
+%added a break here
+Our use case is slightly different \jme{to previous analyses}.
 We will utilize a \ac{GP} as a prior on the functions that describe the primary mass distribution of \acp{CBC}.
@@
 Practically, the only difference in the inference of the population when using a GP versus other modeling choices is that $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is determined directly by a realization of the \ac{GP}, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
-
+%joined paragraphs here
 In other words, when using parametric models, $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is calculated by evaluating a specific functional form described by a small set of hyper-parameters.
@@
 \label{sec:results}
-
 In this Section, we first show that fitting an incorrect functional form to the mass distribution of \acp{CBC} biases the inference of cosmological parameters when using the spectral siren methodology.
@@
 
-We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within \result{X-$\sigma$}.
-However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by \result{X}-$\sigma$.
+We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within \result{X$\sigma$}.
+However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by \result{X}$\sigma$.
+We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an \result{overestimate} of \Ho{} at greater than $1\sigma \variable{BPL_bias.txt}\%$ of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and
+the \plp{} model reaches the same level of bias $\variable{PLP_bias.txt}\%$ of the time.
+
 This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the \ac{GP}-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution.
@@
 
-The mass distribution inferred with the \ac{GP}-based model closely resembles the true, simulated distribution.%, and the injected value of \Ho{} ($\variable{output/H0_FID.txt}\Hunits$) is consistent with its inferred posterior ($\variable{output/nonparh0CI.txt}\Hunits$).
+Additionally, the mass distribution inferred with the \ac{GP}-based model closely resembles the true, simulated distribution.
 At very high and very low masses, there is both a lack of data and little detector sensitivity.
@@
 
-We have repeated this analysis with \result{50} separate simulated catalogs, and found that the broken power law model produces an \result{overestimate} of \Ho{} at greater than \result{X}-$\sigma$, \result{Y}\% of the time, whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and the \plp{} model reaches the same level of bias \result{A}\% of the time.
-This indicates that prior knowledge of the shape of the mass distribution is not required to perform a spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.
+Collectively, these results indicate that prior knowledge of the shape of the mass distribution is not required to perform a spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.
 

### 2024-03-13 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 We do this by constructing a flexible, Gaussian process-based model for the \ac{GW} source population to analyze simulated \ac{GW} data consistent with expectations for the next \acl{LVK} observing run.
-We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hzpercent.txt}\%$  measurement of $H(z=\variable{output/mostsensitivez.txt})$ ($1\sigma$ uncertainties).
+We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$  measurement of $H(z=\variable{output/mostsensitivez.txt})$ ($1\sigma$ uncertainties).
 This astrophysics-agnostic spectral siren technique will be key to deliver precise and unbiased cosmological constraints with future observations and in the presence of redshift evolution of the mass spectrum.
@@
 However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by \result{X}$\sigma$.
-We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an \result{overestimate} of \Ho{} at greater than $1\sigma \variable{BPL_bias.txt}\%$ of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and
-the \plp{} model reaches the same level of bias $\variable{PLP_bias.txt}\%$ of the time.
+We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an \result{overestimate} of \Ho{} at greater than $1\sigma \variable{output/BPL_bias_percent.txt}\%$ of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and
+the \plp{} model reaches the same level of bias $\variable{output/PLP_bias_percent.txt}\%$ of the time.
 

### 2024-03-13 src/scripts/bias_study.py
add sanity check plot

@@ -17,6 +17,7 @@ jax.config.update("jax_enable_x64", True)
 # options
 N_CATALOGS=5
 N_SOURCES = N_SOURCES//4 # cut catalogs in half for this study, for computational feasibility
+plot = True
 
 # random number generators
 jax_rng = jax.random.PRNGKey(42)
@@ -27,6 +28,8 @@ m1zinj_det = np.load(paths.data / "gw_data/m1zinj_det.npy")
 dLinj_det = np.load(paths.data / "gw_data/dLinj_det.npy")
 log_pinj_det = np.load(paths.data / "gw_data/log_pinj_det.npy")
 
+if plot:
+    import matplotlib.pyplot as plt
 bias_PLP = np.zeros(N_CATALOGS)
 bias_BPL = np.zeros(N_CATALOGS)
 for i in trange(N_CATALOGS):
@@ -52,7 +55,8 @@ for i in trange(N_CATALOGS):
     id = az.from_numpyro(mcmc)
     id.to_netcdf(paths.data / f"bias/mcmc_parametric_PLP_{int}.nc4")
     bias_PLP[i] = (id.posterior['H0'][0].mean() - H0_FID)/id.posterior['H0'][0].std()
-
+    if plot:
+        plt.hist(id.posterior['H0'][0],density=True, bins=50,histtype='step',color='green',alpha=0.5,lw=0.5)
     # Inference - broken power law
     nuts_settings = dict(target_accept_prob=0.9, max_tree_depth=10,dense_mass=False)
     nuts_kernel = numpyro.infer.NUTS(BPL,**nuts_settings)
@@ -66,6 +70,8 @@ for i in trange(N_CATALOGS):
     id = az.from_numpyro(mcmc)
     id.to_netcdf(paths.data / f"bias/mcmc_parametric_BPL_{int}.nc4")
     bias_BPL[i] = (id.posterior['H0'][0].mean() - H0_FID)/id.posterior['H0'][0].std()
+    if plot:
+        plt.hist(id.posterior['H0'][0],density=True, bins=50,histtype='step',color='orange',alpha=0.5,lw=0.5)
 
 # calcualte summary statistics and save
 percent_bias_PLP = np.sum(bias_PLP>1)/N_CATALOGS * 100
@@ -75,4 +81,7 @@ with open(paths.output / "PLP_bias_percent.txt", "w") as f:
 with open(paths.output / "BPL_bias_percent.txt", "w") as f:
[…]

### 2024-03-13 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 The details of the data simulation are described in Appendix~\ref{ap:data generation}.
-We then use this catalog to infer \Ho{} using three different models for the source-frame mass distribution:
+We then use this catalog to infer \jme{the cosmic expansion rate, focusing on \Ho{},} using three different models for the source-frame mass distribution:
 \begin{enumerate}
@@
 The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
-Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,60\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
+Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,60\Hunits]$ \jme{[upper bound cannot be correct :)]}, indicating that the data is informative despite the flexibility of the population model.
 

### 2024-03-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-\author{Michael Zevin}
+\author[0000-0002-0147-0835]{Michael Zevin}
 \affiliation{The Adler Planetarium, 1300 South DuSable Lake Shore Drive, Chicago, 60605, IL, USA}
@@
 
-\section{Methods I: Spectral Siren Cosmology}
+\section{Spectral Siren Cosmology}
 \label{sec:ss}
@@
 
-\section{Methods II: Gaussian process-based mass distribution}
+\section{Non-parametric mass distribution}
 \label{sec:model}
@@
 
-We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within \result{X$\sigma$}.
-However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by \result{X}$\sigma$.
-We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an \result{overestimate} of \Ho{} at greater than $1\sigma \variable{output/BPL_bias_percent.txt}\%$ of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and
+We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within %$\variable{output/PLPh0offset.txt}\sigma$.
+However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by %$\variable{output/BPLh0offset.txt}\sigma$.
+We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and
 the \plp{} model reaches the same level of bias $\variable{output/PLP_bias_percent.txt}\%$ of the time.
 
-This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the \ac{GP}-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution.
+This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the \ac{GP}-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution: the true value is recovered at the %$\variable{output/nonparh0offset.txt}\sigma$ level.
 The \ac{GP}-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters. 
@@
 The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
-Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,60\Hunits]$ \jme{[upper bound cannot be correct :)]}, indicating that the data is informative despite the flexibility of the population model.
+Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,120\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
 
@@
     \includegraphics[width=\textwidth]{figures/O5_GP_pm.pdf}
-    \caption{Spectral siren measurement for an \ac{O5}-like catalog with the ``correct'' parametric model (\plp, green), an ``incorrect'' parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{GP}, blue).
+    \caption{Spectral siren measurement for an \ac{O5}-like catalog with the ``correct'' parametric model (\plp, green), an deliberately incorrect parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{GP}, blue).
     The left panel shows the recovered source frame primary mass distribution for each model, and the simultaneously-inferred posteriors on \Ho{} are shown in the right panel.
@@
 \acp{GW} are unique messengers in that they carry both redshift and distance information, making them remarkably clean probes of the Universe's expansion history. 
-However, the current method of determining \ac{GW} redshifts via the distribution of their source masses employs an assumption on the shape of their population.
-This choice of mass distribution, often presumed to be necessary or fundamental to the method, introduces a systematic bias to a measurement that otherwise makes very few assumptions.
+However, \jme{current approaches to the spectral siren method} %the current method of 
+determining \ac{GW} redshifts via the distribution of their source masses employs an assumption on the shape of their population\jme{, which it is typically modeled with simplified parametric functions}.
+This choice of mass distribution, often presumed to be necessary or fundamental to the method, \jme{may introduce} %introduces 
+a systematic bias to a measurement \jme{of the cosmological parameters} that otherwise makes very few assumptions.
 
@@
 
-Additionally, we find spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ at the end of \acf{O5}.
-This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}, meaning upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances.
-Proposed next generation detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to $z\sim40$.
+Additionally, we find \jme{that} spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ at the end of \acf{O5}.
+This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}\jme{, which can only be detected relatively locally}, meaning \jme{that} upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances \jme{by \ac{GW}}.
+Proposed next generation \jme{\ac{GW}} detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to \jme{coalescence binaries across cosmic time}$z\sim40$.
 At that time, cosmological surveys such as those enabled by the Nancy Grace Roman Space Telescope and the Vera Rubin Observatory are expected to be able to precisely measure $H(z)$ to $z\sim3$ \citep{spergel_wide-field_2015}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
@@
 53101 and 37766).
+    %Mike
+    M.Z. gratefully acknowledges funding from the Brinson Foundation in support of astrophysics research at the Adler Planetarium.
     %Daniel

### 2024-03-14 src/scripts/nonparametric_inference.py

@@ -12,7 +12,7 @@ import numpy as np
 
 jax.config.update("jax_enable_x64", True)
 
-NSAMPS=50
+NSAMPS=1000
 
 # random number generators
 jax_rng = jax.random.PRNGKey(425)

### 2024-03-14 src/scripts/submit.sh

@@ -1,7 +1,7 @@
 #!/bin/bash
-#SBATCH --job-name=paper_fitOm0_gwmc
-#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_gwmc.out
-#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_gwmc.err
+#SBATCH --job-name=paper_gwmc
+#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_gwmc.out
+#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_gwmc.err
 #SBATCH --account=kicp
 #SBATCH --time=48:00:00
 #SBATCH --partition=kicp-gpu
@@ -13,4 +13,6 @@ module load cuda/11.5
 echo "loaded cuda"
 module load python/anaconda-2022.05
 echo "loaded anaconda"
-showyourwork build
+source activate jax_gpu
+echo "activated env"
+python /home/afarah/p2/spectral-sirens-with-GPs/src/scripts/nonparametric_inference.py

### 2024-03-14 src/scripts/nonparametric_numbers.py
add offset stats

@@ -1,6 +1,7 @@
 import arviz as az
 import numpy as np
 import paths
+from data_generation import H0_FID
 
 id = az.InferenceData.from_netcdf(paths.data / "mcmc_nonparametric.nc4")
 h0samps = id.posterior['H0'][0]
@@ -11,3 +12,7 @@ lower = np.mean(h0samps)-np.percentile(h0samps,5)
 upper = np.percentile(h0samps,95)-np.mean(h0samps)
 with open(paths.output / "nonparh0CI.txt", "w") as f:
     print(f"${np.mean(h0samps):.1f}"+"^{+"+f"{lower:.1f}"+"}"+"_{-"+f"{upper:.1f}"+"}$", file=f)
+
+nonpar_offset = np.abs(h0samps.mean()-H0_FID)/h0samps.std()
+with open(paths.output / "nonparh0offset.txt","w") as f:
+    print(f"{nonpar_offset:.1f}",file=f)

### 2024-03-14 src/scripts/parametric_inference.py

@@ -4,6 +4,7 @@ import paths
 Clight = jgwcosmo.Clight
 from priors import PLP, BPL
 from nonparametric_inference import NSAMPS
+from data_generation import H0_FID
 
 # inference
 import jax
@@ -46,6 +47,9 @@ if  __name__ == "__main__":
     # save results
     id = az.from_numpyro(mcmc)
     id.to_netcdf(paths.data / "mcmc_parametric_PLP.nc4")
+    offset = np.abs(id.posterior['H0'][0].mean()-H0_FID)/id.posterior['H0'][0].std()
+    with open(paths.output / "PLPh0offset.txt","w") as f:
+        print(f"{offset:.1f}",file=f)
 
     # Inference - broken power law
     nuts_settings = dict(target_accept_prob=0.9, max_tree_depth=10,dense_mass=False)
@@ -56,6 +60,9 @@ if  __name__ == "__main__":
     mcmc = numpyro.infer.MCMC(nuts_kernel,num_warmup=NSAMPS,num_samples=NSAMPS,
                               num_chains=1,progress_bar=True)   
     mcmc.run(jax_rng,**kwargs)
+    offset = np.abs(id.posterior['H0'][0].mean()-H0_FID)/id.posterior['H0'][0].std()
+    with open(paths.output / "BPLh0offset.txt","w") as f:
+        print(f"{offset:.1f}",file=f)
 
     # save results
     id = az.from_numpyro(mcmc)

### 2024-03-14 src/scripts/submit.sh

@@ -1,13 +1,12 @@
 #!/bin/bash
-#SBATCH --job-name=paper_gwmc
-#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_gwmc.out
-#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_gwmc.err
+#SBATCH --job-name=paper_parametric
+#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_parametric.out
+#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_parametric.err
 #SBATCH --account=kicp
 #SBATCH --time=48:00:00
-#SBATCH --partition=kicp-gpu
+#SBATCH --partition=kicp
 #SBATCH --nodes=1
 #SBATCH --mem=128G
-#SBATCH --gres=gpu:1
 
 module load cuda/11.5
 echo "loaded cuda"
@@ -15,4 +14,4 @@ module load python/anaconda-2022.05
 echo "loaded anaconda"
 source activate jax_gpu
 echo "activated env"
-python /home/afarah/p2/spectral-sirens-with-GPs/src/scripts/nonparametric_inference.py
+python /home/afarah/p2/spectral-sirens-with-GPs/src/scripts/parametric_inference.py

### 2024-03-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-After passing the simulated events through projected detector selection effects, the resulting catalog has \variable{output/num_found_events.txt}events, consistent with the numbers projected for \ac{O5} by \citet{kiendrebogo_observing_2023}.
+After passing the simulated events through projected detector selection effects, the resulting catalog has $\variable{output/num_found_events.txt}$ events, consistent with the numbers projected for \ac{O5} by \citet{kiendrebogo_observing_2023}.
 We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from \ac{O5} detectors.

### 2024-03-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-We find that using this model allows for a \variable{output/nonparh0percent.txt}\% measurement of \Ho{} by the end of \ac{O5}, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
+We find that using this model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt}$ by the end of \ac{O5}.%, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
 While constraints on \Ho{} are used in this work to quantify the accuracy and precision of our non-parametric method, we emphasize that spectral sirens are able to constrain $H(z)$ out to redshifts that are inaccessible by electromagnetic observations, especially with next-generation detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn, You:2020wju}\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}.
@@
 
-We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within %$\variable{output/PLPh0offset.txt}\sigma$.
-However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by %$\variable{output/BPLh0offset.txt}\sigma$.
+We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within $\variable{output/PLPh0offset.txt}\sigma$.
+However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by $\variable{output/BPLh0offset.txt}\sigma$.
 We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and
@@
 
-This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the \ac{GP}-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution: the true value is recovered at the %$\variable{output/nonparh0offset.txt}\sigma$ level.
+This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the \ac{GP}-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution: the true value is recovered at the $\variable{output/nonparh0offset.txt}\sigma$ level.
 The \ac{GP}-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters. 
@@
 \subsection{Projections for Future Measurements}
-The fit shown in Figure~\ref{fig:O5_GP} implies that we can expect to measure \Ho{} to \variable{output/nonparh0percent.txt}\% ($1-\sigma$) by the end of \ac{O5} using the \ac{GP}-based spectral siren method.
+The fit shown in Figure~\ref{fig:O5_GP} implies that we can expect to measure \Ho{} to $\variable{output/nonparh0percent.txt}\%$ ($1-\sigma$ uncertainties) by the end of \ac{O5} using the \ac{GP}-based spectral siren method.
 However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\sim 3$, with most sources lying near redshift $\sim 1.2$ \citep{chen_distance_2021}.
@@
 
-We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens and a \result{X\%} %$\variable{output/parh0percent.txt}\%$
+We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens and a $\variable{output/PLPh0percent.txt}\%$ 
 measurement with parametric spectral sirens, demonstrating a \result{negligible/substantial/marginal} increase in precision when using the correct parametric model.
@@
 
-Additionally, we find \jme{that} spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ at the end of \acf{O5}.
+Additionally, we find that spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ at the end of \acf{O5}.
 This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}\jme{, which can only be detected relatively locally}, meaning \jme{that} upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances \jme{by \ac{GW}}.
@@
 The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
-The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho{} ($\variable{output/H0_FID.txt}\Hunits$) is consistent with its inferred posterior (\variable{output/nonparh0CI.txt}$\Hunits$).
+The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho{} ($\variable{output/H0_FID.txt}\Hunits$) is consistent with its inferred posterior ($\variable{output/nonparh0CI.txt}\Hunits$).
 At very high and very low masses, there is both a lack of data and little detector sensitivity.
@@
 The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
-Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,60\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
+Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,120\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
 

### 2024-03-14 src/scripts/nonparametric_inference_fitOm0.py
increase NSAMPs

@@ -11,7 +11,7 @@ import numpyro
 # data
 import numpy as np
 
-NSAMPS = 50
+NSAMPS = 1500
 
 jax.config.update("jax_enable_x64", True)
 # random number generators

### 2024-03-14 src/scripts/nonparametric_inference.py
put statistics calcuation in inference script

@@ -48,3 +48,18 @@ if  __name__ == "__main__":
     # save results
     id = az.from_numpyro(mcmc)
     id.to_netcdf(paths.data / "mcmc_nonparametric.nc4")
+
+    # calculate stats
+    h0samps = id.posterior['H0'][0]
+    from data_generation import H0_FID
+
+    with open(paths.output / "nonparh0percent.txt", "w") as f:
+        print(f"{np.std(h0samps)/np.mean(h0samps)*100:.1f}", file=f)
+    lower = np.mean(h0samps)-np.percentile(h0samps,5)
+    upper = np.percentile(h0samps,95)-np.mean(h0samps)
+    with open(paths.output / "nonparh0CI.txt", "w") as f:
+        print(f"${np.mean(h0samps):.1f}"+"^{+"+f"{lower:.1f}"+"}"+"_{-"+f"{upper:.1f}"+"}$", file=f)
+
+    nonpar_offset = np.abs(h0samps.mean()-H0_FID)/h0samps.std()
+    with open(paths.output / "nonparh0offset.txt","w") as f:
+        print(f"{nonpar_offset:.1f}",file=f)

### 2024-03-14 src/scripts/parametric_inference.py
add percentage calculation

@@ -50,6 +50,8 @@ if  __name__ == "__main__":
     offset = np.abs(id.posterior['H0'][0].mean()-H0_FID)/id.posterior['H0'][0].std()
     with open(paths.output / "PLPh0offset.txt","w") as f:
         print(f"{offset:.1f}",file=f)
+    with open(paths.output / "PLPh0percent.txt","w") as f:
+        print(f"{np.std(id.posterior['H0'][0])/np.mean(id.posterior['H0'][0])*100:.1f}",file=f)
 
     # Inference - broken power law
     nuts_settings = dict(target_accept_prob=0.9, max_tree_depth=10,dense_mass=False)

### 2024-03-14 src/scripts/submit.sh
update

@@ -1,12 +1,13 @@
 #!/bin/bash
-#SBATCH --job-name=paper_parametric
-#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_parametric.out
-#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_parametric.err
+#SBATCH --job-name=paper_fitOm0_gwmc
+#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_gwmc.out
+#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_gwmc.err
 #SBATCH --account=kicp
 #SBATCH --time=48:00:00
-#SBATCH --partition=kicp
+#SBATCH --partition=kicp-gpu
 #SBATCH --nodes=1
 #SBATCH --mem=128G
+#SBATCH --gres=gpu:1
 
 module load cuda/11.5
 echo "loaded cuda"
@@ -14,4 +15,4 @@ module load python/anaconda-2022.05
 echo "loaded anaconda"
 source activate jax_gpu
 echo "activated env"
-python /home/afarah/p2/spectral-sirens-with-GPs/src/scripts/parametric_inference.py
+python /home/afarah/p2/spectral-sirens-with-GPs/src/scripts/nonparametric_inference_fitOm0.py

### 2024-03-14 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within $\variable{output/PLPh0offset.txt}\sigma$.
-However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by $\variable{output/BPLh0offset.txt}\sigma$.
+We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within %$\variable{output/PLPh0offset.txt}\sigma$.
+However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by %$\variable{output/BPLh0offset.txt}\sigma$.
 We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and
@@
 
-This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the \ac{GP}-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution: the true value is recovered at the $\variable{output/nonparh0offset.txt}\sigma$ level.
+This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the \ac{GP}-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution: the true value is recovered at the %$\variable{output/nonparh0offset.txt}\sigma$ level.
 The \ac{GP}-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters. 
@@
 
-We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens and a $\variable{output/PLPh0percent.txt}\%$ 
+We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens and a% $\variable{output/PLPh0percent.txt}\%$ 
 measurement with parametric spectral sirens, demonstrating a \result{negligible/substantial/marginal} increase in precision when using the correct parametric model.

### 2024-03-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 %\author{@afarah18}
-\author{Amanda M. Farah}
+\author[0000-0002-6121-0285]{Amanda M. Farah}
 \email{afarah@uchicago.edu}
@@
 
-\author{Jose Mar\'ia Ezquiaga}
+\author[0000-0002-7213-3211]{Jose Mar\'ia Ezquiaga}
 %\email{jose.ezquiaga@nbi.ku.dk}
@@
 
-\section{}
+d\section{}
 \textcolor{red}{Showyourwork Guidelines:}
@@
 
-Perhaps because of the analogy to electromagnetic spectra in which the location of the source frame spectral features is well-known, or because of the examples given in \cite{farr_future_2019} and \cite{chernoff_gravitational_1993}, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to do this measurement, from e.g.
-theoretical expectations of the location of a pulsational-pair instability pileup, upper mass gap, or maximum neutron star mass.
+Perhaps because of the analogy to electromagnetic spectra in which the location of the source frame spectral features is well-known, or because of the examples given in \cite{farr_future_2019} and \cite{chernoff_gravitational_1993}, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to do this measurement, from theoretical expectations of e.g.
+the location of a pulsational-pair instability pileup, upper mass gap, or maximum neutron star mass.
 This would certainly be problematic, as mismodeling of the specific feature would hinder the accuracy of the method and introduce biases in the cosmological inference \citep{Mukherjee:2021rtw,mastrogiovanni_importance_2021,pierra_study_2023}.
+Such systematic bias is illustrated in Figure~\ref{fig:O5_GP}, where we show the result of fitting simulated data with a model that has the true mass distribution contained within its prior and one that does not.
+The latter does not accurately infer the true value of \Ho{}, whereas the former does.
+
+\comment{below text is WIP}
+This is particularly troubling in light of the fact that there remains much uncertainty over the cause, location, and robustness of currently-identified features in the \ac{GW} source mass distribution.
+While four or more features have been identified in the current data, \citep{viabhav, abbott_population_2023,bruce_aint} only three of these are statistically robust \citep{farah_things_2023}.
+Additionally, the morphology of the mass dist is caused by the interplay of several complex physical processes, such as compact binary evolution, pair instability, globular cluster dynamics, and the metalicity content of massive stars, each of which has large theoretical uncertainties.
+\citep{many ppl}
+These large theoretical uncertainties mean that accurate predictions for the shape of the \ac{GW} source mass distribution is not currently possible.
+If such \emph{a priori} knowledge of the mass distribution's morphology were requisite for a spectral siren method, the effectiveness of such a measurement would be significantly hampered.
+\comment{end WIP}
 
@@
 Therefore the location, shape, and existence of all features in the mass distribution can be inferred simultaneously with $H(z)$.
-\jme{[do we want to comment somewhere about the observed features in the data? and perhaps quickly mention about their robustenss?]}
 
 In this work, we explicitly demonstrate that no \emph{a priori} knowledge about the shape of the \ac{CBC} mass spectrum is necessary to use the spectral siren methodology. 
-We do this by inferring $H(z)$ with a flexible, non-parametric model for the mass distribution of \acp{CBC}. 
+We do this by inferring $H(z)$ with a flexible, non-parametric model for the mass distribution of \acp{CBC} (blue lines in Figure~\ref{fig:O5_GP}). 
 This model makes minimal prior assumptions about the shape of the mass distribution, enabling it to accurately infer a wide range of morphologies.
@@
 
-We find that using this model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt}$ by the end of \ac{O5}.%, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
+We find that using our non-parametric model model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ by the end of \ac{O5} \jme{when the detectors will reach their design sensitivity}.
+%, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
 While constraints on \Ho{} are used in this work to quantify the accuracy and precision of our non-parametric method, we emphasize that spectral sirens are able to constrain $H(z)$ out to redshifts that are inaccessible by electromagnetic observations, especially with next-generation detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn, You:2020wju}\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}.
@@
 Section~\ref{sec:model} describes the non-parametric mass distribution we develop for use within the spectral siren method.
-We show the results of using parametric and non-parametric mass distributions in Section~\ref{sec:results}, as well as projections for future constraints on $H(z)$.
+We present the results of using parametric and non-parametric mass distributions in Section~\ref{sec:results}, as well as projections for future constraints on $H(z)$.
 In Section~\ref{sec:discussion} we discuss the implications of our results and outline future work.
@@
 
-\section{Spectral Siren Cosmology}
+\begin{figure*}
+    \centering
+    \includegraphics[width=\textwidth]{figures/O5_GP_pm.pdf}
+    \caption{Spectral siren measurement for an \ac{O5}-like catalog with the correct parametric model (\plp, green), a deliberately incorrect parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{GP}, blue).
+    The left panel shows the recovered source frame primary mass distribution for each model, and the simultaneously-inferred posteriors on \Ho{} are shown in the right panel.
+    The mass distribution and \Ho{} value used to generate the data are shown by a solid black line in each panel. 
+    The deliberately incorrect parametric model fails to recover the true mass distribution and therefore produces a biased estimate of \Ho{}, whereas both the correct and non-parametric models recover  the mass distribution and \Ho{}.
+    As the true mass distribution is unknown for real observations, using a non-parametric model mitigates systematic uncertainty that would otherwise arise from mismodeling the \ac{CBC} population, though it does introduce additional statistical uncertainty.
+    }
+    \label{fig:O5_GP}
+    \script{nonparametric_twopanel.py}
+\end{figure*}
+
+\section{Cosmology with an astrophysically-agnostic mass distribution model}
+\label{sec:methods}
+\subsection{The Spectral Siren Method}
 \label{sec:ss}
+To simultaneously infer cosmological parameters and the population of GW sources, we employ a hierarchical Bayesian analysis.
+This allows us to ``undo'' the selection effects of \ac{GW} detectors to obtain a true, or astrophysical, population and constrain the cosmic expansion history.
 
-To simultaneously infer cosmological parameters and the population of GW sources, we employ a hierarchical Bayesian analysis.
-This allows us to ``undo'' the selection effects of \ac{GW} detectors to obtain a true, or astrophysical, population \jme{and constrain the cosmic expansion history}.
-
-Given \ac{GW} data $\{d\}$ that contains $N_{\text{evs}}$ detected \ac{GW} signals, each with parameters $\theta_i$, the posterior probability of the \ac{GW} sources being drawn from an underlying population and background cosmology described by hyper-parameters $\Lambda$ is \citep{loredo, taylor, mandel}
+Given \ac{GW} data $\{d\}$ that contains $N_{\text{evs}}$ detected \ac{GW} signals, each with parameters $\theta_i$, the posterior probability of the \ac{GW} sources being drawn from an underlying population and background cosmology described by hyper-parameters $\Lambda$ is \citep{loredo_handling_2009, Taylor:2011fs, mandel_extracting_2019}
 \begin{equation}
@@
 Here, $\Lambda$ is the set of population parameters, or hyper-parameters, and $p(\Lambda)$ is the prior on $\Lambda$.
-$\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ is the detector frame merger rate density conditioned on hyper-parameters $\Lambda$, this quantity is also referred to as the population model.
+$\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ is the detector frame merger rate density of \acp{BBH} conditioned on hyper-parameters $\Lambda$, this quantity is also referred to as the population model.
 Following \citet{callister_parameter-free_2023}, we use a semicolon to explicitly indicate that this is a function of $\Lambda$, not a density over $\Lambda$.
-$N_{\text{exp}}(\Lambda)$ is the expected number of detections given $\Lambda$ and the \ac{GW} detector sensitivity, and is calculated using a Monte Carlo sum over $N_{\text{inj}}$ found signals injected into the data stream \citep[see][for a detailed explanation of this process]{essick}.
+$N_{\text{exp}}(\Lambda)$ is the expected number of detections given $\Lambda$ and the \ac{GW} detector sensitivity, and is calculated using a Monte Carlo sum over $N_{\text{inj}}$ found signals injected into the data stream \citep[see][for a detailed explanation of this process]{essick_estimating_2021, essick_precision_2022}.
+In this work, we have restricted our analysis to the \ac{BBH} mass distribution.
+However, the method can be trivially extended to the full mass distribution of \acp{CBC} \citep[e.g.][]{fishbach_does_2020, ezquiaga_spectral_2022}.
 
@@
 Finally, $p(\Lambda)$ is the prior on the population and parameters on the background cosmology, which we discuss in further detail below.
-We infer the posterior in Equation~\ref{eq:inhomog-poisson} using the no-u-turn sampler for Hamiltonian Monte Carlo within \texttt{numpyro} \citep{hoffman_no-u-turn_2011, numpyro}.
+We infer the posterior in Equation~\ref{eq:inhomog-poisson} using the no-u-turn sampler for Hamiltonian Monte Carlo within \texttt{numpyro} \citep{hoffman_no-u-turn_2011, bingham_pyro_2019,phan_composable_2019}.
 
@@
 masses and redshifts, neglecting spins as they do not contain cosmological information.
-Thus, $\theta = \{m_1,m_2,z\}$. 
-\jme{The population can be modeled with different levels of complexity and flexibility. 
+For simplicity, we additionally assume that the population has equal masses, $m_1=m_2$.
+Thus, $\theta = \{m_1,z\}$. 
+The population can be modeled with different levels of complexity and flexibility. 
 As described in detail in Section~\ref{sec:model}, we model the distribution of primary masses in an agnostic way, using a \acl{GP}. 
 We also compare this non-parametric reconstruction with different parametric models.  
-For simplicity we assume that the population has equal masses, $m_1=m_2$, and that its redshift distribution is uniform-in-comoving volume, as in Equation~\ref{eq:underlying redshift dist}. }
-%We assume a uniform-in-comoving volume redshift distribution, as in Equation~\ref{eq:underlying redshift dist}.
-%We assume $m_1=m_2$ and model the distribution of primary masses using a \acl{GP} (described in Section~\ref{sec:model}).
+For all models considered in this work, we assume a uniform-in-comoving volume redshift distribution, as in Equation~\ref{eq:underlying redshift dist}.
 
-The spectral siren method of measuring cosmological parameters relies on the assumption that all compact binaries follow the same mass distribution at all redshifts\footnote{This assumption does not strictly have to hold, as \citet{ezquiaga_spectral_2022} show that the source frame masses can follow different distributions at different redshifts, so long as this redshift evolution of the mass distribution does not exactly mimic cosmology (i.e.
-all features cannot exhibit a monotonic rightward shift with redshift).
-However, for the redshift range considered in this work, we assume the mass distribution does not evolve at all with redshift.}.
+The spectral siren method of measuring cosmological parameters relies on the assumption that either all compact binaries follow the same mass distribution at all redshifts, or that any redshift evolution of the mass distribution does  not exactly mimic cosmology (i.e.
+all features cannot exhibit a monotonic rightward shift with redshift) \citep{ezquiaga_spectral_2022}.
+For the redshift range considered in this work, we assume the mass distribution has no redshift evolution.
 The likelihood in Equation~\ref{eq:inhomog-poisson} is therefore maximized when the relationship between redshift and luminosity distance makes all source frame masses follow the same distribution.
@@
 The full set of hyper-parameters therefore includes the cosmological parameters that dictate the $D_L$-$z$ relation 
-\jme{such as the local expansion rate \Ho, the present fractional energy density of dark matter \Omm, dark energy $\Omega_Lambda$ and radiation $\Omega_r$, or the equation of state of dark energy $w$.} %: \Ho, \Omm, $\Omega_\Lambda, \Omega_r,$ and $w$.
+\jme{such as the local expansion rate \Ho, the present fractional energy density of dark matter \Omm, dark energy $\Omega_\Lambda$ and radiation $\Omega_r$, or the equation of state of dark energy $w$.} %: \Ho, \Omm, $\Omega_\Lambda, \Omega_r,$ and $w$.
 In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and \jme{$w=-1$} and use uniform priors on \Ho{} and \Omm{}. 
@@
 
-\section{Non-parametric mass distribution}
+\subsection{\Acl{GP}-based mass distribution}
 \label{sec:model}
 
-\jme{Our goal is to model the population of \ac{GW} in an agnostic way. 
-We choose to do it with a \Acfp{GP}. 
-\ac{GP}} 
-%\Acfp{GP} 
-are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
+Our goal is to model the population of \ac{GW} in an astrophysically-agnostic way. 
+We choose to do it with \Acfp{GP}, a common tool for non-parametric inference. 
+\Acp{GP} are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
 Their smoothness properties make them widely useful in \ac{GW} data analysis and beyond for regression problems, such as modeling time-domain waveforms \citep{doctor_statistical_2017, huerta_eccentric_2018} and the neutron star equation of state \citep{landry_nonparametric_2019}, density estimation problems, such as estimating posterior densities of single-event parameters from parameter estimation samples ~\citep{demilio_density_2021}, and as a prior on histogram bin heights for population inference \citep{mandel_model-independent_2017, li_flexible_2021, ray_non-parametric_2023}.
 
-%added a break here
-Our use case is slightly different \jme{to previous analyses}.
-We will utilize a \ac{GP} as a prior on the functions that describe the primary mass distribution of \acp{CBC}.
+Our use case is slightly different from previous analyses:
+we utilize a \ac{GP} as a prior on the functions that describe the primary mass distribution of \acp{CBC}.
 It will therefore replace $p(\Lambda)$ in Equation~\ref{eq:inhomog-poisson}, except in the case of cosmological parameters, which all have uniform priors.
@@
 Practically, the only difference in the inference of the population when using a GP versus other modeling choices is that $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is determined directly by a realization of the \ac{GP}, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
-%joined paragraphs here
+%joined paragraphs here.
+OK sounds good!
 In other words, when using parametric models, $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is calculated by evaluating a specific functional form described by a small set of hyper-parameters.
-With the GP approach, the hyper-parameters describing the mass distribution are the rate at each event-level posterior sample's source frame mass, and the rate at each found injection's source frame mass.
+With the \ac{GP} approach, the hyper-parameters describing the mass distribution are the rate at each event-level posterior sample's source frame mass, and the rate at each found injection's source frame mass.
 % The source frame mass is different for each possible value of \Ho{} and \Omm{}, so the hyper-parameters are defined in different locations for each draw from the hyper-posterior.
@@
 It defines the notion of similarity between adjacent points and thereby encodes our assumptions about the smoothness of the source frame mass distribution \citep{rasmussen_gaussian_2006}.
-We use a Mat\'ern kernel \citep{handcock_and_stein,stein_1999} with $\nu = 5/2$, but have repeated the analysis with $\nu=3/2$ and $\infty$, finding little impact on the results, except that the $\nu=\infty$ case (also called the ``squared exponential'' kernel) produces a slightly more jagged mass distribution.
+We use a Mat\'ern kernel \citep{handcock_bayesian_1993, stein_interpolation_1999} with $\nu = 5/2$, but have repeated the analysis with $\nu=3/2$ and $\infty$, finding little impact on the results, except that the $\nu=\infty$ case (also called the ``squared exponential'' kernel) produces a slightly more jagged mass distribution.
 Mat\'ern kernels have two parameters besides the mean that determine their properties: a length scale $l$ and variance $s$.
@@
 We use ``penalized-complexity'' priors on the hyper-hyper-parameters to enforce that the model does not create small-scale structure uninformed by data, thereby avoiding over-fitting \citep{simpson_penalising_2017,simpson_garcpas_2022}.
-Explicitly, the priors on $l$ and $s$ have less than 5\% support for correlation lengths smaller than the average spacing between event-level posterior means.
+Explicitly, the priors on $l$ and $s$ are Fr\'echet and Gamma distributions, respectively, and are defined to have less than 5\% support for correlation lengths smaller than the average spacing between event-level posterior means.
 
-\ac{GP} evaluations notoriously scale as the cube of the number of data points, making them unwieldy with large data sets such as the ones expected for \ac{O5}.
+The time to evaluate a \acp{GP} notoriously scales as the cube of the number of data points, making \acp{GP} unwieldy with large data sets such as the ones expected for \ac{O5}.
 We therefore make two approximations to a full \ac{GP} to increase computational efficiency.
@@
 
-Algorithmically, each hyper-likelihood \comment{is that a word?} evaluation contains the following steps: \script{priors.py}
+Algorithmically, each likelihood evaluation contains the following steps: \script{priors.py}
 \begin{enumerate}
@@
     \item Draw a single \ac{GP} realization with a kernel defined by $l$ and $s$.
-    This is defined on a regular grid of source-frame masses and evaluated using the \texttt{QuasisepSolver}.
+    This is defined on a regular grid of source-frame masses and evaluated using the \texttt{QuasisepSolver} in \texttt{tinygp}.
     \item Interpolate the \ac{GP} at each event posterior sample and injection source frame mass (from step 3)
@@
 We then demonstrate that our flexible model alleviates this bias without the need to know the mass distribution's morphology \emph{a priori}.
-To illustrate this, we generate a catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
-This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after \ac{O5}.
+To illustrate this, we generate a catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{abbott_population_2023}.
+This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after \jme{one year (?) of observation in} \ac{O5}.
 The details of the data simulation are described in Appendix~\ref{ap:data generation}.
-We then use this catalog to infer \jme{the cosmic expansion rate, focusing on \Ho{},} using three different models for the source-frame mass distribution:
+We then use this catalog to infer the cosmic expansion rate, using three different models for the source-frame mass distribution:
 \begin{enumerate}
     \item the \plp{} model, which includes the true mass distribution within its hyperprior,
-    \item a broken power law, which does not include the true mass distribution within its hyperprior, and 
+    \item a broken power law \jme{[use same font style as \plp{}?]}, which does not include the true mass distribution within its hyperprior, and 
     \item the flexible, \ac{GP}-based model described in Section~\ref{sec:model}, which is able to closely approximate the morphology of the true mass distribution, along with many other morphologies.
@@
 The left panel shows the inferred source frame mass distribution for each of the considered models, and the right panel shows the corresponding posteriors on \Ho{}. 
+\jme{We indicate the true underlying source mass distribution and fiducial \Ho{} value with solid black lines in each panel.}
 
@@
 On the other hand, the lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the \ac{GP} to learn a relatively low merger rate at the edges of the mass distribution.
-Indeed, these features are very informative to the \Ho{} constraint \citep{the_ligo_scientific_collaboration_constraints_2021}.
+Indeed, these features are very informative to the \Ho{} constraint \citep{gwtc3_cosmo}.
 The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
@@
 Collectively, these results indicate that prior knowledge of the shape of the mass distribution is not required to perform a spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=\textwidth]{figures/O5_GP_pm.pdf}
-    \caption{Spectral siren measurement for an \ac{O5}-like catalog with the ``correct'' parametric model (\plp, green), an deliberately incorrect parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{GP}, blue).
-    The left panel shows the recovered source frame primary mass distribution for each model, and the simultaneously-inferred posteriors on \Ho{} are shown in the right panel.
-    The mass distribution and \Ho{} value used to generate the data are shown by a solid black line in each panel. 
-    The ``incorrect'' parametric model fails to recover the true mass distribution and therefore produces a biased estimate of \Ho{}, whereas both the ``correct'' and non-parametric models recover  the mass distribution and \Ho{}.
-    As the true mass distribution is unknown, using a non-parametric model mitigates the systematic uncertainty from mismodeling the \ac{CBC} population, though it does introduce additional statistical uncertainty.
-    }
-    \label{fig:O5_GP}
-    \script{nonparametric_twopanel.py}
-\end{figure*}
 
@@
 This means that future \acp{GW} observations will be more sensitive $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters.
-We demonstrate this by repeating the same analysis while also fitting for the local matter density, \Omm.
-For \ac{O5}, this informs $H(z)$ up to $z \sim 3$.
+We demonstrate this by repeating the same analysis while also fitting for the local matter density, \Omm, but it is equally plausible to constrain other cosmological parameters governing $H(z)$.
 The result is shown in Figure~\ref{fig:O5_corner}.
@@
 
-Next-generation detectors will be sensitive to a much larger range of redshifts \citep{something!} and will therefore break the degeneracy between cosmological parameters, allowing for tighter constraints on each.
-However, the low cosmological volume (and thus low number of mergers) at low redshift will always limit the constraining power of \acp{GW} at $z=0$, making spectral sirens relatively more sensitive to cosmological parameters that affect higher redshifts than to \Ho.
+Next-generation detectors will be sensitive to a much larger range of redshifts \citep{et_steering_committee_einstein_2020, evans_horizon_2021} and will therefore break the degeneracy between cosmological parameters, allowing for tighter constraints on each.
+However, the low cosmological volume (and thus low number of mergers) at low redshift will always limit the constraining power of spectral sirens at $z=0$, potentially making this method more sensitive to cosmological parameters that affect higher redshifts than to \Ho.
+Combining spectral sirens with other methods that are more sensitive to the local expansion rate, such as those that employ electromagnetic counterparts, can increase the precision of \ac{GW} standard sirens at all redshifts \citep[e.g.][]{Chen:2024gdn}.
 
@@
 
-\subsubsection{BONUS: XG}
-To demonstrate the capabilities of next-generation detectors, we repeat the full analysis with \result{1000} events indicative of Cosmic Explorer's projected best-measured events within the first month alone.
-We fit the dark energy equation of state parameter, $w$, along with \Ho, and show results in Figure~\ref{fig:XG_corner}.
-We recover unbiased measurements of each parameter and find a \result{X}\% measurement of $w$ and a \result{X}\% measurement of \Ho.
-In comparison, the most recent constraints on these parameters with \result{X experiment} were \result{X}\% and \result{X}\% , respectively.
-This illustrates the unique ability of \acp{GW} to probe the high-redshift universe.
-
+% \subsubsection{BONUS: XG}
+% To demonstrate the capabilities of next-generation detectors, we repeat the full analysis with \result{1000} events indicative of Cosmic Explorer's projected best-measured events within the first month alone.
+% We fit the dark energy equation of state parameter, $w$, along with \Ho, and show results in Figure~\ref{fig:XG_corner}.
+% We recover unbiased measurements of each parameter and find a \result{X}\% measurement of $w$ and a \result{X}\% measurement of \Ho.
+% In comparison, the most recent constraints on these parameters with \result{X experiment} were \result{X}\% and \result{X}\% , respectively.
+% This illustrates the unique ability of \acp{GW} to probe the high-redshift universe.
+% 
 % \begin{figure*}
@@
 \acp{GW} are unique messengers in that they carry both redshift and distance information, making them remarkably clean probes of the Universe's expansion history. 
-However, \jme{current approaches to the spectral siren method} %the current method of 
-determining \ac{GW} redshifts via the distribution of their source masses employs an assumption on the shape of their population\jme{, which it is typically modeled with simplified parametric functions}.
-This choice of mass distribution, often presumed to be necessary or fundamental to the method, \jme{may introduce} %introduces 
-a systematic bias to a measurement \jme{of the cosmological parameters} that otherwise makes very few assumptions.
+However, the current method of determining \ac{GW} redshifts via the distribution of their source masses (i.e.
+spectral sirens) employs an assumption on the shape of their population, typically enforced with simplified parametric functions.
+This choice of mass distribution, often presumed to be necessary or fundamental to the method, may introduce a systematic bias to a measurement of the cosmological parameters that otherwise makes very few assumptions.
 
@@
 
-Additionally, we find that spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ at the end of \acf{O5}.
-This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}\jme{, which can only be detected relatively locally}, meaning \jme{that} upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances \jme{by \ac{GW}}.
-Proposed next generation \jme{\ac{GW}} detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to \jme{coalescence binaries across cosmic time}$z\sim40$.
-At that time, cosmological surveys such as those enabled by the Nancy Grace Roman Space Telescope and the Vera Rubin Observatory are expected to be able to precisely measure $H(z)$ to $z\sim3$ \citep{spergel_wide-field_2015}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
-When combined with other standard siren methods, the low-redshift expansion will also be well-constrained \citep{Chen:2024gdn}.
-While the sensitivity to high redshifts is a feature of the spectral siren method in general, non-parametric methods such as the one presented here will be imperative to avoid systematic biases as next-generation detectors provide additional data.
-
 Work in preparation will extend the method developed here to mass distributions that are allowed to evolve with redshift.
@@
 
-In this work, we have restricted our analysis to the \ac{BBH} mass distribution, \Ho, and \Omm.
-However, the method can be trivially extended to the full mass distribution of \acp{CBC} \citep[e.g.][]{fishbach_does_2020, ezquiaga_spectral_2022} and to constrain other cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$.
+Because \ac{GW} observations are the only data input to spectral sirens, they are uniquely positioned to probe the expansion history of the universe over a wide range of cosmological redshift, rather than just the local expansion \Ho.
+We have shown this by simultaneously measuring \Omm{} and \Ho{}, but the method can be trivially extended to constrain other cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$.
+We find that spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ at the end of \acf{O5}.
+This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}\jme{, which can only be detected relatively locally}.
+meaning \jme{that} upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances \jme{by \ac{GW}}. 
+Proposed next generation \jme{\ac{GW}} detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to \jme{coalescence binaries across cosmic time}$z\sim40$.
+At that time, cosmological surveys such as those enabled by the Nancy Grace Roman Space Telescope and the Vera Rubin Observatory are expected to be able to precisely measure $H(z)$ to $z\sim3$ \citep{spergel_wide-field_2015}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
+When combined with other standard siren methods, the low-redshift expansion will also be well-constrained \citep{Chen:2024gdn}.
+While the sensitivity to high redshifts is a feature of the spectral siren method in general, non-parametric methods such as the one presented here will be imperative to avoid systematic biases as next-generation detectors provide additional data.
 
@@
     A.M.F. is supported by the National Science Foundation Graduate Research Fellowship Program under Grant No. DGE-1746045.
+    %Tom
     %Jose
@@
 
+\software{\texttt{numpyro} \citep{phan_composable_2019,bingham_pyro_2019}, \texttt{tinygp} \citep{foreman-mackey_tinygp_2021}, \texttt{arviz} \citep{kumar_arviz_2019}, \texttt{jax}, \showyourwork \citep{Luger2021}
+          }
+
 \bibliography{bib}
@@
 We generate a mock catalog of \ac{GW} events characteristic of that expected from \acf{O5}. 
-These events are drawn from an underlying population described by the \plp{} mass distribution presented in \citet{talbot} and used in \citet{abbot_1 abbot2} and the redshift distribution presented in \citet{callister_shouts_2020}.
+These events are drawn from an underlying population described by the \plp{} mass distribution presented in \citet{talbot_measuring_2018} and used in \citet{abbott_binary_2019, abbott_population_2023} and the redshift distribution presented in \citet{callister_shouts_2020}.
 \begin{equation}
@@
 When generating the simulated events, we have fixed $\alpha=-2.7,m_{\max}=78\Msun,m_{\min}=10\Msun, \mu=30\Msun,\sigma=7.0\Msun, f_{\text{peak}}=0.05,   z_p=2.4,\alpha_z=1.,\beta_z=3.4$.
-These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{o3b_pop}.
+These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{abbott_population_2023}.
 We do not fit for secondary masses or spins, as they do not carry additional cosmological information.
@@
 We emphasize, however, that the choice of cosmological parameters for data generation is arbitrary and does not impact the results, since we are concerned only with the ability of our method to recover the injected values.
+Throughout the data generation and inference, we use the approximations presented in \citet{adachi_analytical_2012} to efficiently convert between $d_L$ and $z$ for a given set of cosmological parameters \Omm{} and \Ho.
 
@@
 
-After passing the simulated events through projected detector selection effects, the resulting catalog has $\variable{output/num_found_events.txt}$ events, consistent with the numbers projected for \ac{O5} by \citet{kiendrebogo_observing_2023}.
+After passing the simulated events through projected detector selection effects, the resulting catalog has $\variable{output/num_found_events.txt}$ events, consistent with the numbers projected for \ac{O5} by \citet{kiendrebeogo_updated_2023}.
 We use the software package \texttt{GWMockCat} \citep{farah_things_2023} to simulate posterior samples for these events with measurement uncertainties typical of those expected from \ac{O5} detectors.
 \texttt{GWMockCat} also simulates a set of software injections, which we use to estimate selection effects in the inference.
-To determine the detectability of both injections and simulated events in O5, we use the projected \ac{O5} LIGO power spectral density \citep{dccpage} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
-The full procedure for this mock data generation process is described in \citet{fishbach_where, farah_things_2023, essick_dagnabbit_2023}, and the exact settings used for \texttt{GWMockCat} are made available in the accompanying data release.
+To determine the detectability of both injections and simulated events in O5, we use the projected \ac{O5} LIGO power spectral density \citep{obsscen_noise_curves,abbott_prospects_2020} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
+The full procedure for this mock data generation process is described in \citet{fishbach_where_2017, farah_things_2023, essick_ensuring_2024}.
 
@@
     \caption{Performance of various mass distribution models used for measuring the Hubble constant, \Ho.
-    We consider the correct parametric model, an incorrect parametric model, and the non-parametric, \ac{GP}-based model presented in this work.
+    We consider the correct parametric model, a deliberately incorrect parametric model, and the non-parametric, \ac{GP}-based model presented in this work.
     Bias is defined as the number of standard deviations between the true injected value of \Ho and the recovered posterior mean. 
@@
 \subsection{Option 2: GP only}
-We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated \ac{GW} catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{o3b_pop}.
+We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated \ac{GW} catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{abbott_population_2023}.
 This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after \ac{O5}.
@@
 On the other hand, the lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the GP to learn a relatively low merger rate at the edges of the mass distribution.
-Indeed, these features are very informative to the \Ho{} constraint \citep{the_ligo_scientific_collaboration_constraints_2021}.
+Indeed, these features are very informative to the \Ho{} constraint \citep{gwtc3_cosmo}.
 The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.

### 2024-03-15 src/scripts/nonparametric_twopanel.py
decrease number and width of traces

@@ -26,10 +26,10 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     r_BPL = np.nan_to_num(np.exp(samples_BPL['log_rate'][0]))
 
     fig, axes = plt.subplots(ncols=2,figsize=(7.5,4*.75),facecolor='none',gridspec_kw={'width_ratios': [1.2, 1]})
-    for i in range(NSAMPS):
-        axes[0].plot(TEST_M1S, r[i], lw=0.2, c="blue",alpha=0.05)
-        axes[0].plot(TEST_M1S, r_PLP[i],lw=0.2, c="green",alpha=0.05)
-        axes[0].plot(TEST_M1S, r_BPL[i],lw=0.2, c="orange",alpha=0.05)
+    for i in range(NSAMPS//2):
+        axes[0].plot(TEST_M1S, r[i], lw=0.1, c="blue",alpha=0.05)
+        axes[0].plot(TEST_M1S, r_PLP[i],lw=0.1, c="green",alpha=0.05)
+        axes[0].plot(TEST_M1S, r_BPL[i],lw=0.1, c="orange",alpha=0.05)
     
     axes[0].plot(TEST_M1S, samples_PLP['rate'].mean(axis=1)*powerlaw_peak(TEST_M1S,alpha=-2.7,f_peak=0.05,mMax=78.0,mMin=10.0,mu_m1=30.0,sig_m1=7.0),
                 c='k')

### 2024-03-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by %$\variable{output/BPLh0offset.txt}\sigma$.
-We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and
-the \plp{} model reaches the same level of bias $\variable{output/PLP_bias_percent.txt}\%$ of the time.
+We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ %$\variable{output/BPL_bias_percent.txt}\%$ of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and
+the \plp{} model reaches the same level of bias %$\variable{output/PLP_bias_percent.txt}\%$ of the time.
 

### 2024-03-15 src/scripts/nonparametric_twopanel.py
minor edits

@@ -27,9 +27,9 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
 
     fig, axes = plt.subplots(ncols=2,figsize=(7.5,4*.75),facecolor='none',gridspec_kw={'width_ratios': [1.2, 1]})
     for i in range(NSAMPS//2):
-        axes[0].plot(TEST_M1S, r[i], lw=0.1, c="blue",alpha=0.05)
-        axes[0].plot(TEST_M1S, r_PLP[i],lw=0.1, c="green",alpha=0.05)
-        axes[0].plot(TEST_M1S, r_BPL[i],lw=0.1, c="orange",alpha=0.05)
+        axes[0].plot(TEST_M1S, r[i], lw=0.1, c="blue",alpha=0.03)
+        axes[0].plot(TEST_M1S, r_PLP[i],lw=0.1, c="green",alpha=0.03)
+        axes[0].plot(TEST_M1S, r_BPL[i],lw=0.1, c="orange",alpha=0.03)
     
     axes[0].plot(TEST_M1S, samples_PLP['rate'].mean(axis=1)*powerlaw_peak(TEST_M1S,alpha=-2.7,f_peak=0.05,mMax=78.0,mMin=10.0,mu_m1=30.0,sig_m1=7.0),
                 c='k')

### 2024-03-15 src/scripts/submit.sh

@@ -1,7 +1,7 @@
 #!/bin/bash
-#SBATCH --job-name=paper_fitOm0_gwmc
-#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_gwmc.out
-#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_gwmc.err
+#SBATCH --job-name=paper_gwmc
+#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_gwmc.out
+#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_gwmc.err
 #SBATCH --account=kicp
 #SBATCH --time=48:00:00
 #SBATCH --partition=kicp-gpu
@@ -15,4 +15,4 @@ module load python/anaconda-2022.05
 echo "loaded anaconda"
 source activate jax_gpu
 echo "activated env"
-python /home/afarah/p2/spectral-sirens-with-GPs/src/scripts/nonparametric_inference_fitOm0.py
+python /home/afarah/p2/spectral-sirens-with-GPs/src/scripts/nonparametric_inference.py

### 2024-03-15 src/scripts/nonparametric_inference.py
fit for hyper-hyper parameters

@@ -40,6 +40,7 @@ if  __name__ == "__main__":
     nuts_kernel = numpyro.infer.NUTS(hyper_prior,**nuts_settings)
     kwargs = dict(m1det=m1z_PE,dL=dL_PE, m1det_inj=m1zinj_det,dL_inj=dLinj_det,
                     log_pinj=log_pinj_det, log_PE_prior=log_PE_prior,
+                    PC_params=dict(conc=conc,concentration=concentration,scale=scale,lam_sigma=lam_sigma),
                     remove_low_Neff=remove_low_Neff)
     mcmc = numpyro.infer.MCMC(nuts_kernel,num_warmup=NSAMPS,num_samples=NSAMPS,
                               num_chains=1,progress_bar=True)   

### 2024-03-15 src/scripts/nonparametric_inference_fitOm0.py

@@ -39,6 +39,7 @@ if  __name__ == "__main__":
     nuts_kernel = numpyro.infer.NUTS(hyper_prior,**nuts_settings)
     kwargs = dict(m1det=m1z_PE,dL=dL_PE, m1det_inj=m1zinj_det,dL_inj=dLinj_det,
                     log_pinj=log_pinj_det, log_PE_prior=log_PE_prior,
+                    PC_params=dict(conc=conc,concentration=concentration,scale=scale,lam_sigma=lam_sigma),
                     remove_low_Neff=remove_low_Neff,fit_Om0=True)
     mcmc = numpyro.infer.MCMC(nuts_kernel,num_warmup=NSAMPS,num_samples=NSAMPS,
                               num_chains=1,progress_bar=True)   

### 2024-03-15 src/scripts/priors.py

@@ -112,11 +112,13 @@ def get_sigma_gamma_params(U,alpha=0.05):
     return k, rate
 
 # HYPER PRIOR / POPULATION MODEL
-def hyper_prior(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,remove_low_Neff=False,fit_Om0=False):
+def hyper_prior(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,PC_params=dict(),remove_low_Neff=False,fit_Om0=False):
     """ Non-parametric population inference """
     mean = numpyro.sample("mean",dist.Normal(0,3))
-    sigma = numpyro.deterministic("sigma",2.)
-    rho = numpyro.deterministic("rho",3.)
+    # sigma = numpyro.deterministic("sigma",2.5)
+    sigma = numpyro.sample("sigma",dist.Gamma(concentration=PC_params["conc"], rate=PC_params["lam_sigma"]))
+    # rho = numpyro.deterministic("rho",0.5)
+    rho = numpyro.sample("rho",Frechet(concentration=PC_params["concentration"],scale=PC_params["scale"]))
 
     H0 = numpyro.sample("H0",dist.Uniform(H0_PRIOR_MIN,H0_PRIOR_MAX))
     if fit_Om0:

### 2024-03-15 src/scripts/submit.sh

@@ -1,7 +1,7 @@
 #!/bin/bash
-#SBATCH --job-name=paper_gwmc
-#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_gwmc.out
-#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_gwmc.err
+#SBATCH --job-name=paper_fitOm0_fitrho_gwmc
+#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_fitrho_gwmc.out
+#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_fitrho_gwmc.err
 #SBATCH --account=kicp
 #SBATCH --time=48:00:00
 #SBATCH --partition=kicp-gpu
@@ -15,4 +15,4 @@ module load python/anaconda-2022.05
 echo "loaded anaconda"
 source activate jax_gpu
 echo "activated env"
-python /home/afarah/p2/spectral-sirens-with-GPs/src/scripts/nonparametric_inference.py
+python /home/afarah/p2/spectral-sirens-with-GPs/src/scripts/nonparametric_inference_fitOm0.py

### 2024-03-15 src/scripts/bias_study.py
add more catalogs and more events per catalog

@@ -15,8 +15,8 @@ from priors import PLP, BPL
 jax.config.update("jax_enable_x64", True)
 
 # options
-N_CATALOGS=5
-N_SOURCES = N_SOURCES//4 # cut catalogs in half for this study, for computational feasibility
+N_CATALOGS=50
+N_SOURCES = N_SOURCES//2 # cut catalogs in half for this study, for computational feasibility
 plot = True
 
 # random number generators

### 2024-03-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within %$\variable{output/PLPh0offset.txt}\sigma$.
-However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by %$\variable{output/BPLh0offset.txt}\sigma$.
+We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within $\variable{output/PLPh0offset.txt}\sigma$.
+However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by $\variable{output/BPLh0offset.txt}\sigma$.
 We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ %$\variable{output/BPL_bias_percent.txt}\%$ of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and

### 2024-03-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens and a% $\variable{output/PLPh0percent.txt}\%$ 
+We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens and a $\variable{output/PLPh0percent.txt}\%$ 
 measurement with parametric spectral sirens, demonstrating a \result{negligible/substantial/marginal} increase in precision when using the correct parametric model.

### 2024-03-15 src/scripts/nonparametric_inference.py
try moving nonparametric stat generation to inference script

@@ -12,7 +12,7 @@ import numpy as np
 
 jax.config.update("jax_enable_x64", True)
 
-NSAMPS=1000
+NSAMPS=1
 
 # random number generators
 jax_rng = jax.random.PRNGKey(425)

### 2024-03-15 src/scripts/pm_H0_twopanel_twopanel.py
rename twopanel script to be more descriptive

@@ -61,7 +61,7 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     axes[1].set_ylabel('posterior density')
     fig.legend(ncol=4,framealpha=0,fontsize=9,loc="upper center")
     plt.tight_layout()
-    fig.savefig(paths.figures / "O5_GP_pm.pdf")
+    fig.savefig(paths.figures / "O5_pm.pdf")
     plt.clf()
 
 two_panel(paths.data / "mcmc_nonparametric.nc4", 

### 2024-03-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-We find that using our non-parametric model model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ by the end of \ac{O5} \jme{when the detectors will reach their design sensitivity}.
+We find that using our non-parametric model model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ during \ac{O5}, when the detectors will reach their design sensitivity.
 %, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
@@
     \centering
-    \includegraphics[width=\textwidth]{figures/O5_GP_pm.pdf}
+    \includegraphics[width=\textwidth]{figures/O5_pm.pdf}
     \caption{Spectral siren measurement for an \ac{O5}-like catalog with the correct parametric model (\plp, green), a deliberately incorrect parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{GP}, blue).
@@
     \label{fig:O5_GP}
-    \script{nonparametric_twopanel.py}
+    \script{pm_H0_twopanel.py}
 \end{figure*}
@@
 To illustrate this, we generate a catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{abbott_population_2023}.
-This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after \jme{one year (?) of observation in} \ac{O5}.
+This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after one year of observation in \ac{O5}.
 The details of the data simulation are described in Appendix~\ref{ap:data generation}.
@@
 The left panel shows the inferred source frame mass distribution for each of the considered models, and the right panel shows the corresponding posteriors on \Ho{}. 
-\jme{We indicate the true underlying source mass distribution and fiducial \Ho{} value with solid black lines in each panel.}
+We indicate the true underlying source mass distribution and fiducial \Ho{} value with solid black lines in each panel.
 
@@
 \subsection{Projections for Future Measurements}
-The fit shown in Figure~\ref{fig:O5_GP} implies that we can expect to measure \Ho{} to $\variable{output/nonparh0percent.txt}\%$ ($1-\sigma$ uncertainties) by the end of \ac{O5} using the \ac{GP}-based spectral siren method.
+The fit shown in Figure~\ref{fig:O5_GP} implies that we can expect to measure \Ho{} to $\variable{output/nonparh0percent.txt}\%$ ($1-\sigma$ uncertainties) after one year of \ac{O5} using the \ac{GP}-based spectral siren method.
 However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\sim 3$, with most sources lying near redshift $\sim 1.2$ \citep{chen_distance_2021}.
@@
 This means that future \acp{GW} observations will be more sensitive $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters.
-We demonstrate this by repeating the same analysis while also fitting for the local matter density, \Omm, but it is equally plausible to constrain other cosmological parameters governing $H(z)$.
+We demonstrate this by repeating the same analysis while also fitting for the local matter density, \Omm, but we emphasize that it would also have been possible to constrain other cosmological parameters governing $H(z)$.
 The result is shown in Figure~\ref{fig:O5_corner}.
@@
     \caption{
-    Projected constraints on multiple cosmological parameters at the end of \ac{O5} using the \acl{GP}-based spectral siren method.
+    Projected constraints on multiple cosmological parameters after one year of observing at the \ac{LVK}'s design sensitivity, using the \acl{GP}-based spectral siren method.
     The right panel shows the inferred expansion history of the universe, $H(z)$.
@@
 
-We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} by the end of \ac{O5} using non-parametric spectral sirens and a $\variable{output/PLPh0percent.txt}\%$ 
+We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} after one year of observing with the \ac{LVK} network's design sensitivity using non-parametric spectral sirens and a $\variable{output/PLPh0percent.txt}\%$
 measurement with parametric spectral sirens, demonstrating a \result{negligible/substantial/marginal} increase in precision when using the correct parametric model.
@@
 We have shown this by simultaneously measuring \Omm{} and \Ho{}, but the method can be trivially extended to constrain other cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$.
-We find that spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ at the end of \acf{O5}.
+We find that spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ during \acf{O5}.
 This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}\jme{, which can only be detected relatively locally}.
@@
 
-We generate a mock catalog of \ac{GW} events characteristic of that expected from \acf{O5}. 
+We generate a mock catalog of \ac{GW} events characteristic of that expected from one year of \acf{O5}. 
 These events are drawn from an underlying population described by the \plp{} mass distribution presented in \citet{talbot_measuring_2018} and used in \citet{abbott_binary_2019, abbott_population_2023} and the redshift distribution presented in \citet{callister_shouts_2020}.

### 2024-03-15 src/scripts/nonparametric_inference.py
increase nsamps again

@@ -12,7 +12,7 @@ import numpy as np
 
 jax.config.update("jax_enable_x64", True)
 
-NSAMPS=1
+NSAMPS=1000
 
 # random number generators
 jax_rng = jax.random.PRNGKey(425)

### 2024-03-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 This allows for a precise estimate of each \ac{GW} catalogs' completeness and an unbiased measurement of the true \ac{GW} population. 
-Additionally, the fact that \ac{GW} signals provide direct measurements of the distance to their sources makes them ``standard sirens'': powerful probes of cosmological parameters that circumvent the cosmological distance ladder~\citep{schutz_determining_1986,holz_using_2005}.
+Additionally, the \ac{GW} signals detected by the LIGO \citep{aasi_advanced_2015}, Virgo \citep{acernese_advanced_2014}, and KAGRA \citep{akutsu_overview_2021} detectors provide direct measurements of the distance to their sources.
+This makes them ``standard sirens'': powerful probes of cosmological parameters that circumvent the cosmological distance ladder~\citep{schutz_determining_1986,holz_using_2005}. 
 This is particularly useful in the context of the current ``Hubble tension:'' a disagreement between multiple methods of measuring the local expansion rate of the universe with electromagnetic observations \citep{freedman_measurements_2021}.
@@
 
-\comment{below text is WIP}
 This is particularly troubling in light of the fact that there remains much uncertainty over the cause, location, and robustness of currently-identified features in the \ac{GW} source mass distribution.
-While four or more features have been identified in the current data, \citep{viabhav, abbott_population_2023,bruce_aint} only three of these are statistically robust \citep{farah_things_2023}.
-Additionally, the morphology of the mass dist is caused by the interplay of several complex physical processes, such as compact binary evolution, pair instability, globular cluster dynamics, and the metalicity content of massive stars, each of which has large theoretical uncertainties.
-\citep{many ppl}
-These large theoretical uncertainties mean that accurate predictions for the shape of the \ac{GW} source mass distribution is not currently possible.
+While several features have been identified in the current data \citep{tiwari_exploring_2022,edelman_aint_2022, abbott_population_2023,edelman_cover_2023}, including three overdensities, two underdensities, a maximum and minimum \ac{BH} mass, only three of these are statistically robust \citep{farah_things_2023}.
+Additionally, the morphology of the mass distribution is a result of the interplay between several complex physical processes pertaining to massive stars, such as binary stellar evolution, pair instability, globular cluster dynamics, and stellar winds, each of which has large theoretical uncertainties \citep[e.g.][]{ zevin_constraining_2017, mapelli_binary_2020,marchant_evolution_2023}.
+These large theoretical uncertainties mean that accurate predictions for the shape of the \ac{GW} source mass distribution are a considerable challenge.
 If such \emph{a priori} knowledge of the mass distribution's morphology were requisite for a spectral siren method, the effectiveness of such a measurement would be significantly hampered.
-\comment{end WIP}
 
 However, the information in the spectral siren measurement comes solely from the assumption that all \acp{CBC} follow the same mass distribution.\footnote{More precisely, the spectral siren method is also valid if \acp{CBC} have different mass distributions at different redshifts, so long as the evolution of the mass distribution does not exactly mimic cosmology, and the model for the mass distribution allows for evolution \citep{ezquiaga_spectral_2022}.
-%\jme{[maybe is worth having this comment in the main text?] \comment{True, but it kind of interrupts the flow and I think the current text sets up the question nicely.
-I discuss it more in the discussion.}} OK!
 }
@@
 % The source frame mass is different for each possible value of \Ho{} and \Omm{}, so the hyper-parameters are defined in different locations for each draw from the hyper-posterior.
+In this way, our \ac{GP}-based mass distribution is similar to the autoregressive population models used in ~\citet{callister_parameter-free_2023}.
+Indeed, an autoregressive process is a \ac{GP} with a particular choice of kernel (described below).
 We therefore have $N_{\text{evs}}M + N_{\text{inj}}$ mass hyper-parameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of posterior samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
 We label these $\{\mathcal{R}_{ij},\mathcal{R}_k\}$, and our full set of hyper-parameters is $\Lambda=\{H_0, \Omega_M, \mathcal{R}_{ij},\mathcal{R}_k\}$. 
-In this way, our GP-based mass distribution is similar to the autoregressive population models used in ~\citet{callister_parameter-free_2023}.
-Indeed, an autoregressive model is a \ac{GP} with a particular choice of kernel.
+\comment{Put something here about why we chose to do it this way.
+We were specifically trying to avoid a mass model that has pre-defined features in source frame mass.
+This is why we didn't do the spline or BGP models, as they have to choose source frame masses at which to put their feautres, and they did this by fitting the distribution to the data with a pre-determined H0.
+If we wanted to do a spline, we would want to fit for where the spline nodes lie (which I think does happen in the b-spline paper?).
+Caution that other approaches should be careful about their choice of non-parametric model for this reason.}
 
-A kernel is a function that defines the covariance between input points in the GP (in our case, two source frame mass values). 
+The smoothness of a given \acp{GP} is determined by its kernel, which is a function that defines the covariance between input points in the \ac{GP} (in our case, two source frame mass values). 
 It defines the notion of similarity between adjacent points and thereby encodes our assumptions about the smoothness of the source frame mass distribution \citep{rasmussen_gaussian_2006}.
@@
 We therefore make two approximations to a full \ac{GP} to increase computational efficiency.
-First, for each likelihood evaluation, we evaluate a full \ac{GP} on a regular grid between $0\Msun$ and $250\Msun$ and then interpolate it at each data point.
+First, for each likelihood evaluation, we evaluate a full \ac{GP} on a regular grid between $0.1\Msun$ and $250\Msun$ and then interpolate it at each data point.
 Second, we use the quasi-separability of Mat\'ern kernels to analytically perform the transformation between covariance matrix and \ac{GP} draw.
@@
 However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by $\variable{output/BPLh0offset.txt}\sigma$.
-We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ %$\variable{output/BPL_bias_percent.txt}\%$ of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and
-the \plp{} model reaches the same level of bias %$\variable{output/PLP_bias_percent.txt}\%$ of the time.
+We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ %$\variable{output/BPL_bias_percent.txt}\%$
+of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and 
+the \plp{} model reaches the same level of bias %$\variable{output/PLP_bias_percent.txt}\%$ 
+of the time.
 
@@
 The first condition (multiple features) is known to be met in current data \citep{abbott_population_2023}, with three robust features \citep{farah_things_2023}: a maximum black hole mass and overdensities at $\sim10\Msun$ and $\sim35\Msun$.
-The latter scenario (features identically and monotonically moving to higher masses with increasing redshift) would be astrophysically unlikely as the the locations of features in the mass distribution are each thought to be governed by different physical processes \citep[e.g.][]{mapelli_binary_2020}.
+The latter scenario (features identically and monotonically moving to higher masses with increasing redshift) would be astrophysically unlikely as the the locations of features in the mass distribution are each thought to be governed by different physical processes \citep{mapelli_binary_2020}.
 Since \aclp{GP} naturally scale to multiple data dimensions, \citep{rasmussen_gaussian_2006}, the method presented here can be straightforwardly generalized to fitting the redshift dependence of the mass distribution.
@@
 \begin{acknowledgments}
-    The authors thank Reed Essick, Utkarsh Mali, Ben Farr, Maya Fishbach,  for helpful conversations. 
+    The authors thank Reed Essick, Utkarsh Mali, Ben Farr, Maya Fishbach, Colm Talbot,  for helpful conversations. 
     %Amanda

### 2024-03-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 This is particularly troubling in light of the fact that there remains much uncertainty over the cause, location, and robustness of currently-identified features in the \ac{GW} source mass distribution.
-While several features have been identified in the current data \citep{tiwari_exploring_2022,edelman_aint_2022, abbott_population_2023,edelman_cover_2023}, including three overdensities, two underdensities, a maximum and minimum \ac{BH} mass, only three of these are statistically robust \citep{farah_things_2023}.
+While several features have been identified in the current data \citep{tiwari_exploring_2022,edelman_aint_2022, abbott_population_2023,edelman_cover_2023}, including three overdensities, two underdensities, a maximum and minimum black hole mass, only three of these are statistically robust \citep{farah_things_2023}.
 Additionally, the morphology of the mass distribution is a result of the interplay between several complex physical processes pertaining to massive stars, such as binary stellar evolution, pair instability, globular cluster dynamics, and stellar winds, each of which has large theoretical uncertainties \citep[e.g.][]{ zevin_constraining_2017, mapelli_binary_2020,marchant_evolution_2023}.

### 2024-03-15 src/scripts/priors.py
fit in logm

@@ -21,7 +21,7 @@ import jgwcosmo
 import jgwpop
 
 ## CONSTANTS
-TEST_M1S = jnp.linspace(0,250.,num=500)
+TEST_M1S = jnp.linspace(0.1,250.,num=500)
 
 H0_PRIOR_MIN=30.
 H0_PRIOR_MAX=120.
@@ -115,8 +115,8 @@ def get_sigma_gamma_params(U,alpha=0.05):
 def hyper_prior(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,PC_params=dict(),remove_low_Neff=False,fit_Om0=False):
     """ Non-parametric population inference """
     mean = numpyro.sample("mean",dist.Normal(0,3))
-    # sigma = numpyro.deterministic("sigma",2.5)
-    sigma = numpyro.sample("sigma",dist.Gamma(concentration=PC_params["conc"], rate=PC_params["lam_sigma"]))
+    sigma = numpyro.deterministic("sigma",2.5)
+    # sigma = numpyro.sample("sigma",dist.Gamma(concentration=PC_params["conc"], rate=PC_params["lam_sigma"]))
     # rho = numpyro.deterministic("rho",0.5)
     rho = numpyro.sample("rho",Frechet(concentration=PC_params["concentration"],scale=PC_params["scale"]))
 
@@ -126,46 +126,59 @@ def hyper_prior(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,PC_params=dic
     else:
         Om0=OM0_FID
 
+    # redshift dist
+    alpha_z = 1.0
+    zp = 2.4 
+    beta_z = 3.4 
+
     # construct GP in the source frame
+    logtestm1s = jnp.log(TEST_M1S)
     kernel = sigma**2 * kernels.quasisep.Matern52(rho) # can change kernel type
-    gp = GaussianProcess(kernel,TEST_M1S,mean=mean,diag=0.001,
+    gp = GaussianProcess(kernel,logtestm1s,mean=mean,diag=0.001,
                          solver=QuasisepSolver,assume_sorted=True)
     log_rate_test = numpyro.sample("log_rate_test",gp.numpyro_dist())
 
     # convert event data to source frame
     z = jgwcosmo.z_at_dl_approx(dL,H0,Om0,zmin=ZMIN,zmax=ZMAX+8.)
-    m1source = m1det / (1 + z)
[…]

### 2024-03-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-\jme{For our population model we focus on the} %We only consider 
-masses and redshifts, neglecting spins as they do not contain cosmological information.
+For our population model, we focus on the masses and redshifts, neglecting spins as they do not contain cosmological information.
 For simplicity, we additionally assume that the population has equal masses, $m_1=m_2$.
@@
 This is the basis of the spectral siren method of measuring cosmological parameters; we can simultaneously infer the relationship between redshift and luminosity distance with the parameters of the source frame mass distribution.
-The full set of hyper-parameters therefore includes the cosmological parameters that dictate the $D_L$-$z$ relation 
-\jme{such as the local expansion rate \Ho, the present fractional energy density of dark matter \Omm, dark energy $\Omega_\Lambda$ and radiation $\Omega_r$, or the equation of state of dark energy $w$.} %: \Ho, \Omm, $\Omega_\Lambda, \Omega_r,$ and $w$.
-In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and \jme{$w=-1$} and use uniform priors on \Ho{} and \Omm{}. 
-\jme{This corresponds to a flat $\Lambda$CDM cosmology.}
+The full set of hyper-parameters therefore includes the cosmological parameters that dictate the $D_L$-$z$ relation: the local expansion rate \Ho, the present fractional energy density of dark matter \Omm, dark energy $\Omega_\Lambda$ and radiation $\Omega_r$, and the equation of state of dark energy $w$.
+In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and $w=-1$ and use uniform priors on \Ho{} and \Omm{}. 
+This corresponds to a flat $\Lambda$CDM cosmology.
 
-Algorithmically, we evaluate the likelihood in Equation~\ref{eq:inhomog-poisson} by drawing \Ho{} and \Omm{} from their prior distributions (contained in $p(\Lambda)$), using them to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to \jme{$m_{\det} /(1+z(D_L)) = m_{\source}$}, evaluating the source-frame mass distribution at these transformed values ($\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ for events and $N_{\text{exp}}(\Lambda)$ for injections), and then taking the product of these terms to evaluate the expression in Equation~\ref{eq:inhomog-poisson}.
+Algorithmically, we evaluate the likelihood in Equation~\ref{eq:inhomog-poisson} by drawing \Ho{} and \Omm{} from their prior distributions (contained in $p(\Lambda)$), using them to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to $m_{\det} /(1+z(D_L,H_0,\Omega_M)) = m_{\source}$, evaluating the source-frame mass distribution at these transformed values ($\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ for events and $N_{\text{exp}}(\Lambda)$ for injections), and then taking the product of these terms to evaluate the expression in Equation~\ref{eq:inhomog-poisson}.
 The full process is outlined in Section~\ref{sec:model}.

### 2024-03-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL under an MIT license, and was enabled by the \showyourwork package \citep{Luger2021}.
+The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{} under an MIT license, and was enabled by the \showyourwork package \citep{Luger2021}.
 
@@
 However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by $\variable{output/BPLh0offset.txt}\sigma$.
-We have repeated this analysis with \result{50} separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ %$\variable{output/BPL_bias_percent.txt}\%$
+We have repeated this analysis with 50 separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ %$\variable{output/BPL_bias_percent.txt}\%$
 of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and 
@@
 We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} after one year of observing with the \ac{LVK} network's design sensitivity using non-parametric spectral sirens and a $\variable{output/PLPh0percent.txt}\%$
-measurement with parametric spectral sirens, demonstrating a \result{negligible/substantial/marginal} increase in precision when using the correct parametric model.
-However, the true population of compact objects in the Universe is unknown.
+measurement with parametric spectral sirens, demonstrating an increase in precision when using the correct parametric model.
+However, the true population of merging compact binaries in the Universe is unknown because neither the mechanism from which compact binaries were formed, nor the physical processes within that mechanism has been conclusively determined.
 It is therefore impossible to know the true functional form of the mass distribution, so systematic uncertainties arising from an incorrect choice for the form of the mass distribution are inevitable.

### 2024-03-15 src/scripts/priors.py
add placeholder variable to make a margin icon next to algorithm steps

@@ -19,12 +19,16 @@ from jax.scipy.special import gammaln
 from data_generation import NUM_INJ, OM0_FID, ZMAX, ZMIN, N_SAMPLES_PER_EVENT, TRUEVALS
 import jgwcosmo
 import jgwpop
+import paths
 
 ## CONSTANTS
 TEST_M1S = jnp.linspace(0.1,250.,num=500)
 
 H0_PRIOR_MIN=30.
 H0_PRIOR_MAX=120.
+# make a dummy variable so that the PDF links to this module
+with open(paths.output / "priors_placeholder.txt","w") as f:
+        print(" ",file=f)
 
 ## LENGTH SCALE PRIOR
 ## Applying: https://dansblog.netlify.app/posts/2022-09-07-priors5/priors5.html#rescuing-the-pc-prior-on-ell-or-what-i-recommend-you-do

### 2024-03-17 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 Our goal is to model the population of \ac{GW} in an astrophysically-agnostic way. 
-We choose to do it with \Acfp{GP}, a common tool for non-parametric inference. 
+There exist several non-parametric methods developed for this purpose \citep{tiwari_vamana_2021,edelman_aint_2022,sadiq_flexible_2022,edelman_cover_2023,mandel_extracting_2019,ray_non-parametric_2023}.
+While well-suited to infer the \ac{GW} source population with a fixed cosmology, several of these methods employ fixed features in source frame mass, such as bin edges \citep{mandel_extracting_2019,ray_non-parametric_2023} or spline nodes \citep{edelman_aint_2022}.
+Since these locations were chosen with a fixed cosmology, they risk causing the inference to prefer the cosmological parameters used when choosing the feature locations. 
+Indeed, \citet{mastrogiovanni_importance_2021} show how using fixed features can significantly bias cosmological inference within the spectral siren methodology.
+We therefore opt for a model of the source frame mass distribution that foregoes the need to define such features.
+
+For this purpose, we construct a model with a \acf{GP}, a common tool for non-parametric inference. 
 \Acp{GP} are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
@@
 Practically, the only difference in the inference of the population when using a GP versus other modeling choices is that $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is determined directly by a realization of the \ac{GP}, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
-%joined paragraphs here.
-OK sounds good!
 In other words, when using parametric models, $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is calculated by evaluating a specific functional form described by a small set of hyper-parameters.
@@
 We label these $\{\mathcal{R}_{ij},\mathcal{R}_k\}$, and our full set of hyper-parameters is $\Lambda=\{H_0, \Omega_M, \mathcal{R}_{ij},\mathcal{R}_k\}$. 
-\comment{Put something here about why we chose to do it this way.
-We were specifically trying to avoid a mass model that has pre-defined features in source frame mass.
-This is why we didn't do the spline or BGP models, as they have to choose source frame masses at which to put their feautres, and they did this by fitting the distribution to the data with a pre-determined H0.
-If we wanted to do a spline, we would want to fit for where the spline nodes lie (which I think does happen in the b-spline paper?).
-Caution that other approaches should be careful about their choice of non-parametric model for this reason.}
+As the hyper-parameters differ at each data point, we have successfully fulfilled our goal to construct a model without pre-defined features in source frame mass.
+We note that other non-parametric methods may be adapted to avoid such pre-defined features, such as fitting for the locations of their features simultaneously the rest of the inference \citep[e.g.][]{tiwari_vamana_2021} or by using an appropriate smoothing \citep[e.g.][]{edelman_cover_2023}.
+Such adaptations are necessary to avoid circular inference.
 
@@
 
-Algorithmically, each likelihood evaluation contains the following steps: \script{priors.py}
+Algorithmically, each likelihood evaluation contains the following steps: %\variable{output/priors_placeholder.txt}
 \begin{enumerate}

### 2024-03-17 src/scripts/data_generation.py
test

@@ -8,6 +8,7 @@ from mock_posteriors import gen_snr_scaled_PE
 import os
 import numpy as np
 from GWMockCat.vt_utils import draw_thetas, interpolate_optimal_snr_grid 
+print("test")
 
 # Constants
 H0_FID = 67.66

### 2024-03-18 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff_gravitational_1993,Taylor:2011fs} using the binary neutron star mass distribution, and extended to the \ac{BBH} mass distribution by \cite{farr_future_2019}.
-Spectral siren analyses have since been implemented by the LIGO--Virgo--KAGRA Collaborations using the latest catalog \cite{gwtc3_cosmo}.
+Spectral siren analyses have since been implemented by the LIGO--Virgo--KAGRA Collaborations using the latest \ac{GW} catalog \cite{gwtc3_cosmo}.
 
@@
 On the other hand, the lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the \ac{GP} to learn a relatively low merger rate at the edges of the mass distribution.
-Indeed, these features are very informative to the \Ho{} constraint \citep{gwtc3_cosmo}.
+Indeed, these features are very informative to the \Ho{} constraint, especially at high masses for current detectors \citep{gwtc3_cosmo}.
 The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
@@
 Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 40$.
-This means that future \acp{GW} observations will be more sensitive $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters.
-We demonstrate this by repeating the same analysis while also fitting for the local matter density, \Omm, but we emphasize that it would also have been possible to constrain other cosmological parameters governing $H(z)$.
+This means that future \acp{GW} observations will be more sensitive $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters \citep{Chen:2024gdn}.
+We demonstrate this by repeating the same \ac{GP} spectral siren analysis of the previous section while also fitting for the local matter density, \Omm, but we emphasize that it would also have been possible to constrain other cosmological parameters governing $H(z)$.
 The result is shown in Figure~\ref{fig:O5_corner}.
@@
 We find \ac{O5} observations to be most sensitive to $H(z=\variable{output/mostsensitivez.txt})$, which results in similarly informative constraints on \Omm{} and \Ho.
-The left panel of Figure~\ref{fig:O5_corner} demonstrates a strong anti-correlation between the \Omm{} and \Ho posteriors. 
+The left panel of Figure~\ref{fig:O5_corner} demonstrates a strong anti-correlation between the \Omm{} and \Ho{} posteriors. 
 This is because \Ho{} controls the $y$-intercept of the $H(z)$ curves on the right panel, while \Omm{} informs the slope of those curves.
@@
 
-\acp{GW} are unique messengers in that they carry both redshift and distance information, making them remarkably clean probes of the Universe's expansion history. 
+\acp{GW} are unique cosmic messengers in that they carry both redshift and distance information, making them remarkably clean probes of the Universe's expansion history. 
 However, the current method of determining \ac{GW} redshifts via the distribution of their source masses (i.e.
@@
 \end{equation}
-Here, $V_C(H_0, \Omega_M)$ is the comoving volume for given cosmological parameters \Ho{} and \Omm{}, and $\bar{\Lambda}_m = \{\alpha, m_{\min}, m_{\max}, \mu, \sigma, f_{\text{peak}}\}$ are the parameters describing the power law in primary mass, minimum and maximum black hole mass, Gaussian peak location and width, and fraction of events in the Gaussian peak, respectively.
+Here, $V_C(H_0, \Omega_M)$ is the comoving volume for given cosmological parameters \Ho{} and \Omm{}, and $\bar{\Lambda}_m = \{\alpha, m_{\min}, m_{\max}, \mu, \sigma, f_{\text{peak}}\}$ are the (hyper-)parameters describing the power law in primary mass, minimum and maximum black hole mass, Gaussian peak location and width, and fraction of events in the Gaussian peak, respectively.
 $\mathcal{S}$ is a smoothing function at low and high masses, and $\mathcal{N}_{\text{pl}}$ and $\mathcal{N}_{\text{g}}$ are the normalizations between  $m_{\min}$ and $m_{\max}$ for the power law component and truncated Gaussian component, respectively.

### 2024-03-18 src/scripts/nonparametric_inference.py
nsamps 20 for build

@@ -12,7 +12,7 @@ import numpy as np
 
 jax.config.update("jax_enable_x64", True)
 
-NSAMPS=1000
+NSAMPS=20
 
 # random number generators
 jax_rng = jax.random.PRNGKey(425)

### 2024-03-18 src/scripts/nonparametric_inference_fitOm0.py

@@ -11,7 +11,7 @@ import numpyro
 # data
 import numpy as np
 
-NSAMPS = 1000
+NSAMPS = 20
 
 jax.config.update("jax_enable_x64", True)
 # random number generators

### 2024-03-18 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 We do this by inferring $H(z)$ with a flexible, non-parametric model for the mass distribution of \acp{CBC} (blue lines in Figure~\ref{fig:O5_GP}). 
-This model makes minimal prior assumptions about the shape of the mass distribution, enabling it to accurately infer a wide range of morphologies.
+This model makes minimal prior assumptions about the shape of the mass distribution, enabling it to accurately infer a wide range of morphologies and remain agnostic to the astrophysical processes that give rise to the mass distribution's features.
 Despite its flexibility, it is able to consistently obtain unbiased measurements of cosmological parameters, showing that nonparametric methods are not only sufficient for a spectral siren measurement, they can also mitigate systematic effects in the measurement caused by model misspecification.

### 2024-03-18 src/scripts/data_generation.py
more injections (for consistency with old runs)

@@ -27,7 +27,7 @@ N_SAMPLES_PER_EVENT = 100
 SIGMA_M = 0.5
 SIGMA_DL = 1000
 SNR_THRESH=8.
-NUM_INJ=N_SOURCES*50
+NUM_INJ=N_SOURCES*50*30
 
 # random number generators
 np_rng = np.random.default_rng(516)

### 2024-03-18 src/scripts/pm_H0_twopanel.py
some fixes to tide us over until new runs are done (incl normalizing mass dist)

@@ -5,7 +5,7 @@ from scipy.stats import gaussian_kde
 
 from priors import TEST_M1S
 from nonparametric_inference import NSAMPS
-from data_generation import H0_FID
+from data_generation import H0_FID,TRUEVALS
 from gwpop import powerlaw_peak
 
 import paths
@@ -18,24 +18,31 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     r = np.nan_to_num(np.exp(samples['log_rate_test'][0]))
 
     id_PLP = az.InferenceData.from_netcdf(path_PLP)
-    samples_PLP = id_PLP.posterior
-    r_PLP = np.nan_to_num(np.exp(samples_PLP['log_rate'][0]))
+    samples_PLP = id_PLP.posterior.sel(chain=0).reset_coords("chain",drop=True)
+    r_PLP = np.nan_to_num(np.exp(samples_PLP['log_rate']))
     
     id_BPL = az.InferenceData.from_netcdf(path_BPL)
-    samples_BPL = id_BPL.posterior
-    r_BPL = np.nan_to_num(np.exp(samples_BPL['log_rate'][0]))
+    samples_BPL = id_BPL.posterior.sel(chain=0).reset_coords("chain",drop=True)
+    r_BPL = np.nan_to_num(np.exp(samples_BPL['log_rate']))
+    m1s_plotting_parametric = np.linspace(0,TEST_M1S.max(),len(TEST_M1S)) # TODO: remove this w/ new runs
 
     fig, axes = plt.subplots(ncols=2,figsize=(7.5,4*.75),facecolor='none',gridspec_kw={'width_ratios': [1.2, 1]})
     for i in range(NSAMPS//2):
-        axes[0].plot(TEST_M1S, r[i], lw=0.1, c="blue",alpha=0.03)
-        axes[0].plot(TEST_M1S, r_PLP[i],lw=0.1, c="green",alpha=0.03)
-        axes[0].plot(TEST_M1S, r_BPL[i],lw=0.1, c="orange",alpha=0.03)
+        axes[0].plot(TEST_M1S, r[i]/np.trapz(y=r[i],x=TEST_M1S), lw=0.1, c="blue",alpha=0.03)
+        # TODO: get rid of indexing and plot parametric with TEST_M1S when new parametric runs are done with proper TEST_M1S
[…]

### 2024-03-18 src/scripts/nonparametric_inference.py
nsamps back to 1000

@@ -12,7 +12,7 @@ import numpy as np
 
 jax.config.update("jax_enable_x64", True)
 
-NSAMPS=20
+NSAMPS=1000
 
 # random number generators
 jax_rng = jax.random.PRNGKey(425)

### 2024-03-18 src/scripts/nonparametric_inference_fitOm0.py

@@ -11,7 +11,7 @@ import numpyro
 # data
 import numpy as np
 
-NSAMPS = 20
+NSAMPS = 1000
 
 jax.config.update("jax_enable_x64", True)
 # random number generators

### 2024-03-18 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 On the other hand, the lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the \ac{GP} to learn a relatively low merger rate at the edges of the mass distribution.
-Indeed, these features are very informative to the \Ho{} constraint, especially at high masses for current detectors \citep{gwtc3_cosmo}.
-The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
+% Indeed, the upper edge of the mass distribution is very informative to the \Ho{} constraint with current detections \citep{gwtc3_cosmo}.
+The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but this reversion to the prior is uninformative to the \Ho{} constraint and does not affect inference on cosmological parameters.
 Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,120\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
@@
 However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\sim 3$, with most sources lying near redshift $\sim 1.2$ \citep{chen_distance_2021}.
-Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 40$.
+Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 100$.
 This means that future \acp{GW} observations will be more sensitive $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters \citep{Chen:2024gdn}.
-We demonstrate this by repeating the same \ac{GP} spectral siren analysis of the previous section while also fitting for the local matter density, \Omm, but we emphasize that it would also have been possible to constrain other cosmological parameters governing $H(z)$.
+We demonstrate this by repeating the same \ac{GP}-based spectral siren analysis of the previous subsection while also fitting for the local matter density, \Omm, but we emphasize that it would also have been possible to constrain other cosmological parameters governing $H(z)$.
 The result is shown in Figure~\ref{fig:O5_corner}.
@@
 These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{abbott_population_2023}.
-We do not fit for secondary masses or spins, as they do not carry additional cosmological information.
+We do not fit for spins, as they do not carry additional cosmological information, and additionally neglect secondary masses.
 We use cosmological parameters \Ho$=\variable{output/H0_FID.txt}\Hunits$, \Omm$=0.3$, $\Omega_\Lambda=1-$\Omm, consistent with those found by \citet{planck_collaboration_planck_2016}.

### 2024-03-18 src/scripts/nonparametric_corner.py
get nsamps directly from inferencedata object

@@ -4,7 +4,6 @@ import arviz as az
 from scipy.stats import gaussian_kde
 
 from data_generation import H0_FID, OM0_FID, ZMIN, ZMAX
-from nonparametric_inference_fitOm0 import NSAMPS
 from gwcosmo import E_of_z
 
 import paths
@@ -20,13 +19,14 @@ def Hz(id, ax=None,save=False,inset=True):
     try:
         Om0=samples['Om0'][0].values
         H0=samples['H0'][0].values
+        nsamps = len(H0)
     except KeyError:
         pass
         
     if ax is None:
         fig, ax = plt.subplots(figsize=(8,8))
-    H_z = np.zeros((NSAMPS,len(TEST_Z)))
-    for i in range(NSAMPS):
+    H_z = np.zeros((nsamps,len(TEST_Z)))
+    for i in range(nsamps):
         H_z[i] = calc_Hz(TEST_Z,H0[i],Om0[i])
         ax.plot(TEST_Z,H_z[i],lw=0.2, c="blue",alpha=0.1)
     ax.plot(TEST_Z, calc_Hz(TEST_Z,H0_FID,OM0_FID),c='k', lw=2,label="True value")

### 2024-03-18 src/scripts/parametric_inference.py
calcualte BPL offset after saving inference object

@@ -62,10 +62,10 @@ if  __name__ == "__main__":
     mcmc = numpyro.infer.MCMC(nuts_kernel,num_warmup=NSAMPS,num_samples=NSAMPS,
                               num_chains=1,progress_bar=True)   
     mcmc.run(jax_rng,**kwargs)
-    offset = np.abs(id.posterior['H0'][0].mean()-H0_FID)/id.posterior['H0'][0].std()
-    with open(paths.output / "BPLh0offset.txt","w") as f:
-        print(f"{offset:.1f}",file=f)
 
     # save results
     id = az.from_numpyro(mcmc)
     id.to_netcdf(paths.data / "mcmc_parametric_BPL.nc4")
+    offset = np.abs(id.posterior['H0'][0].mean()-H0_FID)/id.posterior['H0'][0].std()
+    with open(paths.output / "BPLh0offset.txt","w") as f:
+        print(f"{offset:.1f}",file=f)

### 2024-03-18 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-Algorithmically, each likelihood evaluation contains the following steps: %\variable{output/priors_placeholder.txt}
+Algorithmically, each likelihood evaluation contains the following steps: \variable{output/priors_placeholder.txt}
 \begin{enumerate}
@@
 \end{enumerate}
-\jme{[comment: is there a way to summarize this list in a simple diagram?]}
+% TODO: \jme{[comment: is there a way to summarize this list in a simple diagram?]}
 
@@
     \item the \plp{} model, which includes the true mass distribution within its hyperprior,
-    \item a broken power law \jme{[use same font style as \plp{}?]}, which does not include the true mass distribution within its hyperprior, and 
+    \item a broken power law, which does not include the true mass distribution within its hyperprior, and 
     \item the flexible, \ac{GP}-based model described in Section~\ref{sec:model}, which is able to closely approximate the morphology of the true mass distribution, along with many other morphologies.
@@
 However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by $\variable{output/BPLh0offset.txt}\sigma$.
-We have repeated this analysis with 50 separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ %$\variable{output/BPL_bias_percent.txt}\%$
-of the time of the time, whereas %the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and 
-the \plp{} model reaches the same level of bias %$\variable{output/PLP_bias_percent.txt}\%$ 
+We have repeated this analysis with 50 separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$
+of the time of the time, whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and 
+the \plp{} model reaches the same level of bias $\variable{output/PLP_bias_percent.txt}\%$ 
 of the time.
@@
 We find that spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ during \acf{O5}.
-This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}\jme{, which can only be detected relatively locally}.
-meaning \jme{that} upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances \jme{by \ac{GW}}. 
-Proposed next generation \jme{\ac{GW}} detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to \jme{coalescence binaries across cosmic time}$z\sim40$.
+This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}, which can only be detected relatively locally.
+This means that upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances by \acp{GW}. 
+Proposed next generation \ac{GW} detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to \acp{CBC} across cosmic time (up to $z\sim100$).
 At that time, cosmological surveys such as those enabled by the Nancy Grace Roman Space Telescope and the Vera Rubin Observatory are expected to be able to precisely measure $H(z)$ to $z\sim3$ \citep{spergel_wide-field_2015}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.

### 2024-03-18 src/scripts/pm_H0_twopanel.py
remove normalization and multiply GP by logm jac.

@@ -28,21 +28,19 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
 
     fig, axes = plt.subplots(ncols=2,figsize=(7.5,4*.75),facecolor='none',gridspec_kw={'width_ratios': [1.2, 1]})
     for i in range(NSAMPS//2):
-        axes[0].plot(TEST_M1S, r[i]/np.trapz(y=r[i],x=TEST_M1S), lw=0.1, c="blue",alpha=0.03)
+        axes[0].plot(TEST_M1S, r[i]/TEST_M1S, lw=0.1, c="blue",alpha=0.03)
         # TODO: get rid of indexing and plot parametric with TEST_M1S when new parametric runs are done with proper TEST_M1S
-        # TODO: get rid of normalization when new parametric and nonparametric runs are done
-        axes[0].plot(m1s_plotting_parametric, r_PLP[i]/np.trapz(y=r_PLP[i][1:],x=TEST_M1S[1:]),lw=0.1, c="green",alpha=0.03)
-        axes[0].plot(m1s_plotting_parametric, r_BPL[i]/np.trapz(y=r_BPL[i][1:],x=TEST_M1S[1:]),lw=0.1, c="orange",alpha=0.03)
+        axes[0].plot(m1s_plotting_parametric, r_PLP[i],lw=0.1, c="green",alpha=0.03)
+        axes[0].plot(m1s_plotting_parametric, r_BPL[i],lw=0.1, c="orange",alpha=0.03)
     
-    axes[0].plot(TEST_M1S, powerlaw_peak(TEST_M1S,alpha=TRUEVALS['alpha'],f_peak=TRUEVALS['f_peak'],mMax=TRUEVALS['mmax'],
-                                         mMin=TRUEVALS['mmin'],mu_m1=TRUEVALS['mu_m1'],sig_m1=TRUEVALS['sig_m1']
-                                         )/np.trapz(y= powerlaw_peak(TEST_M1S,alpha=TRUEVALS['alpha'],f_peak=TRUEVALS['f_peak'],mMax=TRUEVALS['mmax'],
-                                         mMin=TRUEVALS['mmin'],mu_m1=TRUEVALS['mu_m1'],sig_m1=TRUEVALS['sig_m1']),x=TEST_M1S),
-                c='k')
+    axes[0].plot(TEST_M1S, samples_PLP['rate'].mean(dim='draw').values*\
[…]

### 2024-03-18 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 % Define document class
-\documentclass[]{aastex631}
+\documentclass[preprint2,linenumbers]{aastex631}
 \usepackage{showyourwork}
@@
 
-d\section{}
+\section{}
 \textcolor{red}{Showyourwork Guidelines:}
@@
 \label{sec:intro}
+\begin{figure*}
+    \centering
+    \includegraphics[width=\textwidth]{figures/O5_pm.pdf}
+    \caption{Spectral siren measurement for an \ac{O5}-like catalog with the correct parametric model (\plp, green), a deliberately incorrect parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{GP}, blue).
+    The left panel shows the recovered source frame primary mass distribution for each model, and the simultaneously-inferred posteriors on \Ho{} are shown in the right panel.
+    The mass distribution and \Ho{} value used to generate the data are shown by a solid black line in each panel. 
+    The deliberately incorrect parametric model fails to recover the true mass distribution and therefore produces a biased estimate of \Ho{}, whereas both the correct and non-parametric models recover  the mass distribution and \Ho{}.
+    As the true mass distribution is unknown for real observations, using a non-parametric model mitigates systematic uncertainty that would otherwise arise from mismodeling the \ac{CBC} population, though it does introduce additional statistical uncertainty.
+    }
+    \label{fig:O5_GP}
+    \script{pm_H0_twopanel.py}
+\end{figure*}
+
 Like light, \acp{GW} are redshifted as they propagate across the universe, thereby becoming imprinted with the cosmic expansion history.
@@
 This makes them ``standard sirens'': powerful probes of cosmological parameters that circumvent the cosmological distance ladder~\citep{schutz_determining_1986,holz_using_2005}. 
-This is particularly useful in the context of the current ``Hubble tension:'' a disagreement between multiple methods of measuring the local expansion rate of the universe with electromagnetic observations \citep{freedman_measurements_2021}.
+This is particularly useful in the context of the current ``Hubble tension'': a disagreement between multiple methods of measuring the local expansion rate of the universe with electromagnetic observations \citep{freedman_measurements_2021}.
 
@@
 The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{} under an MIT license, and was enabled by the \showyourwork package \citep{Luger2021}.
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=\textwidth]{figures/O5_pm.pdf}
-    \caption{Spectral siren measurement for an \ac{O5}-like catalog with the correct parametric model (\plp, green), a deliberately incorrect parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{GP}, blue).
-    The left panel shows the recovered source frame primary mass distribution for each model, and the simultaneously-inferred posteriors on \Ho{} are shown in the right panel.
-    The mass distribution and \Ho{} value used to generate the data are shown by a solid black line in each panel. 
-    The deliberately incorrect parametric model fails to recover the true mass distribution and therefore produces a biased estimate of \Ho{}, whereas both the correct and non-parametric models recover  the mass distribution and \Ho{}.
-    As the true mass distribution is unknown for real observations, using a non-parametric model mitigates systematic uncertainty that would otherwise arise from mismodeling the \ac{CBC} population, though it does introduce additional statistical uncertainty.
-    }
-    \label{fig:O5_GP}
-    \script{pm_H0_twopanel.py}
-\end{figure*}
 

### 2024-03-20 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by $\variable{output/BPLh0offset.txt}\sigma$.
-We have repeated this analysis with 50 separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$
-of the time of the time, whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time, and 
-the \plp{} model reaches the same level of bias $\variable{output/PLP_bias_percent.txt}\%$ 
+We have repeated this analysis with 50 separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time,
+and the \plp{} model reaches the same level of bias $\variable{output/PLP_bias_percent.txt}\%$ 
 of the time.

### 2024-03-22 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-\DeclareAcronym{LVK}{short=LVK, long=LIGO Virgo KAGRA}
+\DeclareAcronym{LVK}{short=LVK, long=LIGO-Virgo-KAGRA}
 \DeclareAcronym{O5}{short=O5, long=the fifth LIGO Virgo KAGRA observing run}
@@
 \Acp{GW} from merging compact objects are the only transient sources that encode direct information about their luminosity distances. 
-Paired with their redshifts, a Hubble diagram can be constructed to probe the Universe's expansion.  
-``Spectral sirens'' use features in the mass distribution of \ac{GW} sources to extract this redshift information without the need for accompanying electromagnetic observations.
-So far, this technique has only been applied with simple parametric representations of the mass distribution. 
-However, the use of an incorrect mass model inevitably leads to biases in the cosmological inference, an acute problem given our current lack of understanding of the true source population.
-Furthermore, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to obtain unbiased measurements of cosmological parameters with the spectral siren methodology.
+Once paired with their redshifts, a Hubble diagram can be constructed to directly probe the Universe's expansion.
+This can be done in the absence of any electromagnetic information: features in the mass distribution of \ac{GW} can provide self-calibrating redshift measurements without the need for an accompanying electromagnetic transient or catalog of potential host galaxies.
+\sout{So far, this technique has only been applied with simple parametric representations of the mass distribution. 
+However, the use of an incorrect mass model inevitably leads to biases in the cosmological inference, an acute problem given our current lack of understanding of the true source population.}
+It is commonly presumed, however, that the exact shape of the mass distribution must be known \emph{a priori} to obtain unbiased measurements of cosmological parameters in this fashion.
 We demonstrate that prior knowledge of the \ac{GW} source mass distribution is not necessary in order to accurately infer cosmological parameters with spectral sirens.
-We do this by constructing a flexible, Gaussian process-based model for the \ac{GW} source population to analyze simulated \ac{GW} data consistent with expectations for the next \acl{LVK} observing run.
+By adopting a flexible, Gaussian-process model for the mass distribution of compact binaries, we find that we can accurately and agnostically learn this mass distribution while simultaneously measuring parameters governing cosmological expansion.
+We demonstrate this analyses using a simulated catalog of $10^3$ gravitational-wave events consistent with expectations for the next \acl{LVK} observing run.
 We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$  measurement of $H(z=\variable{output/mostsensitivez.txt})$ ($1\sigma$ uncertainties).
@@
 Unlike light, however, the form of \ac{GW} signals are known from first principles and their selection effects are extremely well understood.
-This allows for a precise estimate of each \ac{GW} catalogs' completeness and an unbiased measurement of the true \ac{GW} population. 
-Additionally, the \ac{GW} signals detected by the LIGO \citep{aasi_advanced_2015}, Virgo \citep{acernese_advanced_2014}, and KAGRA \citep{akutsu_overview_2021} detectors provide direct measurements of the distance to their sources.
-This makes them ``standard sirens'': powerful probes of cosmological parameters that circumvent the cosmological distance ladder~\citep{schutz_determining_1986,holz_using_2005}. 
+This allows for a precise estimate of each \ac{GW} catalogs' completeness and an unbiased measurement of the true \ac{GW} source population~\citep{2023PhRvX..13d1039A,abbott_population_2023,2023PhRvD.108d3011E}.
+Additionally, the \ac{GW} signals observed by the LIGO, Virgo, and KAGRA detectors~\citep{aasi_advanced_2015,acernese_advanced_2014,akutsu_overview_2021}  provide direct measurements of the distance to their sources.
+This makes them ``standard sirens'': direct probes of cosmological parameters that circumvent the need for a cosmological distance ladder~\citep{schutz_determining_1986,holz_using_2005}. 
 This is particularly useful in the context of the current ``Hubble tension'': a disagreement between multiple methods of measuring the local expansion rate of the universe with electromagnetic observations \citep{freedman_measurements_2021}.
 
-A well-known application of standard sirens was the multi-messenger event GW170817 \citep{abbott_multi-messenger_2017}, whose clear association with a host galaxy provided a precise redshift measurement, allowing for a direct measurement of the Hubble constant, \Ho{} \citep{ abbott_gravitational-wave_2017}.
-External redshift information can also come from galaxy catalogs, which provide redshift distributions for each \ac{GW} signal, allowing for a probabilistic measurement of \Ho{} when multiple \ac{GW} detections are combined \citep[e.g.][]{del_pozzo_inference_2012, chen_two_2018, fishbach_standard_2019, gray_cosmological_2020, gwtc3_cosmo, gair_hitchhikers_2023}.
+Cosmological measurements with gravitational waves traditionally require independent electromagnetic measurement of source redshifts.
+A well-known demonstration was the multi-messenger event GW170817 \citep{abbott_multi-messenger_2017,coulter_swope_2017,valenti_discovery_2017,2017ApJ...848L..27T}, whose clear association with a host galaxy provided a precise redshift measurement and allowed for a direct measurement of the Hubble constant, \Ho{} \citep{abbott_gravitational-wave_2017}.
+External redshift information can also come from galaxy catalogs, which provide redshift distributions for each \ac{GW} signal and allow for a probabilistic measurement of \Ho{} when multiple \ac{GW} detections are combined \citep[e.g.][]{del_pozzo_inference_2012, chen_two_2018, fishbach_standard_2019, gray_cosmological_2020, gwtc3_cosmo, gair_hitchhikers_2023}.
 
-However, external information about \ac{GW} source redshifts need not be available in order to use them as standard sirens.
+However, external electromagnetic information about \ac{GW} source redshifts need not be available in order to use them as standard sirens.
 \ac{GW} signals provide direct measurements of each source's luminosity distance, $d_L$, and redshifted, or detector frame, masses, $m_{\det}= m_{\source}(1+z)$ \citep[e.g.][]{chen_mass-redshift_2019}.
-Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the rate of expansion of the universe at the time the \ac{GW} signal was emitted, $H(z)$.
-In practice, unless tidal information is available \citep{messenger_measuring_2012}, the source frame mass is unknown.
-However, it is possible to know the location of features in the source frame mass distribution.
-The whole mass distribution therefore acts similarly to an electromagnetic spectrum, where the location of spectral features provides a redshift measurement.
+Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the expansion of the universe at the time the \ac{GW} signal was emitted, $H(z)$.
+
+In practice, the source-frame masses of individual events are not known~\citep[unless tidal information is available][]{messenger_measuring_2012}.
+It is, however, possible to regard the \textit{population} of compact binaries at large and use known features in their source-frame mass distribution to obtain self-calibrated redshift estimates.
+The full mass distribution therefore acts analogously to an electromagnetic spectrum, in which the apparent locations of spectral features provide a redshift measurement.
 The method of using the mass distribution of \ac{GW} sources to measure cosmological parameters has therefore been coined ``spectral sirens'' \citep{ezquiaga_spectral_2022}.
-Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff_gravitational_1993,Taylor:2011fs} using the binary neutron star mass distribution, and extended to the \ac{BBH} mass distribution by \cite{farr_future_2019}.
-Spectral siren analyses have since been implemented by the LIGO--Virgo--KAGRA Collaborations using the latest \ac{GW} catalog \cite{gwtc3_cosmo}.
 
-Perhaps because of the analogy to electromagnetic spectra in which the location of the source frame spectral features is well-known, or because of the examples given in \cite{farr_future_2019} and \cite{chernoff_gravitational_1993}, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to do this measurement, from theoretical expectations of e.g.
-the location of a pulsational-pair instability pileup, upper mass gap, or maximum neutron star mass.
+Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff_gravitational_1993} and \cite{Taylor:2011fs} using the binary neutron star mass distribution, and extended to the \ac{BBH} mass distribution by \cite{farr_future_2019}.
+Spectral siren analyses have since been implemented by the LIGO--Virgo--KAGRA Collaborations using the latest \ac{GW} catalog \citep{gwtc3_cosmo}.
+
+Perhaps because of the analogy to electromagnetic spectra, in which the locations of source-frame spectral features are often known from first principles, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to perform a spectral siren redshift measurement.
+Indeed, first principles models for the mass distribution of merging compact binaries are not available: significant theoretical uncertainties exist on the shapes, locations, and very existences of predicted features like a ``pile-up'' due to pulsational-pair instability, a pair-instability-driven upper mass gap, or the maximum neutron star mass.
+To this end, spectral siren cosmology\footnote{And also statistical cosmological measurements leveraging galaxy catalogs.} relies on \textit{simultaneously} measuring a source-frame mass spectrum alongside cosmological parameters.
+
 This would certainly be problematic, as mismodeling of the specific feature would hinder the accuracy of the method and introduce biases in the cosmological inference \citep{Mukherjee:2021rtw,mastrogiovanni_importance_2021,pierra_study_2023}.
@@
 This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}, which can only be detected relatively locally.
+\comment{Projected DESI errors are $\sim3\%$ at $z=1$, which are comparable with our estimated error of $\variable{output/Hz_percent.txt}\%$, but those won't be realized until 202X.}
 This means that upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances by \acp{GW}. 

### 2024-03-22 src/scripts/nonparametric_corner.py
[showyourwork] overleaf sync

@@ -108,5 +108,5 @@ def H0_Om_corner(id):
     Hz(id,ax=axsRight,save=False)
     fig.savefig(paths.figures / "O5_GP_corner.pdf")
 
-id = az.InferenceData.from_netcdf(paths.data / "mcmc_nonparametric_fitOm0.nc4")
+id = az.InferenceData.from_netcdf(paths.data / "mcmc_nonparametric_fitOm0_old.nc4")
 H0_Om_corner(id)

### 2024-03-22 src/scripts/pm_H0_twopanel.py

@@ -24,14 +24,12 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     id_BPL = az.InferenceData.from_netcdf(path_BPL)
     samples_BPL = id_BPL.posterior.sel(chain=0).reset_coords("chain",drop=True)
     r_BPL = np.nan_to_num(np.exp(samples_BPL['log_rate']))
-    m1s_plotting_parametric = np.linspace(0,TEST_M1S.max(),len(TEST_M1S)) # TODO: remove this w/ new runs
 
     fig, axes = plt.subplots(ncols=2,figsize=(7.5,4*.75),facecolor='none',gridspec_kw={'width_ratios': [1.2, 1]})
     for i in range(NSAMPS//2):
         axes[0].plot(TEST_M1S, r[i]/TEST_M1S, lw=0.1, c="blue",alpha=0.03)
-        # TODO: get rid of indexing and plot parametric with TEST_M1S when new parametric runs are done with proper TEST_M1S
-        axes[0].plot(m1s_plotting_parametric, r_PLP[i],lw=0.1, c="green",alpha=0.03)
-        axes[0].plot(m1s_plotting_parametric, r_BPL[i],lw=0.1, c="orange",alpha=0.03)
+        axes[0].plot(TEST_M1S, r_PLP[i],lw=0.1, c="green",alpha=0.03)
+        axes[0].plot(TEST_M1S, r_BPL[i],lw=0.1, c="orange",alpha=0.03)
     
     axes[0].plot(TEST_M1S, samples_PLP['rate'].mean(dim='draw').values*\
                  powerlaw_peak(TEST_M1S,f_peak=TRUEVALS['f_peak'],

### 2024-03-26 src/scripts/data_generation.py
add flag for perfect measurements

@@ -29,6 +29,8 @@ SIGMA_DL = 1000
 SNR_THRESH=8.
 NUM_INJ=N_SOURCES*50*30
 
+perf_meas = False
+
 # random number generators
 np_rng = np.random.default_rng(516)
 
@@ -89,13 +91,20 @@ if  __name__ == "__main__":
     np.save(paths.data / "gw_data/log_pinj_det.npy",log_pinj_det)
     
     ## find events and generate mock PE
-    m1z_PE, m2z_PE, dL_PE, log_PE_prior = gen_snr_scaled_PE(np_rng,m1s_true,m1s_true,dL_true/1000,osnr_interp,
+    m1z_PE, m2z_PE, dL_PE, log_PE_prior, det_dict = gen_snr_scaled_PE(np_rng,m1s_true,m1s_true,dL_true/1000,osnr_interp,
                                                             reference_distance,N_SAMPLES_PER_EVENT,H0_FID,OM0_FID,
                                                             # errors taken from Jose's "gwutils.py" for O5
                                                             mc_sigma=3.0e-2,eta_sigma=5.0e-3,theta_sigma=5.0e-2, snr_thresh=SNR_THRESH,
-                                                            return_og=False)
+                                                            return_og=True)
 
     dL_PE *= 1000 # unit matching
+    if perf_meas:
+        # overwrite PE samples with the true values for each event, adding a dimension so that 
+        # later array operations still work
+        m1z_PE = np.expand_dims(det_dict['m1'] * (1 + det_dict['z']),1)
+        m2z_PE = np.expand_dims(det_dict['m2'] * (1 + det_dict['z']),1)
+        dL_PE = np.expand_dims(det_dict['lum_dist'], 1)
+        log_PE_prior = 0
 
     # cut out samples below SNR interpolation range. 
     # It is very unlikely that any samples are down there, but we do this just in case.

### 2024-03-26 src/scripts/priors.py
fix redshift dist parameters in PLP model (already fixed in others)

@@ -207,13 +207,13 @@ def PLP(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,remove_low_Neff=False
     mmin = numpyro.sample("mmin",dist.Uniform(1.0,20.0))
     mmax = numpyro.sample("mmax",dist.Uniform(30.0,100.0))
     alpha = numpyro.sample("alpha",dist.Normal(0,5))
-    mu_m1 = numpyro.sample("mu_m1",dist.Uniform(20,60))
-    sig_m1 = numpyro.sample("sig_m1",dist.Uniform(1,10))
+    mu_m1 = numpyro.sample("mu_m1",dist.Uniform(20,60))# TRUEVALS['mu_m1']
+    sig_m1 = numpyro.sample("sig_m1",dist.Uniform(1,10))# TRUEVALS['sig_m1']
     f_peak = TRUEVALS['f_peak'] #numpyro.sample("f_peak",dist.Uniform(0,1))
     #Fixed
-    alpha_z = numpyro.sample("alpha_z",dist.Uniform(0,2))#TRUEVALS['alpha_z']#
-    zp = numpyro.sample("zp",dist.Uniform(0,5)) #TRUEVALS['zp']#
-    beta = numpyro.sample("beta_z",dist.Uniform(0,5)) #numpyro.deterministic('beta',TRUEVALS['beta_z'])# 
+    alpha_z = TRUEVALS['alpha_z']#numpyro.sample("alpha_z",dist.Uniform(0,2))
+    zp = TRUEVALS['zp']#numpyro.sample("zp",dist.Uniform(0,5)) #
+    beta = TRUEVALS['beta_z'] #numpyro.sample("beta_z",dist.Uniform(0,5)) #numpyro.deterministic('beta',TRUEVALS['beta_z'])# 
 
     # convert event data to source frame
     z = jgwcosmo.z_at_dl_approx(dL,H0,Om0,zmin=ZMIN,zmax=ZMAX+8.)

### 2024-03-26 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 \DeclareAcronym{LVK}{short=LVK, long=LIGO-Virgo-KAGRA}
-\DeclareAcronym{O5}{short=O5, long=the fifth LIGO Virgo KAGRA observing run}
+\DeclareAcronym{O5}{short=O5, long=the fifth LIGO-Virgo-KAGRA observing run}
 
@@
 \title{No need to know: astrophysics-agnostic gravitational wave cosmology}
+% \daniel{No need to know: astrophysically-agnostic gravitational wave cosmology\\
+% No need to know: gravitational-wave cosmology without astrophysics\\
+% No need to know: gravitational-wave cosmology without astrophysical assumptions}}
+% Ignorance is bliss: astrophysics-agnostic gravitational wave cosmology
 %\title{No need to know: astrophysics-agnostic spectral siren cosmology}
@@
 \affiliation{Department of Physics, University of Chicago, Chicago, IL 60637, USA}
-\affiliation{Kavli Institute for Cosmological Physics and Enrico Fermi Institute, The University of Chicago, Chicago, IL 60637, USA}
+\affiliation{Kavli Institute for Cosmological Physics, The University of Chicago, Chicago, IL 60637, USA}
 \affiliation{Department of Astronomy \& Astrophysics, The University of Chicago, Chicago, IL 60637, USA}
+\affiliation{Enrico Fermi Institute, The University of Chicago, Chicago, IL 60637, USA}
 
@@
 \begin{abstract}
-\Acp{GW} from merging compact objects are the only transient sources that encode direct information about their luminosity distances. 
-Once paired with their redshifts, a Hubble diagram can be constructed to directly probe the Universe's expansion.
-This can be done in the absence of any electromagnetic information: features in the mass distribution of \ac{GW} can provide self-calibrating redshift measurements without the need for an accompanying electromagnetic transient or catalog of potential host galaxies.
-\sout{So far, this technique has only been applied with simple parametric representations of the mass distribution. 
-However, the use of an incorrect mass model inevitably leads to biases in the cosmological inference, an acute problem given our current lack of understanding of the true source population.}
-It is commonly presumed, however, that the exact shape of the mass distribution must be known \emph{a priori} to obtain unbiased measurements of cosmological parameters in this fashion.
-We demonstrate that prior knowledge of the \ac{GW} source mass distribution is not necessary in order to accurately infer cosmological parameters with spectral sirens.
-By adopting a flexible, Gaussian-process model for the mass distribution of compact binaries, we find that we can accurately and agnostically learn this mass distribution while simultaneously measuring parameters governing cosmological expansion.
-We demonstrate this analyses using a simulated catalog of $10^3$ gravitational-wave events consistent with expectations for the next \acl{LVK} observing run.
-We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$  measurement of $H(z=\variable{output/mostsensitivez.txt})$ ($1\sigma$ uncertainties).
-This astrophysics-agnostic spectral siren technique will be key to deliver precise and unbiased cosmological constraints with future observations and in the presence of redshift evolution of the mass spectrum.
+\Acp{GW} from merging compact objects encode direct information about the luminosity distance to the binary. 
+When paired with a redshift measurement, this enables standard-siren cosmology: a Hubble diagram can be constructed to directly probe the Universe's expansion.
+This can be done in the absence of electromagnetic measurements: features in the mass distribution of \ac{GW} sources can provide self-calibrating redshift measurements without the need for an accompanying electromagnetic transient or catalog of potential host galaxies.
+% So far, this technique has only been applied with simple parametric representations of the mass distribution. 
+% However, the use of an incorrect mass model inevitably leads to biases in the cosmological inference, an acute problem given our current lack of understanding of the true source population.
+% Furthermore, i
+It is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori}\/ to obtain unbiased measurements of cosmological parameters in this fashion.
+We demonstrate that spectral sirens can accurately infer cosmological parameters without assuming
+prior knowledge of the \ac{GW} mass distribution.
+By adopting a flexible, Gaussian process-based model for the mass distribution of compact binaries, we find that we can accurately and agnostically infer the mass distribution while simultaneously measuring parameters governing cosmological expansion.
+We demonstrate this analysis using a simulated catalog of 1,000 gravitational-wave events consistent with expectations for the next \acl{LVK} observing run.
+We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{}, keeping all other cosmological parameters fixed, and a $\variable{output/Hz_percent.txt}\%$  measurement of $H(z=\variable{output/mostsensitivez.txt})$ when fitting for multiple cosmological parameters ($1\sigma$ uncertainties).
+This astrophysically-agnostic spectral siren technique will be essential to arrive at precise and unbiased cosmological constraints from gravitational-wave source populations.
 \end{abstract}
-
-\section{}
-\textcolor{red}{Showyourwork Guidelines:}
-\begin{itemize}
-    \item Always edit \texttt{ms.tex} in Overleaf
-    \item Always edit \texttt{bib.bib} in Overleaf
-    \item Always edit \texttt{macros.sty} in Overleaf
-    \item Always edit anything in the \texttt{figures} directory using git (with a python script that should generate the figures)
-    \item Always edit anything in the \texttt{output} directory using git (with a python script that should generate the output)
-\end{itemize}
 
@@
 
-Like light, \acp{GW} are redshifted as they propagate across the universe, thereby becoming imprinted with the cosmic expansion history.
-Unlike light, however, the form of \ac{GW} signals are known from first principles and their selection effects are extremely well understood.
+Like light, \acp{GW} are redshifted as they propagate across the universe, thereby bearing imprints of the Universe's cosmic expansion history.
+Unlike light, however, the form of \ac{GW} signals are known from first principles, directly from the theory of general relativity.
+Furthermore, because \acp{GW} propagate across the Universe without attenuation from intervening matter, and because the properties of \ac{GW} detectors are well characterized, \ac{GW} selection effects are extremely well understood.% This is to be contrasted with many electromagnetic observations, where effects such as reddening, selection bias, and $k$-corrections can lead to significant complications.
+[too strong?]
 This allows for a precise estimate of each \ac{GW} catalogs' completeness and an unbiased measurement of the true \ac{GW} source population~\citep{2023PhRvX..13d1039A,abbott_population_2023,2023PhRvD.108d3011E}.
@@
 This makes them ``standard sirens'': direct probes of cosmological parameters that circumvent the need for a cosmological distance ladder~\citep{schutz_determining_1986,holz_using_2005}.
-This is particularly useful in the context of the current ``Hubble tension'': a disagreement between multiple methods of measuring the local expansion rate of the universe with electromagnetic observations \citep{freedman_measurements_2021}.
 
-Cosmological measurements with gravitational waves traditionally require independent electromagnetic measurement of source redshifts.
-A well-known demonstration was the multi-messenger event GW170817 \citep{abbott_multi-messenger_2017,coulter_swope_2017,valenti_discovery_2017,2017ApJ...848L..27T}, whose clear association with a host galaxy provided a precise redshift measurement and allowed for a direct measurement of the Hubble constant, \Ho{} \citep{abbott_gravitational-wave_2017}.
-External redshift information can also come from galaxy catalogs, which provide redshift distributions for each \ac{GW} signal and allow for a probabilistic measurement of \Ho{} when multiple \ac{GW} detections are combined \citep[e.g.][]{del_pozzo_inference_2012, chen_two_2018, fishbach_standard_2019, gray_cosmological_2020, gwtc3_cosmo, gair_hitchhikers_2023}.
+A well-known demonstration of standard siren cosmology was the multi-messenger event GW170817 \citep{abbott_multi-messenger_2017,coulter_swope_2017,valenti_discovery_2017,2017ApJ...848L..27T}, whose clear association with a host galaxy provided a precise redshift measurement and allowed for a direct ``bright siren'' measurement of the Hubble constant, \Ho{} \citep{abbott_gravitational-wave_2017}.
+External redshift information can also come from galaxy catalogs, which provide an ensemble of possible redshifts for each \ac{GW} signal, allowing for a probabilistic ``dark siren'' measurement of \Ho{} when multiple \ac{GW} detections are combined \citep{del_pozzo_inference_2012, chen_two_2018, fishbach_standard_2019, gray_cosmological_2020, gwtc3_cosmo, gair_hitchhikers_2023}.
 
-However, external electromagnetic information about \ac{GW} source redshifts need not be available in order to use them as standard sirens.
+External information about \ac{GW} source redshifts need not be available in order to use them as standard sirens, however.
 \ac{GW} signals provide direct measurements of each source's luminosity distance, $d_L$, and redshifted, or detector frame, masses, $m_{\det}= m_{\source}(1+z)$ \citep[e.g.][]{chen_mass-redshift_2019}.
@@
 
-In practice, the source-frame masses of individual events are not known~\citep[unless tidal information is available][]{messenger_measuring_2012}.
+In practice, the source-frame masses of individual events are not known~\citep[unless tidal information is available, e.g.][]{messenger_measuring_2012, 2021PhRvD.104h3528C}.
 It is, however, possible to regard the \textit{population} of compact binaries at large and use known features in their source-frame mass distribution to obtain self-calibrated redshift estimates.
@@
 The method of using the mass distribution of \ac{GW} sources to measure cosmological parameters has therefore been coined ``spectral sirens'' \citep{ezquiaga_spectral_2022}.
-
 Spectral sirens were first demonstrated to be a feasible method to measure the Hubble constant by \cite{chernoff_gravitational_1993} and \cite{Taylor:2011fs} using the binary neutron star mass distribution, and extended to the \ac{BBH} mass distribution by \cite{farr_future_2019}.
@@
 
-Perhaps because of the analogy to electromagnetic spectra, in which the locations of source-frame spectral features are often known from first principles, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to perform a spectral siren redshift measurement.
-Indeed, first principles models for the mass distribution of merging compact binaries are not available: significant theoretical uncertainties exist on the shapes, locations, and very existences of predicted features like a ``pile-up'' due to pulsational-pair instability, a pair-instability-driven upper mass gap, or the maximum neutron star mass.
-To this end, spectral siren cosmology\footnote{And also statistical cosmological measurements leveraging galaxy catalogs.} relies on \textit{simultaneously} measuring a source-frame mass spectrum alongside cosmological parameters.
+Central to the spectral siren methodology is knowledge of the compact binary mass distribution.
+First principles models for mass distributions of merging compact binaries are, however, not available: significant theoretical uncertainties exist about the shapes, locations, and very existence of predicted features in the mass distribution~\citep[e.g.][]{ zevin_constraining_2017, mapelli_binary_2020,2021ApJ...910..152Z,marchant_evolution_2023}.
+This includes uncertainties about critical aspects, such as the existence of a ``pile-up'' due to pulsational-pair instability, a pair-instability-driven upper mass gap, the maximum neutron star mass, and the existence of a putative lower mass gap between neutron stars and black holes.
+To this end, spectral siren cosmology\footnote{As well as statistical cosmological measurements leveraging galaxy catalogs} relies on \textit{simultaneously} measuring a source-frame mass spectrum alongside cosmological parameters.
+This is typically accomplished by adopting a phenomenological, parametric model for the mass distribution, usually composed of power laws and Gaussians \citep[e.g.][]{gwtc3_cosmo}.
 
-This would certainly be problematic, as mismodeling of the specific feature would hinder the accuracy of the method and introduce biases in the cosmological inference \citep{Mukherjee:2021rtw,mastrogiovanni_importance_2021,pierra_study_2023}.
-Such systematic bias is illustrated in Figure~\ref{fig:O5_GP}, where we show the result of fitting simulated data with a model that has the true mass distribution contained within its prior and one that does not.
-The latter does not accurately infer the true value of \Ho{}, whereas the former does.
+Such parametric modeling of the compact binary mass distribution raises its own set of dangers.
+It is well-known that different parametric models can generically yield very different results~\citep{abbott_population_2021,abbott_population_2023}.
+This is problematic: because the measured mass distribution serves as the template by which to extract redshifts, a mismodeled mass distribution would introduce systematic errors in inferred redshifts and, in turn, systematically bias any resulting cosmological inference~\citep{ezquiaga_spectral_2022,Mukherjee:2021rtw,mastrogiovanni_importance_2021,pierra_study_2023}.
 
-This is particularly troubling in light of the fact that there remains much uncertainty over the cause, location, and robustness of currently-identified features in the \ac{GW} source mass distribution.
-While several features have been identified in the current data \citep{tiwari_exploring_2022,edelman_aint_2022, abbott_population_2023,edelman_cover_2023}, including three overdensities, two underdensities, a maximum and minimum black hole mass, only three of these are statistically robust \citep{farah_things_2023}.
-Additionally, the morphology of the mass distribution is a result of the interplay between several complex physical processes pertaining to massive stars, such as binary stellar evolution, pair instability, globular cluster dynamics, and stellar winds, each of which has large theoretical uncertainties \citep[e.g.][]{ zevin_constraining_2017, mapelli_binary_2020,marchant_evolution_2023}.
-These large theoretical uncertainties mean that accurate predictions for the shape of the \ac{GW} source mass distribution are a considerable challenge.
-If such \emph{a priori} knowledge of the mass distribution's morphology were requisite for a spectral siren method, the effectiveness of such a measurement would be significantly hampered.
+The situation is demonstrated in Fig.~\ref{fig:O5_GP}, in which we perform spectral siren cosmology on a simulated population of binary black holes.
+We reconstruct the mass distribution using two parametric models, one that contains the true simulated mass distribution and one that does not.
+While the former yields a measurement of $H_0$ consistent with the true underlying value, the latter does not.
+Such systematic biases may already be relevant, as cosmological measurements by the LIGO-Virgo-KAGRA are known to depend on the exact choice of mass model used~\citep{gwtc3_cosmo}.
+Furthermore, these biases may become a dominant source of uncertainty in the near future~\citep{pierra_study_2023}.
+The prospects of such a dominant systematic uncertainty is troubling.
+If \emph{a priori}\/ knowledge of the mass distribution's morphology (whether an exact theoretical prediction or  knowledge of the correct parametric family of models) is a pre-requisite for the spectral siren method, the effectiveness of such a measurement would be significantly hampered.
 
-However, the information in the spectral siren measurement comes solely from the assumption that all \acp{CBC} follow the same mass distribution.\footnote{More precisely, the spectral siren method is also valid if \acp{CBC} have different mass distributions at different redshifts, so long as the evolution of the mass distribution does not exactly mimic cosmology, and the model for the mass distribution allows for evolution \citep{ezquiaga_spectral_2022}.
-}
-Therefore the location, shape, and existence of all features in the mass distribution can be inferred simultaneously with $H(z)$.
-
-In this work, we explicitly demonstrate that no \emph{a priori} knowledge about the shape of the \ac{CBC} mass spectrum is necessary to use the spectral siren methodology. 
+In this work, we explicitly demonstrate that no \emph{a priori}\/ knowledge of the shape of the \ac{CBC} mass spectrum is necessary to use the spectral siren methodology. 
 We do this by inferring $H(z)$ with a flexible, non-parametric model for the mass distribution of \acp{CBC} (blue lines in Figure~\ref{fig:O5_GP}). 
 This model makes minimal prior assumptions about the shape of the mass distribution, enabling it to accurately infer a wide range of morphologies and remain agnostic to the astrophysical processes that give rise to the mass distribution's features.
-Despite its flexibility, it is able to consistently obtain unbiased measurements of cosmological parameters, showing that nonparametric methods are not only sufficient for a spectral siren measurement, they can also mitigate systematic effects in the measurement caused by model misspecification.
+Despite its flexibility, our approach is able to consistently obtain unbiased measurements of cosmological parameters, showing that nonparametric methods are not only sufficient for a spectral siren measurement, they can also mitigate systematic effects in the measurement caused by model misspecification.
+The success of the non-parametric mass model in recovering the injected cosmological parameters demonstrates that the information in the spectral siren measurement comes solely from the assumption that all \acp{CBC} follow the same mass distribution, not from the enforcement of specific features.\footnote{More precisely, the spectral siren method is also valid if \acp{CBC} have different mass distributions at different redshifts, so long as the evolution of the mass distribution does not exactly mimic cosmology, and the model for the mass distribution allows for evolution \citep{ezquiaga_spectral_2022}.
+}
 
-We find that using our non-parametric model model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ during \ac{O5}, when the detectors will reach their design sensitivity.
+We find that using our non-parametric model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ during \ac{O5}, when the detectors will reach their design sensitivity.
 %, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
-While constraints on \Ho{} are used in this work to quantify the accuracy and precision of our non-parametric method, we emphasize that spectral sirens are able to constrain $H(z)$ out to redshifts that are inaccessible by electromagnetic observations, especially with next-generation detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn, You:2020wju}\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}.
-Applied to data from next-generation detectors, this method will therefore be able to constrain the dark energy equation of state parameter, $w$, at redshifts $\gtrsim 2$.
+We highlight measurements of \Ho{} in order to benchmark the accuracy and precision of our non-parametric method, as well as explore the role of spectral sirens in elucidating the Hubble tension.
+However, the primary utility of spectral siren measurements will be in constraining $H(z)$ at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn, You:2020wju}\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}.
+Applied to data from next-generation detectors, this method will constrain the dark energy equation of state parameter, $w$, at redshifts $\gtrsim 2$.
 
-This paper is organized as follows: Section~\ref{sec:ss} explains how cosmological parameters are inferred from the mass distribution of \ac{GW} sources (i.e. the spectral siren method).
+This paper is organized as follows: Section~\ref{sec:ss} introduces the spectral siren method, demonstrating how cosmological parameters are inferred from the mass distribution of \ac{GW} sources.
 Section~\ref{sec:model} describes the non-parametric mass distribution we develop for use within the spectral siren method.
@@
 
-The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{} under an MIT license, and was enabled by the \showyourwork package \citep{Luger2021}.
+The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{githubLink}, and was enabled by the \showyourwork package \citep{Luger2021}.
 
-\section{Cosmology with an astrophysically-agnostic mass distribution model}
+\section{Cosmology with an astrophysically-agnostic mass distribution}
 \label{sec:methods}
+
+The spectral siren method of measuring cosmological parameters relies on the assumption that either all compact binaries follow the same mass distribution at all redshifts, or that any redshift evolution of the mass distribution does not perfectly mimic cosmology \citep[i.e.
+all features do not simultaneously exhibit an {\em identical}\/ monotonic rightward shift with redshift;][]{ezquiaga_spectral_2022}.
+Specifically, the method identifies the preferred relationship between luminosity distance and redshift that causes all masses to follow, as closely as possible, this common source-frame distribution.
+We do not need to know this source-frame distribution in advance, but, given some model for the black hole mass distribution, can infer both this mass distribution and cosmology simultaneously.
+
+As described above, though, strongly parametrized models for the mass distribution yield biased measurements of cosmological parameters if they poorly approximate the true mass distribution of compact binaries.
+Our goal is to circumvent this and instead model the population of \acp{GW} sources in a flexible and astrophysically-agnostic way. 
+There exist several non-parametric methods developed for this purpose \citep{tiwari_vamana_2021,edelman_aint_2022,sadiq_flexible_2022,edelman_cover_2023,mandel_extracting_2019,ray_non-parametric_2023}.
+While well-suited to infer the \ac{GW} source population with a fixed cosmology, several of these methods employ fixed features in source frame mass, such as bin edges \citep{mandel_extracting_2019,ray_non-parametric_2023} or spline nodes \citep{edelman_aint_2022}.
+Since these locations were chosen with a fixed cosmology, they risk causing the inference to prefer the cosmological parameters used when choosing the feature locations. 
+Indeed, \citet{mastrogiovanni_importance_2021} show how using fixed features can significantly bias cosmological inference within the spectral siren methodology.
+We therefore opt for a model of the source frame mass distribution that foregoes the need to define such features.
+
+% Intro to GPs
+For this purpose, we construct a model for the with a \acf{GP}, a common tool for non-parametric inference. 
+\Acp{GP} are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
+Their smoothness properties make them widely useful in \ac{GW} data analysis for regression problems, such as modeling time-domain waveforms \citep{doctor_statistical_2017, huerta_eccentric_2018} and the neutron star equation of state \citep{landry_nonparametric_2019}, density estimation problems, such as estimating posterior densities of single-event parameters from parameter estimation samples ~\citep{demilio_density_2021}, and as a prior on histogram bin heights for population inference \citep{mandel_model-independent_2017, li_flexible_2021, ray_non-parametric_2023}.
+
+Our use case is slightly different from previous analyses:
+we utilize a \ac{GP} as a prior on the functions that describe the primary mass distribution of \acp{CBC}.
+This choice adds very little prior information about the shape of the mass distribution, besides enforcing that it must be smooth.
+
+\subsection{Simulated Data}
+%% TRANSITION: Describe how our goal is to demonstrate the utility of GPs and cosmology by constructing and analyzing a simulated catalog. 
+To demonstrate the effectiveness of our \ac{GP}-based mass distribution in inferring both cosmological parameters and the population of \ac{GW} sources without introducing features or biases driven driven by prior assumptions, we apply our methodology to a simulated dataset.
+In generating a catalog of \ac{GW} sources from a known population and cosmological model, we are able to quantify the accuracy of our method.
+The use of simulated data also enables us to make projections for future datasets and safely ignore dimensions such as spin and mass ratio that do not impact cosmological measurements but would otherwise be important to simultaneously fit to avoid biases in inference of a true dataset's population.
+
+%% Describe details of catalog
+We design our simulated catalog to match the characteristics of the data expected from one year of observation in \ac{O5}. 
+The events in this catalog are drawn from an underlying population described by the \plp{} mass distribution presented in \citet{talbot_measuring_2018} and used in \citet{abbott_binary_2019, abbott_population_2023}, as well and the redshift distribution presented in \citet{callister_shouts_2020}, with hyperparameters consistent with those found in \citet{abbott_population_2023}.
+We assume the mass distribution has no redshift evolution for the redshift range to which the \ac{O5} detectors will be sensitive. 
+This assumption is consistent with current data \citep{fishbach_when_2021,van_son_redshift_2022,abbott_population_2023}, but will have to be revisited in future observing runs.
+% \daniel{We should say more about the importance of this assumption?}
+This process results in a catalog of $\variable{output/num_found_events.txt}$ \acp{BBH}.
+
+We use the \texttt{GWMockCat} package to generate consistent simulated data and sensitivity estimates \citep{farah_things_2023}.
+Additional details of the data simulation, including the form of the injected population, are described in Appendix~\ref{ap:data generation}.
+
+%% Go into the details of the likelihoods needed for the data analysis - new subsection?
 \subsection{The Spectral Siren Method}
 \label{sec:ss}
-To simultaneously infer cosmological parameters and the population of GW sources, we employ a hierarchical Bayesian analysis.
+To simultaneously infer cosmological parameters and the population of \ac{GW} sources, we employ a hierarchical Bayesian analysis.
 This allows us to ``undo'' the selection effects of \ac{GW} detectors to obtain a true, or astrophysical, population and constrain the cosmic expansion history.
 
-Given \ac{GW} data $\{d\}$ that contains $N_{\text{evs}}$ detected \ac{GW} signals, each with parameters $\theta_i$, the posterior probability of the \ac{GW} sources being drawn from an underlying population and background cosmology described by hyper-parameters $\Lambda$ is \citep{loredo_handling_2009, Taylor:2011fs, mandel_extracting_2019}
+Given a source population and background cosmology described by hyperparameters $\Lambda$, the likelihood of observing data $\{d\}$ that contains $N_{\text{ev}}$ detected \ac{GW} signals, each with parameters $\theta_i$, is \citep{loredo_handling_2009, Taylor:2011fs, mandel_extracting_2019}
 \begin{equation}
-    p(\Lambda|d) \propto p(\Lambda) e^{-N_{\text{exp}}(\Lambda)} \prod_i^{N_{\text{evs}}} \frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda) \, .
+\begin{aligned}
+    &p(\{d\},\{\theta\}|\Lambda) \propto \\
+    &\hspace{1.2cm} e^{-N_{\text{exp}}(\Lambda)}\prod_i^{N_{\text{ev}}} p(d_i|\theta_i) \frac{\diff N}{\diff t_{\det} \diff \theta} (\theta_i;\Lambda) \, .
+\end{aligned}
 \label{eq:inhomog-poisson}
 \end{equation}
-Here, $\Lambda$ is the set of population parameters, or hyper-parameters, and $p(\Lambda)$ is the prior on $\Lambda$.
+Here, $\Lambda$ is the set of population parameters, or hyper-parameters.
 $\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ is the detector frame merger rate density of \acp{BBH} conditioned on hyper-parameters $\Lambda$, this quantity is also referred to as the population model.
@@
 
-Since the event parameters are not perfectly measured, we marginalize $\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ over the likelihood of each event's parameters given the data.
-Practically, this is done by a Monte Carlo average over each event's posterior samples $\{\theta_j\}_i$, dividing out the prior used when inferring those posterior samples, $\pi_{PE}(\theta)$:
+Since the event parameters are not perfectly measured, we incorporate uncertainty to marginalize over the possible parameters associated with each event.
+Practically, this is done by a Monte Carlo average over each event's posterior samples $\{\theta_j\}_i$, dividing out the prior used when inferring those posterior samples, $\pi_{\rm PE}(\theta)$:
 \begin{equation}
-    \frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda) = \frac{1}{N_{\text{samples}}}\sum_j^{N_{\text{samples}}} \frac{\diff N}{\diff t_{\det} \diff \theta_{i,j}} (\theta_{i,j};\Lambda) \frac{1}{\pi_{PE}(\theta_{i,j})} .
+\begin{aligned}
+    &p(\{d\}|\Lambda) \\
+    &\hspace{3mm} \propto e^{-N_{\text{exp}}(\Lambda)}\prod_i^{N_{\text{ev}}} \int 
+    d\theta_i\, p(d_i|\theta_i) \frac{\diff N}{\diff t_{\det} \diff \theta} (\theta_i;\Lambda) \\
+     &\hspace{3mm} \approx e^{-N_{\text{exp}}(\Lambda)}\prod_i^{N_{\text{ev}}} \frac{1}{N_{\rm samps}} \sum_{j=1}^{N_{\rm samps}} \frac{\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta_{j,i};\Lambda)}{\pi_{\rm PE}(\theta_{j,i})}\,.
+\end{aligned}
 \label{eq:single-event-likelihood}
 \end{equation}
-Finally, $p(\Lambda)$ is the prior on the population and parameters on the background cosmology, which we discuss in further detail below.
-We infer the posterior in Equation~\ref{eq:inhomog-poisson} using the no-u-turn sampler for Hamiltonian Monte Carlo within \texttt{numpyro} \citep{hoffman_no-u-turn_2011, bingham_pyro_2019,phan_composable_2019}.
+This provides a posterior on both the compact binary population and the background cosmology when combined with a prior $p(\Lambda)$ on the population and cosmological parameters, as discussed below.
+We will use the term ``hyper-prior'' to refer to $p(\Lambda)$ for the remainder of this work.
 
-For our population model, we focus on the masses and redshifts, neglecting spins as they do not contain cosmological information.
-For simplicity, we additionally assume that the population has equal masses, $m_1=m_2$.
-Thus, $\theta = \{m_1,z\}$. 
-The population can be modeled with different levels of complexity and flexibility. 
-As described in detail in Section~\ref{sec:model}, we model the distribution of primary masses in an agnostic way, using a \acl{GP}. 
-We also compare this non-parametric reconstruction with different parametric models.  
-For all models considered in this work, we assume a uniform-in-comoving volume redshift distribution, as in Equation~\ref{eq:underlying redshift dist}.
-
-The spectral siren method of measuring cosmological parameters relies on the assumption that either all compact binaries follow the same mass distribution at all redshifts, or that any redshift evolution of the mass distribution does  not exactly mimic cosmology (i.e.
-all features cannot exhibit a monotonic rightward shift with redshift) \citep{ezquiaga_spectral_2022}.
-For the redshift range considered in this work, we assume the mass distribution has no redshift evolution.
-The likelihood in Equation~\ref{eq:inhomog-poisson} is therefore maximized when the relationship between redshift and luminosity distance makes all source frame masses follow the same distribution.
+The likelihood in Equation~\eqref{eq:inhomog-poisson} is maximized when the relationship between redshift and luminosity distance makes all source frame masses follow the same distribution.
 This is the basis of the spectral siren method of measuring cosmological parameters; we can simultaneously infer the relationship between redshift and luminosity distance with the parameters of the source frame mass distribution.
-The full set of hyper-parameters therefore includes the cosmological parameters that dictate the $D_L$-$z$ relation: the local expansion rate \Ho, the present fractional energy density of dark matter \Omm, dark energy $\Omega_\Lambda$ and radiation $\Omega_r$, and the equation of state of dark energy $w$.
-In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and $w=-1$ and use uniform priors on \Ho{} and \Omm{}. 
-This corresponds to a flat $\Lambda$CDM cosmology.
+The full set of hyper-parameters therefore includes the cosmological parameters that dictate the $D_L$--$z$ relation: the local expansion rate \Ho, the present fractional energy densities of dark matter \Omm, dark energy $\Omega_\Lambda$ and radiation $\Omega_r$, and the equation of state of dark energy $w$.
+In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and $w=-1$ and use uniform priors on \Ho{} and \Omm{}, corresponding to a flat $\Lambda$CDM cosmology.
 
 Algorithmically, we evaluate the likelihood in Equation~\ref{eq:inhomog-poisson} by drawing \Ho{} and \Omm{} from their prior distributions (contained in $p(\Lambda)$), using them to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to $m_{\det} /(1+z(D_L,H_0,\Omega_M)) = m_{\source}$, evaluating the source-frame mass distribution at these transformed values ($\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ for events and $N_{\text{exp}}(\Lambda)$ for injections), and then taking the product of these terms to evaluate the expression in Equation~\ref{eq:inhomog-poisson}.
-The full process is outlined in Section~\ref{sec:model}.
+The full process is outlined in the next subsection.
 
+%%  Wrap up with discussion of GP priors and simplifying assumptions.
 \subsection{\Acl{GP}-based mass distribution}
@@
 
-Our goal is to model the population of \ac{GW} in an astrophysically-agnostic way. 
-There exist several non-parametric methods developed for this purpose \citep{tiwari_vamana_2021,edelman_aint_2022,sadiq_flexible_2022,edelman_cover_2023,mandel_extracting_2019,ray_non-parametric_2023}.
-While well-suited to infer the \ac{GW} source population with a fixed cosmology, several of these methods employ fixed features in source frame mass, such as bin edges \citep{mandel_extracting_2019,ray_non-parametric_2023} or spline nodes \citep{edelman_aint_2022}.
-Since these locations were chosen with a fixed cosmology, they risk causing the inference to prefer the cosmological parameters used when choosing the feature locations. 
-Indeed, \citet{mastrogiovanni_importance_2021} show how using fixed features can significantly bias cosmological inference within the spectral siren methodology.
-We therefore opt for a model of the source frame mass distribution that foregoes the need to define such features.
-
-For this purpose, we construct a model with a \acf{GP}, a common tool for non-parametric inference. 
-\Acp{GP} are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
-Their smoothness properties make them widely useful in \ac{GW} data analysis and beyond for regression problems, such as modeling time-domain waveforms \citep{doctor_statistical_2017, huerta_eccentric_2018} and the neutron star equation of state \citep{landry_nonparametric_2019}, density estimation problems, such as estimating posterior densities of single-event parameters from parameter estimation samples ~\citep{demilio_density_2021}, and as a prior on histogram bin heights for population inference \citep{mandel_model-independent_2017, li_flexible_2021, ray_non-parametric_2023}.
-
-Our use case is slightly different from previous analyses:
-we utilize a \ac{GP} as a prior on the functions that describe the primary mass distribution of \acp{CBC}.
-It will therefore replace $p(\Lambda)$ in Equation~\ref{eq:inhomog-poisson}, except in the case of cosmological parameters, which all have uniform priors.
-This choice adds very little prior information about the shape of the mass distribution, besides enforcing that it must be smooth.
-Practically, the only difference in the inference of the population when using a GP versus other modeling choices is that $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is determined directly by a realization of the \ac{GP}, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
+Practically, the only difference in the inference of the population when using a \ac{GP} versus other modeling choices is that $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is determined directly by a realization of the \ac{GP}, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
 In other words, when using parametric models, $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is calculated by evaluating a specific functional form described by a small set of hyper-parameters.
 With the \ac{GP} approach, the hyper-parameters describing the mass distribution are the rate at each event-level posterior sample's source frame mass, and the rate at each found injection's source frame mass.
+The \ac{GP} \emph{is} $p(\Lambda)$, the prior on population parameters (except in the case of cosmological parameters, which all have uniform priors).
+
 % The source frame mass is different for each possible value of \Ho{} and \Omm{}, so the hyper-parameters are defined in different locations for each draw from the hyper-posterior.
@@
 Indeed, an autoregressive process is a \ac{GP} with a particular choice of kernel (described below).
-We therefore have $N_{\text{evs}}M + N_{\text{inj}}$ mass hyper-parameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of posterior samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
+We therefore have $N_{\text{ev}}M + N_{\text{inj}}$ mass hyper-parameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of posterior samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
 We label these $\{\mathcal{R}_{ij},\mathcal{R}_k\}$, and our full set of hyper-parameters is $\Lambda=\{H_0, \Omega_M, \mathcal{R}_{ij},\mathcal{R}_k\}$. 
 As the hyper-parameters differ at each data point, we have successfully fulfilled our goal to construct a model without pre-defined features in source frame mass.
-We note that other non-parametric methods may be adapted to avoid such pre-defined features, such as fitting for the locations of their features simultaneously the rest of the inference \citep[e.g.][]{tiwari_vamana_2021} or by using an appropriate smoothing \citep[e.g.][]{edelman_cover_2023}.
+We note that other non-parametric methods may be adapted to avoid such pre-defined features, such as fitting for the locations of their features simultaneously with the rest of the inference \citep[e.g.][]{tiwari_vamana_2021} or by using an appropriate smoothing \citep[e.g.][]{edelman_cover_2023}.
 Such adaptations are necessary to avoid circular inference.
 
-The smoothness of a given \acp{GP} is determined by its kernel, which is a function that defines the covariance between input points in the \ac{GP} (in our case, two source frame mass values). 
+The smoothness of a given \ac{GP} is determined by its kernel, which is a function that defines the covariance between input points in the \ac{GP} (in our case, two source frame mass values). 
 It defines the notion of similarity between adjacent points and thereby encodes our assumptions about the smoothness of the source frame mass distribution \citep{rasmussen_gaussian_2006}.
 We use a Mat\'ern kernel \citep{handcock_bayesian_1993, stein_interpolation_1999} with $\nu = 5/2$, but have repeated the analysis with $\nu=3/2$ and $\infty$, finding little impact on the results, except that the $\nu=\infty$ case (also called the ``squared exponential'' kernel) produces a slightly more jagged mass distribution.
-Mat\'ern kernels have two parameters besides the mean that determine their properties: a length scale $l$ and variance $s$.
+In addition to the mean, Mat\'ern kernels have two parameters that determine their properties: a length scale $l$ and a variance $s$.
 In our use case, these are one level further removed from hyper-parameters, so we adopt the terminology used in \citet{callister_parameter-free_2023} and call them ``hyper-hyper-parameters.''
@@
 
-Algorithmically, each likelihood evaluation contains the following steps: \variable{output/priors_placeholder.txt}
+Algorithmically, each posterior evaluation contains the following steps: \variable{output/priors_placeholder.txt}
 \begin{enumerate}
@@
 % TODO: \jme{[comment: is there a way to summarize this list in a simple diagram?]}
+We perform these steps within \texttt{numpyro} \citep{bingham_pyro_2019,phan_composable_2019}, sampling the posterior using the no-u-turn sampler for Hamiltonian Monte Carlo \citep{hoffman_no-u-turn_2011}.
 
@@
 We then demonstrate that our flexible model alleviates this bias without the need to know the mass distribution's morphology \emph{a priori}.
-To illustrate this, we generate a catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{abbott_population_2023}.
-This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after one year of observation in \ac{O5}.
-The details of the data simulation are described in Appendix~\ref{ap:data generation}.
-We then use this catalog to infer the cosmic expansion rate, using three different models for the source-frame mass distribution:
+To illustrate this, we use three different models for the source-frame mass distribution to infer the cosmic expansion rate from the simulated catalog described in Section~\ref{sec:data generation} and Appendix~\ref{ap:data generation}.
+The three models are as follows:
 \begin{enumerate}
-    \item the \plp{} model, which includes the true mass distribution within its hyperprior,
-    \item a broken power law, which does not include the true mass distribution within its hyperprior, and 
+    \item the \plp{} model, which includes the true mass distribution within its hyper-prior,
+    \item a broken power law\footnote{This is not the same as the \textsc{Broken Power Law} model presented in \citet{abbott_population_2021}, as we do not employ a high-mass truncation.}, which does not include the true mass distribution within its hyper-prior, and
     \item the flexible, \ac{GP}-based model described in Section~\ref{sec:model}, which is able to closely approximate the morphology of the true mass distribution, along with many other morphologies.
 \end{enumerate}
+For all models considered in this work, we assume the redshift distribution described in Equation~\ref{eq:underlying redshift dist}.
 The results of each of these fits are shown in Figure~\ref{fig:O5_GP}. 
 The left panel shows the inferred source frame mass distribution for each of the considered models, and the right panel shows the corresponding posteriors on \Ho{}. 
-We indicate the true underlying source mass distribution and fiducial \Ho{} value with solid black lines in each panel.
+We indicate the true underlying source mass distribution and \Ho{} value with solid black lines in each panel.
 
@@
 Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 100$.
-This means that future \acp{GW} observations will be more sensitive $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters \citep{Chen:2024gdn}.
-We demonstrate this by repeating the same \ac{GP}-based spectral siren analysis of the previous subsection while also fitting for the local matter density, \Omm, but we emphasize that it would also have been possible to constrain other cosmological parameters governing $H(z)$.
+This means that future \acp{GW} observations will be more sensitive to $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters \citep{Chen:2024gdn}.
+We demonstrate this by repeating the same \ac{GP}-based spectral siren analysis of the previous subsection while also fitting for the local matter density, \Omm.
+However, we emphasize that it would also have been possible to constrain other cosmological parameters governing $H(z)$.
 The result is shown in Figure~\ref{fig:O5_corner}.
 
-We find \ac{O5} observations to be most sensitive to $H(z=\variable{output/mostsensitivez.txt})$, which results in similarly informative constraints on \Omm{} and \Ho.
-The left panel of Figure~\ref{fig:O5_corner} demonstrates a strong anti-correlation between the \Omm{} and \Ho{} posteriors. 
+We find \ac{O5} observations to be most sensitive to $H(z=\variable{output/mostsensitivez.txt})$, which is measured at $\variable{output/Hz_percent.txt}\%$.
+% For comparison, the Dark Energy Spectroscopic Instrument is projected to measure $H(z=1)$ to $\sim1\%$ \citep{aghamousa_desi_2016} by 2026 \citep{schlegel_spectroscopic_2022}, 1 year before the projected start of \ac{O5} \citep{emfollow_user_guide}.
+The left panel of Figure~\ref{fig:O5_corner} demonstrates a strong anti-correlation between the \Omm{} and \Ho{} posteriors, resulting in similarly informative constraints on the two parameters.
 This is because \Ho{} controls the $y$-intercept of the $H(z)$ curves on the right panel, while \Omm{} informs the slope of those curves.
 The same measurement of $H(z\neq0)$ can be obtained by increasing the slope while decreasing the $y$-intercept, and vice versa.
-Similar behavior can be observed in measurements of the \ac{BBH} redshift distribution, which exhibit a ``waist'' at $z\sim0.2$ with current observations \citep{abbott_population_2023, callister_parameter-free_2023}.
+Similar behavior can be observed in current measurements of the \ac{BBH} redshift distribution, which exhibit a ``waist'' at $z\sim0.2$ with current observations \citep{abbott_population_2023, callister_parameter-free_2023}.
 
@@
 
+Spectral sirens are particularly useful in the context of the current ``Hubble tension'': a disagreement between multiple methods of measuring the local expansion rate of the universe with electromagnetic observations \citep{freedman_measurements_2021}.
+As this tension can only be explained by non-standard physics or yet-unknown systematic uncertainties in either the cosmic distance ladder or cosmic microwave background measurements, direct probes of the local expansion that do not rely on electromagnetic observations will help determine whether new physics need be invoked to resolve the current discrepancy.
+
 We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} after one year of observing with the \ac{LVK} network's design sensitivity using non-parametric spectral sirens and a $\variable{output/PLPh0percent.txt}\%$
@@
 Thus, the additional statistical uncertainty introduced by non-parametric approaches may be preferable to the systematic errors associated with choosing a parametric model.
+
+In parallel to this work, \citet{magana_inprep} performed a spectral siren analysis on currently-public \ac{LVK} data using a model that employs histogram bins defined at fixed locations in source-frame mass to flexibly measure the \ac{BBH} mass distribution \citep[originally presented in][]{mandel_model-independent_2017,ray_non-parametric_2023}.
+While both the method presented in this work and that of \citet{magana_inprep} employ \acp{GP}, theirs uses it to modulate the histogram bin heights, leading to a piecewise constant function in log mass, whereas ours defines the \ac{GP} at each data point, constructing a smooth function with no pre-defined features in source-frame mass.
+% Sharp bin edges may lead to optimistic constraints on cosmological parameters, especially as such features are unlikely to exist in the true mass distribution, possibly explaining the increase in precision relative to parametric methods seen in \citet{magana_inprep}.
 
@@
 This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}, which can only be detected relatively locally.
-\comment{Projected DESI errors are $\sim3\%$ at $z=1$, which are comparable with our estimated error of $\variable{output/Hz_percent.txt}\%$, but those won't be realized until 202X.}
 This means that upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances by \acp{GW}. 
@@
 \label{ap:data generation}
-
-We generate a mock catalog of \ac{GW} events characteristic of that expected from one year of \acf{O5}. 
-These events are drawn from an underlying population described by the \plp{} mass distribution presented in \citet{talbot_measuring_2018} and used in \citet{abbott_binary_2019, abbott_population_2023} and the redshift distribution presented in \citet{callister_shouts_2020}.
+The exact form of the injected population is
 \begin{equation}
@@
 These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{abbott_population_2023}.
-We do not fit for spins, as they do not carry additional cosmological information, and additionally neglect secondary masses.
+For simplicity, we additionally assume that the population has equal masses, $m_1=m_2$, and thus, $\theta = \{m_1,z\}$.
+We consistently apply this assumption to the data generation process and the population inference.
+We do not fit for or simulate spins, as they do not redshift and hence do tnot carry additional cosmological information, and additionally neglect secondary masses.
+
 We use cosmological parameters \Ho$=\variable{output/H0_FID.txt}\Hunits$, \Omm$=0.3$, $\Omega_\Lambda=1-$\Omm, consistent with those found by \citet{planck_collaboration_planck_2016}.
@@
 
-\section{Old}
-Old tex that we may end up needing.
+% \section{Old}
+% Old tex that we may end up needing.
 
-\begin{table}[]
-    \centering
-    \begin{tabular}{c|c c}
-         Model & $\sigma_{H_0}$ [km/s/Mpc] & bias [km/s/Mpc]\\
-         \hline
-         & 
-    \end{tabular}
-    \caption{Performance of various mass distribution models used for measuring the Hubble constant, \Ho.
-    We consider the correct parametric model, a deliberately incorrect parametric model, and the non-parametric, \ac{GP}-based model presented in this work.
-    Bias is defined as the number of standard deviations between the true injected value of \Ho and the recovered posterior mean. 
-    The parametric models yield more precise constraints on \Ho{}, as shown by smaller standard deviations ($\sigma_{H_0}$), but the incorrect parametric model is less accurate than the non-parametric model.
-    In reality, it is impossible to know the true functional form of the mass distribution, so the additional statistical uncertainty introduced by non-parametric approaches may be desirable over the systematic errors associated with choosing a parametric model.
+% \begin{table}[]
+%     \centering
+%     \begin{tabular}{c|c c}
+%          Model & $\sigma_{H_0}$ [km/s/Mpc] & bias [km/s/Mpc]\\
+%          \hline
+%          & 
+%     \end{tabular}
+%     \caption{Performance of various mass distribution models used for measuring the Hubble constant, \Ho.
+%     We consider the correct parametric model, a deliberately incorrect parametric model, and the non-parametric, \ac{GP}-based model presented in this work.
+%     Bias is defined as the number of standard deviations between the true injected value of \Ho and the recovered posterior mean. 
+%     The parametric models yield more precise constraints on \Ho{}, as shown by smaller standard deviations ($\sigma_{H_0}$), but the incorrect parametric model is less accurate than the non-parametric model.
+%     In reality, it is impossible to know the true functional form of the mass distribution, so the additional statistical uncertainty introduced by non-parametric approaches may be desirable over the systematic errors associated with choosing a parametric model.
+%     }
+%     \label{tab:bias}
+% \end{table}
+
+% \subsection{Option 2: GP only}
+% We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated \ac{GW} catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{abbott_population_2023}.
+% This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after \ac{O5}.
+% The details of the data simulation are described in Appendix~\ref{ap:data generation}.
+
+% The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
+% The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho{} ($\variable{output/H0_FID.txt}\Hunits$) is consistent with its inferred posterior (\variable{output/nonparh0CI.txt}$\Hunits$).
+% At very high and very low masses, there is both a lack of data and little detector sensitivity.
+% The mass distribution is therefore uninformed in this region and the \ac{GP} reverts to its prior distribution, which resembles random scatter around the mean differential merger rate.
+% Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
+% On the other hand, the lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the GP to learn a relatively low merger rate at the edges of the mass distribution.
+% Indeed, these features are very informative to the \Ho{} constraint \citep{gwtc3_cosmo}.
+% The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
+% Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,120\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
+
+% \subsection{Old intro text}
+% Perhaps because of the analogy to electromagnetic spectra, in which the locations of source-frame spectral features are often known from first principles, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori} to perform a spectral siren redshift measurement.
+% This would certainly be problematic, as mismodeling of the specific features would hinder the accuracy of the method and introduce biases in the cosmological inference \citep{Mukherjee:2021rtw,mastrogiovanni_importance_2021,pierra_study_2023}.
+% Such systematic bias is illustrated in Figure~\ref{fig:O5_GP}, where we show the result of fitting simulated data with a model that has the true mass distribution contained within its prior and one that does not.
+% The latter does not accurately infer the true value of \Ho{}, whereas the former does.
+
+% This is particularly troubling in light of the fact that there remains much uncertainty over the cause, location, and robustness of currently-identified features in the \ac{GW} source mass distribution.
+% While several features have been identified in the current data \citep{tiwari_exploring_2022,edelman_aint_2022, abbott_population_2023,edelman_cover_2023}, including three overdensities, two underdensities, a maximum and minimum black hole mass, only three of these are statistically robust \citep{farah_things_2023}.
+% Additionally, the morphology of the mass distribution is a result of the interplay between several complex physical processes pertaining to massive stars, such as binary stellar evolution, pair instability, globular cluster dynamics, and stellar winds, each of which has large theoretical uncertainties \citep[e.g.][]{ zevin_constraining_2017, mapelli_binary_2020,marchant_evolution_2023}.
+% These large theoretical uncertainties mean that accurate predictions for the shape of the \ac{GW} source mass distribution are a considerable challenge.
+%If such \emph{a priori} knowledge of the mass distribution's morphology were requisite for a spectral siren method, the effectiveness of such a measurement would be significantly hampered.
+
+% However, the information in the spectral siren measurement comes solely from the assumption that all \acp{CBC} follow the same mass distribution.\footnote{More precisely, the spectral siren method is also valid if \acp{CBC} have different mass distributions at different redshifts, so long as the evolution of the mass distribution does not exactly mimic cosmology, and the model for the mass distribution allows for evolution \citep{ezquiaga_spectral_2022}.
 }
-    \label{tab:bias}
-\end{table}
-
-\subsection{Option 2: GP only}
-We fit the non-parametric population model described in Section~\ref{sec:model} to a simulated \ac{GW} catalog of $\variable{output/num_found_events.txt}$ \ac{GW} events from an underlying mass distribution described by the \plp{} model, with hyperparameters consistent with those found in \citet{abbott_population_2023}.
-This simulated catalog is designed to be representative of the \ac{GW}  transient catalog after \ac{O5}.
-The details of the data simulation are described in Appendix~\ref{ap:data generation}.
-
-The resulting inferred mass distribution is shown in Figure~\ref{fig:O5_GP}, along with the corresponding posterior on \Ho.
-The inferred mass distribution closely resembles the true, simulated distribution, and the injected value of \Ho{} ($\variable{output/H0_FID.txt}\Hunits$) is consistent with its inferred posterior (\variable{output/nonparh0CI.txt}$\Hunits$).
-At very high and very low masses, there is both a lack of data and little detector sensitivity.
-The mass distribution is therefore uninformed in this region and the \ac{GP} reverts to its prior distribution, which resembles random scatter around the mean differential merger rate.
-Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
-On the other hand, the lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the GP to learn a relatively low merger rate at the edges of the mass distribution.
-Indeed, these features are very informative to the \Ho{} constraint \citep{gwtc3_cosmo}.
-The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but these features are uninformative and do not affect inference on cosmological parameters.
-Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,120\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
+% Therefore the location, shape, and existence of all features in the mass distribution can be inferred simultaneously with $H(z)$.
 

### 2024-03-26 src/scripts/submit.sh
hack to use old runs/figs

@@ -1,7 +1,7 @@
 #!/bin/bash
-#SBATCH --job-name=paper_fitOm0_gwmc
-#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_gwmc.out
-#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_fitOm0_gwmc.err
+#SBATCH --job-name=paper_parametric
+#SBATCH --output=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_parametric.out
+#SBATCH --error=/home/afarah/p2/spectral-sirens-with-GPs/src/outfiles/paper_parametric.err
 #SBATCH --account=kicp
 #SBATCH --time=48:00:00
 #SBATCH --partition=kicp-gpu
@@ -15,4 +15,4 @@ module load python/anaconda-2022.05
 echo "loaded anaconda"
 source activate jax_gpu
 echo "activated env"
-python /home/afarah/p2/spectral-sirens-with-GPs/src/scripts/nonparametric_inference_fitOm0.py
+python /home/afarah/p2/spectral-sirens-with-GPs/src/scripts/parametric_inference.py

### 2024-03-26 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 % Intro to GPs
-For this purpose, we construct a model for the with a \acf{GP}, a common tool for non-parametric inference. 
+For this purpose, we construct a model with a \acf{GP}, a common tool for non-parametric inference. 
 \Acp{GP} are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
@@
     \item the \plp{} model, which includes the true mass distribution within its hyper-prior,
-    \item a broken power law\footnote{This is not the same as the \textsc{Broken Power Law} model presented in \citet{abbott_population_2021}, as we do not employ a high-mass truncation.}, which does not include the true mass distribution within its hyper-prior, and
+    \item a broken power law\footnote{This differs from the \textsc{Broken Power Law} model presented in \citet{abbott_population_2021}, as we do not employ a high-mass truncation.}, which does not include the true mass distribution within its hyper-prior, and
     \item the flexible, \ac{GP}-based model described in Section~\ref{sec:model}, which is able to closely approximate the morphology of the true mass distribution, along with many other morphologies.
@@
 We have repeated this analysis with 50 separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time,
-and the \plp{} model reaches the same level of bias $\variable{output/PLP_bias_percent.txt}\%$ 
+and the \plp{} model reaches the same level of bias only $\variable{output/PLP_bias_percent.txt}\%$ 
 of the time.
@@
 measurement with parametric spectral sirens, demonstrating an increase in precision when using the correct parametric model.
+\comment{These numbers are placeholders for now and will be updated in the next version.}
 However, the true population of merging compact binaries in the Universe is unknown because neither the mechanism from which compact binaries were formed, nor the physical processes within that mechanism has been conclusively determined.

### 2024-03-26 src/tex/ms.tex
[showyourwork] overleaf sync

@@
     \includegraphics[width=\textwidth]{figures/O5_pm.pdf}
-    \caption{Spectral siren measurement for an \ac{O5}-like catalog with the correct parametric model (\plp, green), a deliberately incorrect parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{GP}, blue).
+    \caption{Spectral siren measurement for a simulated catalog with the correct parametric model (\plp, green), a deliberately incorrect parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{GP}, blue).
     The left panel shows the recovered source frame primary mass distribution for each model, and the simultaneously-inferred posteriors on \Ho{} are shown in the right panel.
@@
 \begin{enumerate}
-    \item the \plp{} model, which includes the true mass distribution within its hyper-prior,
+    \item \plp{}, which includes the true mass distribution within its hyper-prior,
     \item a broken power law\footnote{This differs from the \textsc{Broken Power Law} model presented in \citet{abbott_population_2021}, as we do not employ a high-mass truncation.}, which does not include the true mass distribution within its hyper-prior, and
@@
 measurement with parametric spectral sirens, demonstrating an increase in precision when using the correct parametric model.
-\comment{These numbers are placeholders for now and will be updated in the next version.}
+\comment{[These numbers are placeholders for now and will be updated in the next version.]}
 However, the true population of merging compact binaries in the Universe is unknown because neither the mechanism from which compact binaries were formed, nor the physical processes within that mechanism has been conclusively determined.

### 2024-03-26 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{githubLink}, and was enabled by the \showyourwork package \citep{Luger2021}.
+The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
 

### 2024-03-31 src/scripts/pm_H0_twopanel.py
turn traces into 90% CIs

@@ -12,6 +12,11 @@ import paths
 
 plt.style.use(paths.scripts / "matplotlibrc")
 
+color_GP = "#1f78b4"
+color_GP_light = "#a6cee3"
+color_PLP = "#33a02c"
+color_BPL = "#fb9a99"
+
 def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     id = az.InferenceData.from_netcdf(path)
     samples = id.posterior
@@ -26,10 +31,19 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     r_BPL = np.nan_to_num(np.exp(samples_BPL['log_rate']))
 
     fig, axes = plt.subplots(ncols=2,figsize=(7.5,4*.75),facecolor='none',gridspec_kw={'width_ratios': [1.2, 1]})
-    for i in range(NSAMPS//2):
-        axes[0].plot(TEST_M1S, r[i]/TEST_M1S, lw=0.1, c="blue",alpha=0.03)
-        axes[0].plot(TEST_M1S, r_PLP[i],lw=0.1, c="green",alpha=0.03)
-        axes[0].plot(TEST_M1S, r_BPL[i],lw=0.1, c="orange",alpha=0.03)
+    # for i in range(NSAMPS//2):
+    #     axes[0].plot(TEST_M1S, r[i]/TEST_M1S, lw=0.1, c=color_GP,alpha=0.03)
+    #     axes[0].plot(TEST_M1S, r_PLP[i],lw=0.1, c=color_PLP,alpha=0.03)
+    #     axes[0].plot(TEST_M1S, r_BPL[i],lw=0.1, c=color_BPL,alpha=0.03)
+    axes[0].plot(TEST_M1S,np.percentile(r/TEST_M1S,(5,95),axis=0).T,lw=0.2, c=color_GP,alpha=0.5)
+    # axes[0].plot(TEST_M1S,np.mean(r/TEST_M1S,axis=0),lw=1.5, c=color_GP)
+    axes[0].fill_between(TEST_M1S,*np.percentile(r/TEST_M1S,(5,95),axis=0), color=color_GP_light,alpha=0.3)
+    
+    axes[0].plot(TEST_M1S,np.percentile(r_PLP,(5,95),axis=0).T, c=color_PLP,ls="--")
+    # axes[0].plot(TEST_M1S,np.mean(r_PLP,axis=0),lw=2, c=color_PLP,ls="--")
[…]

### 2024-03-31 src/scripts/nonparametric_inference.py
play around w redshift distribution fitting

@@ -15,7 +15,7 @@ jax.config.update("jax_enable_x64", True)
 NSAMPS=1000
 
 # random number generators
-jax_rng = jax.random.PRNGKey(425)
+jax_rng = jax.random.PRNGKey(42)
 
 # options
 remove_low_Neff=False
@@ -48,7 +48,7 @@ if  __name__ == "__main__":
 
     # save results
     id = az.from_numpyro(mcmc)
-    id.to_netcdf(paths.data / "mcmc_nonparametric.nc4")
+    id.to_netcdf(paths.data / "mcmc_nonparametric_fitz.nc4")
 
     # calculate stats
     h0samps = id.posterior['H0'][0]

### 2024-03-31 src/scripts/parametric_inference.py

@@ -8,7 +8,6 @@ from data_generation import H0_FID
 
 # inference
 import jax
-import jax.numpy as jnp
 import numpyro
 import arviz as az
 
@@ -46,7 +45,7 @@ if  __name__ == "__main__":
 
     # save results
     id = az.from_numpyro(mcmc)
-    id.to_netcdf(paths.data / "mcmc_parametric_PLP.nc4")
+    id.to_netcdf(paths.data / "mcmc_parametric_PLP_fitz.nc4")
     offset = np.abs(id.posterior['H0'][0].mean()-H0_FID)/id.posterior['H0'][0].std()
     with open(paths.output / "PLPh0offset.txt","w") as f:
         print(f"{offset:.1f}",file=f)
@@ -65,7 +64,7 @@ if  __name__ == "__main__":
 
     # save results
     id = az.from_numpyro(mcmc)
-    id.to_netcdf(paths.data / "mcmc_parametric_BPL.nc4")
+    id.to_netcdf(paths.data / "mcmc_parametric_BPL_fitz.nc4")
     offset = np.abs(id.posterior['H0'][0].mean()-H0_FID)/id.posterior['H0'][0].std()
     with open(paths.output / "BPLh0offset.txt","w") as f:
         print(f"{offset:.1f}",file=f)

### 2024-03-31 src/scripts/priors.py

@@ -131,9 +131,9 @@ def hyper_prior(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,PC_params=dic
         Om0=OM0_FID
 
     # redshift dist
-    alpha_z = 1.0
-    zp = 2.4 
-    beta_z = 3.4 
+    alpha_z = numpyro.sample("alpha_z",dist.Uniform(0,2)) #TRUEVALS['alpha_z']#
+    zp = TRUEVALS['zp']#numpyro.sample("zp",dist.Uniform(0,5)) #
+    beta_z =  TRUEVALS['beta_z'] # numpyro.deterministic('beta',TRUEVALS['beta_z'])# numpyro.sample("beta_z",dist.Uniform(0,5)) #
 
     # construct GP in the source frame
     logtestm1s = jnp.log(TEST_M1S)
@@ -211,9 +211,9 @@ def PLP(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,remove_low_Neff=False
     sig_m1 = numpyro.sample("sig_m1",dist.Uniform(1,10))# TRUEVALS['sig_m1']
     f_peak = TRUEVALS['f_peak'] #numpyro.sample("f_peak",dist.Uniform(0,1))
     #Fixed
-    alpha_z = TRUEVALS['alpha_z']#numpyro.sample("alpha_z",dist.Uniform(0,2))
+    alpha_z = numpyro.sample("alpha_z",dist.Uniform(0,2)) #TRUEVALS['alpha_z']#
     zp = TRUEVALS['zp']#numpyro.sample("zp",dist.Uniform(0,5)) #
-    beta = TRUEVALS['beta_z'] #numpyro.sample("beta_z",dist.Uniform(0,5)) #numpyro.deterministic('beta',TRUEVALS['beta_z'])# 
+    beta = TRUEVALS['beta_z'] #numpyro.sample("beta_z",dist.Uniform(0,5)) # numpyro.deterministic('beta',TRUEVALS['beta_z'])# 
 
     # convert event data to source frame
     z = jgwcosmo.z_at_dl_approx(dL,H0,Om0,zmin=ZMIN,zmax=ZMAX+8.)
@@ -286,9 +286,9 @@ def BPL(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,remove_low_Neff=False
     
     #Fixed
[…]

### 2024-03-31 src/scripts/bias_study.py
fix typo

@@ -53,7 +53,7 @@ for i in trange(N_CATALOGS):
     mcmc.run(jax_rng,**kwargs)
 
     id = az.from_numpyro(mcmc)
-    id.to_netcdf(paths.data / f"bias/mcmc_parametric_PLP_{int}.nc4")
+    id.to_netcdf(paths.data / f"bias/mcmc_parametric_PLP_{i}.nc4")
     bias_PLP[i] = (id.posterior['H0'][0].mean() - H0_FID)/id.posterior['H0'][0].std()
     if plot:
         plt.hist(id.posterior['H0'][0],density=True, bins=50,histtype='step',color='green',alpha=0.5,lw=0.5)
@@ -68,7 +68,7 @@ for i in trange(N_CATALOGS):
     mcmc.run(jax_rng,**kwargs)
 
     id = az.from_numpyro(mcmc)
-    id.to_netcdf(paths.data / f"bias/mcmc_parametric_BPL_{int}.nc4")
+    id.to_netcdf(paths.data / f"bias/mcmc_parametric_BPL_{i}.nc4")
     bias_BPL[i] = (id.posterior['H0'][0].mean() - H0_FID)/id.posterior['H0'][0].std()
     if plot:
         plt.hist(id.posterior['H0'][0],density=True, bins=50,histtype='step',color='orange',alpha=0.5,lw=0.5)

### 2024-03-31 src/scripts/data_generation.py
clean up, bugfix

@@ -5,7 +5,6 @@ Clight = gwcosmo.Clight
 from utils import inverse_transform_sample
 from mock_posteriors import gen_snr_scaled_PE
 
-import os
 import numpy as np
 from GWMockCat.vt_utils import draw_thetas, interpolate_optimal_snr_grid 
 
@@ -32,7 +31,7 @@ NUM_INJ=N_SOURCES*50*30
 perf_meas = False
 
 # random number generators
-np_rng = np.random.default_rng(516)
+np_rng = np.random.default_rng(520)
 
 def true_vals_PLP(rng, n_sources=N_SOURCES):
     m1s_true = inverse_transform_sample(gwpop.powerlaw_peak,[1,400],rng,N=n_sources,
@@ -70,7 +69,7 @@ osnr_interp, reference_distance = interpolate_optimal_snr_grid(
 if  __name__ == "__main__":
     # generate data and save 
     m1s_true, zt, m1z_true, dL_true = true_vals_PLP(rng=np_rng)
-    os.mkdir(paths.data / "gw_data")
+    # os.mkdir(paths.data / "gw_data")
     np.save(paths.data / "gw_data/m1s_true_PLP.npy", dL_true)
     np.save(paths.data / "gw_data/z_true_PLP.npy", zt)
     np.save(paths.data / "gw_data/m1z_true_PLP.npy",m1z_true)
@@ -104,7 +103,7 @@ if  __name__ == "__main__":
         m1z_PE = np.expand_dims(det_dict['m1'] * (1 + det_dict['z']),1)
         m2z_PE = np.expand_dims(det_dict['m2'] * (1 + det_dict['z']),1)
         dL_PE = np.expand_dims(det_dict['lum_dist'], 1)
-        log_PE_prior = 0
+        log_PE_prior = np.zeros((len(det_dict['m1']),1))
 
     # cut out samples below SNR interpolation range. 
     # It is very unlikely that any samples are down there, but we do this just in case.

### 2024-03-31 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 % Title
-\title{No need to know: astrophysics-agnostic gravitational wave cosmology}
-% \daniel{No need to know: astrophysically-agnostic gravitational wave cosmology\\
-% No need to know: gravitational-wave cosmology without astrophysics\\
-% No need to know: gravitational-wave cosmology without astrophysical assumptions}}
-% Ignorance is bliss: astrophysics-agnostic gravitational wave cosmology
-%\title{No need to know: astrophysics-agnostic spectral siren cosmology}
-%\title{No need to know: spectral siren cosmology with an astrophysics-agnostic mass distribution of gravitational-wave sources}
-%\title{No need to know: gravitational-wave cosmology for the astrophysically-ignorant}
+\title{No need to know: astrophysics-free gravitational-wave cosmology}
+% Daniel : No need to know: gravitational-wave cosmology without astrophysics [I like this one better than the top one, since it's catchier/more direct, and will get more attention; but are people worried it's too strong?]}
+% Amanda: I put it to a poll and the current one got more votes! No one gave a reason as to why, though
+% Daniel: okay, FWIW i just asked a professional writer, and she votes for the 2nd one: "astrophysics-free GW cosmo".
+'I think the "astrophysics-free" is very sugarfree funny, and matches the No Need to Know wit well.'
+% Amanda: I like astrophysics-free! That wasn't in the poll I sent yesterday so I  will just take the liberty to assume that others are okay with it as well :p
+
+% Other ideas:
+% No need to know: gravitational-wave cosmology without astrophysical assumptions
+% No need to know: gravitational-wave cosmology for the astrophysically-ignorant
 
@@
 When paired with a redshift measurement, this enables standard-siren cosmology: a Hubble diagram can be constructed to directly probe the Universe's expansion.
-This can be done in the absence of electromagnetic measurements: features in the mass distribution of \ac{GW} sources can provide self-calibrating redshift measurements without the need for an accompanying electromagnetic transient or catalog of potential host galaxies.
-% So far, this technique has only been applied with simple parametric representations of the mass distribution. 
-% However, the use of an incorrect mass model inevitably leads to biases in the cosmological inference, an acute problem given our current lack of understanding of the true source population.
-% Furthermore, i
-It is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori}\/ to obtain unbiased measurements of cosmological parameters in this fashion.
-We demonstrate that spectral sirens can accurately infer cosmological parameters without assuming
-prior knowledge of the \ac{GW} mass distribution.
-By adopting a flexible, Gaussian process-based model for the mass distribution of compact binaries, we find that we can accurately and agnostically infer the mass distribution while simultaneously measuring parameters governing cosmological expansion.
-We demonstrate this analysis using a simulated catalog of 1,000 gravitational-wave events consistent with expectations for the next \acl{LVK} observing run.
-We find that both the source mass model and cosmological parameters are correctly reconstructed, and predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{}, keeping all other cosmological parameters fixed, and a $\variable{output/Hz_percent.txt}\%$  measurement of $H(z=\variable{output/mostsensitivez.txt})$ when fitting for multiple cosmological parameters ($1\sigma$ uncertainties).
-This astrophysically-agnostic spectral siren technique will be essential to arrive at precise and unbiased cosmological constraints from gravitational-wave source populations.
+This can be done in the absence of electromagnetic measurements, as features in the mass distribution of \ac{GW} sources provide self-calibrating redshift measurements without the need for a definite (bright siren) or probabilistic (dark siren) host galaxy association.
+This ``spectral siren'' technique has thus far only been applied with simple parametric representations of the mass distribution. 
+However, the use of an inaccurate representation leads to biases in the cosmological inference, an acute problem given the current uncertainties in true source population.
+Furthermore, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori}\/ to obtain unbiased measurements of cosmological parameters in this fashion.
+Here, we demonstrate that spectral sirens can accurately infer cosmological parameters without such prior assumptions.
+We apply a flexible, non-parametric model for the mass distribution of compact binaries to a simulated catalog of 1,000 \ac{GW} signals, consistent with expectations for the next \acl{LVK} observing run.
+We find that, despite our model's flexibility, both the source mass model and cosmological parameters are correctly reconstructed.
+We predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{}, keeping all other cosmological parameters fixed, and a $\variable{output/Hz_percent.txt}\%$  measurement of $H(z=\variable{output/mostsensitivez.txt})$ when fitting for multiple cosmological parameters ($1\sigma$ uncertainties).
+This astrophysically-agnostic spectral siren technique will be essential to arrive at precise and unbiased cosmological constraints from \ac{GW} source populations.
 \end{abstract}
@@
 \label{sec:intro}
+Like light, \acp{GW} are redshifted as they propagate across the universe, thereby bearing imprints of the Universe's cosmic expansion history.
+Unlike light, however, the form of \ac{GW} signals are known from first principles, directly from the theory of general relativity.
+Furthermore, because \acp{GW} propagate across the Universe without attenuation from intervening matter, and because the properties of \ac{GW} detectors are well characterized, \ac{GW} selection effects are extremely well understood.
+% This is to be contrasted with many electromagnetic observations, where effects such as reddening, selection bias, and $k$-corrections can lead to significant complications.
+[too strong?]
+This allows for a precise estimate of each \ac{GW} catalogs' completeness and an unbiased measurement of the true \ac{GW} source population~\citep{2023PhRvX..13d1039A,abbott_population_2023,2023PhRvD.108d3011E}.
+Additionally, the \ac{GW} signals observed by the LIGO, Virgo, and KAGRA detectors~\citep{aasi_advanced_2015,acernese_advanced_2014,akutsu_overview_2021}  provide direct measurements of the distance to their sources.
+This makes them ``standard sirens'': direct probes of cosmological parameters that circumvent the need for a cosmological distance ladder~\citep{schutz_determining_1986,holz_using_2005}.
+
+A well-known demonstration of standard siren cosmology was the multi-messenger event GW170817 \citep{abbott_multi-messenger_2017,coulter_swope_2017,valenti_discovery_2017,2017ApJ...848L..27T}, whose clear association with a host galaxy provided a precise redshift measurement and allowed for a direct ``bright siren'' measurement of the Hubble constant, \Ho{} \citep{abbott_gravitational-wave_2017}.
+External redshift information can also come from galaxy catalogs, which provide an ensemble of possible redshifts for each \ac{GW} signal, allowing for a probabilistic ``dark siren'' measurement of \Ho{} when multiple \ac{GW} detections are combined \citep{del_pozzo_inference_2012, chen_two_2018, fishbach_standard_2019, soares-santos_first_2019,gray_cosmological_2020, abbott_gravitational-wave_2021, gwtc3_cosmo, gray_pixelated_2022,gray_joint_2023,mastrogiovanni_joint_2023, gair_hitchhikers_2023}.
+
+Electromagnetic information about \ac{GW} source redshifts need not be available in order to use them as standard sirens, however.
+\ac{GW} signals provide direct measurements of each source's luminosity distance, $d_L$, and redshifted (detector frame), masses, $m_{\det}= m_{\source}(1+z)$ \citep[e.g.][]{chen_mass-redshift_2019}.
+Therefore, if the source frame mass is known, each \ac{GW} signal provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the expansion of the universe at the time the \ac{GW} signal was emitted, $H(z)$.
 \begin{figure*}
@@
     The mass distribution and \Ho{} value used to generate the data are shown by a solid black line in each panel. 
-    The deliberately incorrect parametric model fails to recover the true mass distribution and therefore produces a biased estimate of \Ho{}, whereas both the correct and non-parametric models recover  the mass distribution and \Ho{}.
+    The deliberately incorrect parametric model fails to recover the true mass distribution and therefore produces an estimate of \Ho{} that is offset from the true value, whereas both the correct and non-parametric models recover the mass distribution and \Ho{}.
     As the true mass distribution is unknown for real observations, using a non-parametric model mitigates systematic uncertainty that would otherwise arise from mismodeling the \ac{CBC} population, though it does introduce additional statistical uncertainty.
@@
 
-Like light, \acp{GW} are redshifted as they propagate across the universe, thereby bearing imprints of the Universe's cosmic expansion history.
-Unlike light, however, the form of \ac{GW} signals are known from first principles, directly from the theory of general relativity.
-Furthermore, because \acp{GW} propagate across the Universe without attenuation from intervening matter, and because the properties of \ac{GW} detectors are well characterized, \ac{GW} selection effects are extremely well understood.% This is to be contrasted with many electromagnetic observations, where effects such as reddening, selection bias, and $k$-corrections can lead to significant complications.
-[too strong?]
-This allows for a precise estimate of each \ac{GW} catalogs' completeness and an unbiased measurement of the true \ac{GW} source population~\citep{2023PhRvX..13d1039A,abbott_population_2023,2023PhRvD.108d3011E}.
-Additionally, the \ac{GW} signals observed by the LIGO, Virgo, and KAGRA detectors~\citep{aasi_advanced_2015,acernese_advanced_2014,akutsu_overview_2021}  provide direct measurements of the distance to their sources.
-This makes them ``standard sirens'': direct probes of cosmological parameters that circumvent the need for a cosmological distance ladder~\citep{schutz_determining_1986,holz_using_2005}.
-
-A well-known demonstration of standard siren cosmology was the multi-messenger event GW170817 \citep{abbott_multi-messenger_2017,coulter_swope_2017,valenti_discovery_2017,2017ApJ...848L..27T}, whose clear association with a host galaxy provided a precise redshift measurement and allowed for a direct ``bright siren'' measurement of the Hubble constant, \Ho{} \citep{abbott_gravitational-wave_2017}.
-External redshift information can also come from galaxy catalogs, which provide an ensemble of possible redshifts for each \ac{GW} signal, allowing for a probabilistic ``dark siren'' measurement of \Ho{} when multiple \ac{GW} detections are combined \citep{del_pozzo_inference_2012, chen_two_2018, fishbach_standard_2019, gray_cosmological_2020, gwtc3_cosmo, gair_hitchhikers_2023}.
-
-External information about \ac{GW} source redshifts need not be available in order to use them as standard sirens, however.
-\ac{GW} signals provide direct measurements of each source's luminosity distance, $d_L$, and redshifted, or detector frame, masses, $m_{\det}= m_{\source}(1+z)$ \citep[e.g.][]{chen_mass-redshift_2019}.
-Therefore, if the source frame mass is known, each event provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the expansion of the universe at the time the \ac{GW} signal was emitted, $H(z)$.
-
-In practice, the source-frame masses of individual events are not known~\citep[unless tidal information is available, e.g.][]{messenger_measuring_2012, 2021PhRvD.104h3528C}.
-It is, however, possible to regard the \textit{population} of compact binaries at large and use known features in their source-frame mass distribution to obtain self-calibrated redshift estimates.
-The full mass distribution therefore acts analogously to an electromagnetic spectrum, in which the apparent locations of spectral features provide a redshift measurement.
+In practice, the source-frame masses of individual \ac{GW} signals are not known~\citep[unless tidal information is available, e.g.][]{messenger_measuring_2012, 2021PhRvD.104h3528C}.
+It is, however, possible to consider the \textit{population} of compact binaries at large and use known features in their source-frame mass distribution to obtain self-calibrated redshift estimates.
+The full mass distribution therefore acts analogously to an electromagnetic spectrum, in which the apparent locations of spectral features relative to their rest-frame locations provide a redshift measurement.
 The method of using the mass distribution of \ac{GW} sources to measure cosmological parameters has therefore been coined ``spectral sirens'' \citep{ezquiaga_spectral_2022}.
@@
 Central to the spectral siren methodology is knowledge of the compact binary mass distribution.
-First principles models for mass distributions of merging compact binaries are, however, not available: significant theoretical uncertainties exist about the shapes, locations, and very existence of predicted features in the mass distribution~\citep[e.g.][]{ zevin_constraining_2017, mapelli_binary_2020,2021ApJ...910..152Z,marchant_evolution_2023}.
-This includes uncertainties about critical aspects, such as the existence of a ``pile-up'' due to pulsational-pair instability, a pair-instability-driven upper mass gap, the maximum neutron star mass, and the existence of a putative lower mass gap between neutron stars and black holes.
-To this end, spectral siren cosmology\footnote{As well as statistical cosmological measurements leveraging galaxy catalogs} relies on \textit{simultaneously} measuring a source-frame mass spectrum alongside cosmological parameters.
-This is typically accomplished by adopting a phenomenological, parametric model for the mass distribution, usually composed of power laws and Gaussians \citep[e.g.][]{gwtc3_cosmo}.
+However, first principles models for mass distributions of merging compact binaries are not available: significant theoretical uncertainties exist about the shapes, locations, and very existence of predicted features in the mass distribution~\citep[e.g.][]{ zevin_constraining_2017, mapelli_binary_2020,2021ApJ...910..152Z,marchant_evolution_2023}.
+This includes uncertainties about potentially dominant features, such as the existence of a ``pile-up'' due to pulsational-pair instability, a pair-instability-driven upper mass gap~\citep{farmer}, the maximum neutron star mass~\citep{fryer_theoretical_2001, alsing_evidence_2018}, and the existence of a putative lower mass gap between neutron stars and black holes~\citep{ozel_black_2010, farr_mass_2011, farah_bridging_2022}.
+To this end, spectral siren cosmology relies on \textit{simultaneously} measuring a source-frame mass spectrum alongside cosmological parameters.
+This is typically accomplished by adopting a phenomenological, parametric model for the mass distribution, usually composed of power laws and Gaussians \citep[e.g.][]{gwtc3_cosmo,mastrogiovanni_icarogw_2023}.
 
 Such parametric modeling of the compact binary mass distribution raises its own set of dangers.
-It is well-known that different parametric models can generically yield very different results~\citep{abbott_population_2021,abbott_population_2023}.
+It is well known that different parametric models can generically yield very different constraints on cosmological parameters ~\citep{abbott_population_2021,abbott_population_2023}.
 This is problematic: because the measured mass distribution serves as the template by which to extract redshifts, a mismodeled mass distribution would introduce systematic errors in inferred redshifts and, in turn, systematically bias any resulting cosmological inference~\citep{ezquiaga_spectral_2022,Mukherjee:2021rtw,mastrogiovanni_importance_2021,pierra_study_2023}.
@@
 While the former yields a measurement of $H_0$ consistent with the true underlying value, the latter does not.
-Such systematic biases may already be relevant, as cosmological measurements by the LIGO-Virgo-KAGRA are known to depend on the exact choice of mass model used~\citep{gwtc3_cosmo}.
+Such systematic biases may already be relevant, as cosmological measurements by the LIGO-Virgo-KAGRA are known to depend on the choice of mass model used~\citep{gwtc3_cosmo}.
 Furthermore, these biases may become a dominant source of uncertainty in the near future~\citep{pierra_study_2023}.
 The prospects of such a dominant systematic uncertainty is troubling.
-If \emph{a priori}\/ knowledge of the mass distribution's morphology (whether an exact theoretical prediction or  knowledge of the correct parametric family of models) is a pre-requisite for the spectral siren method, the effectiveness of such a measurement would be significantly hampered.
+If prior knowledge of the mass distribution's morphology (whether an exact theoretical prediction or knowledge of the correct parametric family of models) is a prerequisite for the spectral siren method, the effectiveness of such a technique would be significantly hampered.
 
-In this work, we explicitly demonstrate that no \emph{a priori}\/ knowledge of the shape of the \ac{CBC} mass spectrum is necessary to use the spectral siren methodology. 
+In this work, we explicitly demonstrate that no prior knowledge of the shape of the \ac{CBC} mass spectrum is necessary to use the spectral siren methodology. 
 We do this by inferring $H(z)$ with a flexible, non-parametric model for the mass distribution of \acp{CBC} (blue lines in Figure~\ref{fig:O5_GP}). 
-This model makes minimal prior assumptions about the shape of the mass distribution, enabling it to accurately infer a wide range of morphologies and remain agnostic to the astrophysical processes that give rise to the mass distribution's features.
+This model makes minimal prior assumptions about the shape of the mass distribution, enabling it to accurately infer a wide range of morphologies and remain agnostic to the astrophysical processes that give rise to features in the mass distribution.
 Despite its flexibility, our approach is able to consistently obtain unbiased measurements of cosmological parameters, showing that nonparametric methods are not only sufficient for a spectral siren measurement, they can also mitigate systematic effects in the measurement caused by model misspecification.
-The success of the non-parametric mass model in recovering the injected cosmological parameters demonstrates that the information in the spectral siren measurement comes solely from the assumption that all \acp{CBC} follow the same mass distribution, not from the enforcement of specific features.\footnote{More precisely, the spectral siren method is also valid if \acp{CBC} have different mass distributions at different redshifts, so long as the evolution of the mass distribution does not exactly mimic cosmology, and the model for the mass distribution allows for evolution \citep{ezquiaga_spectral_2022}.
-}
+The success of the non-parametric mass model in recovering the injected cosmological parameters demonstrates that the information in the spectral siren measurement does not come from the enforcement of specific features in the mass distribution.
+Rather, it is provided by the assumption that either all \acp{CBC} follow a common mass distribution, or that any evolution of the mass distribution with redshift does not exactly mimic cosmology \citep{ezquiaga_spectral_2022}.
 
-We find that using our non-parametric model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ during \ac{O5}, when the detectors will reach their design sensitivity.
+We find that our non-parametric model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ during \ac{O5}, when the detectors will reach their design sensitivity.
 %, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
 We highlight measurements of \Ho{} in order to benchmark the accuracy and precision of our non-parametric method, as well as explore the role of spectral sirens in elucidating the Hubble tension.
-However, the primary utility of spectral siren measurements will be in constraining $H(z)$ at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn, You:2020wju}\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}.
-Applied to data from next-generation detectors, this method will constrain the dark energy equation of state parameter, $w$, at redshifts $\gtrsim 2$.
+However, the primary utility of spectral siren measurements will be in constraining $H(z)$ at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn, You:2020wju}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
+When applied to data from next-generation detectors, this method will measure the dark energy equation of state parameter, $w$, at redshifts $\gtrsim 2$.
 
-This paper is organized as follows: Section~\ref{sec:ss} introduces the spectral siren method, demonstrating how cosmological parameters are inferred from the mass distribution of \ac{GW} sources.
+This paper is organized as follows: Section~\ref{sec:data generation} describes the simulated dataset.
+Section~\ref{sec:ss} introduces the spectral siren method, demonstrating how cosmological parameters are inferred from the mass distribution of \ac{GW} sources.
 Section~\ref{sec:model} describes the non-parametric mass distribution we develop for use within the spectral siren method.
-We present the results of using parametric and non-parametric mass distributions in Section~\ref{sec:results}, as well as projections for future constraints on $H(z)$.
-In Section~\ref{sec:discussion} we discuss the implications of our results and outline future work.
+In Section~\ref{sec:results}, we present the results of using parametric and non-parametric mass distributions, as well as projections for future constraints on $H(z)$.
+We discuss the implications of our results and outline future work in Section~\ref{sec:discussion}.
 
-The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
+The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \url{https://github.com/afarah18/spectral-sirens-with-GPs}.% \GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
 
-\section{Cosmology with an astrophysically-agnostic mass distribution}
+\section{Cosmology with an astrophysically-agnostic mass distribution model}
 \label{sec:methods}
@@
 all features do not simultaneously exhibit an {\em identical}\/ monotonic rightward shift with redshift;][]{ezquiaga_spectral_2022}.
-Specifically, the method identifies the preferred relationship between luminosity distance and redshift that causes all masses to follow, as closely as possible, this common source-frame distribution.
-We do not need to know this source-frame distribution in advance, but, given some model for the black hole mass distribution, can infer both this mass distribution and cosmology simultaneously.
+Specifically, the method identifies the preferred relationship between luminosity distance and redshift that causes all masses to follow this common source-frame distribution.
+We do not need to know this source-frame distribution in advance, but, given some model for the black hole mass distribution, can infer it simultaneously with the cosmological expansion.
 
 As described above, though, strongly parametrized models for the mass distribution yield biased measurements of cosmological parameters if they poorly approximate the true mass distribution of compact binaries.
-Our goal is to circumvent this and instead model the population of \acp{GW} sources in a flexible and astrophysically-agnostic way. 
-There exist several non-parametric methods developed for this purpose \citep{tiwari_vamana_2021,edelman_aint_2022,sadiq_flexible_2022,edelman_cover_2023,mandel_extracting_2019,ray_non-parametric_2023}.
+We aim to circumvent such biases and instead model the population of \acp{GW} sources in a flexible and astrophysically-agnostic way. 
+There exist several non-parametric methods developed for this purpose \citep{tiwari_vamana_2021,edelman_aint_2022,sadiq_flexible_2022,rinaldi_hdpgmm_2022,edelman_cover_2023,mandel_extracting_2019,ray_non-parametric_2023}.
 While well-suited to infer the \ac{GW} source population with a fixed cosmology, several of these methods employ fixed features in source frame mass, such as bin edges \citep{mandel_extracting_2019,ray_non-parametric_2023} or spline nodes \citep{edelman_aint_2022}.
-Since these locations were chosen with a fixed cosmology, they risk causing the inference to prefer the cosmological parameters used when choosing the feature locations. 
+Since these locations were chosen with a fixed cosmology, they risk causing the inference to prefer the cosmological parameters assumed when choosing the feature locations. 
 Indeed, \citet{mastrogiovanni_importance_2021} show how using fixed features can significantly bias cosmological inference within the spectral siren methodology.
@@
 \subsection{Simulated Data}
+\label{sec:data generation}
 %% TRANSITION: Describe how our goal is to demonstrate the utility of GPs and cosmology by constructing and analyzing a simulated catalog. 
-To demonstrate the effectiveness of our \ac{GP}-based mass distribution in inferring both cosmological parameters and the population of \ac{GW} sources without introducing features or biases driven driven by prior assumptions, we apply our methodology to a simulated dataset.
-In generating a catalog of \ac{GW} sources from a known population and cosmological model, we are able to quantify the accuracy of our method.
-The use of simulated data also enables us to make projections for future datasets and safely ignore dimensions such as spin and mass ratio that do not impact cosmological measurements but would otherwise be important to simultaneously fit to avoid biases in inference of a true dataset's population.
+To demonstrate the effectiveness of our \ac{GP}-based mass distribution in inferring both cosmological parameters and the population of \ac{GW} sources without introducing features or biases driven by prior assumptions, we apply our methodology to a simulated dataset.
+By generating a catalog of \ac{GW} sources from a known population and cosmological model, we are able to quantify the accuracy of our method.
+The use of simulated data also enables us to make projections for future datasets and safely ignore dimensions such as spin and mass ratio that do not impact cosmological measurements but would otherwise be important to simultaneously fit to avoid biases in population inference of real data~\citep{biscoveanu_sources_2021}.
 
@@
 We design our simulated catalog to match the characteristics of the data expected from one year of observation in \ac{O5}. 
-The events in this catalog are drawn from an underlying population described by the \plp{} mass distribution presented in \citet{talbot_measuring_2018} and used in \citet{abbott_binary_2019, abbott_population_2023}, as well and the redshift distribution presented in \citet{callister_shouts_2020}, with hyperparameters consistent with those found in \citet{abbott_population_2023}.
+The \acp{BBH} in this catalog are drawn from an underlying population described by the \plp{} mass distribution presented in \citet{talbot_measuring_2018} and used in \citet{abbott_binary_2019, abbott_population_2023}, as well and the redshift distribution presented in \citet{callister_shouts_2020}, with hyperparameters consistent with those found in \citet{abbott_population_2023}.
 We assume the mass distribution has no redshift evolution for the redshift range to which the \ac{O5} detectors will be sensitive. 
 This assumption is consistent with current data \citep{fishbach_when_2021,van_son_redshift_2022,abbott_population_2023}, but will have to be revisited in future observing runs.
-% \daniel{We should say more about the importance of this assumption?}
-This process results in a catalog of $\variable{output/num_found_events.txt}$ \acp{BBH}.
+If we had constructed a catalog with an evolving mass distribution, we would have needed to correspondingly allow for redshift evolution in our population inference.
+We leave a full treatment of redshift evolution for future work, as discussed in Section~\ref{sec:discussion}.
+We use the \texttt{GWMockCat}~ \citep{farah_things_2023} package to apply \ac{O5}-like selection effects to the drawn \acp{BBH}, generate realistic measurement uncertainty, and produce sensitivity estimates that are consistent with the simulated \ac{GW} signals.
+We will use the term ``event'' to refer to \ac{GW} signals that pass the criteria for detection.
+This process results in a catalog of $N_{\text{ev}} = \variable{output/num_found_events.txt}$ \ac{GW} signals that pass the criteria for detection, hereon called events.
 
-We use the \texttt{GWMockCat} package to generate consistent simulated data and sensitivity estimates \citep{farah_things_2023}.
 Additional details of the data simulation, including the form of the injected population, are described in Appendix~\ref{ap:data generation}.
@@
 To simultaneously infer cosmological parameters and the population of \ac{GW} sources, we employ a hierarchical Bayesian analysis.
-This allows us to ``undo'' the selection effects of \ac{GW} detectors to obtain a true, or astrophysical, population and constrain the cosmic expansion history.
+This allows us to undo the selection effects of \ac{GW} detectors to obtain a true, astrophysical population and constrain the cosmic expansion history.
 
-Given a source population and background cosmology described by hyperparameters $\Lambda$, the likelihood of observing data $\{d\}$ that contains $N_{\text{ev}}$ detected \ac{GW} signals, each with parameters $\theta_i$, is \citep{loredo_handling_2009, Taylor:2011fs, mandel_extracting_2019}
+Given a source population and background cosmology described by hyperparameters $\Lambda$, the likelihood of observing data $\{d\}$ that contains $N_{\text{ev}}$ detected \ac{GW} signals, each with parameters $\theta_i$, is \citep{loredo_handling_2009, Taylor:2011fs, mandel_extracting_2019,vitale_inferring_2020}
 \begin{equation}
@@
 \end{equation}
-Here, $\Lambda$ is the set of population parameters, or hyper-parameters.
+Here, $\Lambda$ is the set of population parameters hyper-parameters.
 $\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ is the detector frame merger rate density of \acp{BBH} conditioned on hyper-parameters $\Lambda$, this quantity is also referred to as the population model.
@@
 Since the event parameters are not perfectly measured, we incorporate uncertainty to marginalize over the possible parameters associated with each event.
-Practically, this is done by a Monte Carlo average over each event's posterior samples $\{\theta_j\}_i$, dividing out the prior used when inferring those posterior samples, $\pi_{\rm PE}(\theta)$:
+Practically, this is done by a Monte Carlo average over the posterior samples $\{\theta_j\}_i$ of each event $i$ and dividing out the prior used when inferring those posterior samples, $\pi_{\rm PE}(\theta)$:
 \begin{equation}
@@
 This provides a posterior on both the compact binary population and the background cosmology when combined with a prior $p(\Lambda)$ on the population and cosmological parameters, as discussed below.
-We will use the term ``hyper-prior'' to refer to $p(\Lambda)$ for the remainder of this work.
+We will use the term hyper-prior to refer to $p(\Lambda)$ for the remainder of this work.
 
-The likelihood in Equation~\eqref{eq:inhomog-poisson} is maximized when the relationship between redshift and luminosity distance makes all source frame masses follow the same distribution.
-This is the basis of the spectral siren method of measuring cosmological parameters; we can simultaneously infer the relationship between redshift and luminosity distance with the parameters of the source frame mass distribution.
-The full set of hyper-parameters therefore includes the cosmological parameters that dictate the $D_L$--$z$ relation: the local expansion rate \Ho, the present fractional energy densities of dark matter \Omm, dark energy $\Omega_\Lambda$ and radiation $\Omega_r$, and the equation of state of dark energy $w$.
+The likelihood in Equation~\eqref{eq:inhomog-poisson} is maximized when the relationship between redshift and luminosity distance makes all source-frame masses follow the same distribution.
+This is the basis of the spectral siren method of measuring cosmological parameters; we can simultaneously infer the relationship between redshift and luminosity distance with the parameters of the source-frame mass distribution.
+The full set of hyper-parameters therefore includes the cosmological parameters that dictate the $D_L$--$z$ relation: the local expansion rate \Ho, the present fractional energy densities of dark matter \Omm, dark energy $\Omega_\Lambda$, and radiation $\Omega_r$, and the equation of state of dark energy $w$.
 In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and $w=-1$ and use uniform priors on \Ho{} and \Omm{}, corresponding to a flat $\Lambda$CDM cosmology.
 
-Algorithmically, we evaluate the likelihood in Equation~\ref{eq:inhomog-poisson} by drawing \Ho{} and \Omm{} from their prior distributions (contained in $p(\Lambda)$), using them to define a relationship between luminosity distance and redshift, then transforming detector frame masses to source frame masses according to $m_{\det} /(1+z(D_L,H_0,\Omega_M)) = m_{\source}$, evaluating the source-frame mass distribution at these transformed values ($\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ for events and $N_{\text{exp}}(\Lambda)$ for injections), and then taking the product of these terms to evaluate the expression in Equation~\ref{eq:inhomog-poisson}.
+Algorithmically, we evaluate the likelihood in Equation~\ref{eq:inhomog-poisson} by drawing \Ho{} and \Omm{} from their prior distributions (contained in $p(\Lambda)$), use them to define a relationship between luminosity distance and redshift, transform detector-frame masses to source-frame masses according to $m_{\det} /(1+z(D_L,H_0,\Omega_M)) = m_{\source}$, evaluate the source-frame mass distribution at these transformed values ($\diff N/\diff t_{\det} \diff \theta_i (\theta_i;\Lambda)$ for events and $N_{\text{exp}}(\Lambda)$ for injections), and then take the product of these terms to evaluate the expression in Equation~\ref{eq:inhomog-poisson}.
 The full process is outlined in the next subsection.
@@
 
-Practically, the only difference in the inference of the population when using a \ac{GP} versus other modeling choices is that $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is determined directly by a realization of the \ac{GP}, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
-In other words, when using parametric models, $\frac{\diff N}{\diff t_{\det} \diff \theta_i} (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson} is calculated by evaluating a specific functional form described by a small set of hyper-parameters.
+Practically, the difference in the inference of the population when using a \ac{GP} versus other modeling choices is that the population model ($\diff N/\diff t_{\det} \diff \theta_i (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson}) is determined directly by a realization of the \ac{GP}, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
+In other words, when using parametric models, $\diff N/\diff t_{\det} \diff \theta_i (\theta_i;\Lambda)$ is calculated by evaluating a specific functional form described by a small set of hyper-parameters.
 With the \ac{GP} approach, the hyper-parameters describing the mass distribution are the rate at each event-level posterior sample's source frame mass, and the rate at each found injection's source frame mass.
@@
 % The source frame mass is different for each possible value of \Ho{} and \Omm{}, so the hyper-parameters are defined in different locations for each draw from the hyper-posterior.
-In this way, our \ac{GP}-based mass distribution is similar to the autoregressive population models used in ~\citet{callister_parameter-free_2023}.
-Indeed, an autoregressive process is a \ac{GP} with a particular choice of kernel (described below).
-We therefore have $N_{\text{ev}}M + N_{\text{inj}}$ mass hyper-parameters, where $N_{\text{ev}}$ is the number of events, $M$ is the number of posterior samples per event, and $N_{\text{inj}}$ is the number of injections used to calculate the selection function $\xi$.
+In this way, our \ac{GP}-based mass distribution is similar to the autoregressive population models used in~\citet{callister_parameter-free_2023}.
+Indeed, an autoregressive process is a \ac{GP} with a particular choice of kernel, which we describe below.
+We therefore have $N_{\text{ev}}M + N_{\text{inj}}$ mass hyper-parameters, where $M$ is the number of posterior samples per event and $N_{\text{inj}}$ is the number of injections used to calculate the selection function~\citep[see e.g.][]{vitale_inferring_2020,talbot,essick_estimating_2021}.
 We label these $\{\mathcal{R}_{ij},\mathcal{R}_k\}$, and our full set of hyper-parameters is $\Lambda=\{H_0, \Omega_M, \mathcal{R}_{ij},\mathcal{R}_k\}$. 
-As the hyper-parameters differ at each data point, we have successfully fulfilled our goal to construct a model without pre-defined features in source frame mass.
-We note that other non-parametric methods may be adapted to avoid such pre-defined features, such as fitting for the locations of their features simultaneously with the rest of the inference \citep[e.g.][]{tiwari_vamana_2021} or by using an appropriate smoothing \citep[e.g.][]{edelman_cover_2023}.
-Such adaptations are necessary to avoid circular inference.
+As the hyper-parameters differ at each data point, we have successfully fulfilled our goal to construct a model without predefined features in source-frame mass.
+We note that other non-parametric methods may be adapted to avoid such predefined features, such as fitting for the locations of their features simultaneously with the rest of the inference \citep[e.g.][]{tiwari_vamana_2021} or, in the case of splines, by using a smoothing function that allows for features to occur at arbitrary locations appropriate smoothing \citep[e.g.][]{edelman_cover_2023}.
+% Such adaptations are necessary to avoid circular inference.
 
@@
 It defines the notion of similarity between adjacent points and thereby encodes our assumptions about the smoothness of the source frame mass distribution \citep{rasmussen_gaussian_2006}.
-We use a Mat\'ern kernel \citep{handcock_bayesian_1993, stein_interpolation_1999} with $\nu = 5/2$, but have repeated the analysis with $\nu=3/2$ and $\infty$, finding little impact on the results, except that the $\nu=\infty$ case (also called the ``squared exponential'' kernel) produces a slightly more jagged mass distribution.
+We use a Mat\'ern kernel \citep{handcock_bayesian_1993, stein_interpolation_1999} with $\nu = 5/2$, but have repeated the analysis with $\nu=3/2$ and $\infty$, finding little impact on the results, except that the $\nu=\infty$ case (also called the squared exponential kernel) produces a slightly more jagged mass distribution.
 In addition to the mean, Mat\'ern kernels have two parameters that determine their properties: a length scale $l$ and a variance $s$.
@@
 We fit these hyper-hyper-parameters along with the hyper-parameters $\Lambda$ to minimize prior assumptions about the form of the mass distribution.
-We use ``penalized-complexity'' priors on the hyper-hyper-parameters to enforce that the model does not create small-scale structure uninformed by data, thereby avoiding over-fitting \citep{simpson_penalising_2017,simpson_garcpas_2022}.
+We use penalized-complexity priors on the hyper-hyper-parameters to enforce that the model does not create small-scale structure uninformed by data, thereby avoiding over-fitting \citep{simpson_penalising_2017,simpson_garcpas_2022}.
 Explicitly, the priors on $l$ and $s$ are Fr\'echet and Gamma distributions, respectively, and are defined to have less than 5\% support for correlation lengths smaller than the average spacing between event-level posterior means.
 
-The time to evaluate a \acp{GP} notoriously scales as the cube of the number of data points, making \acp{GP} unwieldy with large data sets such as the ones expected for \ac{O5}.
+The time to evaluate a \ac{GP} is notorious for scaling as the cube of the number of data points, making \acp{GP} unwieldy with large data sets, such as the $\mathcal{O}(10^9)$ posterior samples and software injections expected for \ac{O5}.
 We therefore make two approximations to a full \ac{GP} to increase computational efficiency.
@@
     \item Draw cosmological parameters \Ho{} and \Omm{} from uniform prior distributions.
-    \item Transform the luminosity distances and detector frame masses of each event posterior sample and injection into redshifts and source frame masses according to the cosmology specified by step 1.
+    \item Convert the luminosity distances and detector-frame masses of each event posterior sample to redshifts and source-frame masses according to the cosmology specified by step 1.
     \item Draw hyper-hyper-parameters $l$ and $s$ from the penalized-complexity priors described above.
@@
     This is defined on a regular grid of source-frame masses and evaluated using the \texttt{QuasisepSolver} in \texttt{tinygp}.
-    \item Interpolate the \ac{GP} at each event posterior sample and injection source frame mass (from step 3)
+    \item Interpolate the \ac{GP} at each event posterior sample and injection source-frame mass (from step 3).
     \item Calculate the population likelihood according to Equation~\ref{eq:inhomog-poisson}.
@@
 \label{sec:results}
-In this Section, we first show that fitting an incorrect functional form to the mass distribution of \acp{CBC} biases the inference of cosmological parameters when using the spectral siren methodology.
-We then demonstrate that our flexible model alleviates this bias without the need to know the mass distribution's morphology \emph{a priori}.
-To illustrate this, we use three different models for the source-frame mass distribution to infer the cosmic expansion rate from the simulated catalog described in Section~\ref{sec:data generation} and Appendix~\ref{ap:data generation}.
+In this Section we show that fitting an incorrect functional form to the mass distribution of \acp{CBC} biases the inference of cosmological parameters when using the spectral siren methodology.
+We then demonstrate that our flexible model alleviates this bias without the need to know the morphology of the mass distribution \emph{a priori}.
+We illustrate this explicitly by using three different models for the source-frame mass distribution to infer the cosmic expansion rate from the simulated catalog described in Section~\ref{sec:data generation} and Appendix~\ref{ap:data generation}.
 The three models are as follows:
@@
     \item \plp{}, which includes the true mass distribution within its hyper-prior,
-    \item a broken power law\footnote{This differs from the \textsc{Broken Power Law} model presented in \citet{abbott_population_2021}, as we do not employ a high-mass truncation.}, which does not include the true mass distribution within its hyper-prior, and
+    \item the \textsc{Broken Power Law} model presented in \citet{abbott_population_2021}, as we do not employ a high-mass truncation, which does not include the true mass distribution within its hyper-prior, and
     \item the flexible, \ac{GP}-based model described in Section~\ref{sec:model}, which is able to closely approximate the morphology of the true mass distribution, along with many other morphologies.
 \end{enumerate}
-For all models considered in this work, we assume the redshift distribution described in Equation~\ref{eq:underlying redshift dist}.
+For all models considered in this work, we assume the redshift distribution used to generated the data, described in Equation~\ref{eq:underlying redshift dist}.
 The results of each of these fits are shown in Figure~\ref{fig:O5_GP}.  
-The left panel shows the inferred source frame mass distribution for each of the considered models, and the right panel shows the corresponding posteriors on \Ho{}. 
+The left panel shows the inferred source-frame mass distribution for each of the considered models, and the right panel shows the corresponding posteriors on \Ho{}. 
 We indicate the true underlying source mass distribution and \Ho{} value with solid black lines in each panel.
 
-We find that using a parametric model of the same form as the true simulated distribution (\plp) recovers the injected value of \Ho{} within $\variable{output/PLPh0offset.txt}\sigma$.
-However, when using a model that cannot accurately represent the true shape of the mass distribution, we recover a biased estimate of \Ho: the true value is offset from the mean of the posterior by $\variable{output/BPLh0offset.txt}\sigma$.
-We have repeated this analysis with 50 separate simulated catalogs of $\sim500$ events, and found that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time,
+The fits presented in Figure~\ref{fig:O5_GP} are representative results from a single simulated catalog.
+These provide insight into the full statistical results presented below.
+In particular, it can be seen that the broken power law (orange curve) is inconsistent with the true value of $H_0$: in the run shown in Figure~\ref{fig:O5_GP}, the true value of \Ho{} is offset from the mean of the posterior by $\variable{output/BPLh0offset.txt}\sigma$.
+In contrast, the \plp{} and \ac{GP}-based models (green and blue curves) are consistent with the underlying truth.
+These models recover mean values of \Ho{} that are offset from the true value at $\variable{output/PLPh0offset.txt}\sigma$ and $\variable{output/nonparh0offset.txt}\sigma$, respectively. 
+Additionally, the mass distribution inferred with the \ac{GP}-based model closely resembles the true, simulated distribution.
+This indicates that using models that cannot accurately approximate the true mass distribution will lead to a significant systematic bias in the estimation of cosmological parameters.
+
+This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the \ac{GP}-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution.
+The \ac{GP}-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters. 
+In reality, we do not know the true functional form of the mass distribution, so the additional statistical uncertainty introduced by non-parametric approaches may be desirable over the possible systematic errors introduced by choosing a parametric model that likely does not contain the true mass distribution within its hyper-prior.
+
+% The lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects at those masses, causes the \ac{GP} to learn a relatively low merger rate at the edges of the mass distribution.
+% On the other hand, there is both a lack of data and little detector sensitivity at masses above $\sim100\Msun$ and below $\sim5\Msun$.
+% The mass distribution is therefore uninformed in this region and the \ac{GP} reverts to its prior distribution, which resembles random scatter around the mean differential merger rate.
+% Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
+% % Indeed, the upper edge of the mass distribution is very informative to the \Ho{} constraint with current detections \citep{gwtc3_cosmo}.
+% The combination of these two effects results in what appears as an uptick in the merger rate below $\sim5\Msun$ and above $\sim100\Msun$.
+% However, this reversion to the prior is uninformative for the \Ho{} constraint and does not affect inference on cosmological parameters.
+% Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,120\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
+
+To obtain a quantitative measure of the systematic bias introduced by mismodeling the mass distribution, we repeat the parametric analyses with 50 separate simulated catalogs of $\sim500$ events each.
+We find that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time,
 and the \plp{} model reaches the same level of bias only $\variable{output/PLP_bias_percent.txt}\%$ 
@@
 
-This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the \ac{GP}-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution: the true value is recovered at the $\variable{output/nonparh0offset.txt}\sigma$ level.
-The \ac{GP}-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters. 
-In reality, it is impossible to know the true functional form of the mass distribution, so the additional statistical uncertainty introduced by non-parametric approaches may be desirable over the systematic errors introduced by choosing a parametric model.
-
-Additionally, the mass distribution inferred with the \ac{GP}-based model closely resembles the true, simulated distribution.
-At very high and very low masses, there is both a lack of data and little detector sensitivity.
-The mass distribution is therefore uninformed in this region and the \ac{GP} reverts to its prior distribution, which resembles random scatter around the mean differential merger rate.
-Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
-On the other hand, the lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects in that mass range causes the \ac{GP} to learn a relatively low merger rate at the edges of the mass distribution.
-% Indeed, the upper edge of the mass distribution is very informative to the \Ho{} constraint with current detections \citep{gwtc3_cosmo}.
-The combination of these two effects results in what appears as an uptick in the merger rate at very low and very high masses, but this reversion to the prior is uninformative to the \Ho{} constraint and does not affect inference on cosmological parameters.
-Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,120\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
-
-Collectively, these results indicate that prior knowledge of the shape of the mass distribution is not required to perform a spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.
+Collectively, these results indicate that prior knowledge of the shape of the mass distribution is not required to perform an unbiased spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.
 
 \subsection{Projections for Future Measurements}
-The fit shown in Figure~\ref{fig:O5_GP} implies that we can expect to measure \Ho{} to $\variable{output/nonparh0percent.txt}\%$ ($1-\sigma$ uncertainties) after one year of \ac{O5} using the \ac{GP}-based spectral siren method.
-However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\sim 3$, with most sources lying near redshift $\sim 1.2$ \citep{chen_distance_2021}.
+Figure~\ref{fig:O5_GP} demonstrates an expected $\variable{output/nonparh0percent.txt}\%$ ($1\sigma$ uncertainties) measurement of \Ho{} after one year of \ac{O5} using the \ac{GP}-based spectral siren method.
+However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\approx 3$, with most sources lying near redshift $\simeq 1.2$ \citep{chen_distance_2021}.
 Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 100$.
-This means that future \acp{GW} observations will be more sensitive to $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters \citep{Chen:2024gdn}.
-We demonstrate this by repeating the same \ac{GP}-based spectral siren analysis of the previous subsection while also fitting for the local matter density, \Omm.
-However, we emphasize that it would also have been possible to constrain other cosmological parameters governing $H(z)$.
+This means that future \ac{GW} observations will be more sensitive to $H(z\gtrsim1)$ than to \Ho, and can therefore constrain several additional cosmological parameters \citep{Chen:2024gdn}.
+We demonstrate this by repeating the same \ac{GP}-based spectral siren analysis while also simultaneously fitting for the local matter density, \Omm.
 The result is shown in Figure~\ref{fig:O5_corner}.
+We emphasize that these precise measurements over a wide range in redshift enable precision estimation of additional cosmological parameters governing $H(z)$.
 
@@
 The left panel of Figure~\ref{fig:O5_corner} demonstrates a strong anti-correlation between the \Omm{} and \Ho{} posteriors, resulting in similarly informative constraints on the two parameters.
-This is because \Ho{} controls the $y$-intercept of the $H(z)$ curves on the right panel, while \Omm{} informs the slope of those curves.
-The same measurement of $H(z\neq0)$ can be obtained by increasing the slope while decreasing the $y$-intercept, and vice versa.
-Similar behavior can be observed in current measurements of the \ac{BBH} redshift distribution, which exhibit a ``waist'' at $z\sim0.2$ with current observations \citep{abbott_population_2023, callister_parameter-free_2023}.
+This is because \Ho{} controls the $y$-intercept of the $H(z)$ curves on the right panel, while \Omm{} informs the slope of those curves; the same measurement of $H(z\neq0)$ can be obtained by increasing the slope while decreasing the $y$-intercept, and vice versa.
+Similar behavior can be observed in current measurements of the \ac{BBH} redshift distribution, which exhibits a tightening of the posterior at $z\sim0.2$ with current observations \citep{abbott_population_2023, callister_parameter-free_2023}.
 
-Next-generation detectors will be sensitive to a much larger range of redshifts \citep{et_steering_committee_einstein_2020, evans_horizon_2021} and will therefore break the degeneracy between cosmological parameters, allowing for tighter constraints on each.
-However, the low cosmological volume (and thus low number of mergers) at low redshift will always limit the constraining power of spectral sirens at $z=0$, potentially making this method more sensitive to cosmological parameters that affect higher redshifts than to \Ho.
-Combining spectral sirens with other methods that are more sensitive to the local expansion rate, such as those that employ electromagnetic counterparts, can increase the precision of \ac{GW} standard sirens at all redshifts \citep[e.g.][]{Chen:2024gdn}.
+Next-generation detectors will be sensitive to a larger range of redshifts \citep{et_steering_committee_einstein_2020, evans_horizon_2021}, and will therefore break the degeneracy between cosmological parameters and allow for tighter constraints on both \Omm{} and \Ho{}.
+However, the small cosmological volume (and thus low number of mergers) at low redshift will generally limit the constraining power of spectral sirens at $z=0$, potentially making this method more sensitive to cosmological parameters that affect higher redshifts.
+% than to \Ho.
+Combining spectral sirens with other methods that are sensitive to the local expansion rate, such as those that employ electromagnetic counterparts, may increase the precision of \ac{GW} standard sirens at all redshifts \citep[e.g.][]{Chen:2024gdn}.
 
@@
     The two parameters are strongly degenerate because of the multiple ways of measuring $z=\variable{output/mostsensitivez.txt}$.
-    Spectral sirens are particularly well-suited to measuring cosmological parameters that affect the mid-to-high redshift universe. 
+    Spectral sirens are particularly well suited for measuring cosmological parameters that affect the Universe at $z\gtrsim0.2$. 
     }
@@
 
-\acp{GW} are unique cosmic messengers in that they carry both redshift and distance information, making them remarkably clean probes of the Universe's expansion history. 
+\acp{GW} are unique cosmic messengers in that they carry both redshift and distance information, making them remarkably clean probes of the Universe's expansion history calibrated directly by the theory of general relativity.
 However, the current method of determining \ac{GW} redshifts via the distribution of their source masses (i.e.
-spectral sirens) employs an assumption on the shape of their population, typically enforced with simplified parametric functions.
-This choice of mass distribution, often presumed to be necessary or fundamental to the method, may introduce a systematic bias to a measurement of the cosmological parameters that otherwise makes very few assumptions.
+spectral sirens) employs an assumption on the shape of their population, typically encapsulated by simplified parametric functions.
+This choice, often presumed to be necessary or fundamental to the method, may introduce a systematic bias to a measurement of the cosmological parameters that is otherwise appealing for its unique elegance and simplicity.
 
-In this work, we show that such a choice is not in fact necessary.
-We do this by accurately measuring cosmological parameters with a highly flexible model for the mass distribution. 
-The success of the flexible model in recovering an informative and accurate posteriors on \Ho{} and \Omm{} demonstrates that there is no need to know the \ac{CBC} mass distribution's morphology in order to constrain cosmological parameters with it.
+In this work we show that a specific choice of mass distribution is unnecessary to arrive at informative and accurate posteriors on \Ho{} and \Omm{}.
+We do this by accurately measuring these cosmological parameters with a highly flexible model for the mass distribution. 
+%The success of the flexible model in recovering informative and accurate posteriors on \Ho{} and \Omm{} demonstrates that there is no need to know the morphology of the \ac{CBC} mass distribution in order to constrain cosmological parameters.
 This reinforces the notion that the information in the spectral siren measurement comes from the assumption that all \ac{GW} sources come from the same population, a far less stringent statement than the assumption that we understand the exact astrophysical processes which give rise to that population (i.e.
@@
 
-Spectral sirens are particularly useful in the context of the current ``Hubble tension'': a disagreement between multiple methods of measuring the local expansion rate of the universe with electromagnetic observations \citep{freedman_measurements_2021}.
-As this tension can only be explained by non-standard physics or yet-unknown systematic uncertainties in either the cosmic distance ladder or cosmic microwave background measurements, direct probes of the local expansion that do not rely on electromagnetic observations will help determine whether new physics need be invoked to resolve the current discrepancy.
+Spectral sirens are particularly useful in the context of the current Hubble tension: a disagreement between multiple methods of measuring the local expansion rate of the universe \citep{freedman_measurements_2021}.
+As this tension can only be explained by non-standard physics or yet-unknown systematic uncertainties in either the cosmic distance ladder or cosmic microwave background measurements, direct and independent probes of the local expansion---such as those presented here---may help determine whether the current discrepancy represents a fundamental crack in our standard model of physics and/or cosmology.
 
-We have also predicted a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} after one year of observing with the \ac{LVK} network's design sensitivity using non-parametric spectral sirens and a $\variable{output/PLPh0percent.txt}\%$
-measurement with parametric spectral sirens, demonstrating an increase in precision when using the correct parametric model.
-\comment{[These numbers are placeholders for now and will be updated in the next version.]}
-However, the true population of merging compact binaries in the Universe is unknown because neither the mechanism from which compact binaries were formed, nor the physical processes within that mechanism has been conclusively determined.
-It is therefore impossible to know the true functional form of the mass distribution, so systematic uncertainties arising from an incorrect choice for the form of the mass distribution are inevitable.
-With current observations, these systematic effects are dwarfed by statistical uncertainties.
-However, next-generation detectors will herald in enough \ac{GW} observations to substantially decrease statistical uncertainty in these measurements: \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to $\sim3\sigma$ systematic biases in \Ho{} when catalogs contain 2,000 events, though this bias may be an over-estimate as it does not include measurement uncertainty of \ac{GW} parameters.
+We predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} with one year of observing at the \ac{LVK} network's design sensitivity using the non-parametric spectral siren approach, and a $\variable{output/PLPh0percent.txt}\%$ measurement with parametric spectral sirens.
+% Discussion on this! why are parametric and nonparametric comparable?? 
+However, the true population of merging compact binaries in the Universe is unknown because neither the mechanism from which compact binaries were formed, nor the physical processes within the mechanism, have been conclusively determined.
+The true functional form of the mass distribution therefore remains elusive, and systematic uncertainties arising from an incorrect choice for the form of the mass distribution are inevitable.
+With current observations, these systematic effects are likely smaller than statistical uncertainties.
+However, next-generation detectors will herald sufficient \ac{GW} observations to substantially decrease statistical uncertainty in these measurements; for example, \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to $\sim3\sigma$ systematic biases in \Ho{} with catalogs of 2,000 events, although this bias may be an overestimate as it does not include measurement uncertainty of the \ac{GW} parameters.
 Thus, the additional statistical uncertainty introduced by non-parametric approaches may be preferable to the systematic errors associated with choosing a parametric model.
 
-In parallel to this work, \citet{magana_inprep} performed a spectral siren analysis on currently-public \ac{LVK} data using a model that employs histogram bins defined at fixed locations in source-frame mass to flexibly measure the \ac{BBH} mass distribution \citep[originally presented in][]{mandel_model-independent_2017,ray_non-parametric_2023}.
-While both the method presented in this work and that of \citet{magana_inprep} employ \acp{GP}, theirs uses it to modulate the histogram bin heights, leading to a piecewise constant function in log mass, whereas ours defines the \ac{GP} at each data point, constructing a smooth function with no pre-defined features in source-frame mass.
+In parallel with this work, \citet{magana_inprep} performed a spectral siren analysis on currently-public \ac{LVK} data using a model that employs histogram bins defined at fixed locations in source-frame mass to flexibly measure the \ac{BBH} mass distribution \citep[originally presented in][]{mandel_model-independent_2017,ray_non-parametric_2023}.
+While both the method presented in this work and that of \citet{magana_inprep} employ \acp{GP}, \citet{magana_inprep} uses \acp{GP} to modulate histogram bin heights, leading to a piecewise constant function in log mass, whereas we define the \ac{GP} at each data point, constructing a smooth function with no predefined features in source-frame mass.
 % Sharp bin edges may lead to optimistic constraints on cosmological parameters, especially as such features are unlikely to exist in the true mass distribution, possibly explaining the increase in precision relative to parametric methods seen in \citet{magana_inprep}.
 
-Work in preparation will extend the method developed here to mass distributions that are allowed to evolve with redshift.
-Non-parametric methods will allow for arbitrary redshift evolution, but must constrain that evolution to avoid exactly mimicking the effects of cosmological redshift.
-As demonstrated in \citet{ezquiaga_spectral_2022}, the degeneracy between an evolving mass distribution and the expansion of the universe can be broken provided there are multiple features present in the mass distribution and all features do not identically and monotonically move to higher source frame mass with increasing redshift.
+As demonstrated in \citet{ezquiaga_spectral_2022}, the degeneracy between an evolving mass distribution and the expansion of the universe can be broken provided that there are multiple features present in the mass distribution, and that we do not live in a fine-tuned universe where the evolutionary effects governing the \ac{CBC} mass distribution perfectly mimick the effects of cosmological redshift.
 The first condition (multiple features) is known to be met in current data \citep{abbott_population_2023}, with three robust features \citep{farah_things_2023}: a maximum black hole mass and overdensities at $\sim10\Msun$ and $\sim35\Msun$.
-The latter scenario (features identically and monotonically moving to higher masses with increasing redshift) would be astrophysically unlikely as the the locations of features in the mass distribution are each thought to be governed by different physical processes \citep{mapelli_binary_2020}.
-Since \aclp{GP} naturally scale to multiple data dimensions, \citep{rasmussen_gaussian_2006}, the method presented here can be straightforwardly generalized to fitting the redshift dependence of the mass distribution.
+The latter scenario---with features identically and monotonically shifting to higher masses with increasing redshift---would be astrophysically unlikely as the locations of features in the mass distribution are each thought to be governed by fundamentally different physical processes \citep{mapelli_binary_2020}.
+The non-parametric methods presented here will allow for arbitrary redshift evolution.
+Additionally, since \aclp{GP} naturally scale to multiple data dimensions \citep{rasmussen_gaussian_2006}, the method presented here can easily be generalized to fit the redshift dependence of the mass distribution.
+Future work will extend the method developed here to mass distributions that are allowed to evolve with redshift.
 
 Because \ac{GW} observations are the only data input to spectral sirens, they are uniquely positioned to probe the expansion history of the universe over a wide range of cosmological redshift, rather than just the local expansion \Ho.
-We have shown this by simultaneously measuring \Omm{} and \Ho{}, but the method can be trivially extended to constrain other cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$.
-We find that spectral sirens will be most sensitive to $z=\variable{output/mostsensitivez.txt}$ during \acf{O5}.
-This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers ~\citep{kiendrebeogo_updated_2023}, which can only be detected relatively locally.
-This means that upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances by \acp{GW}. 
-Proposed next generation \ac{GW} detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to \acp{CBC} across cosmic time (up to $z\sim100$).
-At that time, cosmological surveys such as those enabled by the Nancy Grace Roman Space Telescope and the Vera Rubin Observatory are expected to be able to precisely measure $H(z)$ to $z\sim3$ \citep{spergel_wide-field_2015}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
-When combined with other standard siren methods, the low-redshift expansion will also be well-constrained \citep{Chen:2024gdn}.
-While the sensitivity to high redshifts is a feature of the spectral siren method in general, non-parametric methods such as the one presented here will be imperative to avoid systematic biases as next-generation detectors provide additional data.
+We have shown this by simultaneously measuring \Omm{} and \Ho{}; the method can be trivially expanded to constrain additional cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$.
+We find that for \acf{O5}, spectral sirens will be most constraining at $z=\variable{output/mostsensitivez.txt}$.
+This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers~\citep{kiendrebeogo_updated_2023}, implying that upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances by \acp{GW}.
+Proposed next-generation \ac{GW} detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to \acp{CBC} across cosmic time (out to $z\sim100$)~\citep{et_steering_committee_einstein_2020, evans_horizon_2021}.
+Future cosmological surveys, such as those enabled by the Nancy Grace Roman Space Telescope and the Vera Rubin Observatory, are expected to be able to precisely measure $H(z)$ to $z\sim3$ \citep{spergel_wide-field_2015}, making the spectral siren method uniquely positioned to measure the expansion of the universe at high redshift.
+When combined with bright and dark siren methods, the low-redshift expansion history will also be well-constrained \citep{Chen:2024gdn}.
+Sensitivity to high redshifts is a feature of the spectral siren method in general, and non-parametric methods such as the one presented here will be imperative to avoid systematic biases in spectral siren cosmology.
 
 \begin{acknowledgments}
-    The authors thank Reed Essick, Utkarsh Mali, Ben Farr, Maya Fishbach, Colm Talbot,  for helpful conversations. 
+    The authors thank Reed Essick, Ben Farr, Maya Fishbach, Utkarsh Mali, and Colm Talbot for helpful conversations. 
     %Amanda
@@
 $\mathcal{S}$ is a smoothing function at low and high masses, and $\mathcal{N}_{\text{pl}}$ and $\mathcal{N}_{\text{g}}$ are the normalizations between  $m_{\min}$ and $m_{\max}$ for the power law component and truncated Gaussian component, respectively.
-$\bar{\Lambda}_z = \{z_p,\alpha_z,\beta_z\}$ are the parameters governing the redshift distribution's peak redshift, low-$z$ power law slope, and high-$z$ power law slope.
-When generating the simulated events, we have fixed $\alpha=-2.7,m_{\max}=78\Msun,m_{\min}=10\Msun, \mu=30\Msun,\sigma=7.0\Msun, f_{\text{peak}}=0.05,   z_p=2.4,\alpha_z=1.,\beta_z=3.4$.
+$\bar{\Lambda}_z = \{z_p,\alpha_z,\beta_z\}$ are the parameters governing the peak of the redshift distribution, low-$z$ power law slope, and high-$z$ power law slope.
+When generating the simulated events, we have fixed $\alpha=-2.7$, $m_{\max}=78\Msun$, $m_{\min}=10~\Msun$, $\mu=30~\Msun$, $\sigma=7.0~\Msun$, $f_{\text{peak}}=0.05$, $z_p=2.4$, $\alpha_z=1.0$, and $\beta_z=3.4$.
 These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{abbott_population_2023}.
@@
 We consistently apply this assumption to the data generation process and the population inference.
-We do not fit for or simulate spins, as they do not redshift and hence do tnot carry additional cosmological information, and additionally neglect secondary masses.
+We do not fit for or simulate spins, as they do not redshift and hence do not carry additional cosmological information, and additionally neglect secondary masses.
 
-We use cosmological parameters \Ho$=\variable{output/H0_FID.txt}\Hunits$, \Omm$=0.3$, $\Omega_\Lambda=1-$\Omm, consistent with those found by \citet{planck_collaboration_planck_2016}.
+We use cosmological parameters \Ho$=\variable{output/H0_FID.txt}\Hunits$, \Omm$=0.3$, and $\Omega_\Lambda=1-$\Omm, consistent with those found by \citet{planck_collaboration_planck_2016}.
 We emphasize, however, that the choice of cosmological parameters for data generation is arbitrary and does not impact the results, since we are concerned only with the ability of our method to recover the injected values.
@@
 We do not include neutron stars in our simulation set, as their contribution to the spectral siren measurement is expected to be subdominant in \ac{O5}.
-However, if a lower mass gap between the heaviest neutron stars and lightest black holes exists, it will provide an additional feature with which to inform the measurement, and will be the most informative feature for spectral siren measurements with next generation detectors \citep{ezquiaga_spectral_2022}.
+However, if a lower mass gap between the heaviest neutron stars and lightest black holes exists, it will provide an additional feature with which to inform the measurement, and will be the most informative feature for spectral siren measurements with next-generation detectors \citep{ezquiaga_spectral_2022}.
 
@@
 \texttt{GWMockCat} also simulates a set of software injections, which we use to estimate selection effects in the inference.
-To determine the detectability of both injections and simulated events in O5, we use the projected \ac{O5} LIGO power spectral density \citep{obsscen_noise_curves,abbott_prospects_2020} to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with $\rho_{\text{obs}}>8$ to be detectable.
+To determine the detectability of both injections and simulated events in O5, we use the projected \ac{O5} LIGO power spectral density \citep{obsscen_noise_curves,abbott_prospects_2020} for a single detector to calculate observed signal-to-noise ratios $\rho_{\text{obs}}$, and we consider events and injections with a single-detector SNR $\rho_{\text{obs}}>8$ to be detectable.
 The full procedure for this mock data generation process is described in \citet{fishbach_where_2017, farah_things_2023, essick_ensuring_2024}.
+
+% \section{\Acf{GP}-based mass distribution}
+% \label{ap:GP}
+
+% \begin{itemize}
+%     \item fit of GP to various mass dist morphologies, showing how flexible it is (fixing H0)
+%     \item zoom-in on GP traces 
+%     \item details from section 2.3
+%     \item discussion of reversion to prior at extrema of mass dist (move from section 3)
+% \end{itemize}
 
@@
 
+% synonyms
+
+% a parametric model of the same form as the true simulated distribution (\plp) 
+% a model that cannot accurately represent the true shape of the mass distribution (BPL)
+
 \end{document}

### 2024-03-31 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \url{https://github.com/afarah18/spectral-sirens-with-GPs}.% \GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
+The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
 

### 2024-03-31 src/scripts/bias_study.py
add nonparametric analysis to bias study:  - get all summary stats from there, incl parametric ones  - double number of events used in each catalog  - basically the full analysis is done in this script now

@@ -10,7 +10,7 @@ from mock_posteriors import gen_snr_scaled_PE
 from data_generation import (true_vals_PLP, SNR_THRESH, N_SAMPLES_PER_EVENT,N_SOURCES,
                               H0_FID, OM0_FID, osnr_interp, reference_distance)
 from parametric_inference import NSAMPS
-from priors import PLP, BPL
+from priors import PLP, BPL, hyper_prior, get_ell_frechet_params, get_sigma_gamma_params
 
 jax.config.update("jax_enable_x64", True)
 
@@ -20,7 +20,7 @@ N_SOURCES = N_SOURCES#//2 # cut catalogs in half for this study, for computation
 plot = True
 
 # random number generators
-jax_rng = jax.random.PRNGKey(42)
+jax_rng = jax.random.PRNGKey(42) # these numbers are arbitrary, I think I just copied them from a tutorial
 np_rng = np.random.default_rng(516)
 
  # load injection set
@@ -52,26 +52,64 @@ for i in trange(N_CATALOGS):
                               num_chains=1,progress_bar=False)   
     mcmc.run(jax_rng,**kwargs)
 
-    id = az.from_numpyro(mcmc)
-    id.to_netcdf(paths.data / f"bias/mcmc_parametric_PLP_{i}.nc4")
-    bias_PLP[i] = (id.posterior['H0'][0].mean() - H0_FID)/id.posterior['H0'][0].std()
+    id_PLP = az.from_numpyro(mcmc)
+    id_PLP.to_netcdf(paths.data / f"bias/mcmc_parametric_{i}.nc4")
+    bias_PLP[i] = np.abs(id_PLP.posterior['H0'][0].mean() - H0_FID)/id_PLP.posterior['H0'][0].std()
     if plot:
-        plt.hist(id.posterior['H0'][0],density=True, bins=50,histtype='step',color='green',alpha=0.5,lw=0.5)
+        plt.hist(id_PLP.posterior['H0'][0],density=True, bins=50,histtype='step',color='green',alpha=0.5,lw=0.5)
+    
     # Inference - broken power law
     nuts_settings = dict(target_accept_prob=0.9, max_tree_depth=10,dense_mass=False)
     nuts_kernel = numpyro.infer.NUTS(BPL,**nuts_settings)
-    kwargs = dict(m1det=m1z_PE,dL=dL_PE, m1det_inj=m1zinj_det,dL_inj=dLinj_det,
[…]

### 2024-03-31 src/scripts/pm_H0_twopanel.py

@@ -83,5 +83,5 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     plt.clf()
 
 two_panel(paths.data / "mcmc_nonparametric.nc4", 
-          paths.data / "mcmc_parametric_PLP_fitz.nc4",
-          paths.data / "mcmc_parametric_BPL_fitz.nc4")
+          paths.data / "bias/mcmc_parametric_PLP_16.nc4",
+          paths.data / "bias/mcmc_parametric_BPL_16.nc4")

### 2024-03-31 src/scripts/priors.py
minor change (Add comment)

@@ -279,7 +279,7 @@ def BPL(m1det,dL,m1det_inj,dL_inj,log_pinj,log_PE_prior=0.,remove_low_Neff=False
 
     # mass dist
     mmin = numpyro.sample("mmin",dist.Uniform(1.0,20.0))
-    mmax = numpyro.sample("mmax",dist.Uniform(30.0,100.0))
+    mmax = numpyro.sample("mmax",dist.Uniform(30.0,100.0)) # 225
     alpha1 = numpyro.sample("alpha1",dist.Normal(0,5))
     alpha2 = numpyro.sample("alpha2",dist.Normal(0,5))
     b_m1 = numpyro.sample("b_m1",dist.Uniform(20,100))

### 2024-03-31 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 However, the use of an inaccurate representation leads to biases in the cosmological inference, an acute problem given the current uncertainties in true source population.
-Furthermore, it is commonly presumed that the exact shape of the mass distribution must be known \emph{a priori}\/ to obtain unbiased measurements of cosmological parameters in this fashion.
+Furthermore, it is commonly presumed that the form of the mass distribution must be known \emph{a priori}\/ to obtain unbiased measurements of cosmological parameters in this fashion.
 Here, we demonstrate that spectral sirens can accurately infer cosmological parameters without such prior assumptions.
@@
     The deliberately incorrect parametric model fails to recover the true mass distribution and therefore produces an estimate of \Ho{} that is offset from the true value, whereas both the correct and non-parametric models recover the mass distribution and \Ho{}.
-    As the true mass distribution is unknown for real observations, using a non-parametric model mitigates systematic uncertainty that would otherwise arise from mismodeling the \ac{CBC} population, though it does introduce additional statistical uncertainty.
+    As the true mass distribution is unknown for real observations, using a non-parametric model mitigates systematic uncertainty that would otherwise arise from mismodeling the \ac{CBC} population.
     }
@@
 
-The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
+The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \url{https://github.com/afarah18/spectral-sirens-with-GPs}.
+%\GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
 
-\section{Cosmology with an astrophysically-agnostic mass distribution model}
+\section{Cosmology with an astrophysically-agnostic mass model}
 \label{sec:methods}
@@
 Algorithmically, we evaluate the likelihood in Equation~\ref{eq:inhomog-poisson} by drawing \Ho{} and \Omm{} from their prior distributions (contained in $p(\Lambda)$), use them to define a relationship between luminosity distance and redshift, transform detector-frame masses to source-frame masses according to $m_{\det} /(1+z(D_L,H_0,\Omega_M)) = m_{\source}$, evaluate the source-frame mass distribution at these transformed values ($\diff N/\diff t_{\det} \diff \theta_i (\theta_i;\Lambda)$ for events and $N_{\text{exp}}(\Lambda)$ for injections), and then take the product of these terms to evaluate the expression in Equation~\ref{eq:inhomog-poisson}.
-The full process is outlined in the next subsection.
+The full process is outlined in Appendix~\ref{ap:GP}.
 
@@
 
-Practically, the difference in the inference of the population when using a \ac{GP} versus other modeling choices is that the population model ($\diff N/\diff t_{\det} \diff \theta_i (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson}) is determined directly by a realization of the \ac{GP}, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
-In other words, when using parametric models, $\diff N/\diff t_{\det} \diff \theta_i (\theta_i;\Lambda)$ is calculated by evaluating a specific functional form described by a small set of hyper-parameters.
+In this Section, we give an overview of the non-parametric mass model developed for this work.
+Further details on this model, including an introduction to \acp{GP} and a discussion of their properties is given in Appendix~\ref{ap:GP}.
+
 With the \ac{GP} approach, the hyper-parameters describing the mass distribution are the rate at each event-level posterior sample's source frame mass, and the rate at each found injection's source frame mass.
 The \ac{GP} \emph{is} $p(\Lambda)$, the prior on population parameters (except in the case of cosmological parameters, which all have uniform priors).
+This is demonstrated in Figure~\ref{fig:GP example}, where the left panel shows draws from a \ac{GP}, which are prior draws for the population inference.
+The smooth appearance of individual draws from the population prior, as well as the absence of overdensities at specific source frame mass values in the full prior distribution shown in Figure~\ref{fig:GP example} demonstrate that we have successfully fulfilled our goal to construct a model without predefined features in source-frame mass.
+Combined with the population likelihood in Equation~\ref{eq:inhomog-poisson} the prior illustrated in the left panel of Figure~\ref{fig:GP example} gives the population posterior in the right panel.
 
-% The source frame mass is different for each possible value of \Ho{} and \Omm{}, so the hyper-parameters are defined in different locations for each draw from the hyper-posterior.
-In this way, our \ac{GP}-based mass distribution is similar to the autoregressive population models used in~\citet{callister_parameter-free_2023}.
-Indeed, an autoregressive process is a \ac{GP} with a particular choice of kernel, which we describe below.
-We therefore have $N_{\text{ev}}M + N_{\text{inj}}$ mass hyper-parameters, where $M$ is the number of posterior samples per event and $N_{\text{inj}}$ is the number of injections used to calculate the selection function~\citep[see e.g.][]{vitale_inferring_2020,talbot,essick_estimating_2021}.
-We label these $\{\mathcal{R}_{ij},\mathcal{R}_k\}$, and our full set of hyper-parameters is $\Lambda=\{H_0, \Omega_M, \mathcal{R}_{ij},\mathcal{R}_k\}$. 
-As the hyper-parameters differ at each data point, we have successfully fulfilled our goal to construct a model without predefined features in source-frame mass.
-We note that other non-parametric methods may be adapted to avoid such predefined features, such as fitting for the locations of their features simultaneously with the rest of the inference \citep[e.g.][]{tiwari_vamana_2021} or, in the case of splines, by using a smoothing function that allows for features to occur at arbitrary locations appropriate smoothing \citep[e.g.][]{edelman_cover_2023}.
+The lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects at those masses, causes the \ac{GP} to learn a relatively low merger rate at the edges of the mass distribution.
+On the other hand, there is both a lack of data and little detector sensitivity at masses above $\sim100\Msun$ and below $\sim5\Msun$.
+The mass distribution is therefore uninformed in this region and the \ac{GP} reverts to its prior distribution, which resembles random scatter around the mean differential merger rate.
+Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
+% Indeed, the upper edge of the mass distribution is very informative to the \Ho{} constraint with current detections \citep{gwtc3_cosmo}.
+The combination of these two effects results in what appears as an uptick in the merger rate below $\sim5\Msun$ and above $\sim100\Msun$.
+However, this reversion to the prior is uninformative for the \Ho{} constraint and does not affect inference on cosmological parameters.
+Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,120\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
+
+We note that other non-parametric methods may be adapted to avoid predefined features, such as fitting for the locations of their features simultaneously with the rest of the inference \citep[e.g.][]{tiwari_vamana_2021} or, in the case of splines, by using a smoothing function that allows for features to occur at arbitrary locations appropriate smoothing \citep[e.g.][]{edelman_cover_2023}.
 % Such adaptations are necessary to avoid circular inference.
@@
 It defines the notion of similarity between adjacent points and thereby encodes our assumptions about the smoothness of the source frame mass distribution \citep{rasmussen_gaussian_2006}.
-We use a Mat\'ern kernel \citep{handcock_bayesian_1993, stein_interpolation_1999} with $\nu = 5/2$, but have repeated the analysis with $\nu=3/2$ and $\infty$, finding little impact on the results, except that the $\nu=\infty$ case (also called the squared exponential kernel) produces a slightly more jagged mass distribution.
-In addition to the mean, Mat\'ern kernels have two parameters that determine their properties: a length scale $l$ and a variance $s$.
+Kernels themselves have parameters that determine their properties.
 In our use case, these are one level further removed from hyper-parameters, so we adopt the terminology used in \citet{callister_parameter-free_2023} and call them ``hyper-hyper-parameters.''
 We fit these hyper-hyper-parameters along with the hyper-parameters $\Lambda$ to minimize prior assumptions about the form of the mass distribution.
-We use penalized-complexity priors on the hyper-hyper-parameters to enforce that the model does not create small-scale structure uninformed by data, thereby avoiding over-fitting \citep{simpson_penalising_2017,simpson_garcpas_2022}.
-Explicitly, the priors on $l$ and $s$ are Fr\'echet and Gamma distributions, respectively, and are defined to have less than 5\% support for correlation lengths smaller than the average spacing between event-level posterior means.
 
-The time to evaluate a \ac{GP} is notorious for scaling as the cube of the number of data points, making \acp{GP} unwieldy with large data sets, such as the $\mathcal{O}(10^9)$ posterior samples and software injections expected for \ac{O5}.
-We therefore make two approximations to a full \ac{GP} to increase computational efficiency.
-First, for each likelihood evaluation, we evaluate a full \ac{GP} on a regular grid between $0.1\Msun$ and $250\Msun$ and then interpolate it at each data point.
-Second, we use the quasi-separability of Mat\'ern kernels to analytically perform the transformation between covariance matrix and \ac{GP} draw.
-This second step is done using the \texttt{QuasisepSolver} module \citep{foreman-mackey_fast_2017} in the \texttt{tinygp} code base \citep{foreman-mackey_tinygp_2021}, and requires data to be sortable (i.e.
-one-dimensional).
-
-Algorithmically, each posterior evaluation contains the following steps: \variable{output/priors_placeholder.txt}
-\begin{enumerate}
-\script{nonparametric_inference.py}
-    \item Draw cosmological parameters \Ho{} and \Omm{} from uniform prior distributions.
-    \item Convert the luminosity distances and detector-frame masses of each event posterior sample to redshifts and source-frame masses according to the cosmology specified by step 1.
-    \item Draw hyper-hyper-parameters $l$ and $s$ from the penalized-complexity priors described above.
-    \item Draw a single \ac{GP} realization with a kernel defined by $l$ and $s$.
-    This is defined on a regular grid of source-frame masses and evaluated using the \texttt{QuasisepSolver} in \texttt{tinygp}.
-    \item Interpolate the \ac{GP} at each event posterior sample and injection source-frame mass (from step 3).
-    \item Calculate the population likelihood according to Equation~\ref{eq:inhomog-poisson}.
-\end{enumerate}
-% TODO: \jme{[comment: is there a way to summarize this list in a simple diagram?]}
-We perform these steps within \texttt{numpyro} \citep{bingham_pyro_2019,phan_composable_2019}, sampling the posterior using the no-u-turn sampler for Hamiltonian Monte Carlo \citep{hoffman_no-u-turn_2011}.
+\begin{figure}
+    \centering
+    \includegraphics{figures/GP_example.pdf}
+    \caption{Draws from the \acf{GP} used to model the mass distribution.
+    The left panel shows prior draws from the \ac{GP} and the right panel shows the posterior draws once the population inference is performed on the simulated data.
+    The posterior draws in the right panel are a subset of those used to create the 90\% credible intervals in Figure~\ref{fig:O5_GP}.}
+    \label{fig:GP example}
+    \script{GP_example_plot.py}
+\end{figure}
 
@@
 This bias is not due to the need to know the morphology of the mass distribution \emph{a priori}, as the \ac{GP}-based model recovers the correct value of \Ho{} despite making minimal assumptions about the mass distribution.
-The \ac{GP}-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters. 
-In reality, we do not know the true functional form of the mass distribution, so the additional statistical uncertainty introduced by non-parametric approaches may be desirable over the possible systematic errors introduced by choosing a parametric model that likely does not contain the true mass distribution within its hyper-prior.
-
-% The lack of data just below the minimum black hole mass and just above the maximum black hole mass, combined with the fact that \ac{GW} detectors are sensitive to objects at those masses, causes the \ac{GP} to learn a relatively low merger rate at the edges of the mass distribution.
-% On the other hand, there is both a lack of data and little detector sensitivity at masses above $\sim100\Msun$ and below $\sim5\Msun$.
-% The mass distribution is therefore uninformed in this region and the \ac{GP} reverts to its prior distribution, which resembles random scatter around the mean differential merger rate.
-% Similar effects can be seen in other non-parametric methods \citep{edelman_cover_2023, callister_parameter-free_2023}.
-% % Indeed, the upper edge of the mass distribution is very informative to the \Ho{} constraint with current detections \citep{gwtc3_cosmo}.
-% The combination of these two effects results in what appears as an uptick in the merger rate below $\sim5\Msun$ and above $\sim100\Msun$.
-% However, this reversion to the prior is uninformative for the \Ho{} constraint and does not affect inference on cosmological parameters.
-% Additionally, the posterior on \Ho{} is distinct from its prior distribution (uniform in the range $[30\Hunits,120\Hunits]$, indicating that the data is informative despite the flexibility of the population model.
+% The \ac{GP}-based model does, however, obtain larger uncertainties on the inferred value of \Ho, as it has many more free parameters. 
+In reality, we do not know the true functional form of the mass distribution, %so the additional statistical uncertainty introduced by non-parametric approaches may be desirable over the possible systematic errors introduced by choosing a parametric model that likely does not contain the true mass distribution within its hyper-prior.
+so it may be desirable to use a non-parametric approach to avoid potential systematic errors introduced by choosing a parametric model that likely does not contain the true mass distribution within its hyper-prior.
 
@@
 \subsection{Projections for Future Measurements}
-Figure~\ref{fig:O5_GP} demonstrates an expected $\variable{output/nonparh0percent.txt}\%$ ($1\sigma$ uncertainties) measurement of \Ho{} after one year of \ac{O5} using the \ac{GP}-based spectral siren method.
-However, by the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\approx 3$, with most sources lying near redshift $\simeq 1.2$ \citep{chen_distance_2021}.
+Figure~\ref{fig:O5_GP} demonstrates an expected $\variable{output/nonparh0percent.txt}\%$ ($1\sigma$ uncertainties) measurement of \Ho{} after one year of \ac{O5} using the \ac{GP}-based spectral siren method, and a $\variable{output/PLPh0percent.txt}\%$ measurement with parametric spectral sirens, demonstrating comparable statistical uncertainties.
+By the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\approx 3$, with most sources lying near redshift $\simeq 1.2$ \citep{chen_distance_2021}.
 Additionally, next-generation detectors will be sensitive to sources up to redshift $\sim 100$.
@@
 
-We predict a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} with one year of observing at the \ac{LVK} network's design sensitivity using the non-parametric spectral siren approach, and a $\variable{output/PLPh0percent.txt}\%$ measurement with parametric spectral sirens.
-% Discussion on this! why are parametric and nonparametric comparable?? 
-However, the true population of merging compact binaries in the Universe is unknown because neither the mechanism from which compact binaries were formed, nor the physical processes within the mechanism, have been conclusively determined.
-The true functional form of the mass distribution therefore remains elusive, and systematic uncertainties arising from an incorrect choice for the form of the mass distribution are inevitable.
+Neither the mechanism from which compact binaries were formed, nor the physical processes within the mechanism, have been conclusively determined.
+The true functional form of the mass distribution therefore remains elusive, meaning that systematic uncertainties arising from an incorrect choice for the form of the mass distribution are inevitable.
 With current observations, these systematic effects are likely smaller than statistical uncertainties.
 However, next-generation detectors will herald sufficient \ac{GW} observations to substantially decrease statistical uncertainty in these measurements; for example, \citet{pierra_study_2023} show that incorrect assumptions about the shape of the mass distribution can lead to $\sim3\sigma$ systematic biases in \Ho{} with catalogs of 2,000 events, although this bias may be an overestimate as it does not include measurement uncertainty of the \ac{GW} parameters.
-Thus, the additional statistical uncertainty introduced by non-parametric approaches may be preferable to the systematic errors associated with choosing a parametric model.
+Thus, non-parametric approaches may be preferable to avoid the systematic errors associated with choosing a parametric model.
 
@@
 We have shown this by simultaneously measuring \Omm{} and \Ho{}; the method can be trivially expanded to constrain additional cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$.
-We find that for \acf{O5}, spectral sirens will be most constraining at $z=\variable{output/mostsensitivez.txt}$.
+We find that for \acl{O5}, spectral sirens will be most constraining at $z=\variable{output/mostsensitivez.txt}$.
 This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers~\citep{kiendrebeogo_updated_2023}, implying that upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances by \acp{GW}.
@@
     The Tycho supercomputer hosted at the SCIENCE HPC center at the University of Copenhagen was used for supporting this work.
+    %LIGO
+    This material is based upon work supported by NSF's LIGO Laboratory which is a major facility fully funded by the National Science Foundation.
 \end{acknowledgments}
@@
 
-% \section{\Acf{GP}-based mass distribution}
-% \label{ap:GP}
+\section{\Acl{GP}-based mass distribution}
+\label{ap:GP}
+
+In this Section, we discuss the properties of the \ac{GP}-based mass distribution and describe our modeling choices in more detail.
+% we also show how it is able to recover a wide range of morphologies and is not tuned to the PLP mass distribution.
+
+Practically, the difference in the inference of the population when using a \ac{GP} versus other modeling choices is that the population model ($\diff N/\diff t_{\det} \diff \theta_i (\theta_i;\Lambda)$ in Equation~\ref{eq:inhomog-poisson}) is determined directly by a realization of the \ac{GP}, rather than by a handful of hyper-parameters $\Lambda$ and evaluated on an analytical function.
+In other words, when using parametric models, $\diff N/\diff t_{\det} \diff \theta_i (\theta_i;\Lambda)$ is calculated by evaluating a specific functional form described by a small set of hyper-parameters.
+With the \ac{GP} approach, the hyper-parameters describing the mass distribution are the rate at each event-level posterior sample's source frame mass, and the rate at each found injection's source frame mass.
+The \ac{GP} \emph{is} $p(\Lambda)$, the prior on population parameters (except in the case of cosmological parameters, which all have uniform priors).
+
+Because the \ac{GP} is defined only at specific data points, we have $N_{\text{ev}}M + N_{\text{inj}}$ mass hyper-parameters, where $M$ is the number of posterior samples per event and $N_{\text{inj}}$ is the number of injections used to calculate the selection function~\citep[see e.g.][]{vitale_inferring_2020,talbot,essick_estimating_2021}.
+In this way, our \ac{GP}-based mass distribution is similar to the autoregressive population models used in~\citet{callister_parameter-free_2023}.
+Indeed, an autoregressive process is a \ac{GP} with a specific choice of kernel.
+
+The kernel is a function that defines the covariance between input points in the \ac{GP} (in our case, two source frame mass values). 
+It defines the notion of similarity between adjacent points and thereby encodes our assumptions about the smoothness of the source frame mass distribution \citep{rasmussen_gaussian_2006}.
+We use a Mat\'ern kernel \citep{handcock_bayesian_1993, stein_interpolation_1999} with $\nu = 5/2$, but have repeated the analysis with $\nu=3/2$ and $\infty$, finding little impact on the results, except that the $\nu=\infty$ case (also called the squared exponential kernel) produces a slightly more jagged mass distribution.
+In addition to the mean, Mat\'ern kernels have two parameters that determine their properties: a length scale $l$ and a variance $s$.
+In our use case, these are one level further removed from hyper-parameters, so we adopt the terminology used in \citet{callister_parameter-free_2023} and call them ``hyper-hyper-parameters.''
+We fit these hyper-hyper-parameters along with the hyper-parameters $\Lambda$ to minimize prior assumptions about the form of the mass distribution.
+We use penalized-complexity priors on the hyper-hyper-parameters to enforce that the model does not create small-scale structure uninformed by data, thereby avoiding over-fitting \citep{simpson_penalising_2017,simpson_garcpas_2022}.
+Explicitly, the priors on $l$ and $s$ are Fr\'echet and Gamma distributions, respectively, and are defined to have less than 5\% support for correlation lengths smaller than the average spacing between event-level posterior means.
+
+The time to evaluate a \ac{GP} is notorious for scaling as the cube of the number of data points, making \acp{GP} unwieldy with large data sets, such as the $\mathcal{O}(10^9)$ posterior samples and software injections expected for \ac{O5}.
+We therefore make two approximations to a full \ac{GP} to increase computational efficiency.
+First, for each likelihood evaluation, we evaluate a full \ac{GP} on a regular grid between $0.1\Msun$ and $250\Msun$ and then interpolate it at each data point.
+Second, we use the quasi-separability of Mat\'ern kernels to analytically perform the transformation between covariance matrix and \ac{GP} draw.
+This second step is done using the \texttt{QuasisepSolver} module \citep{foreman-mackey_fast_2017} in the \texttt{tinygp} code base \citep{foreman-mackey_tinygp_2021}, and requires data to be sortable (i.e.
+one-dimensional).
+
+Algorithmically, each posterior evaluation contains the following steps: \variable{output/priors_placeholder.txt}
+\begin{enumerate}
+\script{nonparametric_inference.py}
+    \item Draw cosmological parameters \Ho{} and \Omm{} from uniform prior distributions.
+    \item Convert the luminosity distances and detector-frame masses of each event posterior sample to redshifts and source-frame masses according to the cosmology specified by step 1.
+    \item Draw hyper-hyper-parameters $l$ and $s$ from the penalized-complexity priors described above.
+    \item Draw a single \ac{GP} realization with a kernel defined by $l$ and $s$.
+    This is defined on a regular grid of source-frame masses and evaluated using the \texttt{QuasisepSolver} in \texttt{tinygp}.
+    \item Interpolate the \ac{GP} at each event posterior sample and injection source-frame mass (from step 3).
+    \item Calculate the population likelihood according to Equation~\ref{eq:inhomog-poisson}.
+\end{enumerate}
+% TODO: \jme{[comment: is there a way to summarize this list in a simple diagram?]}
+We perform these steps within \texttt{numpyro} \citep{bingham_pyro_2019,phan_composable_2019}, sampling the posterior using the no-u-turn sampler for Hamiltonian Monte Carlo \citep{hoffman_no-u-turn_2011}.
+This can be seen explicitly in the source code accompanying this paper, in the \texttt{scripts/priors.py} \variable{output/priors_placeholder.txt}script.
 
@@
 %     \item fit of GP to various mass dist morphologies, showing how flexible it is (fixing H0)
-%     \item zoom-in on GP traces 
-%     \item details from section 2.3
-%     \item discussion of reversion to prior at extrema of mass dist (move from section 3)
 % \end{itemize}

### 2024-03-31 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \url{https://github.com/afarah18/spectral-sirens-with-GPs}.
-%\GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
+The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
 
@@
 
-To obtain a quantitative measure of the systematic bias introduced by mismodeling the mass distribution, we repeat the parametric analyses with 50 separate simulated catalogs of $\sim500$ events each.
+To obtain a quantitative measure of the systematic bias introduced by mismodeling the mass distribution, we repeat the parametric analyses with 50 separate simulated catalogs of $\sim1,000$ events each.
 We find that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time,

### 2024-04-02 src/scripts/bias_study.py
minor aesthetic changes

@@ -53,7 +53,7 @@ for i in trange(N_CATALOGS):
     mcmc.run(jax_rng,**kwargs)
 
     id_PLP = az.from_numpyro(mcmc)
-    id_PLP.to_netcdf(paths.data / f"bias/mcmc_parametric_{i}.nc4")
+    id_PLP.to_netcdf(paths.data / f"bias/mcmc_parametric_PLP_{i}.nc4")
     bias_PLP[i] = np.abs(id_PLP.posterior['H0'][0].mean() - H0_FID)/id_PLP.posterior['H0'][0].std()
     if plot:
         plt.hist(id_PLP.posterior['H0'][0],density=True, bins=50,histtype='step',color='green',alpha=0.5,lw=0.5)

### 2024-04-02 src/scripts/pm_H0_twopanel.py

@@ -35,9 +35,9 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     #     axes[0].plot(TEST_M1S, r[i]/TEST_M1S, lw=0.1, c=color_GP,alpha=0.03)
     #     axes[0].plot(TEST_M1S, r_PLP[i],lw=0.1, c=color_PLP,alpha=0.03)
     #     axes[0].plot(TEST_M1S, r_BPL[i],lw=0.1, c=color_BPL,alpha=0.03)
-    axes[0].plot(TEST_M1S,np.percentile(r/TEST_M1S,(5,95),axis=0).T,lw=0.2, c=color_GP,alpha=0.5)
+    axes[0].plot(TEST_M1S,np.percentile(r/TEST_M1S,(5,95),axis=0).T,lw=0.2, c=color_GP,alpha=0.7)
     # axes[0].plot(TEST_M1S,np.mean(r/TEST_M1S,axis=0),lw=1.5, c=color_GP)
-    axes[0].fill_between(TEST_M1S,*np.percentile(r/TEST_M1S,(5,95),axis=0), color=color_GP_light,alpha=0.3)
+    axes[0].fill_between(TEST_M1S,*np.percentile(r/TEST_M1S,(5,95),axis=0), color=color_GP_light,alpha=0.5)
     
     axes[0].plot(TEST_M1S,np.percentile(r_PLP,(5,95),axis=0).T, c=color_PLP,ls="--")
     # axes[0].plot(TEST_M1S,np.mean(r_PLP,axis=0),lw=2, c=color_PLP,ls="--")

### 2024-04-02 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2024-03-31 src/tex/ms.tex
@@
 
-\author{Daniel E. Holz}
+\author[0000-0002-0175-5064]{Daniel E. Holz}
 \affiliation{Department of Physics, University of Chicago, Chicago, IL 60637, USA}
@@
 Electromagnetic information about \ac{GW} source redshifts need not be available in order to use them as standard sirens, however.
-\ac{GW} signals provide direct measurements of each source's luminosity distance, $d_L$, and redshifted (detector frame), masses, $m_{\det}= m_{\source}(1+z)$ \citep[e.g.][]{chen_mass-redshift_2019}.
+\ac{GW} signals provide direct measurements of each source's luminosity distance, $D_L$, and redshifted (detector frame) masses, $m_{\det}= m_{\source}(1+z)$ \citep[e.g.][]{chen_mass-redshift_2019}.
 Therefore, if the source frame mass is known, each \ac{GW} signal provides a direct mapping between luminosity distance and redshift, allowing for a measurement of the expansion of the universe at the time the \ac{GW} signal was emitted, $H(z)$.
@@
     \includegraphics[width=\textwidth]{figures/O5_pm.pdf}
-    \caption{Spectral siren measurement for a simulated catalog with the correct parametric model (\plp, green), a deliberately incorrect parametric model (broken power law, orange) and the non-parametric model presented in this work (\acl{GP}, blue).
+    \caption{Spectral siren measurement for a simulated catalog with the correct parametric model (\plp, green dashed), a deliberately incorrect parametric model (\textsc{Broken Power Law}, pink solid) and the non-parametric model presented in this work (\acl{GP}, blue filled).
     The left panel shows the recovered source frame primary mass distribution for each model, and the simultaneously-inferred posteriors on \Ho{} are shown in the right panel.
+    The bands in the left panel represent the 90\% credible intervals.
     The mass distribution and \Ho{} value used to generate the data are shown by a solid black line in each panel. 
@@
 However, first principles models for mass distributions of merging compact binaries are not available: significant theoretical uncertainties exist about the shapes, locations, and very existence of predicted features in the mass distribution~\citep[e.g.][]{ zevin_constraining_2017, mapelli_binary_2020,2021ApJ...910..152Z,marchant_evolution_2023}.
-This includes uncertainties about potentially dominant features, such as the existence of a ``pile-up'' due to pulsational-pair instability, a pair-instability-driven upper mass gap~\citep{farmer}, the maximum neutron star mass~\citep{fryer_theoretical_2001, alsing_evidence_2018}, and the existence of a putative lower mass gap between neutron stars and black holes~\citep{ozel_black_2010, farr_mass_2011, farah_bridging_2022}.
+This includes uncertainties about potentially dominant features, such as the existence of a ``pile-up'' due to pulsational-pair instability, a pair-instability-driven upper mass gap~\citep{farmer_mind_2019}, the maximum neutron star mass~\citep{fryer_theoretical_2001, alsing_evidence_2018}, and the existence of a putative lower mass gap between neutron stars and black holes~\citep{ozel_black_2010, farr_mass_2011, farah_bridging_2022}.
 To this end, spectral siren cosmology relies on \textit{simultaneously} measuring a source-frame mass spectrum alongside cosmological parameters.
@@
 We highlight measurements of \Ho{} in order to benchmark the accuracy and precision of our non-parametric method, as well as explore the role of spectral sirens in elucidating the Hubble tension.
-However, the primary utility of spectral siren measurements will be in constraining $H(z)$ at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{ezquiaga_spectral_2022, Chen:2024gdn, You:2020wju}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
+However, a primary utility of spectral siren measurements will be in constraining $H(z)$ at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{You:2020wju, Ezquiaga:2020tns, ezquiaga_spectral_2022, Chen:2024gdn}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
 When applied to data from next-generation detectors, this method will measure the dark energy equation of state parameter, $w$, at redshifts $\gtrsim 2$.
@@
 
-The spectral siren method of measuring cosmological parameters relies on the assumption that either all compact binaries follow the same mass distribution at all redshifts, or that any redshift evolution of the mass distribution does not perfectly mimic cosmology \citep[i.e.
-all features do not simultaneously exhibit an {\em identical}\/ monotonic rightward shift with redshift;][]{ezquiaga_spectral_2022}.
-Specifically, the method identifies the preferred relationship between luminosity distance and redshift that causes all masses to follow this common source-frame distribution.
-We do not need to know this source-frame distribution in advance, but, given some model for the black hole mass distribution, can infer it simultaneously with the cosmological expansion.
-
+The spectral siren method functions by identifying the relationship between luminosity distance and redshift that causes all source frame masses to follow a distribution that smoothly varies as a function of time.
+This allows for the simultaneous inference of both a mass distribution and cosmological parameters, even if the form of the source-frame mass distribution is not known in advance.
 As described above, though, strongly parametrized models for the mass distribution yield biased measurements of cosmological parameters if they poorly approximate the true mass distribution of compact binaries.
 We aim to circumvent such biases and instead model the population of \acp{GW} sources in a flexible and astrophysically-agnostic way.
-There exist several non-parametric methods developed for this purpose \citep{tiwari_vamana_2021,edelman_aint_2022,sadiq_flexible_2022,rinaldi_hdpgmm_2022,edelman_cover_2023,mandel_extracting_2019,ray_non-parametric_2023}.
+
+There exist several non-parametric methods developed for this purpose \citep{tiwari_vamana_2021,edelman_aint_2022,sadiq_flexible_2022,rinaldi_hdpgmm_2022,edelman_cover_2023,mandel_extracting_2019,ray_non-parametric_2023,callister_parameter-free_2023}.
 While well-suited to infer the \ac{GW} source population with a fixed cosmology, several of these methods employ fixed features in source frame mass, such as bin edges \citep{mandel_extracting_2019,ray_non-parametric_2023} or spline nodes \citep{edelman_aint_2022}.
@@
 For this purpose, we construct a model with a \acf{GP}, a common tool for non-parametric inference. 
-\Acp{GP} are random processes for which any linear combination of outcomes are Gaussian distributed \citep{rasmussen_gaussian_2006}.
+\Acp{GP} define a random space of functions in which any subset of function values are jointly Gaussian-distributed \citep{rasmussen_gaussian_2006}.
 Their smoothness properties make them widely useful in \ac{GW} data analysis for regression problems, such as modeling time-domain waveforms \citep{doctor_statistical_2017, huerta_eccentric_2018} and the neutron star equation of state \citep{landry_nonparametric_2019}, density estimation problems, such as estimating posterior densities of single-event parameters from parameter estimation samples ~\citep{demilio_density_2021}, and as a prior on histogram bin heights for population inference \citep{mandel_model-independent_2017, li_flexible_2021, ray_non-parametric_2023}.
@@
 we utilize a \ac{GP} as a prior on the functions that describe the primary mass distribution of \acp{CBC}.
-This choice adds very little prior information about the shape of the mass distribution, besides enforcing that it must be smooth.
+This choice encodes very little prior information about the shape of the mass distribution, besides enforcing that it must be smooth.
 
@@
 %% TRANSITION: Describe how our goal is to demonstrate the utility of GPs and cosmology by constructing and analyzing a simulated catalog. 
-To demonstrate the effectiveness of our \ac{GP}-based mass distribution in inferring both cosmological parameters and the population of \ac{GW} sources without introducing features or biases driven by prior assumptions, we apply our methodology to a simulated dataset.
+To demonstrate the effectiveness of our \ac{GP}-based mass distribution in agnostically inferring both cosmological parameters and the population properties of \ac{GW} sources, we apply our methodology to a simulated dataset.
 By generating a catalog of \ac{GW} sources from a known population and cosmological model, we are able to quantify the accuracy of our method.
@@
 We design our simulated catalog to match the characteristics of the data expected from one year of observation in \ac{O5}. 
-The \acp{BBH} in this catalog are drawn from an underlying population described by the \plp{} mass distribution presented in \citet{talbot_measuring_2018} and used in \citet{abbott_binary_2019, abbott_population_2023}, as well and the redshift distribution presented in \citet{callister_shouts_2020}, with hyperparameters consistent with those found in \citet{abbott_population_2023}.
-We assume the mass distribution has no redshift evolution for the redshift range to which the \ac{O5} detectors will be sensitive. 
-This assumption is consistent with current data \citep{fishbach_when_2021,van_son_redshift_2022,abbott_population_2023}, but will have to be revisited in future observing runs.
-If we had constructed a catalog with an evolving mass distribution, we would have needed to correspondingly allow for redshift evolution in our population inference.
-We leave a full treatment of redshift evolution for future work, as discussed in Section~\ref{sec:discussion}.
+The \acp{BBH} in this catalog are drawn from an underlying population described by the \plp{} mass distribution presented in \citet{talbot_measuring_2018} and used in \citet{abbott_binary_2019, abbott_population_2023}, and follow the redshift distribution presented in \citet{callister_shouts_2020}, with hyperparameters consistent with those found in \citet{abbott_population_2023}.
+We assume the mass distribution does not evolve across the redshift range to which the \ac{O5} detectors will be sensitive. 
+This assumption is consistent with current data \citep{fishbach_when_2021,van_son_redshift_2022,abbott_population_2023}, in which no redshift evolution of the black hole mass function is detected.
+At the same time, a redshift-dependent mass function is a generic astrophysical prediction, due either to changing evolutionary environments or evolving mixture fractions between distinct compact binary formation channels~\citep{neijssel_effect_2019,van_son_redshift_2022,2024arXiv240114837T,2024arXiv240212444Y}.
+We discuss this possibility further in Section~\ref{sec:discussion}, but leave it primarily for future work.
+
 We use the \texttt{GWMockCat}~ \citep{farah_things_2023} package to apply \ac{O5}-like selection effects to the drawn \acp{BBH}, generate realistic measurement uncertainty, and produce sensitivity estimates that are consistent with the simulated \ac{GW} signals.
@@
 This process results in a catalog of $N_{\text{ev}} = \variable{output/num_found_events.txt}$ \ac{GW} signals that pass the criteria for detection, hereon called events.
-
 Additional details of the data simulation, including the form of the injected population, are described in Appendix~\ref{ap:data generation}.
@@
 \end{equation}
-Here, $\Lambda$ is the set of population parameters hyper-parameters.
-$\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ is the detector frame merger rate density of \acp{BBH} conditioned on hyper-parameters $\Lambda$, this quantity is also referred to as the population model.
+Here, $\Lambda$ is the set of population parameters hyper-parameters and
+$\frac{\diff N}{\diff t_{\det} \diff \theta} (\theta;\Lambda)$ is the detector frame merger rate density of \acp{BBH}, conditioned on hyper-parameters $\Lambda$. 
 Following \citet{callister_parameter-free_2023}, we use a semicolon to explicitly indicate that this is a function of $\Lambda$, not a density over $\Lambda$.
 $N_{\text{exp}}(\Lambda)$ is the expected number of detections given $\Lambda$ and the \ac{GW} detector sensitivity, and is calculated using a Monte Carlo sum over $N_{\text{inj}}$ found signals injected into the data stream \citep[see][for a detailed explanation of this process]{essick_estimating_2021, essick_precision_2022}.
-In this work, we have restricted our analysis to the \ac{BBH} mass distribution.
+
+In this work, we have restricted our analysis to the \ac{BBH} primary mass distribution.
 However, the method can be trivially extended to the full mass distribution of \acp{CBC} \citep[e.g.][]{fishbach_does_2020, ezquiaga_spectral_2022}.
 
-Since the event parameters are not perfectly measured, we incorporate uncertainty to marginalize over the possible parameters associated with each event.
+Since parameters of individual events are not perfectly measured, we marginalize over the possible properties of each event.
 Practically, this is done by a Monte Carlo average over the posterior samples $\{\theta_j\}_i$ of each event $i$ and dividing out the prior used when inferring those posterior samples, $\pi_{\rm PE}(\theta)$:
@@
 \end{equation}
-This provides a posterior on both the compact binary population and the background cosmology when combined with a prior $p(\Lambda)$ on the population and cosmological parameters, as discussed below.
-We will use the term hyper-prior to refer to $p(\Lambda)$ for the remainder of this work.
-
-The likelihood in Equation~\eqref{eq:inhomog-poisson} is maximized when the relationship between redshift and luminosity distance makes all source-frame masses follow the same distribution.
-This is the basis of the spectral siren method of measuring cosmological parameters; we can simultaneously infer the relationship between redshift and luminosity distance with the parameters of the source-frame mass distribution.
-The full set of hyper-parameters therefore includes the cosmological parameters that dictate the $D_L$--$z$ relation: the local expansion rate \Ho, the present fractional energy densities of dark matter \Omm, dark energy $\Omega_\Lambda$, and radiation $\Omega_r$, and the equation of state of dark energy $w$.
+When combined with a prior $p(\Lambda)$ (to be discussed below) on the population and cosmological parameters, the result is a posterior on both the compact binary population and the background cosmology.
+The full set of hyper-parameters $\Lambda$ therefore includes the shape of the mass distribution as well as all cosmological parameters that dictate the $D_L$--$z$ relation: the local expansion rate \Ho, the present fractional energy densities of dark matter \Omm, dark energy $\Omega_\Lambda$, and radiation $\Omega_r$, and the equation of state of dark energy $w$.
 In this work, we fix $\Omega_\Lambda=1-\Omega_M, \Omega_r=0$ and $w=-1$ and use uniform priors on \Ho{} and \Omm{}, corresponding to a flat $\Lambda$CDM cosmology.
 
-Algorithmically, we evaluate the likelihood in Equation~\ref{eq:inhomog-poisson} by drawing \Ho{} and \Omm{} from their prior distributions (contained in $p(\Lambda)$), use them to define a relationship between luminosity distance and redshift, transform detector-frame masses to source-frame masses according to $m_{\det} /(1+z(D_L,H_0,\Omega_M)) = m_{\source}$, evaluate the source-frame mass distribution at these transformed values ($\diff N/\diff t_{\det} \diff \theta_i (\theta_i;\Lambda)$ for events and $N_{\text{exp}}(\Lambda)$ for injections), and then take the product of these terms to evaluate the expression in Equation~\ref{eq:inhomog-poisson}.
-The full process is outlined in Appendix~\ref{ap:GP}.
+The process by which we sample the likelihood in Equation~\ref{eq:inhomog-poisson} is outlined in Appendix~\ref{ap:GP}.
 
@@
 
-\begin{figure}
+\begin{figure*}
     \centering
-    \includegraphics{figures/GP_example.pdf}
+    \includegraphics[width=\textwidth]{figures/GP_example.pdf}
     \caption{Draws from the \acf{GP} used to model the mass distribution.
-    The left panel shows prior draws from the \ac{GP} and the right panel shows the posterior draws once the population inference is performed on the simulated data.
+    The left panel shows prior draws from the \ac{GP} and the right panel shows posterior draws once the population inference is performed on the simulated data.
     The posterior draws in the right panel are a subset of those used to create the 90\% credible intervals in Figure~\ref{fig:O5_GP}.}
@@
     \script{GP_example_plot.py}
-\end{figure}
+\end{figure*}
 
@@
 We emphasize, however, that the choice of cosmological parameters for data generation is arbitrary and does not impact the results, since we are concerned only with the ability of our method to recover the injected values.
-Throughout the data generation and inference, we use the approximations presented in \citet{adachi_analytical_2012} to efficiently convert between $d_L$ and $z$ for a given set of cosmological parameters \Omm{} and \Ho.
+Throughout the data generation and inference, we use the approximations presented in \citet{adachi_analytical_2012} to efficiently convert between $D_L$ and $z$ for a given set of cosmological parameters \Omm{} and \Ho.
 
@@
 
-\section{\Acl{GP}-based mass distribution}
+\section{Gaussian process-based mass distribution}
 \label{ap:GP}
@@
 
-Because the \ac{GP} is defined only at specific data points, we have $N_{\text{ev}}M + N_{\text{inj}}$ mass hyper-parameters, where $M$ is the number of posterior samples per event and $N_{\text{inj}}$ is the number of injections used to calculate the selection function~\citep[see e.g.][]{vitale_inferring_2020,talbot,essick_estimating_2021}.
+Because the \ac{GP} is defined only at specific data points, we have $N_{\text{ev}}M + N_{\text{inj}}$ mass hyper-parameters, where $M$ is the number of posterior samples per event and $N_{\text{inj}}$ is the number of injections used to calculate the selection function~\citep[see e.g.][]{vitale_inferring_2020,essick_estimating_2021}.
 In this way, our \ac{GP}-based mass distribution is similar to the autoregressive population models used in~\citet{callister_parameter-free_2023}.

### 2024-04-02 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \url{https://github.com/afarah18/spectral-sirens-with-GPs}.
-%\GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
+The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
+%\url{https://github.com/afarah18/spectral-sirens-with-GPs}.
+%
 

### 2024-04-04 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 % Define document class
-\documentclass[preprint2,linenumbers]{aastex631}
+\documentclass[preprint2]{aastex631}
 \usepackage{showyourwork}
@@
 
-\author{Thomas A. Callister}
+\author[0000-0001-9892-177X]{Thomas A. Callister}
 \affiliation{Kavli Institute for Cosmological Physics, The University of Chicago, Chicago, IL 60637, USA}
@@
 
-The situation is demonstrated in Fig.~\ref{fig:O5_GP}, in which we perform spectral siren cosmology on a simulated population of binary black holes.
+The situation is demonstrated in Figure~\ref{fig:O5_GP}, in which we perform spectral siren cosmology on a simulated population of binary black holes.
 We reconstruct the mass distribution using two parametric models, one that contains the true simulated mass distribution and one that does not.
@@
 \end{enumerate}
-For all models considered in this work, we assume the redshift distribution used to generated the data, described in Equation~\ref{eq:underlying redshift dist}.
-The results of each of these fits are shown in Figure~\ref{fig:O5_GP}.  
+For all models considered in this work, we assume the form of the redshift distribution used to generate the data, described in Equation~\ref{eq:underlying redshift dist}.
+We have performed the analysis both with a fixed redshift distribution and while simultaneously fitting for the redshift distribution and find no qualitative differences in our conclusions: fitting for the redshift distribution broadens the inferred posteriors on \Ho{} equally for all mass models, but does not affect their mean values.
+
+The results of fitting each model to the same dataset are shown in Figure~\ref{fig:O5_GP}.
 The left panel shows the inferred source-frame mass distribution for each of the considered models, and the right panel shows the corresponding posteriors on \Ho{}. 
@@
 
-In parallel with this work, \citet{magana_inprep} performed a spectral siren analysis on currently-public \ac{LVK} data using a model that employs histogram bins defined at fixed locations in source-frame mass to flexibly measure the \ac{BBH} mass distribution \citep[originally presented in][]{mandel_model-independent_2017,ray_non-parametric_2023}.
-While both the method presented in this work and that of \citet{magana_inprep} employ \acp{GP}, \citet{magana_inprep} uses \acp{GP} to modulate histogram bin heights, leading to a piecewise constant function in log mass, whereas we define the \ac{GP} at each data point, constructing a smooth function with no predefined features in source-frame mass.
+In parallel with this work, \citet{magana_inprep} performed a spectral siren analysis on public \ac{LVK} data using histogram bins defined at fixed locations in source-frame mass to flexibly measure the \ac{BBH} mass distribution \citep[as originally presented in][]{mandel_model-independent_2017,ray_non-parametric_2023}.
+Both our method and theirs employ \acp{GP}.
+In our approach, the black hole mass function is \textit{itself} described by a Gaussian process, with no predefined features.
+In \citet{magana_inprep}, on the other hand, the mass distribution is fundamentally a binned piecewise constant function, with a \ac{GP} prior governing the relative heights of these bins.
+While this binned model is flexible, sharp bin edges at predefined locations nevertheless constitute unphysical features.
+These features may yield overly optimistic constraints on cosmological parameters, possibly explaining the increase in precision that \citet{magana_inprep} find relative to the parametric analysis on the exact same dataset performed in \citet{gwtc3_cosmo}.
+\citet{magana_inprep} also include a  binned reconstruction of the \ac{BBH} merger rate as a function of redshift, which we fix or model parametrically in our own work.
+We stress, though, that this is not an intrinsic limitation of our method; one could straightforwardly adopt a second Gaussian process to govern the overall scaling of the merger rate with redshift.
+
+% While both the method presented in this work and that of \citet{magana_inprep} employ \acp{GP}, \citet{magana_inprep} uses \acp{GP} to modulate histogram bin heights, leading to a piecewise constant function in log mass, whereas we define the \ac{GP} at each data point, constructing a smooth function with no predefined features in source-frame mass.
 % Sharp bin edges may lead to optimistic constraints on cosmological parameters, especially as such features are unlikely to exist in the true mass distribution, possibly explaining the increase in precision relative to parametric methods seen in \citet{magana_inprep}.
 
+More interesting is the possibility that the black hole mass distribution itself evolves with redshift.
 As demonstrated in \citet{ezquiaga_spectral_2022}, the degeneracy between an evolving mass distribution and the expansion of the universe can be broken provided that there are multiple features present in the mass distribution, and that we do not live in a fine-tuned universe where the evolutionary effects governing the \ac{CBC} mass distribution perfectly mimick the effects of cosmological redshift.
@@
 We fit these hyper-hyper-parameters along with the hyper-parameters $\Lambda$ to minimize prior assumptions about the form of the mass distribution.
-We use penalized-complexity priors on the hyper-hyper-parameters to enforce that the model does not create small-scale structure uninformed by data, thereby avoiding over-fitting \citep{simpson_penalising_2017,simpson_garcpas_2022}.
+We use penalized-complexity priors on the hyper-hyper-parameters to enforce that the model does not create small-scale structure uninformed by data, thereby avoiding over-fitting \citep{simpson_penalising_2017, simpson_garcpas_2022}.
 Explicitly, the priors on $l$ and $s$ are Fr\'echet and Gamma distributions, respectively, and are defined to have less than 5\% support for correlation lengths smaller than the average spacing between event-level posterior means.

### 2024-04-04 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 % Define document class
-\documentclass[preprint2]{aastex631}
+\documentclass[preprint2,linenumbers]{aastex631}
 \usepackage{showyourwork}
@@
 When paired with a redshift measurement, this enables standard-siren cosmology: a Hubble diagram can be constructed to directly probe the Universe's expansion.
-This can be done in the absence of electromagnetic measurements, as features in the mass distribution of \ac{GW} sources provide self-calibrating redshift measurements without the need for a definite (bright siren) or probabilistic (dark siren) host galaxy association.
-This ``spectral siren'' technique has thus far only been applied with simple parametric representations of the mass distribution. 
+This can be done in the absence of electromagnetic measurements, as features in the mass distribution of \ac{GW} sources provide self-calibrating redshift measurements without the need for a definite or probabilistic host galaxy association.
+This ``spectral siren'' technique has thus far only been applied with simple parametric representations of the mass distribution, and theoretical predictions for features in the mass distribution are commonly presumed to be fundamental to the measurement.
 However, the use of an inaccurate representation leads to biases in the cosmological inference, an acute problem given the current uncertainties in true source population.

### 2024-04-09 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-In parallel with this work, \citet{magana_inprep} performed a spectral siren analysis on public \ac{LVK} data using histogram bins defined at fixed locations in source-frame mass to flexibly measure the \ac{BBH} mass distribution \citep[as originally presented in][]{mandel_model-independent_2017,ray_non-parametric_2023}.
+In parallel with this work, \citet{magana_hernandez_beyond_2024} performed a spectral siren analysis on public \ac{LVK} data using histogram bins defined at fixed locations in source-frame mass to flexibly measure the \ac{BBH} mass distribution \citep[as originally presented in][]{mandel_model-independent_2017,ray_non-parametric_2023}.
 Both our method and theirs employ \acp{GP}.
 In our approach, the black hole mass function is \textit{itself} described by a Gaussian process, with no predefined features.
-In \citet{magana_inprep}, on the other hand, the mass distribution is fundamentally a binned piecewise constant function, with a \ac{GP} prior governing the relative heights of these bins.
+In \citet{magana_hernandez_beyond_2024}, on the other hand, the mass distribution is fundamentally a binned piecewise constant function, with a \ac{GP} prior governing the relative heights of these bins.
 While this binned model is flexible, sharp bin edges at predefined locations nevertheless constitute unphysical features.
-These features may yield overly optimistic constraints on cosmological parameters, possibly explaining the increase in precision that \citet{magana_inprep} find relative to the parametric analysis on the exact same dataset performed in \citet{gwtc3_cosmo}.
-\citet{magana_inprep} also include a  binned reconstruction of the \ac{BBH} merger rate as a function of redshift, which we fix or model parametrically in our own work.
+These features may yield overly optimistic constraints on cosmological parameters, possibly explaining the increase in precision that \citet{magana_hernandez_beyond_2024} find relative to the parametric analysis on the exact same dataset performed in \citet{gwtc3_cosmo}.
+\citet{magana_hernandez_beyond_2024} also include a  binned reconstruction of the \ac{BBH} merger rate as a function of redshift, which we fix or model parametrically in our own work.
 We stress, though, that this is not an intrinsic limitation of our method; one could straightforwardly adopt a second Gaussian process to govern the overall scaling of the merger rate with redshift.
-
-% While both the method presented in this work and that of \citet{magana_inprep} employ \acp{GP}, \citet{magana_inprep} uses \acp{GP} to modulate histogram bin heights, leading to a piecewise constant function in log mass, whereas we define the \ac{GP} at each data point, constructing a smooth function with no predefined features in source-frame mass.
-% Sharp bin edges may lead to optimistic constraints on cosmological parameters, especially as such features are unlikely to exist in the true mass distribution, possibly explaining the increase in precision relative to parametric methods seen in \citet{magana_inprep}.
 

### 2024-04-09 src/scripts/GP_example_plot.py
changes to make SYW workflow actually run

@@ -23,7 +23,8 @@ id = az.InferenceData.from_netcdf(path_fit)
 samples = id.posterior
 r = samples['log_rate_test'][0]
 
-for i in range(25):
+n = min(25,len(r))
+for i in range(n):
     axes[1].plot(logtestm1s, r[i], lw=0.7, c=color_GP,alpha=0.5)
 
 # prior

### 2024-04-09 src/scripts/data_generation.py

@@ -1,13 +1,13 @@
 import gwcosmo
 import gwpop
 import paths
+import os
 Clight = gwcosmo.Clight
 from utils import inverse_transform_sample
 from mock_posteriors import gen_snr_scaled_PE
 
 import numpy as np
 from GWMockCat.vt_utils import draw_thetas, interpolate_optimal_snr_grid 
-print("test")
 
 # Constants
 H0_FID = 67.66
@@ -70,8 +70,11 @@ osnr_interp, reference_distance = interpolate_optimal_snr_grid(
 if  __name__ == "__main__":
     # generate data and save 
     m1s_true, zt, m1z_true, dL_true = true_vals_PLP(rng=np_rng)
-    # os.mkdir(paths.data / "gw_data")
-    np.save(paths.data / "gw_data/m1s_true_PLP.npy", dL_true)
+    try:
+        np.save(paths.data / "gw_data/m1s_true_PLP.npy", dL_true)
+    except FileNotFoundError:
+        os.mkdir(paths.data / "gw_data")
+        np.save(paths.data / "gw_data/m1s_true_PLP.npy", dL_true)
     np.save(paths.data / "gw_data/z_true_PLP.npy", zt)
     np.save(paths.data / "gw_data/m1z_true_PLP.npy",m1z_true)
     np.save(paths.data / "gw_data/dL_true_PLP.npy",dL_true)

### 2024-04-09 src/scripts/nonparametric_corner.py

@@ -108,5 +108,5 @@ def H0_Om_corner(id):
     Hz(id,ax=axsRight,save=False)
     fig.savefig(paths.figures / "O5_GP_corner.pdf")
 
-id = az.InferenceData.from_netcdf(paths.data / "mcmc_nonparametric_fitOm0_old.nc4")
+id = az.InferenceData.from_netcdf(paths.data / "mcmc_nonparametric_fitOm0.nc4")
 H0_Om_corner(id)

### 2024-04-09 src/scripts/nonparametric_inference_fitOm0.py

@@ -11,7 +11,7 @@ import numpyro
 # data
 import numpy as np
 
-NSAMPS = 20
+NSAMPS = 5
 
 jax.config.update("jax_enable_x64", True)
 # random number generators

### 2024-04-09 src/scripts/pm_H0_twopanel.py

@@ -72,8 +72,10 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     except np.linalg.LinAlgError:
         prior = H0_FID
         kde = lambda x: 1.
-    axes[1].plot(prior,kde(prior),c=color_GP,lw=1)
-    axes[1].fill_between(prior,np.zeros_like(prior),kde(prior),color=color_GP,alpha=0.3,label='Gaussian process')
+        kde_PLP = kde
+        kde_BPL = kde    
+    axes[1].plot(prior,kde(prior),c=color_GP,lw=1,label='Gaussian process')
+    # axes[1].fill_between(prior,np.zeros_like(prior),kde(prior),color=color_GP,alpha=0.3)
     axes[1].plot(prior,kde_PLP(prior),c=color_PLP,ls="--",label=r'\textsc{Power Law + Peak}')
     axes[1].plot(prior,kde_BPL(prior),c=color_BPL,label=r'\textsc{Broken Power Law}')
     axes[1].set_ylabel('posterior density')

### 2024-04-09 src/scripts/pm_H0_twopanel.py
more SYW debugging

@@ -84,6 +84,6 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     fig.savefig(paths.figures / "O5_pm.pdf")
     plt.clf()
 
-two_panel(paths.data / "mcmc_nonparametric.nc4", 
+two_panel(paths.data / "bias/mcmc_nonparametric_16.nc4", 
           paths.data / "bias/mcmc_parametric_PLP_16.nc4",
           paths.data / "bias/mcmc_parametric_BPL_16.nc4")

### 2024-04-09 src/scripts/bias_study.py
mkdir if not there

@@ -3,6 +3,7 @@ import arviz as az
 import numpy as np
 from tqdm import trange
 import jax
+import os
 
 # custom
 import paths
@@ -28,6 +29,10 @@ m1zinj_det = np.load(paths.data / "gw_data/m1zinj_det.npy")
 dLinj_det = np.load(paths.data / "gw_data/dLinj_det.npy")
 log_pinj_det = np.load(paths.data / "gw_data/log_pinj_det.npy")
 
+try:
+    os.mkdir(paths.data / "bias")
+except FileExistsError:
+    pass
 if plot:
     import matplotlib.pyplot as plt
 bias_PLP = np.zeros(N_CATALOGS)

### 2024-05-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 In this work, we explicitly demonstrate that no prior knowledge of the shape of the \ac{CBC} mass spectrum is necessary to use the spectral siren methodology. 
-We do this by inferring $H(z)$ with a flexible, non-parametric model for the mass distribution of \acp{CBC} (blue lines in Figure~\ref{fig:O5_GP}). 
+We do this by inferring $H(z)$ with a flexible, non-parametric model for the mass distribution of \acp{CBC} (blue shaded band in Figure~\ref{fig:O5_GP}). 
 This model makes minimal prior assumptions about the shape of the mass distribution, enabling it to accurately infer a wide range of morphologies and remain agnostic to the astrophysical processes that give rise to features in the mass distribution.
-Despite its flexibility, our approach is able to consistently obtain unbiased measurements of cosmological parameters, showing that nonparametric methods are not only sufficient for a spectral siren measurement, they can also mitigate systematic effects in the measurement caused by model misspecification.
-The success of the non-parametric mass model in recovering the injected cosmological parameters demonstrates that the information in the spectral siren measurement does not come from the enforcement of specific features in the mass distribution.
+Despite its flexibility, our approach is able to consistently obtain unbiased measurements of cosmological parameters, showing that non-parametric methods are not only sufficient for a spectral siren measurement, they can also mitigate systematic effects in the measurement caused by model misspecification.
+
+\new{The non-parametric mass model's ability to recover injected cosmological parameters demonstrates} that the information in the spectral siren measurement does not come from the enforcement of specific features in the mass distribution.
 Rather, it is provided by the assumption that either all \acp{CBC} follow a common mass distribution, or that any evolution of the mass distribution with redshift does not exactly mimic cosmology \citep{ezquiaga_spectral_2022}.
@@
 We find that our non-parametric model allows for a $\variable{output/nonparh0percent.txt}\%$ measurement of \Ho{} and a $\variable{output/Hz_percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ during \ac{O5}, when the detectors will reach their design sensitivity.
-%, and a \result{X}\% measurement with only \result{X} weeks of observing with a next-generation detector network.
-We highlight measurements of \Ho{} in order to benchmark the accuracy and precision of our non-parametric method, as well as explore the role of spectral sirens in elucidating the Hubble tension.
-However, a primary utility of spectral siren measurements will be in constraining $H(z)$ at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{You:2020wju, Ezquiaga:2020tns, ezquiaga_spectral_2022, Chen:2024gdn}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
-When applied to data from next-generation detectors, this method will measure the dark energy equation of state parameter, $w$, at redshifts $\gtrsim 2$.
+% \new{This is to be compared with a $\variable{output/PLPh0percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ with the correct parametric model, which does not represent a statistically-significant difference.}
+We highlight measurements of \Ho{} \new{within a $\Lambda$CDM universe} in order to benchmark the accuracy and precision of our non-parametric method, as well as explore the role of spectral sirens in elucidating the Hubble tension.
+However, a primary utility of spectral siren measurements will be in constraining $H(z)$ \new{under different cosmological models, and} at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{You:2020wju, Ezquiaga:2020tns, ezquiaga_spectral_2022, Chen:2024gdn}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
+When applied to data from next-generation detectors, this method will measure \new{all cosmological parameters that govern $H(z)$, such as the curvature parameter, $\kappa$, matter density, \Omm, dark energy density, $\Omega_\Lambda$, and dark energy equation of state parameter, $w$.}
 
@@
 
-The code used to generate all simulated data, perform the population inference, and create all figures in this paper is made publicly available at \GitHubURL{}, and was enabled by the \showyourwork package \citep{Luger2021}.
-%\url{https://github.com/afarah18/spectral-sirens-with-GPs}.
-%
+This study was carried out using the reproducibility software \href{https://github.com/showyourwork/showyourwork}{\showyourwork} \citep{Luger2021}, which leverages continuous integration to programmatically download the data from \href{https://zenodo.org/}{zenodo.org}, create the figures, and compile the manuscript.
+The icons next to each Figure caption are hyperlinks that lead to the code used to make that Figure (github icons), and to the data behind the Figure (cylindrical icons).
+Some programatically-generated numbers in the text also have these icons, which function in the same way.
+We encourage readers to click these hyperlinks to verify or reproduce the results and methods of this work.
+The git repository associated to this study is publicly available at \GitHubURL{}, which allows anyone to re-build the entire manuscript.
 
@@
 By generating a catalog of \ac{GW} sources from a known population and cosmological model, we are able to quantify the accuracy of our method.
-The use of simulated data also enables us to make projections for future datasets and safely ignore dimensions such as spin and mass ratio that do not impact cosmological measurements but would otherwise be important to simultaneously fit to avoid biases in population inference of real data~\citep{biscoveanu_sources_2021}.
+The use of simulated data also enables us to make projections for future datasets and safely ignore dimensions such as spin that do not impact cosmological measurements but would otherwise be important to simultaneously fit to avoid biases in population inference of real data, \new{as real data exhibit correlations between distance, mass, and spin measurements}~\citep{biscoveanu_sources_2021}.
 
@@
     \item \plp{}, which includes the true mass distribution within its hyper-prior,
-    \item the \textsc{Broken Power Law} model presented in \citet{abbott_population_2021}, as we do not employ a high-mass truncation, which does not include the true mass distribution within its hyper-prior, and
+    \item the \bpl{} model presented in \citet{abbott_population_2021}, as we do not employ a high-mass truncation, which does not include the true mass distribution within its hyper-prior, and 
     \item the flexible, \ac{GP}-based model described in Section~\ref{sec:model}, which is able to closely approximate the morphology of the true mass distribution, along with many other morphologies.
@@
 These provide insight into the full statistical results presented below.
-In particular, it can be seen that the broken power law (orange curve) is inconsistent with the true value of $H_0$: in the run shown in Figure~\ref{fig:O5_GP}, the true value of \Ho{} is offset from the mean of the posterior by $\variable{output/BPLh0offset.txt}\sigma$.
+In particular, it can be seen that the \bpl{} (orange curve) is inconsistent with the true value of $H_0$: in the run shown in Figure~\ref{fig:O5_GP}, the true value of \Ho{} is offset from the mean of the posterior by $\variable{output/BPLh0offset.txt}\sigma$.
 In contrast, the \plp{} and \ac{GP}-based models (green and blue curves) are consistent with the underlying truth.
@@
 Additionally, the mass distribution inferred with the \ac{GP}-based model closely resembles the true, simulated distribution.
+\new{The \plp{} fit appears to be inconsistent with the true distribution at low masses, but this inconsistency is not observed in the majority of fits to other realizations of simulated catalogs, so is likely a statistical fluctuation in the particular catalog arbitrarily chosen for this plot.}
 This indicates that using models that cannot accurately approximate the true mass distribution will lead to a significant systematic bias in the estimation of cosmological parameters.
@@
 To obtain a quantitative measure of the systematic bias introduced by mismodeling the mass distribution, we repeat the parametric analyses with 50 separate simulated catalogs of $\sim1,000$ events each.
-We find that the broken power law model produces an overestimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time,
+We find that the \bpl{} model produces an \new{over- or under}estimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time,
 and the \plp{} model reaches the same level of bias only $\variable{output/PLP_bias_percent.txt}\%$ 
 of the time.
+\new{Additionally, the \bpl{} model typically overestimates \Ho{}, while the \plp{} model produces a roughly equal number of over- and underestimates of \Ho{}, implying a systematic bias in the incorrect but not the correct parametric model.}\footnote{\new{
+We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.}}
 
@@
 Figure~\ref{fig:O5_GP} demonstrates an expected $\variable{output/nonparh0percent.txt}\%$ ($1\sigma$ uncertainties) measurement of \Ho{} after one year of \ac{O5} using the \ac{GP}-based spectral siren method, and a $\variable{output/PLPh0percent.txt}\%$ measurement with parametric spectral sirens, demonstrating comparable statistical uncertainties.
+\new{However, we note that the precision reached in \ac{O5} may lessen depending on the actual level of measurement uncertainty in individual events, and the existence (or lack thereof) of a maximum mass feature, which we have assumed to be present in our simulated dataset.}
+\new{These numbers are estimated from a fit to a single simulated catalog, but we find similar levels of statistical uncertainty from fits to different catalog realizations.}
 By the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\approx 3$, with most sources lying near redshift $\simeq 1.2$ \citep{chen_distance_2021}.
@@
 The first condition (multiple features) is known to be met in current data \citep{abbott_population_2023}, with three robust features \citep{farah_things_2023}: a maximum black hole mass and overdensities at $\sim10\Msun$ and $\sim35\Msun$.
+\new{It is also predicted by many population synthesis studies, \citep[e.g.][]{mapelli_cosmic_2018,giacobbo_merging_2018,van_son_locations_2023}}
 The latter scenario---with features identically and monotonically shifting to higher masses with increasing redshift---would be astrophysically unlikely as the locations of features in the mass distribution are each thought to be governed by fundamentally different physical processes \citep{mapelli_binary_2020}.
-The non-parametric methods presented here will allow for arbitrary redshift evolution.
+Extensions to the non-parametric methods presented here will allow for arbitrary redshift evolution.
 Additionally, since \aclp{GP} naturally scale to multiple data dimensions \citep{rasmussen_gaussian_2006}, the method presented here can easily be generalized to fit the redshift dependence of the mass distribution.
 Future work will extend the method developed here to mass distributions that are allowed to evolve with redshift.
+\new{When applying an evolving mass distribution model to a dataset that exhibits evolution, we expect errors on cosmological parameters to broaden.
+Noticeable redshift evolution in the mass distribution is expected to occur above $z\simeq 1.5$, \citep{van_son_redshift_2022}, which is not yet relevant in current data or in expected O5 data, but will certainly be visible in next-generation detectors.}
 
-Because \ac{GW} observations are the only data input to spectral sirens, they are uniquely positioned to probe the expansion history of the universe over a wide range of cosmological redshift, rather than just the local expansion \Ho.
+Because \ac{GW} observations are the only data input to spectral sirens, they are sensitive to the expansion history of the universe over a wide range of cosmological redshift, rather than just the local expansion \Ho.
 We have shown this by simultaneously measuring \Omm{} and \Ho{}; the method can be trivially expanded to constrain additional cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$.
+\new{It is also possible to use a non-$\Lambda$CDM cosmological model for $H(z)$, with different parameters entirely.
+It may also be possible to forego the need for a parametric representation of $H(z)$ altogether, for example by using bins in luminosity distance.
+A similar technique using redshift bins is commonly used in cosmological measurements using galaxy surveys and type 1a supernovae \citep[e.g.][]{perlmutter_measuring_2003,anderson_clustering_2014,aghamousa_desi_2016}, but we have not pursued that possibility in this work.
+Such a method would likely increase uncertainties on $H(z)$, especially for $z\gtrsim 1.5$.}
 We find that for \acl{O5}, spectral sirens will be most constraining at $z=\variable{output/mostsensitivez.txt}$.
 This redshift is larger than the expected redshifts of detectable electromagnetic counterparts of binary neutron star mergers~\citep{kiendrebeogo_updated_2023}, implying that upgrades to current \ac{GW} detectors will allow the spectral siren method to probe $H(z)$ at otherwise unexplored distances by \acp{GW}.
+
 Proposed next-generation \ac{GW} detectors such as Cosmic Explorer and Einstein Telescope will be sensitive to \acp{CBC} across cosmic time (out to $z\sim100$)~\citep{et_steering_committee_einstein_2020, evans_horizon_2021}.
@@
 \begin{equation}
-    p(z|H_0, \Omega_M) \propto \frac{\diff V_C}{\diff z} \frac{1}{1+z} \frac{(1+z)^\alpha}{1+\left(\frac{1+z}{1+z_p}\right)^{\alpha+\beta}}.
+    p(z|H_0, \Omega_M) \propto \frac{\diff V_C}{\diff z} \frac{1}{1+z} \frac{(1+z)^\alpha_z}{1+\left(\frac{1+z}{1+z_p}\right)^{\alpha_z+\beta_z}}.
     \label{eq:underlying redshift dist}
@@
 $\mathcal{S}$ is a smoothing function at low and high masses, and $\mathcal{N}_{\text{pl}}$ and $\mathcal{N}_{\text{g}}$ are the normalizations between  $m_{\min}$ and $m_{\max}$ for the power law component and truncated Gaussian component, respectively.
+\new{The smoothing creates support for masses below $m_{\min}$ and above $m_{\max}$.}
 $\bar{\Lambda}_z = \{z_p,\alpha_z,\beta_z\}$ are the parameters governing the peak of the redshift distribution, low-$z$ power law slope, and high-$z$ power law slope.
@@
 
+%GW datasets have only just become large enough to support non-parametric methods, and we have shown that by the next LVK observing run, datasets will have become large enough that systmatic uncertainties in the mass distribution are relevant to cosmological constraints.
+Removing the need to specify a mass model parametrization eliminates the largest systematic in GW spectral sirens, and thus directly informs the cause of the "Hubble tension," one of the most pressing problems in cosmology today.
+
+% Because of this systematic uncertaintiy, spectral siren cosmology would be severely hampered in the era of next-generation GW detectors without the techniques presented in this work.
+Spectral sirens are the only method that can make use of the high-redshift data in next generation detectors.
+We therefore believe that both the method of non-parmetric spectral sirens as well as the result that theoretical knowledge of the mass distribution is not necessary for it to inform cosmological measurements will continue to impact the field of GW cosmology into the next-generation detector era.
+
+% A secondary effect may be their reconstructed mass distribution.
+Firstly, the rate at ~35 solar masses is likely lower to allow for a lack of mergers near 100 solar masses to be acheived while also maintaining a second power law slope that is not too steep.
+However, it is hard to verify this hypothesis as the plotted distributions in 2312.11627 are normalized probability distributions rather than differential merger rates.
+
 \end{document}

### 2024-05-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 The first condition (multiple features) is known to be met in current data \citep{abbott_population_2023}, with three robust features \citep{farah_things_2023}: a maximum black hole mass and overdensities at $\sim10\Msun$ and $\sim35\Msun$.
-\new{It is also predicted by many population synthesis studies, \citep[e.g.][]{mapelli_cosmic_2018,giacobbo_merging_2018,van_son_locations_2023}}
+\new{It is also predicted by many population synthesis studies, \citep[e.g.][]{mapelli_cosmic_2018,giacobbo_merging_2018,van_son_locations_2023}}.
 The latter scenario---with features identically and monotonically shifting to higher masses with increasing redshift---would be astrophysically unlikely as the locations of features in the mass distribution are each thought to be governed by fundamentally different physical processes \citep{mapelli_binary_2020}.

### 2024-05-15 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 We encourage readers to click these hyperlinks to verify or reproduce the results and methods of this work.
-The git repository associated to this study is publicly available at \GitHubURL{}, which allows anyone to re-build the entire manuscript.
+The git repository associated to this study allows anyone to re-build the entire manuscript, and is publicly available at \GitHubURL{} and \url{https://zenodo.org/records/11199928}.
 

### 2024-06-28 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 However, a primary utility of spectral siren measurements will be in constraining $H(z)$ \new{under different cosmological models, and} at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{You:2020wju, Ezquiaga:2020tns, ezquiaga_spectral_2022, Chen:2024gdn}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
-When applied to data from next-generation detectors, this method will measure \new{all cosmological parameters that govern $H(z)$, such as the curvature parameter, $\kappa$, matter density, \Omm, dark energy density, $\Omega_\Lambda$, and dark energy equation of state parameter, $w$.}
+%When applied to data from next-generation detectors, this method will measure \new{cosmological parameters that govern $H(z)$.}
 
@@
 However, the method can be trivially extended to the full mass distribution of \acp{CBC} \citep[e.g.][]{fishbach_does_2020, ezquiaga_spectral_2022}.
+\new{Additionally, Appendix~\ref{ap:mass ratio} demonstrates that including secondary masses in the fit does not change the results of the analysis presented here.}
 
@@
 These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{abbott_population_2023}.
-For simplicity, we additionally assume that the population has equal masses, $m_1=m_2$, and thus, $\theta = \{m_1,z\}$.
-We consistently apply this assumption to the data generation process and the population inference.
-We do not fit for or simulate spins, as they do not redshift and hence do not carry additional cosmological information, and additionally neglect secondary masses.
+For simplicity, we additionally assume that \new{the population of mass ratios is peaked at $q=0.9$, which is consistent with current observations.}
+Thus, $\theta = \{m_1,z\}$.
+We consistently apply these assumptions to the data generation process and the population inference.
+We do not fit for or simulate spins, as they do not redshift and hence do not carry additional cosmological information, and additionally \new{fix the distribution of mass ratios.}
 
@@
 The full procedure for this mock data generation process is described in \citet{fishbach_where_2017, farah_things_2023, essick_ensuring_2024}.
+
+\section{Effects of fitting for secondary mass}
+\label{ap:mass ratio}
+We examine the effect of only fitting the distribution of primary masses on our results.
+To do so, we perform two spectral siren analyses: one that includes a fit to secondary mass, and one that does not.
+Both analyses use the parametric, \plp{} model for the distribution of primary masses.
+We model the distribution of secondary masses with a power law in mass ratio, similarly to the majority of analyses presented in \citet{abbott_population_2021,abbott_population_2023}.
+
+We use the same set of simulated events for both analyses.
+These are generated in the same way as described in Section~\ref{sec:data generation}, but with a uniform mass ratio distribution (rather than assuming all mass ratios are unity) in order to understand the impact of our assumptions on the results over the full range of possible mass ratios.
+We correspondingly modify our injection set to cover the same mass ratio range.
+
+The results are shown in Figure~\ref{fig:mass ratio}.
+We find the posteriors on \Ho{} to be similar between the two cases.
+Additionally, the recovered mass spectra are nearly identical.
+We therefore conclude that fitting for secondary masses does not significantly impact our main conclusions, as is to be expected from a self-consistent hierarchical analysis when the data do not exhibit correlations.
+
+\begin{figure*}
+    \includegraphics[width=\textwidth]{figures/q_comparison.pdf}
+    \caption{Comparison of a parametric spectral siren analysis performed while fitting for the distributions of primary and secondary masses (magenta solid) and primary masses only (green dashed).
+    The posterior on \Ho{} is relatively unchanged between the two, and only minor changes are observed in the mass spectrum.}
+    \label{fig:mass ratio}
+    \script{fit_q_comparison.py}
+\end{figure*}
 

### 2024-06-28 src/scripts/mmin_hists.py
make the plot more readable

@@ -5,38 +5,40 @@ from tqdm import trange
 import paths
 from data_generation import TRUEVALS
 
-CIs = np.zeros(50)
-for i in trange(2):
-    id = az.from_netcdf(paths.data / f"bias/mcmc_parametric_PLP_unifq_actually_{i}.nc4")
+fig, axs = plt.subplots(nrows=1,ncols=2,figsize=(8,4))
+CIs = np.zeros(17)
+for i in trange(17):
+    id = az.from_netcdf(paths.data / f"bias/mcmc_parametric_PLP_{i}.nc4")
     num_less = np.sum(id.posterior['mmin'] < TRUEVALS['mmin'])
     frac_less = num_less/len(id.posterior['mmin'].T)
-    # id_BPL = az.from_netcdf(paths.data / f"bias/mcmc_parametric_BPL_{i}.nc4")
     if i==16:
         a=1
         l=3
         c='g'
         label="Run displayed in Figure 1"
         zorder=100
-        # plt.axvline(frac_less,color=c,alpha=a,lw=l,zorder=zorder,label=label)
+        axs[1].axvline(frac_less,color=c,alpha=a,lw=l,zorder=zorder,label=label)
     else:
         a=0.3
         l=1
         c='k'
-        label=None
         zorder=None
+        if i==0:
+            label=label="non-displayed runs"
+        else:
+            label=None
     CIs[i] = frac_less
-    # plt.hist(id_BPL.posterior['mmin'].T,density=True,histtype='step',color='orange',alpha=a,lw=l)
-    plt.hist(id.posterior['mmin'].T,density=True,histtype='step',color=c,alpha=a,lw=l,zorder=zorder,label=label)
-plt.axvline(TRUEVALS['mmin'],c='k',label='Injected value')
-# plt.hist(CIs,color=c,density=False,bins=15,histtype='step',alpha=a,lw=l,label="non-displayed runs")
-plt.xlabel("$m_{\\min}\, [M_\\odot]$")
-plt.legend()
+    axs[0].hist(id.posterior['mmin'].T,density=True,histtype='step',color=c,alpha=a,lw=l,zorder=zorder,label=label)
+axs[0].axvline(TRUEVALS['mmin'],c='k',label='Injected value')
+axs[0].set_xlabel("$m_{\\min}\, [M_\\odot]$")
+axs[1].hist(CIs,color='k',density=False,bins=15,histtype='step')
[…]

### 2024-06-28 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 However, the method can be trivially extended to the full mass distribution of \acp{CBC} \citep[e.g.][]{fishbach_does_2020, ezquiaga_spectral_2022}.
-\new{Additionally, Appendix~\ref{ap:mass ratio} demonstrates that including secondary masses in the fit does not change the results of the analysis presented here.}
+\new{Additionally, Appendix~\ref{ap:mass ratio} demonstrates that including the mass ratio distribution in the fit does not change the results of the analysis presented here.}
 
@@
 These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{abbott_population_2023}.
-For simplicity, we additionally assume that \new{the population of mass ratios is peaked at $q=0.9$, which is consistent with current observations.}
+For simplicity, we additionally assume \new{a uniform mass ratio distribution.}
 Thus, $\theta = \{m_1,z\}$.
@@
 \label{ap:mass ratio}
-We examine the effect of only fitting the distribution of primary masses on our results.
-To do so, we perform two spectral siren analyses: one that includes a fit to secondary mass, and one that does not.
+\new{We examine the effect of only fitting the distribution of primary masses on our results.
+To do so, we perform two spectral siren analyses: one that includes a fit to mass ratio, and one that does not.
 Both analyses use the parametric, \plp{} model for the distribution of primary masses.
-We model the distribution of secondary masses with a power law in mass ratio, similarly to the majority of analyses presented in \citet{abbott_population_2021,abbott_population_2023}.
+We model the distribution of mass ratios with a power law, similarly to the majority of analyses presented in \citet{abbott_population_2021,abbott_population_2023}.
 
 We use the same set of simulated events for both analyses.
-These are generated in the same way as described in Section~\ref{sec:data generation}, but with a uniform mass ratio distribution (rather than assuming all mass ratios are unity) in order to understand the impact of our assumptions on the results over the full range of possible mass ratios.
-We correspondingly modify our injection set to cover the same mass ratio range.
+These are generated in the same way as described in Section~\ref{sec:data generation}.
 
@@
 Additionally, the recovered mass spectra are nearly identical.
-We therefore conclude that fitting for secondary masses does not significantly impact our main conclusions, as is to be expected from a self-consistent hierarchical analysis when the data do not exhibit correlations.
-
+We therefore conclude that fitting for mass ratios does not significantly impact our main conclusions.
+}
 \begin{figure*}

### 2024-06-28 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 Both analyses use the parametric, \plp{} model for the distribution of primary masses.
-We model the distribution of mass ratios with a power law, similarly to the majority of analyses presented in \citet{abbott_population_2021,abbott_population_2023}.
+We model the distribution of mass ratios with a power law, similarly to the majority of analyses presented in \citet{abbott_population_2021,abbott_population_2023}.}
 
-We use the same set of simulated events for both analyses.
+\new{We use the same set of simulated events for both analyses.
 These are generated in the same way as described in Section~\ref{sec:data generation}.
+}
 
-The results are shown in Figure~\ref{fig:mass ratio}.
+\new{The results are shown in Figure~\ref{fig:mass ratio}.
 We find the posteriors on \Ho{} to be similar between the two cases.

### 2024-08-20 src/scripts/fit_q_comparison.py
remove whitespace

@@ -88,4 +88,3 @@ axes[1].set_ylabel('posterior density')
 fig.legend(ncol=4,framealpha=0,loc="outside upper center")
 plt.tight_layout()
 plt.savefig(paths.figures / "q_comparison.pdf")
-plt.show()

### 2024-08-20 src/scripts/pm_H0_twopanel.py
only make plots if explicitly run

@@ -75,6 +75,7 @@ def two_panel(path, path_PLP, path_BPL, hyperparam='H0'):
     fig.savefig(paths.figures / "O5_pm.pdf")
     plt.clf()
 
-two_panel(paths.data / "bias/mcmc_nonparametric_16.nc4", 
-          paths.data / "bias/mcmc_parametric_PLP_16.nc4",
-          paths.data / "bias/mcmc_parametric_BPL_16.nc4")
+if  __name__ == "__main__":
+    two_panel(paths.data / "bias/mcmc_nonparametric_16.nc4", 
+            paths.data / "bias/mcmc_parametric_PLP_16.nc4",
+            paths.data / "bias/mcmc_parametric_BPL_16.nc4")

### 2024-08-20 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2024-06-28 src/tex/ms.tex
@@
 However, a primary utility of spectral siren measurements will be in constraining $H(z)$ \new{under different cosmological models, and} at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{You:2020wju, Ezquiaga:2020tns, ezquiaga_spectral_2022, Chen:2024gdn}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
-%When applied to data from next-generation detectors, this method will measure \new{cosmological parameters that govern $H(z)$.}
+When applied to data from next-generation detectors, this method will measure \new{all cosmological parameters that govern $H(z)$, such as the curvature parameter, $\kappa$, matter density, \Omm, dark energy density, $\Omega_\Lambda$, and dark energy equation of state parameter, $w$.}
 
@@
 However, the method can be trivially extended to the full mass distribution of \acp{CBC} \citep[e.g.][]{fishbach_does_2020, ezquiaga_spectral_2022}.
-\new{Additionally, Appendix~\ref{ap:mass ratio} demonstrates that including the mass ratio distribution in the fit does not change the results of the analysis presented here.}
 
@@
 These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{abbott_population_2023}.
-For simplicity, we additionally assume \new{a uniform mass ratio distribution.}
-Thus, $\theta = \{m_1,z\}$.
-We consistently apply these assumptions to the data generation process and the population inference.
-We do not fit for or simulate spins, as they do not redshift and hence do not carry additional cosmological information, and additionally \new{fix the distribution of mass ratios.}
+For simplicity, we additionally assume that the population has equal masses, $m_1=m_2$, and thus, $\theta = \{m_1,z\}$.
+We consistently apply this assumption to the data generation process and the population inference.
+We do not fit for or simulate spins, as they do not redshift and hence do not carry additional cosmological information, and additionally neglect secondary masses.
 
@@
 The full procedure for this mock data generation process is described in \citet{fishbach_where_2017, farah_things_2023, essick_ensuring_2024}.
-
-\section{Effects of fitting for secondary mass}
-\label{ap:mass ratio}
-\new{We examine the effect of only fitting the distribution of primary masses on our results.
-To do so, we perform two spectral siren analyses: one that includes a fit to mass ratio, and one that does not.
-Both analyses use the parametric, \plp{} model for the distribution of primary masses.
-We model the distribution of mass ratios with a power law, similarly to the majority of analyses presented in \citet{abbott_population_2021,abbott_population_2023}.
-
-We use the same set of simulated events for both analyses.
-These are generated in the same way as described in Section~\ref{sec:data generation}.
-
-The results are shown in Figure~\ref{fig:mass ratio}.
-We find the posteriors on \Ho{} to be similar between the two cases.
-Additionally, the recovered mass spectra are nearly identical.
-We therefore conclude that fitting for mass ratios does not significantly impact our main conclusions.
-}
-\begin{figure*}
-    \includegraphics[width=\textwidth]{figures/q_comparison.pdf}
-    \caption{Comparison of a parametric spectral siren analysis performed while fitting for the distributions of primary and secondary masses (magenta solid) and primary masses only (green dashed).
-    The posterior on \Ho{} is relatively unchanged between the two, and only minor changes are observed in the mass spectrum.}
-    \label{fig:mass ratio}
-    \script{fit_q_comparison.py}
-\end{figure*}
 

### 2024-08-20 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2024-06-28 src/tex/ms.tex
@@
 and the \plp{} model reaches the same level of bias only $\variable{output/PLP_bias_percent.txt}\%$ 
-of the time.
-\new{Additionally, the \bpl{} model typically overestimates \Ho{}, while the \plp{} model produces a roughly equal number of over- and underestimates of \Ho{}, implying a systematic bias in the incorrect but not the correct parametric model.}\footnote{\new{
-We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.}}
+of the time, \new{meaning that the \bpl{} model produces a bias more than twice as often as the \plp{} model.}
+\new{Additionally, we show in Appendix~\ref{ap:parametric bias} that the \bpl{} model typically overestimates \Ho{} whereas the \plp{} model produces a roughly equal number of over- and underestimates of \Ho{}.
+This implies a systematic bias in the incorrect but not the correct parametric model.}
 
@@
 \new{We use the same set of simulated events for both analyses.
-These are generated in the same way as described in Section~\ref{sec:data generation}.
-}
-
-\new{The results are shown in Figure~\ref{fig:mass ratio}.
+These are generated in the same way as described in Section~\ref{sec:data generation}. The results are shown in Figure~\ref{fig:mass ratio}.
 We find the posteriors on \Ho{} to be similar between the two cases.
@@
 This can be seen explicitly in the source code accompanying this paper, in the \texttt{scripts/priors.py} \variable{output/priors_placeholder.txt}script.
+
+\section{Biases observed in parametric models}
+\label{ap:parametric bias}
+\new{Figure~\ref{fig:parametric bias} shows recovered posteriors on \Ho{} as inferred from 50 mock catalogs using both the \bpl{} and \plp{} parametric models.
+We find that the \plp{} model recovers \Ho{} posteriors that are largely symmetric about the true value, whereas the \bpl{} model typically finds more support for larger \Ho{} values.
+We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.
+
+Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
+Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.
+Therefore, while we are able to identify the shortcomings of \bpl{} when \plp{} describes the true mass distribution, it is not possible in practice to know which direction an incorrect parametric model will bias \Ho{}.
+Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\linewidth]{bias.pdf}
+    \caption{Caption}
+    \label{fig:parametric bias}
+    \script{bias_numbers.py}
+\end{figure}
 

### 2024-08-20 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 We find that the \plp{} model recovers \Ho{} posteriors that are largely symmetric about the true value, whereas the \bpl{} model typically finds more support for larger \Ho{} values.
-We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.
+We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.}
 
-Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
+\new{Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
 Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.

### 2024-08-20 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2024-08-20 src/tex/ms.tex
@@
 \label{ap:parametric bias}
-\new{Figure~\ref{fig:parametric bias} shows recovered posteriors on \Ho{} as inferred from 50 mock catalogs using both the \bpl{} and \plp{} parametric models.
-We find that the \plp{} model recovers \Ho{} posteriors that are largely symmetric about the true value, whereas the \bpl{} model typically finds more support for larger \Ho{} values.
-We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.
+% \new{Figure~\ref{fig:parametric bias} shows recovered posteriors on \Ho{} as inferred from 50 mock catalogs using both the \bpl{} and \plp{} parametric models.
+% We find that the \plp{} model recovers \Ho{} posteriors that are largely symmetric about the true value, whereas the \bpl{} model typically finds more support for larger \Ho{} values.
+% We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.}
 
-Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
-Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.
-Therefore, while we are able to identify the shortcomings of \bpl{} when \plp{} describes the true mass distribution, it is not possible in practice to know which direction an incorrect parametric model will bias \Ho{}.
-Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.}
+% \new{Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
+% Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.
+% Therefore, while we are able to identify the shortcomings of \bpl{} when \plp{} describes the true mass distribution, it is not possible in practice to know which direction an incorrect parametric model will bias \Ho{}.
+% Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.}
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.5\linewidth]{bias.pdf}
-    \caption{Caption}
-    \label{fig:parametric bias}
-    \script{bias_numbers.py}
-\end{figure}
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=0.5\linewidth]{bias.pdf}
+%     \caption{Recovered posteriors on \Ho{} from spectral siren analyses on 50 mock catalogs with the \plp{} (green) and \bpl{} () parametric models.
+The true, injected value is shown by a vertical black line.
+}
+%     \label{fig:parametric bias}
+%     \script{bias_numbers.py}
+% \end{figure}
 

### 2024-08-20 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-% \new{Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
-% Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.
-% Therefore, while we are able to identify the shortcomings of \bpl{} when \plp{} describes the true mass distribution, it is not possible in practice to know which direction an incorrect parametric model will bias \Ho{}.
-% Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.}
+\new{Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
+Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.
+Therefore, while we are able to identify the shortcomings of \bpl{} when \plp{} describes the true mass distribution, it is not possible in practice to know which direction an incorrect parametric model will bias \Ho{}.
+Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.}
 

### 2024-08-20 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 \new{Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
-Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.
+Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.}
 Therefore, while we are able to identify the shortcomings of \bpl{} when \plp{} describes the true mass distribution, it is not possible in practice to know which direction an incorrect parametric model will bias \Ho{}.
-Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.}
+\new{Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.}
 
@@
 %     \includegraphics[width=0.5\linewidth]{bias.pdf}
-%     \caption{Recovered posteriors on \Ho{} from spectral siren analyses on 50 mock catalogs with the \plp{} (green) and \bpl{} () parametric models.
-The true, injected value is shown by a vertical black line.
-}
+%     \caption{Recovered posteriors on \Ho{} from spectral siren analyses on 50 mock catalogs with the \plp{} (green) and \bpl{} (pink) parametric models.
+The true, injected value is shown by a vertical black line.}
 %     \label{fig:parametric bias}

### 2024-08-20 src/tex/ms.tex
[showyourwork] overleaf sync

--- 2024-08-20 src/tex/ms.tex
@@
 
-\new{Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
+Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
 Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.
 Therefore, while we are able to identify the shortcomings of \bpl{} when \plp{} describes the true mass distribution, it is not possible in practice to know which direction an incorrect parametric model will bias \Ho{}.
-Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.}
+Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.
 
@@
 %     \centering
-%     \includegraphics[width=0.5\linewidth]{bias.pdf}
-%     \caption{Recovered posteriors on \Ho{} from spectral siren analyses on 50 mock catalogs with the \plp{} (green) and \bpl{} () parametric models.
-The true, injected value is shown by a vertical black line.
-}
+%     \includegraphics[width=0.5\linewidth]{figures/bias.pdf}
+%     \caption{Recovered posteriors on \Ho{} from spectral siren analyses on 50 mock catalogs with the \plp{} (green) and \bpl{} (pink) parametric models.
+The true, injected value is shown by a vertical black line.}
 %     \label{fig:parametric bias}

### 2024-08-20 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 \label{ap:parametric bias}
-% \new{Figure~\ref{fig:parametric bias} shows recovered posteriors on \Ho{} as inferred from 50 mock catalogs using both the \bpl{} and \plp{} parametric models.
-% We find that the \plp{} model recovers \Ho{} posteriors that are largely symmetric about the true value, whereas the \bpl{} model typically finds more support for larger \Ho{} values.
-% We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.}
+Figure~\ref{fig:parametric bias} shows recovered posteriors on \Ho{} as inferred from 50 mock catalogs using both the \bpl{} and \plp{} parametric models.
+We find that the \plp{} model recovers \Ho{} posteriors that are largely symmetric about the true value, whereas the \bpl{} model typically finds more support for larger \Ho{} values.
+We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.
 
@@
 
-% \begin{figure}
-%     \centering
-%     \includegraphics[width=0.5\linewidth]{figures/bias.pdf}
-%     \caption{Recovered posteriors on \Ho{} from spectral siren analyses on 50 mock catalogs with the \plp{} (green) and \bpl{} (pink) parametric models.
+\begin{figure}
+    \centering
+    \includegraphics[width=0.5\linewidth]{figures/bias.pdf}
+    \caption{Recovered posteriors on \Ho{} from spectral siren analyses on 50 mock catalogs with the \plp{} (green) and \bpl{} (pink) parametric models.
 The true, injected value is shown by a vertical black line.}
-%     \label{fig:parametric bias}
-%     \script{bias_numbers.py}
-% \end{figure}
+    \label{fig:parametric bias}
+    \script{bias_numbers.py}
+\end{figure}
 

### 2024-08-20 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 % Author list
-%\author{@afarah18}
 \author[0000-0002-6121-0285]{Amanda M. Farah}
@@
 \label{ap:parametric bias}
-Figure~\ref{fig:parametric bias} shows recovered posteriors on \Ho{} as inferred from 50 mock catalogs using both the \bpl{} and \plp{} parametric models.
+\new{Figure~\ref{fig:parametric bias} shows recovered posteriors on \Ho{} as inferred from 50 mock catalogs using both the \bpl{} and \plp{} parametric models.
 We find that the \plp{} model recovers \Ho{} posteriors that are largely symmetric about the true value, whereas the \bpl{} model typically finds more support for larger \Ho{} values.
-We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.
+We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.}
 
-Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
+\new{Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
 Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.
 Therefore, while we are able to identify the shortcomings of \bpl{} when \plp{} describes the true mass distribution, it is not possible in practice to know which direction an incorrect parametric model will bias \Ho{}.
-Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.
+Mitigating the bias by using more flexible models may thus be the simplest solution to this problem.}
 

### 2024-08-20 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 
-\new{The non-parametric mass model's ability to recover injected cosmological parameters demonstrates} that the information in the spectral siren measurement does not come from the enforcement of specific features in the mass distribution.
+The non-parametric mass model's ability to recover injected cosmological parameters demonstrates that the information in the spectral siren measurement does not come from the enforcement of specific features in the mass distribution.
 Rather, it is provided by the assumption that either all \acp{CBC} follow a common mass distribution, or that any evolution of the mass distribution with redshift does not exactly mimic cosmology \citep{ezquiaga_spectral_2022}.
@@
 % \new{This is to be compared with a $\variable{output/PLPh0percent.txt}\%$ measurement of $H(z=\variable{output/mostsensitivez.txt})$ with the correct parametric model, which does not represent a statistically-significant difference.}
-We highlight measurements of \Ho{} \new{within a $\Lambda$CDM universe} in order to benchmark the accuracy and precision of our non-parametric method, as well as explore the role of spectral sirens in elucidating the Hubble tension.
-However, a primary utility of spectral siren measurements will be in constraining $H(z)$ \new{under different cosmological models, and} at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{You:2020wju, Ezquiaga:2020tns, ezquiaga_spectral_2022, Chen:2024gdn}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
+We highlight measurements of \Ho{} within a $\Lambda$CDM universe in order to benchmark the accuracy and precision of our non-parametric method, as well as explore the role of spectral sirens in elucidating the Hubble tension.
+However, a primary utility of spectral siren measurements will be in constraining $H(z)$ under different cosmological models, and at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{You:2020wju, Ezquiaga:2020tns, ezquiaga_spectral_2022, Chen:2024gdn}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
 %When applied to data from next-generation detectors, this method will measure \new{cosmological parameters that govern $H(z)$.}
@@
 By generating a catalog of \ac{GW} sources from a known population and cosmological model, we are able to quantify the accuracy of our method.
-The use of simulated data also enables us to make projections for future datasets and safely ignore dimensions such as spin that do not impact cosmological measurements but would otherwise be important to simultaneously fit to avoid biases in population inference of real data, \new{as real data exhibit correlations between distance, mass, and spin measurements}~\citep{biscoveanu_sources_2021}.
+The use of simulated data also enables us to make projections for future datasets and safely ignore dimensions such as spin that do not impact cosmological measurements but would otherwise be important to simultaneously fit to avoid biases in population inference of real data, as real data exhibit correlations between distance, mass, and spin measurements~\citep{biscoveanu_sources_2021}.
 
@@
 Additionally, the mass distribution inferred with the \ac{GP}-based model closely resembles the true, simulated distribution.
-\new{The \plp{} fit appears to be inconsistent with the true distribution at low masses, but this inconsistency is not observed in the majority of fits to other realizations of simulated catalogs, so is likely a statistical fluctuation in the particular catalog arbitrarily chosen for this plot.}
-This indicates that using models that cannot accurately approximate the true mass distribution will lead to a significant systematic bias in the estimation of cosmological parameters.
+This indicates that using models that cannot accurately approximate the true mass distribution will lead to a noticeable systematic bias in the estimation of cosmological parameters.
 
@@
 To obtain a quantitative measure of the systematic bias introduced by mismodeling the mass distribution, we repeat the parametric analyses with 50 separate simulated catalogs of $\sim1,000$ events each.
-We find that the \bpl{} model produces an \new{over- or under}estimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time,
+We find that the \bpl{} model produces an over- or under-estimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time,
 and the \plp{} model reaches the same level of bias only $\variable{output/PLP_bias_percent.txt}\%$ 
@@
 Figure~\ref{fig:O5_GP} demonstrates an expected $\variable{output/nonparh0percent.txt}\%$ ($1\sigma$ uncertainties) measurement of \Ho{} after one year of \ac{O5} using the \ac{GP}-based spectral siren method, and a $\variable{output/PLPh0percent.txt}\%$ measurement with parametric spectral sirens, demonstrating comparable statistical uncertainties.
-\new{However, we note that the precision reached in \ac{O5} may lessen depending on the actual level of measurement uncertainty in individual events, and the existence (or lack thereof) of a maximum mass feature, which we have assumed to be present in our simulated dataset.}
-\new{These numbers are estimated from a fit to a single simulated catalog, but we find similar levels of statistical uncertainty from fits to different catalog realizations.}
+However, we note that the precision reached in \ac{O5} may lessen depending on the actual level of measurement uncertainty in individual events, and the existence (or lack thereof) of a maximum mass feature, which we have assumed to be present in our simulated dataset.
+These numbers are estimated from a fit to a single simulated catalog, but we find similar levels of statistical uncertainty from fits to different catalog realizations.
 By the time of \ac{O5}, the \ac{GW} detector network is projected to detect \acp{BBH} up to redshift $\approx 3$, with most sources lying near redshift $\simeq 1.2$ \citep{chen_distance_2021}.
@@
 The first condition (multiple features) is known to be met in current data \citep{abbott_population_2023}, with three robust features \citep{farah_things_2023}: a maximum black hole mass and overdensities at $\sim10\Msun$ and $\sim35\Msun$.
-\new{It is also predicted by many population synthesis studies, \citep[e.g.][]{mapelli_cosmic_2018,giacobbo_merging_2018,van_son_locations_2023}}.
+It is also predicted by many population synthesis studies, \citep[e.g.][]{mapelli_cosmic_2018,giacobbo_merging_2018,van_son_locations_2023}.
 The latter scenario---with features identically and monotonically shifting to higher masses with increasing redshift---would be astrophysically unlikely as the locations of features in the mass distribution are each thought to be governed by fundamentally different physical processes \citep{mapelli_binary_2020}.
@@
 Future work will extend the method developed here to mass distributions that are allowed to evolve with redshift.
-\new{When applying an evolving mass distribution model to a dataset that exhibits evolution, we expect errors on cosmological parameters to broaden.
-Noticeable redshift evolution in the mass distribution is expected to occur above $z\simeq 1.5$, \citep{van_son_redshift_2022}, which is not yet relevant in current data or in expected O5 data, but will certainly be visible in next-generation detectors.}
+When applying an evolving mass distribution model to a dataset that exhibits evolution, we expect errors on cosmological parameters to broaden.
+Noticeable redshift evolution in the mass distribution is expected to occur above $z\simeq 1.5$, \citep{van_son_redshift_2022}, which is not yet relevant in current data or in expected O5 data, but will certainly be visible in next-generation detectors.
 
@@
 We have shown this by simultaneously measuring \Omm{} and \Ho{}; the method can be trivially expanded to constrain additional cosmological parameters that govern $H(z)$, such as the dark energy equation of state parameter, $w$.
-\new{It is also possible to use a non-$\Lambda$CDM cosmological model for $H(z)$, with different parameters entirely.
+It is also possible to use a non-$\Lambda$CDM cosmological model for $H(z)$, with different parameters entirely.
 It may also be possible to forego the need for a parametric representation of $H(z)$ altogether, for example by using bins in luminosity distance.
 A similar technique using redshift bins is commonly used in cosmological measurements using galaxy surveys and type 1a supernovae \citep[e.g.][]{perlmutter_measuring_2003,anderson_clustering_2014,aghamousa_desi_2016}, but we have not pursued that possibility in this work.
-Such a method would likely increase uncertainties on $H(z)$, especially for $z\gtrsim 1.5$.}
+Such a method would likely increase uncertainties on $H(z)$, especially for $z\gtrsim 1.5$.
 We find that for \acl{O5}, spectral sirens will be most constraining at $z=\variable{output/mostsensitivez.txt}$.
@@
 $\mathcal{S}$ is a smoothing function at low and high masses, and $\mathcal{N}_{\text{pl}}$ and $\mathcal{N}_{\text{g}}$ are the normalizations between  $m_{\min}$ and $m_{\max}$ for the power law component and truncated Gaussian component, respectively.
-\new{The smoothing creates support for masses below $m_{\min}$ and above $m_{\max}$.}
+The smoothing creates support for masses below $m_{\min}$ and above $m_{\max}$.
 $\bar{\Lambda}_z = \{z_p,\alpha_z,\beta_z\}$ are the parameters governing the peak of the redshift distribution, low-$z$ power law slope, and high-$z$ power law slope.

### 2024-08-20 src/scripts/bias_numbers.py
plot formatting

@@ -48,25 +48,24 @@ with open(paths.output / "nonparh0offset.txt","w") as f:
 # make a plot of recovered H0 posteriors
 fig = plt.figure(figsize=(7.5/2,4*.75),facecolor='none')
 prior=np.linspace(30,110,num=100)
+a=0.3
+l=1
 for i in range(N_CATALOGS):
     id_PLP = az.from_netcdf(paths.data / f"bias/mcmc_parametric_PLP_{i}.nc4")
     id_BPL = az.from_netcdf(paths.data / f"bias/mcmc_parametric_BPL_{i}.nc4")
     kde_PLP = gaussian_kde(id_PLP.posterior['H0'][0].values)
     kde_BPL = gaussian_kde(id_BPL.posterior['H0'][0].values)
-    a=0.3
-    l=1
-    label=None
-    zorder=None
-    plt.plot(prior,kde_PLP(prior),color=color_PLP,alpha=a,lw=l,zorder=zorder)
-    plt.plot(prior,kde_BPL(prior),color=color_BPL,alpha=a,lw=l,zorder=zorder)
+    plt.plot(prior,kde_PLP(prior),color=color_PLP,alpha=a,lw=l)
+    plt.plot(prior,kde_BPL(prior),color=color_BPL,alpha=a,lw=l)
 plt.plot([],c=color_PLP,label=r'\textsc{Power Law + Peak}')
 plt.plot([],c=color_BPL,label=r'\textsc{Broken Power Law}')
 # plt.plot([],c='k',lw=l,alpha=a,label="all other runs")
 plt.ylabel('posterior density')
-plt.legend(framealpha=0.9)
+leg=plt.legend(framealpha=1)
+leg.get_frame().set_linewidth(0.0)
 plt.xlabel('$H_0$ [km/s/Mpc]')
-plt.axvline(H0_FID, c='k',zorder=100)
-plt.ylim(0,0.14)
+plt.axvline(H0_FID, c='k')
+plt.ylim(0,0.12)
 plt.xlim(30,110)
 plt.savefig(paths.figures / "bias.pdf")
 plt.clf()

### 2024-08-21 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 and the \plp{} model reaches the same level of bias only $\variable{output/PLP_bias_percent.txt}\%$ 
-of the time, \new{meaning that the \bpl{} model produces a bias more than twice as often as the \plp{} model.}
+of the time, \new{meaning that the \bpl{} model produces a bias more than three times as often as the \plp{} model.}
 \new{Additionally, we show in Appendix~\ref{ap:parametric bias} that the \bpl{} model typically overestimates \Ho{} whereas the \plp{} model produces a roughly equal number of over- and underestimates of \Ho{}.

### 2024-08-22 src/tex/ms.tex
[showyourwork] overleaf sync

@@
 However, a primary utility of spectral siren measurements will be in constraining $H(z)$ under different cosmological models, and at redshifts that are relatively inaccessible by electromagnetic observations, especially with next-generation gravitational-wave detectors \citep{You:2020wju, Ezquiaga:2020tns, ezquiaga_spectral_2022, Chen:2024gdn}.\footnote{See example spectral siren cosmological inference using parametric mass spectrum models: \href{https://github.com/ezquiaga/spectral_sirens}{https://github.com/ezquiaga/spectral\_sirens}.}
-%When applied to data from next-generation detectors, this method will measure \new{cosmological parameters that govern $H(z)$.}
+%When applied to data from next-generation detectors, this method will measure cosmological parameters that govern $H(z)$.
 
@@
 
-To obtain a quantitative measure of the systematic bias introduced by mismodeling the mass distribution, we repeat the parametric analyses with 50 separate simulated catalogs of $\sim1,000$ events each.
+To obtain a quantitative measure of the systematic bias introduced by mis-modeling the mass distribution, we repeat the parametric analyses with 50 separate simulated catalogs of $\sim1,000$ events each.
 We find that the \bpl{} model produces an over- or under-estimate of \Ho{} at greater than $1\sigma$ $\variable{output/BPL_bias_percent.txt}\%$ of the time, %whereas the \ac{GP}-based model reaches that level of bias only \result{Z}\% of the time,
@@
 \new{Additionally, we show in Appendix~\ref{ap:parametric bias} that the \bpl{} model typically overestimates \Ho{} whereas the \plp{} model produces a roughly equal number of over- and underestimates of \Ho{}.
-This implies a systematic bias in the incorrect but not the correct parametric model.}
+This demonstrates that mis-modeling the mass distribution can introduce statistically significant systematic biases into measurements of cosmological parameters.}
 
-Collectively, these results indicate that prior knowledge of the shape of the mass distribution is not required to perform an unbiased spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made.
+Collectively, our results indicate that \emph{prior knowledge of the shape of the mass distribution is not required to perform an unbiased spectral siren measurement, so long as strong assumptions about the shape of the mass distribution are not made}.
 
@@
 These choices correspond to the maximum \emph{a posteriori} values obtained by an analysis of GWTC-3 data using the \plp{} model \citep{abbott_population_2023}.
-For simplicity, we additionally assume \new{a uniform mass ratio distribution.}
+For simplicity, we assume \new{a uniform mass ratio distribution.}
 Thus, $\theta = \{m_1,z\}$.
 We consistently apply these assumptions to the data generation process and the population inference.
-We do not fit for or simulate spins, as they do not redshift and hence do not carry additional cosmological information, and additionally \new{fix the distribution of mass ratios.}
+We do not fit for or simulate spins, as they do not redshift and hence do not carry additional cosmological information, and \new{we fix the distribution of mass ratios.}
 
@@
 \new{We use the same set of simulated events for both analyses.
-These are generated in the same way as described in Section~\ref{sec:data generation}. The results are shown in Figure~\ref{fig:mass ratio}.
+These are generated in the same way as described in Section~\ref{sec:data generation}.
+The inference results are shown in Figure~\ref{fig:mass ratio}.
 We find the posteriors on \Ho{} to be similar between the two cases.
@@
     \caption{Comparison of a parametric spectral siren analysis performed while fitting for the distributions of primary and secondary masses (magenta solid) and primary masses only (green dashed).
-    The posterior on \Ho{} is relatively unchanged between the two, and only minor changes are observed in the mass spectrum.}
+    The posterior on \Ho{} is relatively unchanged between the two, and only minor differences are observed in the mass spectra.}
     \label{fig:mass ratio}
@@
 
-\section{Biases observed in parametric models}
+\section{Biases induced by the \bpl{} model}
 \label{ap:parametric bias}
@@
 We find that the \plp{} model recovers \Ho{} posteriors that are largely symmetric about the true value, whereas the \bpl{} model typically finds more support for larger \Ho{} values.
-We note that \citep{pierra_study_2023} report an underestimate of \Ho{} with a fit of their Broken Power Law model to a single dataset generated with \plp, but their prior bounds on the maximum mass parameter do not extend below the injected value of that parameter, which in turn does not allow for the possibility of an overestimated \Ho{}.}
+We note that \citep{pierra_study_2023} report an underestimate of \Ho{} when fitting their Broken Power Law model to a single dataset generated with \plp, but the prior bounds on the maximum mass parameter of their model do not extend below the injected value of that parameter, which in turn does
+not allow for the possibility of an overestimated \Ho{}.
+It is therefore not possible to directly compare the direction of our observed \Ho{} offset with theirs, despite the similarities bewteen our chosen population models.}
 
-\new{Regardless, this systematic offset further demonstrates that incorrect parametric models induce a systematic bias in cosmological inference rather than an increased statistical uncertainty. 
+\new{Regardless, this systematic offset demonstrates that incorrect parametric models unfortunately induce a systematic bias in cosmological inference, not an increase in statistical uncertainty. 
 Other parametric models may show different trends depending on which parts of the mass distribution they incorrectly model and how.
